---
title: "Argelaguet: cell embeddings from genomic windows entropy"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
params:
  features_dir: ""
  win_cell_feature: ""
  win_feature: ""
  win_bed: ""
  windows_annotation: ""
  manifest: ""
  out_dir: ""
  log_path: ""
  threads: 1
  i_max_lag: 3
---

```{r logging_early, include = FALSE}
source(Sys.getenv("AMET_RENDER_HELPERS"))
param <- amet_setup_render_logging(params$log_path, params$threads)
```

```{r load_window_annotation}
window_annotation <- amet_load_annotation_matrix(params$windows_annotation)
if (is.null(window_annotation)) {
  message("[annotation] no per-window annotation matrix; per-locus annotation ",
          "UMAPs will be skipped")
  annotation_cols <- character(0)
} else {
  annotation_cols <- setdiff(names(window_annotation),
                             c("chrom", "start", "end", "feature_id"))
  cat("Annotations attached to windows:", length(annotation_cols), "\n")
  cat("Windows in annotation matrix:", nrow(window_annotation), "\n")
}
```

```{r setup}
suppressPackageStartupMessages({
  library(data.table)
  library(readr)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(BiocParallel)
  library(uwot)
  library(irlba)
  library(matrixStats)

  library(cluster)
  library(viridis)
  library(cowplot)
})

repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))
source(file.path(repo_root, "workflow", "scripts", "palettes.R"))
source(file.path(repo_root, "workflow", "scripts", "embedding_utils.R"))


knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  cache = FALSE,
  fig.path = "argelaguet_embeddings_plots/"
)
```

```{r constants}
N_HVW <- 1000
N_PCS <- 10

## amet exports i_total and mean_meth per (cell, window); both are carried
## forward as separate assays for embedding.
assay_map <- c(
  "i_total" = "i_total",
  "meth" = "methylation"
)
```

## Cell selection and NA-aware embedding

The cells embedded here are not a uniformly random subsample. For each
(stage, lineage) annotation we keep up to a fixed number of cells, picking the
highest-coverage cells (file size on disk is monotonic in CpG count) and
distributing the slots round-robin across plates. Stratifying by plate first
keeps a single high-coverage plate from dominating an annotation. The visible
cost is that any residual plate effect (cell cycle, batch) is preserved rather
than averaged out across plates; the alternative of random sampling produced
embeddings collapsed to a handful of cells per group because of how few windows
were complete.

scBS-seq methylation and per-cell entropy values have a meaningful zero
(unmethylated CpG, zero local entropy) which is distinct from missing coverage.
Imputing NAs to zero would bias both the variance signal and the embedding
geometry, so the pipeline keeps NAs as NAs throughout: HVF selection scores
features by variance times coverage fraction, and when too few windows have
complete coverage the PCA step uses NIPALS (`pcaMethods`) which iterates over
the observed entries without filling in missing values. Cells are dropped only
when their per-cell NA fraction is above the QC threshold; once they pass that
filter they are never dropped to make a feature block dense.

# Load data

```{r load_amet}
## amet ships per-cell-per-window i_total and per-cell mean_meth.
## Build cells x windows matrices, one per assay.
win_cf <- fread(params$win_cell_feature)
man <- fread(params$manifest)

col_data <- as.data.frame(unique(win_cf[, .(cell_id)]))
col_data <- merge(col_data, as.data.frame(man),
                  by = "cell_id", all.x = TRUE)
col_data$stage <- factor(col_data$stage)
col_data$lineage <- factor(col_data$lineage10x)
rownames(col_data) <- col_data$cell_id

build_assay_mat <- function(value_col) {
  m <- dcast(win_cf, feature_id ~ cell_id, value.var = value_col,
             fun.aggregate = mean)
  rn <- m$feature_id
  m <- as.matrix(m[, -1])
  rownames(m) <- rn
  m[!is.finite(m)] <- NA
  m[, col_data$cell_id, drop = FALSE]
}

assays_list <- list(
  i_total = build_assay_mat("i_total"),
  meth = build_assay_mat("mean_meth")
)

cat("Cells:", nrow(col_data), "\n")
cat("Windows:", nrow(assays_list$i_total), "\n")
cat("Assays:", paste(names(assays_list), collapse = ", "), "\n")
cat("Stages:\n")
print(table(col_data$stage, useNA = "ifany"))
cat("Top lineages:\n")
print(sort(table(col_data$lineage, useNA = "ifany"), decreasing = TRUE))
```

# Shared helpers

```{r helpers}
hvw_pca_umap <- function(mat) {
  run_embedding(mat, n_hvf = N_HVW, n_pcs = N_PCS,
                n_neighbors = 15L, min_dist = 0.3, seed = 42L)
}

plot_umap_stage <- function(coords, kept_cols, title_prefix) {
  df <- data.frame(
    umap1 = coords[, 1], umap2 = coords[, 2],
    stage = col_data$stage[kept_cols]
  )
  ggplot(df, aes(x = umap1, y = umap2, color = stage)) +
    geom_point(size = 0.6, alpha = 0.7) +
    scale_color_manual(values = argelaguet_stage_pal) +
    labs(title = paste(title_prefix, "- stage"), x = "UMAP 1", y = "UMAP 2") +
    guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
    theme_ng()
}

plot_umap_lineage <- function(coords, kept_cols, title_prefix) {
  df <- data.frame(
    umap1 = coords[, 1], umap2 = coords[, 2],
    lineage = col_data$lineage[kept_cols]
  )
  ggplot(df, aes(x = umap1, y = umap2, color = lineage)) +
    geom_point(size = 0.6, alpha = 0.7) +
    scale_color_manual(values = argelaguet_lineage_pal) +
    labs(title = paste(title_prefix, "- lineage"), x = "UMAP 1", y = "UMAP 2") +
    guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1),
                                ncol = 2)) +
    theme_ng()
}
```

# Per-window embeddings (HVW -> PCA -> UMAP)

HVW selection and NA handling: see `embedding_utils.R`.
Top `r N_HVW` most variable windows retained per assay before PCA and UMAP.

```{r win_hvw_pca_umap}
win_embeds <- lapply(names(assay_map), function(akey) {
  if (!akey %in% names(assays_list)) return(NULL)
  hvw_pca_umap(assays_list[[akey]])
})
names(win_embeds) <- assay_map

failed <- names(which(sapply(win_embeds, is.null)))
if (length(failed) > 0)
  warning("run_embedding returned NULL for: ", paste(failed, collapse = ", "))

cat("\nHVF selection summary:\n")
for (nm in names(win_embeds)) {
  em <- win_embeds[[nm]]
  if (is.null(em)) {
    cat(nm, ": NULL\n")
  } else {
    cat(nm, ": HVF =", length(em$hvf_idx),
        "| cells =", sum(em$kept_cols), "/", nrow(col_data), "\n")
  }
}

saveRDS(
  list(col_data = col_data,
       n_windows = nrow(assays_list$i_total),
       n_cells = nrow(col_data),
       assay_names = names(assays_list),
       win_embeds = win_embeds,
       assay_map = assay_map),
  "argelaguet_umap_windows_i_total.rds", compress = "xz"
)
```

## By stage {.tabset .tabset-pills}

```{r win_stage_i_total, fig.width = ng_fig_size(1, 1, panel_mm = 60)$w, fig.height = ng_fig_size(1, 1, panel_mm = 60)$h}
### i_total
if (!is.null(win_embeds[["i_total"]])) {
  print(plot_umap_stage(win_embeds[["i_total"]]$umap,
                        win_embeds[["i_total"]]$kept_cols,
                        "i_total (windows)"))
}
```

```{r win_stage_meth, fig.width = ng_fig_size(1, 1, panel_mm = 60)$w, fig.height = ng_fig_size(1, 1, panel_mm = 60)$h}
### methylation
if (!is.null(win_embeds[["methylation"]])) {
  print(plot_umap_stage(win_embeds[["methylation"]]$umap,
                        win_embeds[["methylation"]]$kept_cols,
                        "methylation (windows)"))
}
```

## By lineage {.tabset .tabset-pills}

```{r win_lineage_i_total, fig.width = ng_fig_size(1, 1, panel_mm = 60, legend_mm = 55)$w, fig.height = ng_fig_size(1, 1, panel_mm = 60, legend_mm = 55)$h}
### i_total
if (!is.null(win_embeds[["i_total"]])) {
  print(plot_umap_lineage(win_embeds[["i_total"]]$umap,
                          win_embeds[["i_total"]]$kept_cols,
                          "i_total (windows)"))
}
```

```{r win_lineage_meth, fig.width = ng_fig_size(1, 1, panel_mm = 60, legend_mm = 55)$w, fig.height = ng_fig_size(1, 1, panel_mm = 60, legend_mm = 55)$h}
### methylation
if (!is.null(win_embeds[["methylation"]])) {
  print(plot_umap_lineage(win_embeds[["methylation"]]$umap,
                          win_embeds[["methylation"]]$kept_cols,
                          "methylation (windows)"))
}
```

## Two-panel comparison (stage)

```{r win_panel_stage, fig.width = ng_fig_size(2, 1, panel_mm = 55)$w, fig.height = ng_fig_size(2, 1, panel_mm = 55)$h}
ok_embeds <- Filter(Negate(is.null), win_embeds)
df_panel <- bind_rows(lapply(names(ok_embeds), function(lbl) {
  em <- ok_embeds[[lbl]]
  data.frame(umap1 = em$umap[, 1], umap2 = em$umap[, 2],
             stage = col_data$stage[em$kept_cols],
             lineage = col_data$lineage[em$kept_cols],
             assay = lbl)
}))
df_panel$assay <- factor(df_panel$assay, levels = assay_map)

ggplot(df_panel, aes(x = umap1, y = umap2, color = stage)) +
  geom_point(size = 0.4, alpha = 0.6) +
  scale_color_manual(values = argelaguet_stage_pal) +
  facet_wrap(~ assay, nrow = 1) +
  labs(x = "UMAP 1", y = "UMAP 2",
       title = "Per-window embeddings (HVW -> PCA -> UMAP)") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

## Two-panel comparison (lineage)

```{r win_panel_lineage, fig.width = ng_fig_size(2, 1, panel_mm = 55, legend_mm = 55)$w, fig.height = ng_fig_size(2, 1, panel_mm = 55, legend_mm = 55)$h}
ggplot(df_panel, aes(x = umap1, y = umap2, color = lineage)) +
  geom_point(size = 0.4, alpha = 0.6) +
  scale_color_manual(values = argelaguet_lineage_pal) +
  facet_wrap(~ assay, nrow = 1) +
  labs(x = "UMAP 1", y = "UMAP 2",
       title = "Per-window embeddings (HVW -> PCA -> UMAP)") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1),
                              ncol = 2)) +
  theme_ng()
```

## Variance explained

```{r win_varexp, fig.width = ng_fig_size(2, 1)$w, fig.height = ng_fig_size(2, 1)$h}
ve_win <- bind_rows(lapply(names(assay_map), function(akey) {
  lbl <- assay_map[[akey]]
  if (is.null(win_embeds[[lbl]]) || !akey %in% names(assays_list)) return(NULL)
  kept <- win_embeds[[lbl]]$kept_cols
  hvw_mat <- assays_list[[akey]][win_embeds[[lbl]]$hvf_idx, kept, drop = FALSE]
  data.frame(
    assay = lbl,
    stage = median(row_variance_explained(hvw_mat, col_data$stage[kept]), na.rm = TRUE),
    lineage = median(row_variance_explained(hvw_mat, col_data$lineage[kept]), na.rm = TRUE))
}))

ve_long <- pivot_longer(ve_win, c("stage", "lineage"),
                        names_to = "variable", values_to = "median_ve")
ve_long$assay <- factor(ve_long$assay, levels = assay_map)

write.csv(ve_long, "argelaguet_win_varexp.csv", row.names = FALSE)

ggplot(ve_long, aes(x = assay, y = median_ve, fill = variable)) +
  geom_col(position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = c("stage" = "#0072B2", "lineage" = "#E69F00")) +
  labs(x = NULL, y = "median % variance explained",
       title = "Variance explained (per-window HVW)") +
  theme_ng()
```

## Silhouette scores (by stage)

```{r win_sil, fig.width = ng_fig_size(2, 1)$w, fig.height = ng_fig_size(1, 1)$h}
sil_win <- data.frame(
  assay = names(ok_embeds),
  silhouette_stage = sapply(names(ok_embeds), function(lbl)
    sil_score(ok_embeds[[lbl]]$umap,
              col_data$stage[ok_embeds[[lbl]]$kept_cols])),
  silhouette_lineage = sapply(names(ok_embeds), function(lbl)
    sil_score(ok_embeds[[lbl]]$umap,
              col_data$lineage[ok_embeds[[lbl]]$kept_cols]))
)

sil_long <- pivot_longer(sil_win, c("silhouette_stage", "silhouette_lineage"),
                         names_to = "grouping", values_to = "silhouette")
sil_long$grouping <- sub("silhouette_", "", sil_long$grouping)
sil_long$assay <- factor(sil_long$assay, levels = assay_map)

ggplot(sil_long, aes(x = assay, y = silhouette, fill = grouping)) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c("stage" = "#0072B2", "lineage" = "#E69F00")) +
  labs(x = NULL, y = "mean silhouette",
       title = "Cluster separation by assay") +
  theme_ng()
```

# Per-cell entropy summaries

```{r cell_df}
cell_df <- data.frame(
  cell_id = col_data$cell_id,
  stage = col_data$stage,
  lineage = col_data$lineage,
  mean_i_total = colMeans(assays_list$i_total, na.rm = TRUE),
  mean_meth = colMeans(assays_list$meth, na.rm = TRUE)
)

write.csv(cell_df, "argelaguet_embeddings_per_cell_summary.csv", row.names = FALSE)
```

```{r violins, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(4, 2)$h}
cell_long <- pivot_longer(cell_df, c("mean_i_total", "mean_meth"),
                          names_to = "metric", values_to = "value")

ggplot(cell_long, aes(x = stage, y = value, fill = stage)) +
  geom_violin(trim = FALSE, scale = "width", alpha = 0.7) +
  geom_boxplot(width = 0.1, outlier.size = 0.3, alpha = 0.6) +
  scale_fill_manual(values = argelaguet_stage_pal) +
  facet_grid(metric ~ ., scales = "free_y") +
  labs(x = "stage", y = "per-cell mean") +
  theme_ng() +
  theme(legend.position = "none")
```

# Per-locus (per-window) embeddings

Per-locus UMAPs are computed on the windows x cells matrix transposed to
windows x (i_total). Each point is one window; colouring uses the fractional
overlap with each annotation. Provides a complementary view to the cell-level
UMAPs above.

```{r per_locus_setup}
## Build a per-locus feature matrix: rows = windows, cols = stages
## (mean i_total across cells in the stage). Reduces noise vs raw per-cell
## values and produces a square enough matrix for PCA on common cases.
have_annotation <- length(annotation_cols) > 0L
per_locus_ok <- FALSE
if (have_annotation) {
  ## Aggregate per (window, stage) so the embedding sees the stage signal,
  ## not single-cell noise. win_cf was loaded in load_amet; reuse it after
  ## attaching stage from col_data.
  cell_stage <- as.data.frame(unique(col_data[, c("cell_id", "stage")]))
  win_cf_stage <- merge(win_cf, cell_stage, by = "cell_id", all.x = TRUE)
  win_agg <- win_cf_stage[!is.na(stage),
                          .(mean_i_total = mean(i_total, na.rm = TRUE)),
                          by = .(feature_id, stage)]
  m_long <- dcast(win_agg, feature_id ~ stage, value.var = "mean_i_total")
  feat_ids <- m_long$feature_id
  m <- as.matrix(m_long[, -1])
  rownames(m) <- feat_ids
  m[!is.finite(m)] <- NA
  keep <- rowSums(!is.na(m)) >= max(2L, ncol(m) - 1L)
  m <- m[keep, , drop = FALSE]
  if (nrow(m) >= 50L && ncol(m) >= 2L) per_locus_ok <- TRUE
  cat("Per-locus matrix:", nrow(m), "windows x", ncol(m), "stages\n")
}
```

```{r per_locus_umap}
## We want UMAP coords per window. run_embedding expects features x cells and
## treats cells (columns) as observations, so feed it stages x windows: each
## window becomes a "cell" and gets one UMAP point.
if (per_locus_ok) {
  pca_in <- t(m)
  keep_cols <- apply(pca_in, 2, function(x) sd(x, na.rm = TRUE) > 0)
  pca_in <- pca_in[, keep_cols, drop = FALSE]
  per_locus_embed <- run_embedding(pca_in,
                                   n_hvf = min(2000L, nrow(pca_in)),
                                   n_pcs = min(10L, nrow(pca_in) - 1L),
                                   n_neighbors = 15L, min_dist = 0.3,
                                   seed = 42L)
} else {
  per_locus_embed <- NULL
}

if (is.null(per_locus_embed)) {
  cat("Skipping per-locus UMAP: insufficient data.\n")
} else {
  cat("Per-locus UMAP windows kept:",
      sum(per_locus_embed$kept_cols), "\n")
}
```

```{r per_locus_umap_i_total, fig.width = ng_fig_size(1, 1, panel_mm = 60)$w, fig.height = ng_fig_size(1, 1, panel_mm = 60)$h}
if (!is.null(per_locus_embed)) {
  ## kept_cols of run_embedding indexes the columns it received (= windows in
  ## pca_in). Map back to window names and take per-window mean i_total over
  ## stages.
  kept_feats <- colnames(pca_in)[per_locus_embed$kept_cols]
  per_locus_i_total <- rowMeans(m[kept_feats, , drop = FALSE], na.rm = TRUE)
  per_locus_df <- data.frame(
    feature_id = kept_feats,
    umap1 = per_locus_embed$umap[, 1],
    umap2 = per_locus_embed$umap[, 2],
    mean_i_total = per_locus_i_total
  )
  print(ggplot(per_locus_df, aes(x = umap1, y = umap2, color = mean_i_total)) +
    geom_point(size = 0.4, alpha = 0.7) +
    scale_color_viridis_c() +
    labs(title = expression("Per-locus UMAP coloured by mean " * i[total]),
         x = "UMAP 1", y = "UMAP 2") +
    theme_ng(base_size = 8))
}
```

```{r per_locus_umap_i_norm, fig.width = ng_fig_size(1, 1, panel_mm = 60)$w, fig.height = ng_fig_size(1, 1, panel_mm = 60)$h}
## i_norm = i_total / (k_max * H(mean_meth)) per (window, cell), then take the
## per-window mean across cells for colouring. Same UMAP coordinates as i_total.
if (!is.null(per_locus_embed) && !is.null(assays_list$meth)) {
  i_total_mat <- as.matrix(assays_list$i_total)
  meth_mat <- as.matrix(assays_list$meth)
  i_norm_mat <- matrix(NA_real_, nrow(i_total_mat), ncol(i_total_mat),
                       dimnames = dimnames(i_total_mat))
  for (j in seq_len(ncol(i_total_mat))) {
    i_norm_mat[, j] <- compute_i_norm(i_total_mat[, j], meth_mat[, j],
                                      params$i_max_lag)
  }
  rows_in_sce <- intersect(kept_feats, rownames(i_norm_mat))
  per_locus_i_norm <- rep(NA_real_, length(kept_feats))
  names(per_locus_i_norm) <- kept_feats
  if (length(rows_in_sce) > 0) {
    per_locus_i_norm[rows_in_sce] <- rowMeans(
      i_norm_mat[rows_in_sce, , drop = FALSE], na.rm = TRUE)
  }
  per_locus_df_norm <- data.frame(
    feature_id = kept_feats,
    umap1 = per_locus_embed$umap[, 1],
    umap2 = per_locus_embed$umap[, 2],
    mean_i_norm = per_locus_i_norm
  )
  print(ggplot(per_locus_df_norm, aes(x = umap1, y = umap2, color = mean_i_norm)) +
    geom_point(size = 0.4, alpha = 0.7) +
    scale_color_viridis_c() +
    labs(title = expression("Per-locus UMAP coloured by mean " * i[norm]),
         x = "UMAP 1", y = "UMAP 2") +
    theme_ng(base_size = 8))
}
```

```{r per_locus_umap_jsd, fig.width = ng_fig_size(1, 1, panel_mm = 60)$w, fig.height = ng_fig_size(1, 1, panel_mm = 60)$h}
## Per-locus jsd from amet's win_feature table (per group). Aggregate to a
## per-window median across groups for colouring.
if (!is.null(per_locus_embed)) {
  feat_tbl <- fread(params$win_feature)
  jsd_by_win <- feat_tbl[, .(median_jsd = median(jsd, na.rm = TRUE)),
                         by = feature_id]
  jsd_df <- merge(per_locus_df, jsd_by_win, by = "feature_id", all.x = TRUE)
  if (any(is.finite(jsd_df$median_jsd))) {
    print(ggplot(jsd_df, aes(x = umap1, y = umap2, color = median_jsd)) +
      geom_point(size = 0.4, alpha = 0.7) +
      scale_color_viridis_c(option = "magma") +
      labs(title = "Per-locus UMAP coloured by median jsd",
           x = "UMAP 1", y = "UMAP 2") +
      theme_ng(base_size = 8))
  } else {
    cat("median_jsd is all NA; skipping jsd panel.\n")
  }
}
```

```{r per_locus_umap_annotation, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(4, 2)$h}
## Colour the per-locus UMAP by each annotation's overlap fraction.
if (!is.null(per_locus_embed) && have_annotation) {
  ann_df <- merge(per_locus_df,
                  as.data.frame(window_annotation[, c("feature_id",
                                                      annotation_cols),
                                                  with = FALSE]),
                  by = "feature_id", all.x = TRUE)
  ann_long <- pivot_longer(ann_df, all_of(annotation_cols),
                           names_to = "annotation", values_to = "frac")
  ann_long <- ann_long[!is.na(ann_long$frac), ]
  if (nrow(ann_long) > 0L) {
    print(ggplot(ann_long, aes(x = umap1, y = umap2, color = frac)) +
      geom_point(size = 0.2, alpha = 0.6) +
      facet_wrap(~ annotation, ncol = 4) +
      scale_color_viridis_c(option = "plasma", limits = c(0, 1)) +
      labs(title = "Per-locus UMAP, overlap fraction per annotation",
           x = "UMAP 1", y = "UMAP 2", color = "frac") +
      theme_ng(base_size = 6) +
      theme(strip.text = element_text(size = 5.5)))
  } else {
    cat("No non-NA annotation values to plot.\n")
  }
}
```

```{r session}
sessionInfo()
```
