---
title: "Argelaguet: single-cell genomic windows summaries"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
params:
  features_dir: ""
  win_cell_feature: ""
  win_feature: ""
  win_bed: ""
  windows_annotation: ""
  manifest: ""
  out_dir: ""
  log_path: ""
  threads: 1
  i_max_lag: 3
---

```{r logging_early, include = FALSE}
source(Sys.getenv("AMET_RENDER_HELPERS"))
param <- amet_setup_render_logging(params$log_path, params$threads)
```

```{r load_window_annotation}
## Per-window annotation matrix (windows x annotations, fractional overlap).
## Carried alongside the windows long table so downstream chunks can label or
## colour windows by genomic context.
window_annotation <- amet_load_annotation_matrix(params$windows_annotation)
if (is.null(window_annotation)) {
  message("[annotation] no per-window annotation matrix; annotation-dependent ",
          "chunks will be skipped")
  annotation_cols <- character(0)
} else {
  annotation_cols <- setdiff(names(window_annotation),
                             c("chrom", "start", "end", "feature_id"))
  cat("Annotations attached to windows:", length(annotation_cols), "\n")
  cat("Windows in annotation matrix:", nrow(window_annotation), "\n")
}
```

```{r setup}
suppressPackageStartupMessages({
  library(data.table)
  library(readr)
  library(dplyr)
  library(ggplot2)
  library(BiocParallel)
})

repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))
source(file.path(repo_root, "workflow", "scripts", "palettes.R"))


knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  cache = FALSE,
  fig.path = "argelaguet_windows_plots/"
)
```

## Cell selection upstream of this Rmd:
##   Per (stage, lineage10x) stratum, cells are picked by cpg_level tsv.gz
##   size (proxy for observed-CpG coverage) with plate-stratified
##   round-robin so a single plate cannot dominate the pick. Cap =
##   max_cells_per_combo (config/datasets.yaml: prototype.cells_per_group in
##   proto, full.max_cells_per_combo otherwise).

```{r load_meta}
man <- fread(params$manifest)
cat("Cells in manifest:", nrow(man), "\n")
cat("Stages:", paste(sort(unique(man$stage)), collapse = ", "), "\n")
cat("Lineages:", length(unique(man$lineage10x)), "\n")
```

```{r load_amet_windows}
## amet's window-scope output: two long tables. per-cell-per-window i_total
## + mean_meth, and per-window-per-group jsd (loaded but not used in this
## Rmd, kept for symmetry with the per-feature reports).
win_cf <- fread(params$win_cell_feature)
win_cf <- merge(win_cf, man[, .(cell_id, stage, lineage10x)],
                by = "cell_id", all.x = TRUE)
setnames(win_cf, "lineage10x", "lineage")

cat("Cell-window rows:", nrow(win_cf), "\n")
cat("Unique windows:", uniqueN(win_cf$feature_id), "\n")
cat("Unique cells:", uniqueN(win_cf$cell_id), "\n")
cat("Stages in data:\n")
print(table(win_cf$stage, useNA = "ifany"))
cat("Lineages in data:\n")
print(table(win_cf$lineage, useNA = "ifany"))
```

```{r filter_na}
## NA filter: drop windows missing in > 30% of cells. Applied here on the
## long table.
n_cells <- uniqueN(win_cf$cell_id)
win_cov <- win_cf[, .(cov_frac = .N / n_cells), by = feature_id]
keep_win <- win_cov[cov_frac >= 0.7, feature_id]
win_cf <- win_cf[feature_id %in% keep_win]
cat("Windows after NA filter (>= 70% cells covered):", uniqueN(win_cf$feature_id), "\n")
```

# QC

```{r cell_qc_df}
cell_df <- win_cf[, .(
  mean_i_total = mean(i_total, na.rm = TRUE),
  mean_meth = mean(mean_meth, na.rm = TRUE)
), by = .(cell_id, stage, lineage)]

write.csv(cell_df, "argelaguet_windows_per_cell_summary.csv", row.names = FALSE)
```

```{r itotal_by_stage, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(4, 1)$h}
ggplot(cell_df, aes(x = stage, y = mean_i_total, fill = stage)) +
  geom_violin(trim = FALSE, scale = "width") +
  geom_boxplot(width = 0.1, outlier.size = 0.5, alpha = 0.5) +
  theme_ng(base_size = 8) +
  theme(legend.position = "none") +
  labs(y = expression("mean " * i[total]),
       title = expression("Per-cell mean " * i[total] * " by stage"))
```

```{r meth_vs_itotal, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(4, 1)$h}
ggplot(cell_df[is.finite(mean_meth) & is.finite(mean_i_total)],
       aes(x = mean_meth, y = mean_i_total, color = stage)) +
  geom_point(size = 0.5, alpha = 0.6) +
  scale_color_manual(values = c("E4.5" = "#440154", "E5.5" = "#31688e",
                                "E6.5" = "#35b779", "E7.5" = "#fde725")) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  theme_ng(base_size = 8) +
  labs(x = "mean methylation", y = expression("mean " * i[total]),
       title = expression("Global " * i[total] * " vs methylation per cell"))
```

# Per-annotation window summaries

```{r per_window_itotal, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(2, 1)$h}
## Per-window mean i_total across all cells, joined to the per-window annotation
## matrix. For each annotation, summarise the i_total distribution restricted to
## windows whose annotation coverage exceeds 50% (binary "in / out" cut on the
## continuous overlap fraction).
if (length(annotation_cols) == 0L) {
  cat("Skipping per-annotation window summary: empty annotation matrix.\n")
} else {
  win_mean <- win_cf[, .(mean_i_total = mean(i_total, na.rm = TRUE),
                         mean_meth_win = mean(mean_meth, na.rm = TRUE),
                         n_cells_observed = .N), by = feature_id]
  win_mean <- merge(win_mean, window_annotation, by = "feature_id",
                    all.x = TRUE)

  ann_rows <- lapply(annotation_cols, function(a) {
    frac <- win_mean[[a]]
    in_ann <- !is.na(frac) & frac > 0.5
    if (sum(in_ann) < 5L) return(NULL)
    data.frame(
      annotation = a,
      n_windows = sum(in_ann),
      median_i_total = median(win_mean$mean_i_total[in_ann], na.rm = TRUE),
      iqr_i_total = IQR(win_mean$mean_i_total[in_ann], na.rm = TRUE),
      median_meth = median(win_mean$mean_meth_win[in_ann], na.rm = TRUE)
    )
  })
  ann_summary <- do.call(rbind, ann_rows)

  if (is.null(ann_summary) || nrow(ann_summary) == 0L) {
    cat("No annotation contains enough overlapping windows for a summary.\n")
  } else {
    ann_summary <- ann_summary[order(-ann_summary$median_i_total), ]
    ann_summary$annotation <- factor(ann_summary$annotation,
                                     levels = ann_summary$annotation)
    write.csv(ann_summary, "argelaguet_windows_annotation_summary.csv",
              row.names = FALSE)

    print(ggplot(ann_summary, aes(x = annotation, y = median_i_total)) +
      geom_col(fill = "#35b779") +
      coord_flip() +
      labs(x = NULL, y = expression("median " * i[total]),
           title = "Per-annotation median i_total (windows >50% covered)") +
      theme_ng(base_size = 8))
  }
}
```

```{r itotal_by_annotation_violin, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(3, 1)$h}
## Distribution of per-window mean i_total, faceted by annotation. Skipped on
## empty matrices.
if (length(annotation_cols) > 0L && exists("win_mean")) {
  long_rows <- lapply(annotation_cols, function(a) {
    frac <- win_mean[[a]]
    in_ann <- !is.na(frac) & frac > 0.5 & is.finite(win_mean$mean_i_total)
    if (sum(in_ann) < 5L) return(NULL)
    data.frame(annotation = a,
               i_total = win_mean$mean_i_total[in_ann])
  })
  long_df <- do.call(rbind, long_rows)
  if (!is.null(long_df) && nrow(long_df) > 0L) {
    print(ggplot(long_df, aes(x = annotation, y = i_total)) +
      geom_violin(fill = "#bbbbbb", scale = "width", trim = FALSE) +
      geom_boxplot(width = 0.1, outlier.size = 0.3, fill = "white") +
      coord_flip() +
      labs(x = NULL, y = expression("per-window mean " * i[total])) +
      theme_ng(base_size = 8))
  } else {
    cat("Annotation summary empty after >50% cut.\n")
  }
} else {
  cat("Skipping per-annotation violins.\n")
}
```

```{r session}
sessionInfo()
```
