---
title: "Argelaguet: cell embeddings from genomic windows entropy"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
params:
  features_dir: ""
  win_cell_feature: ""
  win_feature: ""
  win_bed: ""
  manifest: ""
  out_dir: ""
---

```{r setup}
suppressPackageStartupMessages({
  library(data.table)
  library(readr)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(BiocParallel)
  library(uwot)
  library(irlba)
  library(matrixStats)

  library(cluster)
  library(viridis)
  library(cowplot)
})

repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))
source(file.path(repo_root, "workflow", "scripts", "palettes.R"))
source(file.path(repo_root, "workflow", "scripts", "embedding_utils.R"))

param <- SerialParam()

knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  cache = FALSE,
  fig.path = "argelaguet_embeddings_plots/"
)
```

```{r constants}
N_HVW <- 1000
N_PCS <- 10

## Only the methylation-stable amet score (i_total) and methylation are
## carried forward. Raw S and yamet's adjS variant are dropped because amet
## does not export them.
assay_map <- c(
  "i_total" = "i_total",
  "meth"    = "methylation"
)
```

## Cell selection and NA-aware embedding

The cells embedded here are not a uniformly random subsample. For each
(stage, lineage) annotation we keep up to a fixed number of cells, picking the
highest-coverage cells (file size on disk is monotonic in CpG count) and
distributing the slots round-robin across plates. Stratifying by plate first
keeps a single high-coverage plate from dominating an annotation. The visible
cost is that any residual plate effect (cell cycle, batch) is preserved rather
than averaged out across plates; the alternative of random sampling produced
embeddings collapsed to a handful of cells per group because of how few windows
were complete.

scBS-seq methylation and per-cell entropy values have a meaningful zero
(unmethylated CpG, zero local entropy) which is distinct from missing coverage.
Imputing NAs to zero would bias both the variance signal and the embedding
geometry, so the pipeline keeps NAs as NAs throughout: HVF selection scores
features by variance times coverage fraction, and when too few windows have
complete coverage the PCA step uses NIPALS (`pcaMethods`) which iterates over
the observed entries without filling in missing values. Cells are dropped only
when their per-cell NA fraction is above the QC threshold; once they pass that
filter they are never dropped to make a feature block dense.

# Load data

```{r load_amet}
## amet ships per-cell-per-window i_total and per-cell mean_meth.
## Build cells x windows matrices, one per assay, in place of yamet's SCE.
win_cf <- fread(params$win_cell_feature)
man <- fread(params$manifest)

col_data <- as.data.frame(unique(win_cf[, .(cell_id)]))
col_data <- merge(col_data, as.data.frame(man),
                  by = "cell_id", all.x = TRUE)
col_data$stage   <- factor(col_data$stage)
col_data$lineage <- factor(col_data$lineage10x)
rownames(col_data) <- col_data$cell_id

build_assay_mat <- function(value_col) {
  m <- dcast(win_cf, feature_id ~ cell_id, value.var = value_col,
             fun.aggregate = mean)
  rn <- m$feature_id
  m <- as.matrix(m[, -1])
  rownames(m) <- rn
  m[!is.finite(m)] <- NA
  m[, col_data$cell_id, drop = FALSE]
}

assays_list <- list(
  i_total = build_assay_mat("i_total"),
  meth    = build_assay_mat("mean_meth")
)

cat("Cells:", nrow(col_data), "\n")
cat("Windows:", nrow(assays_list$i_total), "\n")
cat("Assays:", paste(names(assays_list), collapse = ", "), "\n")
cat("Stages:\n")
print(table(col_data$stage, useNA = "ifany"))
cat("Top lineages:\n")
print(sort(table(col_data$lineage, useNA = "ifany"), decreasing = TRUE))
```

# Shared helpers

```{r helpers}
hvw_pca_umap <- function(mat) {
  run_embedding(mat, n_hvf = N_HVW, n_pcs = N_PCS,
                n_neighbors = 15L, min_dist = 0.3, seed = 42L)
}

plot_umap_stage <- function(coords, kept_cols, title_prefix) {
  df <- data.frame(
    umap1 = coords[, 1], umap2 = coords[, 2],
    stage = col_data$stage[kept_cols]
  )
  ggplot(df, aes(x = umap1, y = umap2, color = stage)) +
    geom_point(size = 0.6, alpha = 0.7) +
    scale_color_manual(values = argelaguet_stage_pal) +
    labs(title = paste(title_prefix, "- stage"), x = "UMAP 1", y = "UMAP 2") +
    guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
    theme_ng()
}

plot_umap_lineage <- function(coords, kept_cols, title_prefix) {
  df <- data.frame(
    umap1   = coords[, 1], umap2 = coords[, 2],
    lineage = col_data$lineage[kept_cols]
  )
  ggplot(df, aes(x = umap1, y = umap2, color = lineage)) +
    geom_point(size = 0.6, alpha = 0.7) +
    scale_color_manual(values = argelaguet_lineage_pal) +
    labs(title = paste(title_prefix, "- lineage"), x = "UMAP 1", y = "UMAP 2") +
    guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1),
                                ncol = 2)) +
    theme_ng()
}
```

# Per-window embeddings (HVW -> PCA -> UMAP)

HVW selection and NA handling: see `embedding_utils.R`.
Top `r N_HVW` most variable windows retained per assay before PCA and UMAP.

```{r win_hvw_pca_umap}
win_embeds <- lapply(names(assay_map), function(akey) {
  if (!akey %in% names(assays_list)) return(NULL)
  hvw_pca_umap(assays_list[[akey]])
})
names(win_embeds) <- assay_map

failed <- names(which(sapply(win_embeds, is.null)))
if (length(failed) > 0)
  warning("run_embedding returned NULL for: ", paste(failed, collapse = ", "))

cat("\nHVF selection summary:\n")
for (nm in names(win_embeds)) {
  em <- win_embeds[[nm]]
  if (is.null(em)) {
    cat(nm, ": NULL\n")
  } else {
    cat(nm, ": HVF =", length(em$hvf_idx),
        "| cells =", sum(em$kept_cols), "/", nrow(col_data), "\n")
  }
}

saveRDS(
  list(col_data = col_data,
       n_windows = nrow(assays_list$i_total),
       n_cells = nrow(col_data),
       assay_names = names(assays_list),
       win_embeds = win_embeds,
       assay_map = assay_map),
  "argelaguet_umap_windows_i_total.rds", compress = "xz"
)
```

## By stage {.tabset .tabset-pills}

<!-- Section dropped: amet has no raw S, so the win_stage_S chunk (yamet/argelaguet_embeddings.Rmd lines 184-190) is gone. -->

```{r win_stage_i_total, fig.width = ng_fig_size(1, 1, panel_mm = 60)$w, fig.height = ng_fig_size(1, 1, panel_mm = 60)$h}
### i_total
if (!is.null(win_embeds[["i_total"]])) {
  print(plot_umap_stage(win_embeds[["i_total"]]$umap,
                        win_embeds[["i_total"]]$kept_cols,
                        "i_total (windows)"))
}
```

```{r win_stage_meth, fig.width = ng_fig_size(1, 1, panel_mm = 60)$w, fig.height = ng_fig_size(1, 1, panel_mm = 60)$h}
### methylation
if (!is.null(win_embeds[["methylation"]])) {
  print(plot_umap_stage(win_embeds[["methylation"]]$umap,
                        win_embeds[["methylation"]]$kept_cols,
                        "methylation (windows)"))
}
```

## By lineage {.tabset .tabset-pills}

<!-- Section dropped: amet has no raw S, so win_lineage_S (yamet/argelaguet_embeddings.Rmd lines 211-217) is gone. -->

```{r win_lineage_i_total, fig.width = ng_fig_size(1, 1, panel_mm = 60, legend_mm = 55)$w, fig.height = ng_fig_size(1, 1, panel_mm = 60, legend_mm = 55)$h}
### i_total
if (!is.null(win_embeds[["i_total"]])) {
  print(plot_umap_lineage(win_embeds[["i_total"]]$umap,
                          win_embeds[["i_total"]]$kept_cols,
                          "i_total (windows)"))
}
```

```{r win_lineage_meth, fig.width = ng_fig_size(1, 1, panel_mm = 60, legend_mm = 55)$w, fig.height = ng_fig_size(1, 1, panel_mm = 60, legend_mm = 55)$h}
### methylation
if (!is.null(win_embeds[["methylation"]])) {
  print(plot_umap_lineage(win_embeds[["methylation"]]$umap,
                          win_embeds[["methylation"]]$kept_cols,
                          "methylation (windows)"))
}
```

## Two-panel comparison (stage)

```{r win_panel_stage, fig.width = ng_fig_size(2, 1, panel_mm = 55)$w, fig.height = ng_fig_size(2, 1, panel_mm = 55)$h}
ok_embeds <- Filter(Negate(is.null), win_embeds)
df_panel <- bind_rows(lapply(names(ok_embeds), function(lbl) {
  em <- ok_embeds[[lbl]]
  data.frame(umap1 = em$umap[, 1], umap2 = em$umap[, 2],
             stage = col_data$stage[em$kept_cols],
             lineage = col_data$lineage[em$kept_cols],
             assay = lbl)
}))
df_panel$assay <- factor(df_panel$assay, levels = assay_map)

ggplot(df_panel, aes(x = umap1, y = umap2, color = stage)) +
  geom_point(size = 0.4, alpha = 0.6) +
  scale_color_manual(values = argelaguet_stage_pal) +
  facet_wrap(~ assay, nrow = 1) +
  labs(x = "UMAP 1", y = "UMAP 2",
       title = "Per-window embeddings (HVW -> PCA -> UMAP)") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

## Two-panel comparison (lineage)

```{r win_panel_lineage, fig.width = ng_fig_size(2, 1, panel_mm = 55, legend_mm = 55)$w, fig.height = ng_fig_size(2, 1, panel_mm = 55, legend_mm = 55)$h}
ggplot(df_panel, aes(x = umap1, y = umap2, color = lineage)) +
  geom_point(size = 0.4, alpha = 0.6) +
  scale_color_manual(values = argelaguet_lineage_pal) +
  facet_wrap(~ assay, nrow = 1) +
  labs(x = "UMAP 1", y = "UMAP 2",
       title = "Per-window embeddings (HVW -> PCA -> UMAP)") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1),
                              ncol = 2)) +
  theme_ng()
```

## Variance explained

```{r win_varexp, fig.width = ng_fig_size(2, 1)$w, fig.height = ng_fig_size(2, 1)$h}
ve_win <- bind_rows(lapply(names(assay_map), function(akey) {
  lbl <- assay_map[[akey]]
  if (is.null(win_embeds[[lbl]]) || !akey %in% names(assays_list)) return(NULL)
  kept <- win_embeds[[lbl]]$kept_cols
  hvw_mat <- assays_list[[akey]][win_embeds[[lbl]]$hvf_idx, kept, drop = FALSE]
  data.frame(
    assay   = lbl,
    stage   = median(row_variance_explained(hvw_mat, col_data$stage[kept]),   na.rm = TRUE),
    lineage = median(row_variance_explained(hvw_mat, col_data$lineage[kept]), na.rm = TRUE))
}))

ve_long <- pivot_longer(ve_win, c("stage", "lineage"),
                        names_to = "variable", values_to = "median_ve")
ve_long$assay <- factor(ve_long$assay, levels = assay_map)

write.csv(ve_long, "argelaguet_win_varexp.csv", row.names = FALSE)

ggplot(ve_long, aes(x = assay, y = median_ve, fill = variable)) +
  geom_col(position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = c("stage" = "#0072B2", "lineage" = "#E69F00")) +
  labs(x = NULL, y = "median % variance explained",
       title = "Variance explained (per-window HVW)") +
  theme_ng()
```

## Silhouette scores (by stage)

```{r win_sil, fig.width = ng_fig_size(2, 1)$w, fig.height = ng_fig_size(1, 1)$h}
sil_win <- data.frame(
  assay = names(ok_embeds),
  silhouette_stage = sapply(names(ok_embeds), function(lbl)
    sil_score(ok_embeds[[lbl]]$umap,
              col_data$stage[ok_embeds[[lbl]]$kept_cols])),
  silhouette_lineage = sapply(names(ok_embeds), function(lbl)
    sil_score(ok_embeds[[lbl]]$umap,
              col_data$lineage[ok_embeds[[lbl]]$kept_cols]))
)

sil_long <- pivot_longer(sil_win, c("silhouette_stage", "silhouette_lineage"),
                         names_to = "grouping", values_to = "silhouette")
sil_long$grouping <- sub("silhouette_", "", sil_long$grouping)
sil_long$assay    <- factor(sil_long$assay, levels = assay_map)

ggplot(sil_long, aes(x = assay, y = silhouette, fill = grouping)) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c("stage" = "#0072B2", "lineage" = "#E69F00")) +
  labs(x = NULL, y = "mean silhouette",
       title = "Cluster separation by assay") +
  theme_ng()
```

# Per-cell entropy summaries

```{r cell_df}
cell_df <- data.frame(
  cell_id   = col_data$cell_id,
  stage     = col_data$stage,
  lineage   = col_data$lineage,
  mean_i_total = colMeans(assays_list$i_total, na.rm = TRUE),
  mean_meth = colMeans(assays_list$meth,    na.rm = TRUE)
)

write.csv(cell_df, "argelaguet_per_cell_summary.csv", row.names = FALSE)
```

```{r violins, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(4, 2)$h}
cell_long <- pivot_longer(cell_df, c("mean_i_total", "mean_meth"),
                          names_to = "metric", values_to = "value")

ggplot(cell_long, aes(x = stage, y = value, fill = stage)) +
  geom_violin(trim = FALSE, scale = "width", alpha = 0.7) +
  geom_boxplot(width = 0.1, outlier.size = 0.3, alpha = 0.6) +
  scale_fill_manual(values = argelaguet_stage_pal) +
  facet_grid(metric ~ ., scales = "free_y") +
  labs(x = "stage", y = "per-cell mean") +
  theme_ng() +
  theme(legend.position = "none")
```

```{r session}
sessionInfo()
```
