---
title: "amet on Argelaguet 2019 mouse gastrulation scNMT-seq"
author: "Izaskun Mallona"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    number_sections: true
    df_print: default
params:
  feat_cell_feature: ""
  feat_feature: ""
  feat_bed: ""
  win_cell_feature: ""
  win_feature: ""
  win_bed: ""
  manifest: ""
  out_dir: ""
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 5, fig.height = 4)

suppressPackageStartupMessages({
    library(data.table)
    library(ggplot2)
    library(ggridges)
    library(ggrepel)
    library(viridis)
    library(patchwork)
    library(dplyr)
    library(tidyr)
    library(uwot)
})

repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))

panels_dir <- file.path(params$out_dir, "panels")
feat_dir <- file.path(panels_dir, "feature")
win_dir <- file.path(panels_dir, "window")
dir.create(feat_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(win_dir, showWarnings = FALSE, recursive = TRUE)

recover_annotation <- function(fid) sub("_\\d+$", "", fid)
```

# Inputs and design

Argelaguet et al. 2019 mouse gastrulation scNMT-seq, prototype run on chr19 with four lineages (Epiblast, Primitive_endoderm, Nascent_mesoderm, Rostral_neurectoderm). Per-cell raw files are scNMT cpg_level tsv.gz read by amet's scnmt parser.

Two analyses with disjoint feature definitions:

- **Feature-based**: amet outputs over gastrulation-specific BEDs (H3K27ac distal peaks for E7.5 ectoderm, endoderm, mesoderm; ESC p300 binding sites). Each feature carries an `<annotation>_<index>` id; the annotation prefix is recovered for stratified plots. This section asks where lineage-specific regulatory regions show distinct entropy.
- **Window-based**: amet outputs over fixed-size genome tiles on chr19. One feature = one window. Used for cell-level summaries, embeddings, and stage-trajectory views without locus bias.

Score mapping: amet `i_total` for adjS-like questions; amet `jsd` for adjH-like questions.

```{r load}
man <- fread(params$manifest)
feat_cf <- fread(params$feat_cell_feature)
feat_fe <- fread(params$feat_feature)
win_cf <- fread(params$win_cell_feature)
win_fe <- fread(params$win_feature)
feat_bed <- fread(params$feat_bed, header = FALSE,
                  col.names = c("chrom", "start", "end", "feature_id"))
win_bed <- fread(params$win_bed, header = FALSE,
                 col.names = c("chrom", "start", "end", "feature_id"))

extra_cols <- intersect(c("stage", "embryo"), colnames(man))
keep <- c("cell_id", "group", extra_cols)
man <- man[, ..keep]

feat_bed[, annotation := recover_annotation(feature_id)]
feat_cf[, annotation := recover_annotation(feature_id)]
feat_fe[, annotation := recover_annotation(feature_id)]

feat_cf <- merge(feat_cf, man[, .(cell_id, stage = if ("stage" %in% colnames(man)) stage else NA)],
                 by = "cell_id", all.x = TRUE)
win_cf <- merge(win_cf, man[, .(cell_id, stage = if ("stage" %in% colnames(man)) stage else NA)],
                by = "cell_id", all.x = TRUE)

group_levels <- sort(unique(man$group))
ng <- length(group_levels)
lineage_pal <- setNames(viridis::viridis(ng, end = 0.92), group_levels)

ann_levels <- sort(unique(feat_cf$annotation))
```

# Sample sizes and coverage

```{r sizes, fig.width = 6, fig.height = 4}
sz <- man[, .(cells = .N), by = group][order(group)]
print(sz)
p_n <- ggplot(sz, aes(x = group, y = cells, fill = group)) +
    geom_col(width = 0.7) +
    scale_fill_manual(values = lineage_pal) +
    labs(x = NULL, y = "cells") +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")
save_eval(p_n, sz, file.path(panels_dir, "ncells_per_group"),
          width_mm = 90, height_mm = 60)
p_n
```

```{r sizes_stage, fig.width = 7, fig.height = 4}
if ("stage" %in% colnames(man)) {
    ss <- man[!is.na(stage), .N, by = .(group, stage)]
    p_ss <- ggplot(ss, aes(x = stage, y = N, fill = group)) +
        geom_col(position = position_dodge(width = 0.85)) +
        scale_fill_manual(values = lineage_pal) +
        labs(x = NULL, y = "cells", fill = "lineage") +
        theme_ng() +
        theme(axis.text.x = element_text(angle = 30, hjust = 1))
    save_eval(p_ss, ss, file.path(panels_dir, "ncells_per_stage_lineage"),
              width_mm = 130, height_mm = 70)
    p_ss
}
```

```{r coverage, fig.width = 7, fig.height = 4}
cov_feat <- feat_cf[, .(n_features = .N), by = .(cell_id, group)]
cov_win <- win_cf[, .(n_windows = .N), by = .(cell_id, group)]
cov_long <- rbind(
    cov_feat[, .(cell_id, group, n = n_features, scope = "feature (regulatory)")],
    cov_win[,  .(cell_id, group, n = n_windows,  scope = "window")]
)
p_cov <- ggplot(cov_long, aes(x = group, y = n, fill = group)) +
    geom_violin(alpha = 0.6, scale = "width") +
    geom_boxplot(width = 0.18, outlier.size = 0.3, fill = "white") +
    scale_fill_manual(values = lineage_pal) +
    facet_wrap(~ scope, scales = "free_y") +
    labs(x = NULL, y = "covered features per cell") +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")
save_eval(p_cov, cov_long, file.path(panels_dir, "coverage_per_cell"),
          width_mm = 130, height_mm = 70)
p_cov
```

# Feature-based analysis: regulatory regions

```{r feat_filter}
fcf <- feat_cf[!is.na(i_total)]
ffe <- feat_fe[!is.na(jsd)]
ffe[, group := factor(group, levels = group_levels)]
```

## i_total per (lineage, annotation)

```{r feat_itotal_facet, fig.width = 9, fig.height = 4}
p_itotal_facet <- ggplot(fcf, aes(x = group, y = i_total, fill = group)) +
    geom_violin(alpha = 0.55, scale = "width", linewidth = 0.3) +
    geom_boxplot(width = 0.18, outlier.size = 0.3, fill = "white",
                 linewidth = 0.3) +
    scale_fill_manual(values = lineage_pal) +
    facet_wrap(~ annotation, nrow = 1) +
    labs(x = NULL, y = expression(i[total])) +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none")
save_eval(p_itotal_facet,
          fcf[, .(cell_id, group, annotation, feature_id, i_total)],
          file.path(feat_dir, "itotal_by_group_annotation"),
          width_mm = 200, height_mm = 70)
p_itotal_facet
```

## i_total density ridges per lineage and annotation

```{r feat_itotal_ridge, fig.width = 9, fig.height = 4}
p_ridge_ann <- ggplot(fcf, aes(x = i_total, y = group, fill = group)) +
    geom_density_ridges(alpha = 0.7, scale = 1.2, linewidth = 0.3) +
    scale_fill_manual(values = lineage_pal) +
    facet_wrap(~ annotation, nrow = 1) +
    labs(x = expression(i[total]), y = NULL) +
    theme_ng() +
    theme(legend.position = "none", aspect.ratio = NULL)
save_eval(p_ridge_ann, fcf[, .(group, annotation, i_total)],
          file.path(feat_dir, "itotal_ridge_group_annotation"),
          width_mm = 200, height_mm = 70)
p_ridge_ann
```

## i_total per cell, per (lineage, annotation)

```{r feat_per_cell, fig.width = 9, fig.height = 4}
cell_feat <- fcf[, .(mean_i_total = mean(i_total, na.rm = TRUE),
                     mean_meth = mean(mean_meth, na.rm = TRUE),
                     n_features = .N),
                 by = .(cell_id, group, annotation)]
if ("stage" %in% colnames(man))
    cell_feat <- merge(cell_feat, man[, .(cell_id, stage)], by = "cell_id")

p_per_cell_ann <- ggplot(cell_feat,
                         aes(x = group, y = mean_i_total, fill = group)) +
    geom_boxplot(outlier.size = 0.3, linewidth = 0.3, alpha = 0.75) +
    geom_jitter(width = 0.18, size = 0.5, alpha = 0.7) +
    scale_fill_manual(values = lineage_pal) +
    facet_wrap(~ annotation, nrow = 1) +
    labs(x = NULL, y = expression("mean " * i[total] * " per cell")) +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none")
save_eval(p_per_cell_ann, cell_feat,
          file.path(feat_dir, "itotal_per_cell_annotation"),
          width_mm = 200, height_mm = 70)
p_per_cell_ann
```

## Stage trajectory of i_total per (lineage, annotation)

```{r feat_stage, fig.width = 9, fig.height = 4}
if ("stage" %in% colnames(cell_feat) && any(!is.na(cell_feat$stage))) {
    p_stage_ann <- ggplot(cell_feat[!is.na(stage)],
                          aes(x = stage, y = mean_i_total, fill = group)) +
        geom_boxplot(outlier.size = 0.3, linewidth = 0.3, alpha = 0.6,
                     position = position_dodge(width = 0.85)) +
        scale_fill_manual(values = lineage_pal) +
        facet_wrap(~ annotation, nrow = 1) +
        labs(x = NULL, y = expression("mean " * i[total] * " per cell"),
             fill = "lineage") +
        theme_ng() +
        theme(axis.text.x = element_text(angle = 30, hjust = 1))
    save_eval(p_stage_ann, cell_feat[!is.na(stage)],
              file.path(feat_dir, "itotal_per_cell_stage_annotation"),
              width_mm = 220, height_mm = 70)
    p_stage_ann
}
```

## Methylation per (lineage, annotation)

```{r feat_meth, fig.width = 9, fig.height = 4}
p_meth_ann <- ggplot(fcf, aes(x = group, y = mean_meth, fill = group)) +
    geom_boxplot(outlier.size = 0.3, linewidth = 0.3, alpha = 0.7) +
    scale_fill_manual(values = lineage_pal) +
    facet_wrap(~ annotation, nrow = 1) +
    labs(x = NULL, y = "mean methylation (per cell, per region)") +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none")
save_eval(p_meth_ann, fcf[, .(cell_id, group, annotation, mean_meth)],
          file.path(feat_dir, "meanmeth_by_group_annotation"),
          width_mm = 200, height_mm = 70)
p_meth_ann
```

## Decoupling: i_total vs methylation

```{r feat_decouple, fig.width = 9, fig.height = 4}
p_dec_ann <- ggplot(fcf, aes(x = mean_meth, y = i_total, colour = group)) +
    geom_point(size = 0.4, alpha = 0.5) +
    geom_smooth(method = "loess", se = FALSE, linewidth = 0.4) +
    scale_colour_manual(values = lineage_pal) +
    facet_wrap(~ annotation, nrow = 1) +
    labs(x = "mean methylation", y = expression(i[total])) +
    theme_ng()
save_eval(p_dec_ann,
          fcf[, .(cell_id, group, annotation, mean_meth, i_total)],
          file.path(feat_dir, "itotal_vs_meth_by_annotation"),
          width_mm = 200, height_mm = 70)
p_dec_ann
```

## jsd per (lineage, annotation)

```{r feat_jsd, fig.width = 9, fig.height = 4}
if (nrow(ffe)) {
    p_jsd_ann <- ggplot(ffe, aes(x = group, y = jsd, fill = group)) +
        geom_violin(alpha = 0.55, scale = "width", linewidth = 0.3) +
        geom_boxplot(width = 0.18, outlier.size = 0.3, fill = "white",
                     linewidth = 0.3) +
        scale_fill_manual(values = lineage_pal) +
        facet_wrap(~ annotation, nrow = 1) +
        labs(x = NULL, y = "jsd") +
        theme_ng() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1),
              legend.position = "none")
    save_eval(p_jsd_ann, ffe[, .(feature_id, group, annotation, n_cells, jsd)],
              file.path(feat_dir, "jsd_by_group_annotation"),
              width_mm = 200, height_mm = 70)
    p_jsd_ann
}
```

## jsd density ridges per (lineage, annotation)

```{r feat_jsd_ridge, fig.width = 9, fig.height = 4}
if (nrow(ffe)) {
    p_jsd_ridge_ann <- ggplot(ffe, aes(x = jsd, y = group, fill = group)) +
        geom_density_ridges(alpha = 0.7, scale = 1.1, linewidth = 0.3) +
        scale_fill_manual(values = lineage_pal) +
        facet_wrap(~ annotation, nrow = 1) +
        labs(x = "jsd", y = NULL) +
        theme_ng() +
        theme(legend.position = "none", aspect.ratio = NULL)
    save_eval(p_jsd_ridge_ann, ffe[, .(group, annotation, jsd)],
              file.path(feat_dir, "jsd_ridge_group_annotation"),
              width_mm = 200, height_mm = 70)
    p_jsd_ridge_ann
}
```

## Joint structure: i_total vs jsd per (feature, lineage)

```{r feat_joint, fig.width = 9, fig.height = 4}
fe_joint <- fcf[, .(median_i_total = median(i_total, na.rm = TRUE)),
                by = .(feature_id, annotation, group)]
fe_joint <- merge(fe_joint, ffe[, .(feature_id, group, jsd, n_cells)],
                  by = c("feature_id", "group"))
fe_joint <- fe_joint[!is.na(jsd) & !is.na(median_i_total)]

p_joint_ann <- ggplot(fe_joint,
                      aes(x = jsd, y = median_i_total, colour = group)) +
    geom_point(size = 0.7, alpha = 0.85) +
    scale_colour_manual(values = lineage_pal) +
    facet_wrap(~ annotation, nrow = 1) +
    labs(x = "jsd (per region, per lineage)",
         y = expression("median " * i[total]),
         colour = "lineage") +
    theme_ng()
save_eval(p_joint_ann, fe_joint,
          file.path(feat_dir, "itotal_vs_jsd_by_annotation"),
          width_mm = 200, height_mm = 70)
p_joint_ann
```

## Summary heatmap: median i_total per (annotation, lineage)

```{r feat_summary_hm, fig.width = 6, fig.height = 4}
sum_dt <- fcf[, .(median_i_total = median(i_total, na.rm = TRUE)),
              by = .(annotation, group)]
p_sum_hm <- ggplot(sum_dt, aes(x = group, y = annotation, fill = median_i_total)) +
    geom_tile() +
    scale_fill_viridis(option = "magma") +
    labs(x = NULL, y = NULL, fill = expression("median " * i[total])) +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          aspect.ratio = NULL)
save_eval(p_sum_hm, sum_dt, file.path(feat_dir, "summary_heatmap_itotal"),
          width_mm = 110, height_mm = 70)
p_sum_hm
```

## Summary heatmap: median jsd per (annotation, lineage)

```{r feat_summary_jsd, fig.width = 6, fig.height = 4}
if (nrow(ffe)) {
    jsd_sum <- ffe[, .(median_jsd = median(jsd, na.rm = TRUE)),
                   by = .(annotation, group)]
    p_jsd_sum <- ggplot(jsd_sum,
                        aes(x = group, y = annotation, fill = median_jsd)) +
        geom_tile() +
        scale_fill_viridis(option = "viridis") +
        labs(x = NULL, y = NULL, fill = "median jsd") +
        theme_ng() +
        theme(axis.text.x = element_text(angle = 30, hjust = 1),
              aspect.ratio = NULL)
    save_eval(p_jsd_sum, jsd_sum,
              file.path(feat_dir, "summary_heatmap_jsd"),
              width_mm = 110, height_mm = 70)
    p_jsd_sum
}
```

# Window-based analysis: chr19 genome tiles

```{r win_filter}
wcf <- win_cf[!is.na(i_total)]
wfe <- win_fe[!is.na(jsd)]
wfe[, group := factor(group, levels = group_levels)]
```

## Per-cell coverage

```{r win_coverage, fig.width = 6, fig.height = 4}
p_win_cov <- ggplot(win_cf[, .(n_windows = .N), by = .(cell_id, group)],
                    aes(x = group, y = n_windows, fill = group)) +
    geom_violin(alpha = 0.55, scale = "width", linewidth = 0.3) +
    scale_fill_manual(values = lineage_pal) +
    labs(x = NULL, y = "covered windows per cell") +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")
save_eval(p_win_cov, win_cf[, .(cell_id, group)],
          file.path(win_dir, "coverage_windows_per_cell"),
          width_mm = 110, height_mm = 70)
p_win_cov
```

## Per-cell mean methylation

```{r win_meanmeth, fig.width = 6, fig.height = 4}
cell_meth <- win_cf[!is.na(mean_meth),
                    .(mean_meth = mean(mean_meth, na.rm = TRUE),
                      n_windows = .N), by = .(cell_id, group)]
if ("stage" %in% colnames(man))
    cell_meth <- merge(cell_meth, man[, .(cell_id, stage)], by = "cell_id")

p_wmeth <- ggplot(cell_meth, aes(x = group, y = mean_meth, fill = group)) +
    geom_violin(alpha = 0.55, scale = "width", linewidth = 0.3) +
    geom_boxplot(width = 0.18, outlier.size = 0.3, fill = "white",
                 linewidth = 0.3) +
    scale_fill_manual(values = lineage_pal) +
    labs(x = NULL, y = "mean methylation per cell (windows)") +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")
save_eval(p_wmeth, cell_meth, file.path(win_dir, "meanmeth_per_cell"),
          width_mm = 110, height_mm = 70)
p_wmeth
```

## Per-cell mean i_total

```{r win_itotal, fig.width = 6, fig.height = 4}
cell_it <- wcf[, .(mean_i_total = mean(i_total, na.rm = TRUE),
                   n_windows = .N), by = .(cell_id, group)]
if ("stage" %in% colnames(man))
    cell_it <- merge(cell_it, man[, .(cell_id, stage)], by = "cell_id")

p_wit <- ggplot(cell_it, aes(x = group, y = mean_i_total, fill = group)) +
    geom_violin(alpha = 0.55, scale = "width", linewidth = 0.3) +
    geom_boxplot(width = 0.18, outlier.size = 0.3, fill = "white",
                 linewidth = 0.3) +
    scale_fill_manual(values = lineage_pal) +
    labs(x = NULL, y = expression("mean " * i[total] * " per cell")) +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")
save_eval(p_wit, cell_it, file.path(win_dir, "mean_itotal_per_cell"),
          width_mm = 110, height_mm = 70)
p_wit
```

## Methylation vs i_total per cell

```{r win_corr, fig.width = 5, fig.height = 4}
cell_full <- merge(cell_meth, cell_it[, .(cell_id, mean_i_total)], by = "cell_id")
p_wcorr <- ggplot(cell_full,
                  aes(x = mean_meth, y = mean_i_total, colour = group)) +
    geom_point(size = 1.6, alpha = 0.9) +
    scale_colour_manual(values = lineage_pal) +
    labs(x = "mean methylation per cell",
         y = expression("mean " * i[total] * " per cell"),
         colour = "lineage") +
    theme_ng()
save_eval(p_wcorr, cell_full, file.path(win_dir, "itotal_vs_meth_per_cell"))
p_wcorr
```

## Stage trajectory at the cell level

```{r win_stage, fig.width = 7, fig.height = 4}
if ("stage" %in% colnames(cell_it) && any(!is.na(cell_it$stage))) {
    p_stage_win <- ggplot(cell_it[!is.na(stage)],
                          aes(x = stage, y = mean_i_total, fill = group)) +
        geom_boxplot(outlier.size = 0.3, linewidth = 0.3, alpha = 0.65,
                     position = position_dodge(width = 0.85)) +
        scale_fill_manual(values = lineage_pal) +
        labs(x = NULL, y = expression("mean " * i[total] * " per cell"),
             fill = "lineage") +
        theme_ng() +
        theme(axis.text.x = element_text(angle = 30, hjust = 1))
    save_eval(p_stage_win, cell_it[!is.na(stage)],
              file.path(win_dir, "itotal_per_cell_stage"),
              width_mm = 140, height_mm = 70)
    p_stage_win
}
```

## Window-level i_total ridges per lineage

```{r win_ridges, fig.width = 6, fig.height = 4}
p_w_it_r <- ggplot(wcf, aes(x = i_total, y = group, fill = group)) +
    geom_density_ridges(alpha = 0.7, scale = 1.2, linewidth = 0.3) +
    scale_fill_manual(values = lineage_pal) +
    labs(x = expression(i[total] * " per (cell, window)"), y = NULL) +
    theme_ng() +
    theme(legend.position = "none", aspect.ratio = NULL)
save_eval(p_w_it_r, wcf[, .(group, i_total)],
          file.path(win_dir, "itotal_ridge_group"),
          width_mm = 110, height_mm = 70)
p_w_it_r
```

## Window-level jsd ridges per lineage

```{r win_jsd_ridges, fig.width = 6, fig.height = 4}
if (nrow(wfe)) {
    p_w_jsd_r <- ggplot(wfe, aes(x = jsd, y = group, fill = group)) +
        geom_density_ridges(alpha = 0.7, scale = 1.1, linewidth = 0.3) +
        scale_fill_manual(values = lineage_pal) +
        labs(x = "jsd per (window, lineage)", y = NULL) +
        theme_ng() +
        theme(legend.position = "none", aspect.ratio = NULL)
    save_eval(p_w_jsd_r, wfe[, .(group, jsd)],
              file.path(win_dir, "jsd_ridge_group"),
              width_mm = 110, height_mm = 70)
    p_w_jsd_r
}
```

## Cell embedding: PCA on cell-by-window i_total

```{r win_build_mat}
mat <- dcast(wcf, cell_id ~ feature_id, value.var = "i_total",
             fun.aggregate = mean)
mat_cells <- mat$cell_id
mat <- as.matrix(mat[, -1])
rownames(mat) <- mat_cells
mat <- mat[, colSums(!is.na(mat)) >= 0.5 * nrow(mat), drop = FALSE]
mat[is.na(mat)] <- 0
cat("embedding matrix:", nrow(mat), "cells x", ncol(mat), "windows\n")
```

```{r win_pca, fig.width = 9, fig.height = 4}
emb <- NULL
if (ncol(mat) >= 2 && nrow(mat) >= 3) {
    pcs <- prcomp(mat, scale. = FALSE)
    emb <- data.table(cell_id = rownames(mat),
                      PC1 = pcs$x[, 1], PC2 = pcs$x[, 2],
                      PC3 = if (ncol(pcs$x) >= 3) pcs$x[, 3] else NA_real_)
    emb <- merge(emb, man[, .(cell_id, group)], by = "cell_id")
    if ("stage" %in% colnames(man))
        emb <- merge(emb, man[, .(cell_id, stage)], by = "cell_id")

    p_pca_lin <- ggplot(emb, aes(x = PC1, y = PC2, colour = group)) +
        geom_point(size = 1.8, alpha = 0.95) +
        scale_colour_manual(values = lineage_pal) +
        labs(title = "PCA: lineage", colour = "lineage") +
        theme_ng()
    save_eval(p_pca_lin, emb, file.path(win_dir, "pca_lineage"))

    if ("stage" %in% colnames(emb) && any(!is.na(emb$stage))) {
        p_pca_stg <- ggplot(emb[!is.na(stage)],
                            aes(x = PC1, y = PC2, colour = stage)) +
            geom_point(size = 1.8, alpha = 0.95) +
            labs(title = "PCA: stage") + theme_ng()
        save_eval(p_pca_stg, emb[!is.na(stage)],
                  file.path(win_dir, "pca_stage"))
        print(p_pca_lin + p_pca_stg)
    } else {
        p_pca_lin
    }
}
```

```{r win_pca_scree, fig.width = 5, fig.height = 4}
if (exists("pcs")) {
    var_explained <- (pcs$sdev^2) / sum(pcs$sdev^2)
    scree <- data.table(PC = seq_along(var_explained),
                        var_explained = var_explained)[1:min(20, .N)]
    p_scree <- ggplot(scree, aes(x = PC, y = var_explained)) +
        geom_col(fill = "#444444") +
        labs(x = "PC", y = "variance explained") +
        theme_ng()
    save_eval(p_scree, scree, file.path(win_dir, "pca_scree"))
    p_scree
}
```

## Cell embedding: UMAP

```{r win_umap, fig.width = 9, fig.height = 4}
if (!is.null(emb) && nrow(mat) >= 5 && ncol(mat) >= 5) {
    set.seed(1L)
    n_neigh <- max(2L, min(15L, nrow(mat) - 1L))
    um <- tryCatch(uwot::umap(mat, n_neighbors = n_neigh, min_dist = 0.3,
                              n_components = 2, verbose = FALSE),
                   error = function(e) { cat(conditionMessage(e), "\n"); NULL })
    if (!is.null(um)) {
        umd <- data.table(cell_id = rownames(mat),
                          UMAP1 = um[, 1], UMAP2 = um[, 2])
        umd <- merge(umd, man[, .(cell_id, group)], by = "cell_id")
        if ("stage" %in% colnames(man))
            umd <- merge(umd, man[, .(cell_id, stage)], by = "cell_id")

        p_um_lin <- ggplot(umd, aes(x = UMAP1, y = UMAP2, colour = group)) +
            geom_point(size = 1.8, alpha = 0.95) +
            scale_colour_manual(values = lineage_pal) +
            labs(title = "UMAP: lineage", colour = "lineage") + theme_ng()
        save_eval(p_um_lin, umd, file.path(win_dir, "umap_lineage"))

        if ("stage" %in% colnames(umd) && any(!is.na(umd$stage))) {
            p_um_stg <- ggplot(umd[!is.na(stage)],
                               aes(x = UMAP1, y = UMAP2, colour = stage)) +
                geom_point(size = 1.8, alpha = 0.95) +
                labs(title = "UMAP: stage") + theme_ng()
            save_eval(p_um_stg, umd[!is.na(stage)],
                      file.path(win_dir, "umap_stage"))
            print(p_um_lin + p_um_stg)
        } else {
            p_um_lin
        }
    }
}
```

## Top variable windows

```{r win_top_var, fig.width = 6, fig.height = 4}
win_var <- wcf[, .(sd_i_total = sd(i_total, na.rm = TRUE), n = .N),
               by = feature_id][n >= 5]
top_win <- head(win_var[order(-sd_i_total)], 20)
p_win_top <- ggplot(top_win,
                    aes(x = reorder(feature_id, sd_i_total), y = sd_i_total)) +
    geom_col(fill = "#444444") +
    coord_flip() +
    labs(x = NULL, y = expression("sd(" * i[total] * ") across cells")) +
    theme_ng() +
    theme(axis.text.y = element_text(size = 6), aspect.ratio = NULL)
save_eval(p_win_top, top_win, file.path(win_dir, "top_variable_windows"),
          width_mm = 110, height_mm = 80)
p_win_top
```

## Heatmap of top-variable windows by lineage

```{r win_heatmap, fig.width = 5, fig.height = 5}
top_ids <- top_win$feature_id
hm_w <- dcast(wcf[feature_id %in% top_ids,
                  .(median_i = median(i_total, na.rm = TRUE)),
                  by = .(feature_id, group)],
              feature_id ~ group, value.var = "median_i")
hm_w_mat <- as.matrix(hm_w[, -1])
rownames(hm_w_mat) <- hm_w$feature_id
hm_w_long <- as.data.table(reshape2::melt(hm_w_mat,
                                          varnames = c("feature_id", "group")))
p_win_hm <- ggplot(hm_w_long, aes(x = group, y = feature_id, fill = value)) +
    geom_tile() +
    scale_fill_viridis(option = "magma", na.value = "grey90") +
    labs(x = NULL, y = "window", fill = expression(i[total])) +
    theme_ng() +
    theme(axis.text.y = element_text(size = 6),
          axis.text.x = element_text(angle = 30, hjust = 1),
          aspect.ratio = NULL)
save_eval(p_win_hm, hm_w_long,
          file.path(win_dir, "heatmap_top_variable_windows"),
          width_mm = 100, height_mm = 130)
p_win_hm
```

# Combined paper figure

```{r paper, fig.width = 12, fig.height = 8}
panels <- list(
    A = if (exists("p_itotal_facet")) p_itotal_facet + labs(title = "A. i_total / regions") else NULL,
    B = if (exists("p_jsd_ann")) p_jsd_ann + labs(title = "B. jsd / regions") else NULL,
    C = if (exists("p_wit")) p_wit + labs(title = "C. mean i_total per cell / windows") else NULL,
    D = if (exists("p_pca_lin")) p_pca_lin + labs(title = "D. PCA cells / windows") else NULL,
    E = if (exists("p_um_lin")) p_um_lin + labs(title = "E. UMAP cells / windows") else NULL,
    F = if (exists("p_sum_hm")) p_sum_hm + labs(title = "F. summary / regions") else NULL
)
panels <- Filter(Negate(is.null), panels)
if (length(panels) >= 2) {
    combined <- Reduce(`+`, panels) + plot_layout(ncol = 3)
    ggsave(file.path(panels_dir, "paper_figure.pdf"),
           combined, width = 280, height = 180, units = "mm",
           device = grDevices::cairo_pdf)
    combined
}
```

# Session

```{r session}
sessionInfo()
```
