---
title: "Argelaguet 2019 mouse gastrulation: methylation entropy by feature (CpG context)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Izaskun Mallona"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
params:
  features_dir: ""
  win_cell_feature: ""
  win_feature: ""
  win_bed: ""
  manifest: ""
  out_dir: ""
---

```{r setup}
suppressPackageStartupMessages({
  library(data.table)
  library(readr)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(BiocParallel)
  library(ggrepel)
  library(ggridges)
  library(ComplexHeatmap)
  library(circlize)
  library(uwot)
  library(irlba)
})

repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))
source(file.path(repo_root, "workflow", "scripts", "palettes.R"))
source(file.path(repo_root, "workflow", "scripts", "driver_utils.R"))
source(file.path(repo_root, "workflow", "scripts", "embedding_utils.R"))
param <- SerialParam()
```

```{r opts}
knitr::opts_chunk$set(
  echo       = TRUE,
  fig.width  = 5,
  fig.height = 5,
  cache      = FALSE,
  include    = TRUE,
  fig.path   = "argelaguet_plots/",
  dev        = c("png", "svg"),
  dpi        = 500,
  cache.lazy = FALSE,
  warning    = TRUE,
  message    = TRUE
)
```

Mouse (mm10 / GRCm38), CpG context. TET TKO cells (Plate11-16) excluded.
Only cells passing methylation QC included.

```{r helpers}
ann_labels <- c(
  "genes"            = "Genes",
  "promoters"        = "Promoters",
  "lines"            = "LINEs",
  "sines"            = "SINEs",
  "h3k4me3"          = "H3K4me3 (ENCODE)",
  "h3k9me3"          = "H3K9me3 (ENCODE)",
  "h3k27me3"         = "H3K27me3 (ENCODE)",
  "h3k4me1"          = "H3K4me1 (ENCODE)",
  "h3k27ac"          = "H3K27ac (ENCODE)",
  ## gastrulation-specific from GSE125318 (E7.5 ChIP-seq)
  "enh-E75-Ect"      = "Enh E7.5 Ect",
  "enh-E75-End"      = "Enh E7.5 End",
  "enh-E75-Mes"      = "Enh E7.5 Mes",
  "enh-E75-union"    = "Enh E7.5 union",
  "h3k4me3-E75-Ect"      = "H3K4me3 E7.5 Ect",
  "h3k4me3-E75-End"      = "H3K4me3 E7.5 End",
  "h3k4me3-E75-Mes"      = "H3K4me3 E7.5 Mes",
  "h3k4me3-E75-common"   = "H3K4me3 E7.5 common",
  ## ESC marks from ENCODE
  "esc-p300"         = "ESC p300",
  "esc-dhs"          = "ESC DHS"
)

harmonize_levels <- function(df) {
  missing_ann <- setdiff(unique(df$annotation), names(ann_labels))
  if (length(missing_ann) > 0)
    warning("unknown annotations: ", paste(missing_ann, collapse = ", "))
  df$annotation <- factor(df$annotation,
                           levels = names(ann_labels),
                           labels = unname(ann_labels))
  if ("stage" %in% names(df))
    df$stage <- factor(df$stage, levels = sort(unique(df$stage)))
  df
}

## Sanitize metadata strings the same way the smk does for wildcard values.
sanitize <- function(x) gsub("[ ._]", "-", x)

## Recover annotation prefix from amet feature_id (<annotation>_<index>).
recover_annotation <- function(fid) sub("_\\d+$", "", fid)

## Filename parsers (mirror yamet/argelaguet.Rmd):
##   filename = "<annotation>_<stage>_<lineage>.cell_feature.tsv.gz"
##   annotation may itself contain hyphens; stage and lineage are the trailing
##   two underscore-separated tokens before the extension.
get_annotation <- function(fn) sub("^(.*)_[^_]+_[^_]+\\..*$", "\\1", fn)
get_stage      <- function(fn) sub("^.*_([^_]+)_[^_]+\\..*$", "\\1", fn)
get_lineage    <- function(fn) sub("^.*_([^_.]+)\\.[^_]*$",   "\\1", fn)
```

```{r load_metadata}
man <- fread(params$manifest)
meta_cells <- as.data.frame(man) %>%
  mutate(
    stage_san   = sanitize(stage),
    lineage_san = sanitize(lineage10x)
  )

meta_grp <- meta_cells %>%
  group_by(stage_san, lineage_san) %>%
  summarise(
    stage         = first(stage),
    lineage       = first(lineage10x),
    lineage_class = {
      if ("lineage10x_2" %in% names(meta_cells)) {
        vals <- lineage10x_2[!is.na(lineage10x_2) & lineage10x_2 != ""]
        if (length(vals) == 0) NA_character_ else first(vals)
      } else NA_character_
    },
    n_meta        = n(),
    .groups       = "drop"
  )

cat("Groups (stage x lineage):", nrow(meta_grp), "\n")
cat("Stages:", paste(sort(unique(meta_grp$stage)), collapse = ", "), "\n")
cat("Lineage classes:", paste(sort(unique(meta_grp$lineage_class)), collapse = ", "), "\n")
```

```{r load_amet}
## Each amet run produced one TSV pair per (annotation, stage, lineage) combo,
## mirroring yamet's per-(annotation, stage, lineage) yamet runs. Enumerate
## files, parse the wildcard values from filenames, concatenate.
features_dir <- params$features_dir
cf_files <- list.files(features_dir,
                       pattern = "\\.cell_feature\\.tsv\\.gz$",
                       full.names = TRUE)
fe_files <- list.files(features_dir,
                       pattern = "\\.feature\\.tsv\\.gz$",
                       full.names = TRUE)
fe_files <- fe_files[!grepl("\\.cell_feature\\.tsv\\.gz$", fe_files)]

read_combo_cf <- function(fp) {
  fn <- basename(fp)
  base <- sub("\\.cell_feature\\.tsv\\.gz$", "", fn)
  ann <- get_annotation(paste0(base, ".x"))
  stg <- get_stage(paste0(base, ".x"))
  lin <- get_lineage(paste0(base, ".x"))
  dt <- fread(fp)
  dt[, annotation := ann]
  dt[, stage_san := stg]
  dt[, lineage_san := lin]
  dt
}

read_combo_fe <- function(fp) {
  fn <- basename(fp)
  base <- sub("\\.feature\\.tsv\\.gz$", "", fn)
  ann <- get_annotation(paste0(base, ".x"))
  stg <- get_stage(paste0(base, ".x"))
  lin <- get_lineage(paste0(base, ".x"))
  dt <- fread(fp)
  dt[, annotation := ann]
  dt[, stage_san := stg]
  dt[, lineage_san := lin]
  dt
}

feat_cf <- rbindlist(lapply(cf_files, read_combo_cf), fill = TRUE)
feat_fe <- rbindlist(lapply(fe_files, read_combo_fe), fill = TRUE)

cat("loaded", length(cf_files), "cell_feature TSVs and",
    length(fe_files), "feature TSVs from", features_dir, "\n")

## Attach stage, lineage, lineage_class from manifest, joining on the sanitized
## stage/lineage that are now columns on the data tables.
meta_join <- as.data.table(meta_cells)[, .(cell_id,
                                            stage_san, lineage_san,
                                            stage, lineage = lineage10x,
                                            lineage_class = if ("lineage10x_2" %in% names(meta_cells))
                                                              lineage10x_2 else NA_character_)]
feat_cf <- merge(feat_cf, meta_join,
                 by = c("cell_id", "stage_san", "lineage_san"),
                 all.x = TRUE)
```

<!-- Section dropped: amet ships only methylation-stable i_total, no raw sampen analogue. yamet/argelaguet.Rmd lines 148-203 (S sample-entropy block, scatter and violin). -->

# adjS - methylation-adjusted sample entropy {.tabset .tabset-pills}

Per-cell normalised sample entropy from amet's per-cell-per-feature output
(amet `i_total`, the within-cell mutual-information score).

```{r import_adjS}
adjsampens <- as.data.frame(feat_cf)
adjsampens$sampen   <- adjsampens$i_total
adjsampens$avg_meth <- adjsampens$mean_meth
adjsampens <- harmonize_levels(adjsampens)
```

## adjS vs methylation

```{r scatter_adjS, fig.width = ng_fig_size(5, 9)$w, fig.height = ng_fig_size(5, 9)$h}
ggplot(adjsampens %>% filter(!is.na(lineage_class)),
       aes(x = avg_meth, y = sampen, color = lineage_class)) +
  geom_point(size = 0.3, alpha = 0.15) +
  scale_color_manual(values = argelaguet_lineage_class_pal, na.value = "grey70") +
  facet_grid(annotation ~ stage) +
  labs(x = "average methylation", y = expression(i[total]),
       color = "lineage class") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

## adjS distributions

```{r ridges_adjS, fig.width = ng_fig_size(4, 9)$w, fig.height = ng_fig_size(4, 9)$h}
ggplot(adjsampens %>% filter(!is.na(lineage_class)),
       aes(x = sampen, y = lineage, fill = lineage_class)) +
  geom_density_ridges(alpha = 0.6, scale = 0.85) +
  scale_fill_manual(values = argelaguet_lineage_class_pal) +
  facet_grid(annotation ~ stage, scales = "free") +
  labs(x = expression(i[total])) +
  theme_ng() +
  theme(legend.position = "none")
```

<!-- Section dropped: amet ships only methylation-stable jsd, no raw Shannon entropy analogue. yamet/argelaguet.Rmd lines 256-298 (H Shannon-entropy median block, scatter). -->

# adjH - methylation-adjusted Shannon entropy {.tabset .tabset-pills}

Per-feature jsd from amet's feature TSV (amet `jsd`, the multi-distribution
Jensen-Shannon divergence across cells in a group).

```{r import_adjH}
## yamet's adjshannons holds one median per (annotation, stage_san,
## lineage_san) — see yamet/argelaguet.Rmd 260-285. Aggregate amet's per-feature
## jsd to the same shape so the downstream group_medians join doesn't explode.
adjshannons <- as.data.frame(
  as.data.table(feat_fe)[
    , .(median_shannon  = median(jsd, na.rm = TRUE),
        median_avg_meth = median(mean_meth_mean, na.rm = TRUE)),
    by = .(annotation, stage_san, lineage_san)
  ]
)
adjshannons <- harmonize_levels(adjshannons)
adjshannons <- adjshannons %>%
  left_join(meta_grp %>% select(stage_san, lineage_san, stage, lineage,
                                 lineage_class),
            by = c("stage_san", "lineage_san"))
```

## Median adjH vs methylation

```{r scatter_adjH, fig.width = ng_fig_size(4, 9)$w, fig.height = ng_fig_size(4, 9)$h}
ggplot(adjshannons %>% filter(!is.na(lineage_class)),
       aes(x = median_avg_meth, y = median_shannon, color = lineage_class)) +
  geom_point(size = 1.5, alpha = 0.8) +
  scale_color_manual(values = argelaguet_lineage_class_pal, na.value = "grey70") +
  facet_grid(annotation ~ stage) +
  labs(x = "median avg. methylation", y = "jsd",
       color = "lineage class") +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  theme_ng()
```

# Assembly {.tabset .tabset-pills}

adjS (i_total) is per-cell. adjH (jsd) is per-feature per-group.

```{r group_medians}
saveRDS(
  list(adjsampens  = adjsampens,
       adjshannons = adjshannons),
  file = "argelaguet_entropy.rds"
)

adjS_grp <- adjsampens %>%
  group_by(annotation, stage_san, lineage_san) %>%
  summarise(median_adjS = median(sampen,    na.rm = TRUE),
            median_meth = median(avg_meth,  na.rm = TRUE),
            .groups     = "drop")

all_grp <- adjS_grp %>%
  left_join(adjshannons %>% select(annotation, stage_san, lineage_san,
                                    median_adjH = median_shannon),
            by = c("annotation", "stage_san", "lineage_san"))

all_grp_meta <- all_grp %>%
  left_join(meta_grp, by = c("stage_san", "lineage_san"))

saveRDS(
  list(all_grp      = all_grp,
       meta_grp     = meta_grp,
       all_grp_meta = all_grp_meta),
  file = "argelaguet_groups_meta.rds"
)

cat("Groups with entropy data:", nrow(all_grp_meta), "\n")
```

## adjS vs adjH

```{r adjS_vs_adjH, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(4, 1)$h}
df_cmp <- all_grp_meta %>% filter(!is.na(lineage_class))
if (nrow(df_cmp) > 0) {
  ggplot(df_cmp, aes(x = median_adjH, y = median_adjS, color = annotation)) +
    geom_point(size = 1.5, alpha = 0.8) +
    scale_color_manual(values = argelaguet_annotation_pal) +
    facet_grid(. ~ lineage_class) +
    labs(x = "median jsd", y = expression("median " * i[total]),
         color = "annotation") +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme_ng()
}
```

<!-- Section dropped: amet has no raw H/S, so the S vs H scatter (yamet/argelaguet.Rmd lines 406-418) is gone. -->

# Time course {.tabset .tabset-pills}

Entropy trajectories across developmental stages. Each point is the median for one
(stage, lineage) group. The gastrulation time series spans E4.5 through E7.5, covering
the emergence of epiblast, primitive streak, and the three germ layers. Entropy at
enhancers and active chromatin marks is expected to rise in progenitor states and
fall upon lineage commitment.

```{r timecourse_data}
stage_order <- sort(unique(all_grp_meta$stage))
tc <- all_grp_meta %>%
  filter(!is.na(lineage_class), !is.na(stage)) %>%
  mutate(stage = factor(stage, levels = stage_order))
```

## adjS by stage and lineage class

```{r timecourse_adjS, fig.width = ng_fig_size(length(stage_order), 9)$w, fig.height = ng_fig_size(length(stage_order), 9)$h}
ggplot(tc, aes(x = stage, y = median_adjS,
               color = lineage_class, group = interaction(lineage_san, annotation))) +
  geom_line(alpha = 0.5, linewidth = 0.4) +
  geom_point(size = 1.2, alpha = 0.8) +
  scale_color_manual(values = argelaguet_lineage_class_pal) +
  facet_wrap(~ annotation, ncol = 3, scales = "free_y") +
  labs(x = "developmental stage", y = expression("median " * i[total]),
       color = "lineage class") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1,
                                                   linewidth = 1))) +
  theme_ng()
```

## adjH by stage and lineage class

```{r timecourse_adjH, fig.width = ng_fig_size(length(stage_order), 9)$w, fig.height = ng_fig_size(length(stage_order), 9)$h}
ggplot(tc, aes(x = stage, y = median_adjH,
               color = lineage_class, group = interaction(lineage_san, annotation))) +
  geom_line(alpha = 0.5, linewidth = 0.4) +
  geom_point(size = 1.2, alpha = 0.8) +
  scale_color_manual(values = argelaguet_lineage_class_pal) +
  facet_wrap(~ annotation, ncol = 3, scales = "free_y") +
  labs(x = "developmental stage", y = "median jsd",
       color = "lineage class") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1,
                                                   linewidth = 1))) +
  theme_ng()
```

## Methylation by stage and lineage class

```{r timecourse_meth, fig.width = ng_fig_size(length(stage_order), 9)$w, fig.height = ng_fig_size(length(stage_order), 9)$h}
ggplot(tc, aes(x = stage, y = median_meth,
               color = lineage_class, group = interaction(lineage_san, annotation))) +
  geom_line(alpha = 0.5, linewidth = 0.4) +
  geom_point(size = 1.2, alpha = 0.8) +
  scale_color_manual(values = argelaguet_lineage_class_pal) +
  facet_wrap(~ annotation, ncol = 3, scales = "free_y") +
  labs(x = "developmental stage", y = "median avg. methylation",
       color = "lineage class") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1,
                                                   linewidth = 1))) +
  theme_ng()
```

# Enhancers vs promoters {.tabset .tabset-pills}

Comparing entropy at active enhancer marks (H3K27ac, H3K4me1) against promoter marks
(H3K4me3, Promoters). The gastrulation paper reports that meso- and endoderm-defining
enhancers undergo concerted demethylation upon germ-layer specification; adjS at
these marks captures the heterogeneity of that process. H3K27ac and H3K4me1
here come from ENCODE bulk mm10 data, not gastrulation-specific ChIP.

```{r enh_vs_prom_data}
enh_marks  <- c("Enh E7.5 Ect", "Enh E7.5 End", "Enh E7.5 Mes",
                "Enh E7.5 union", "ESC p300")
prom_marks <- c("H3K4me3 E7.5 common", "Promoters",
                "H3K4me3 (ENCODE)")

ep_df <- all_grp_meta %>%
  filter(annotation %in% c(enh_marks, prom_marks),
         !is.na(lineage_class)) %>%
  mutate(
    mark_type = case_when(
      annotation %in% enh_marks  ~ "enhancer",
      annotation %in% prom_marks ~ "promoter"
    ),
    stage = factor(stage, levels = stage_order)
  )
```

## adjS enhancers vs promoters across stage

```{r enh_vs_prom_adjS, fig.width = ng_fig_size(length(stage_order), 2)$w, fig.height = ng_fig_size(length(stage_order), 2)$h}
ggplot(ep_df, aes(x = stage, y = median_adjS,
                   color = lineage_class, shape = mark_type,
                   group = interaction(lineage_san, annotation))) +
  geom_line(alpha = 0.4, linewidth = 0.4) +
  geom_point(size = 1.5, alpha = 0.85) +
  scale_color_manual(values = argelaguet_lineage_class_pal) +
  scale_shape_manual(values = mark_type_shapes) +
  facet_grid(mark_type ~ lineage_class, scales = "free_y") +
  labs(x = "developmental stage", y = expression("median " * i[total]),
       color = "lineage class", shape = "mark type") +
  guides(
    color = guide_legend(override.aes = list(size = 2.5, alpha = 1, linewidth = 1)),
    shape = guide_legend(override.aes = list(size = 2.5))
  ) +
  theme_ng()
```

## adjH enhancers vs promoters across stage

```{r enh_vs_prom_adjH, fig.width = ng_fig_size(length(stage_order), 2)$w, fig.height = ng_fig_size(length(stage_order), 2)$h}
ggplot(ep_df, aes(x = stage, y = median_adjH,
                   color = lineage_class, shape = mark_type,
                   group = interaction(lineage_san, annotation))) +
  geom_line(alpha = 0.4, linewidth = 0.4) +
  geom_point(size = 1.5, alpha = 0.85) +
  scale_color_manual(values = argelaguet_lineage_class_pal) +
  scale_shape_manual(values = mark_type_shapes) +
  facet_grid(mark_type ~ lineage_class, scales = "free_y") +
  labs(x = "developmental stage", y = "median jsd",
       color = "lineage class", shape = "mark type") +
  guides(
    color = guide_legend(override.aes = list(size = 2.5, alpha = 1, linewidth = 1)),
    shape = guide_legend(override.aes = list(size = 2.5))
  ) +
  theme_ng()
```

## Paired scatter: adjS enhancer vs adjS promoter

Each point is one (stage, lineage) group. Groups above the diagonal have higher
within-cell entropy at enhancers than at promoters.

```{r enh_prom_scatter, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h}
ep_wide <- all_grp_meta %>%
  filter(annotation %in% c("Enh E7.5 union", "Promoters"),
         !is.na(lineage_class)) %>%
  select(stage_san, lineage_san, lineage_class, stage, annotation, median_adjS) %>%
  pivot_wider(names_from = annotation, values_from = median_adjS) %>%
  rename(adjS_enh = `Enh E7.5 union`, adjS_prom = Promoters) %>%
  mutate(stage = factor(stage, levels = stage_order))

if (all(c("adjS_enh", "adjS_prom") %in% names(ep_wide))) {
  ggplot(ep_wide, aes(x = adjS_prom, y = adjS_enh,
                       color = lineage_class, shape = stage)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                color = "grey60", linewidth = 0.4) +
    geom_point(size = 2, alpha = 0.8) +
    scale_color_manual(values = argelaguet_lineage_class_pal) +
    scale_shape_manual(values = seq_along(levels(ep_wide$stage))) +
    labs(x = expression("median " * i[total] * " at promoters"),
         y = expression("median " * i[total] * " at E7.5 enhancers (union)"),
         color = "lineage class", shape = "stage") +
    guides(
      color = guide_legend(override.aes = list(size = 3)),
      shape = guide_legend(override.aes = list(size = 3))
    ) +
    theme_ng()
}
```

# Variation across lineages {.tabset .tabset-pills}

SD and range of group medians per annotation, broken down by lineage class.

```{r variation_helpers}
safe_range <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  diff(range(x))
}
```

```{r variation_lineage, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(4, 1)$h}
var_lineage <- all_grp_meta %>%
  filter(!is.na(lineage_class)) %>%
  group_by(annotation, lineage_class) %>%
  summarise(
    adjS_sd    = sd(median_adjS, na.rm = TRUE),
    adjH_sd    = sd(median_adjH, na.rm = TRUE),
    adjS_range = safe_range(median_adjS),
    adjH_range = safe_range(median_adjH),
    .groups = "drop"
  )

if (any(!is.na(var_lineage$adjH_sd))) {
  ggplot(var_lineage, aes(x = adjH_sd, y = adjS_sd,
                           color = annotation, label = annotation)) +
    geom_point(size = 3) +
    geom_text_repel(size = 3) +
    scale_color_manual(values = argelaguet_annotation_pal) +
    facet_grid(. ~ lineage_class) +
    labs(x = "SD of jsd across groups",
         y = expression("SD of " * i[total] * " across groups")) +
    theme_ng()
}
```

# Driver categorization

For each annotation, we compute the SD of group-level median adjH (across-cell
heterogeneity) and median adjS (within-cell heterogeneity) across all
(stage, lineage) groups. If one SD is at least 1.5x the other, that entropy
component dominates. Annotations where both SDs fall below the 30th percentile
are labelled "neither".

```{r drivers, fig.width = ng_fig_size(3, 3)$w, fig.height = ng_fig_size(3, 3)$h}
driver_df <- categorize_drivers(all_grp_meta, "lineage_class")

cat("Driver table:\n")
print(table(driver_df$driver))

plot_driver_scatter(driver_df,
                    x_label = "SD of jsd across lineage class groups",
                    y_label = expression("SD of " * i[total] * " across lineage class groups"))
```

# Heatmaps {.tabset .tabset-pills}

Rows = lineage (clustered). Columns = annotation. Row bars show lineage class.

```{r heatmap_helpers}
make_heatmap <- function(df, value_col, title,
                          palette = c("navy", "white", "firebrick")) {
  wide <- df %>%
    select(lineage, annotation, lineage_class,
           value = all_of(value_col)) %>%
    filter(!is.na(lineage_class)) %>%
    group_by(lineage, annotation, lineage_class) %>%
    summarise(value = median(value, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = annotation, values_from = value)

  if (nrow(wide) < 2) {
    message("Too few rows for heatmap")
    return(invisible(NULL))
  }

  mat <- wide %>% select(-lineage, -lineage_class) %>% as.matrix()
  rownames(mat) <- wide$lineage

  ## Drop rows / columns that are entirely NA so hclust gets a finite distance.
  keep_rows <- rowSums(!is.na(mat)) >= 2
  keep_cols <- colSums(!is.na(mat)) >= 2
  mat  <- mat[keep_rows, keep_cols, drop = FALSE]
  wide <- wide[keep_rows, , drop = FALSE]

  if (nrow(mat) < 2 || ncol(mat) < 2) {
    message("Skipping heatmap: not enough non-NA rows/cols")
    return(invisible(NULL))
  }

  breaks <- c(min(mat, na.rm = TRUE),
              mean(range(mat, na.rm = TRUE)),
              max(mat, na.rm = TRUE))
  if (any(!is.finite(breaks)) || breaks[1] == breaks[3]) {
    message("Skipping heatmap: degenerate data (all-NA or constant matrix)")
    return(invisible(NULL))
  }

  row_ann <- rowAnnotation(
    lineage_class = wide$lineage_class,
    col = list(lineage_class = argelaguet_lineage_class_pal),
    annotation_legend_param = list(
      lineage_class = list(title = "lineage class")
    )
  )

  col_fun <- colorRamp2(breaks, palette)

  Heatmap(mat,
          name              = title,
          col               = col_fun,
          row_split         = wide$lineage_class,
          cluster_rows      = TRUE,
          cluster_columns   = FALSE,
          show_row_names    = TRUE,
          show_column_names = TRUE,
          row_names_gp      = gpar(fontsize = 7),
          right_annotation  = row_ann,
          column_title      = title,
          na_col            = "grey90")
}
```

## adjS

```{r heatmap_adjS, fig.width = 10, fig.height = 8}
heat_df_adjs <- adjsampens %>%
  group_by(annotation, stage_san, lineage_san, lineage, lineage_class) %>%
  summarise(median_adjS = median(sampen, na.rm = TRUE), .groups = "drop")

make_heatmap(heat_df_adjs, "median_adjS", "median i_total",
             palette = c("navy", "white", "firebrick"))
```

## adjH

```{r heatmap_adjH, fig.width = 10, fig.height = 8}
heat_df_adjh <- adjshannons %>%
  group_by(annotation, stage_san, lineage_san, lineage, lineage_class) %>%
  summarise(median_adjH = median(median_shannon, na.rm = TRUE), .groups = "drop")

make_heatmap(heat_df_adjh, "median_adjH", "median jsd",
             palette = c("navy", "white", "firebrick"))
```

<!-- Section dropped: amet ships no paper-MOFA UMAP overlay (yamet/argelaguet.Rmd lines 715-885). The MOFA UMAP is a yamet-specific external dependency; revisit if amet ever wires that file in. -->

# Cell-level UMAP {.tabset .tabset-pills}

```{r umap_helpers}

plot_umap <- function(df, color_col, title, subtitle = NULL) {
  if (is.null(df) || nrow(df) == 0) {
    message("No UMAP result for: ", title)
    return(invisible(NULL))
  }
  ggplot(df %>% filter(!is.na(.data[[color_col]])),
         aes(x = UMAP1, y = UMAP2, color = .data[[color_col]])) +
    geom_point(size = 1.5, alpha = 0.7) +
    scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
    labs(title = title, subtitle = subtitle,
         color = color_col) +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme_ng()
}
```

```{r cell_umap_load}
## amet's per-cell-per-feature i_total replaces yamet's per-cell sampen_norm.
## Build the same wide cells x annotation matrix used downstream.
cells_adjS_long <- as.data.frame(feat_cf)[, c("cell_id", "stage_san",
                                               "lineage_san", "annotation",
                                               "i_total")] %>%
  rename(value = i_total) %>%
  group_by(cell_id, stage_san, lineage_san, annotation) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop")
cells_adjS_long$annotation <- factor(cells_adjS_long$annotation,
                                      levels = names(ann_labels),
                                      labels = unname(ann_labels))

cells_meth_long <- as.data.frame(feat_cf)[, c("cell_id", "stage_san",
                                               "lineage_san", "annotation",
                                               "mean_meth")] %>%
  rename(value = mean_meth) %>%
  group_by(cell_id, stage_san, lineage_san, annotation) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop")
cells_meth_long$annotation <- factor(cells_meth_long$annotation,
                                      levels = names(ann_labels),
                                      labels = unname(ann_labels))

meta_cols_cell <- c("stage_san", "lineage_san", "cell_id",
                    "stage", "lineage", "lineage_class")

cells_adjS_wide <- pivot_wider(cells_adjS_long,
                                id_cols    = c(cell_id, stage_san, lineage_san),
                                names_from = annotation,
                                values_from = value) %>%
  left_join(meta_grp %>% select(stage_san, lineage_san, stage, lineage,
                                 lineage_class),
            by = c("stage_san", "lineage_san"))

cells_meth_wide <- pivot_wider(cells_meth_long,
                                id_cols    = c(cell_id, stage_san, lineage_san),
                                names_from = annotation,
                                values_from = value) %>%
  left_join(meta_grp %>% select(stage_san, lineage_san, stage, lineage,
                                 lineage_class),
            by = c("stage_san", "lineage_san"))

saveRDS(list(cells_adjS_wide = cells_adjS_wide,
             cells_meth_wide = cells_meth_wide),
        "argelaguet_cell_matrices.rds")
```

## Cell-level: adjS across annotations

```{r cell_umap_adjS, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h}
umap_cell_adjS <- run_umap_wide(cells_adjS_wide, meta_cols_cell,
                                 n_neighbors = 15L)
saveRDS(umap_cell_adjS, "argelaguet_umap_cell_i_total.rds")

plot_umap(umap_cell_adjS, "lineage_class",
          title    = "Cell-level UMAP - i_total features",
          subtitle = "per-cell i_total per annotation as feature vector")
```

## Cell-level: adjS colored by stage

```{r cell_umap_adjS_stage, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h}
plot_umap(umap_cell_adjS, "stage",
          title    = "Cell-level UMAP - i_total features",
          subtitle = "colored by developmental stage")
```

## Cell-level: avg methylation across annotations

```{r cell_umap_meth, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h}
umap_cell_meth <- run_umap_wide(cells_meth_wide, meta_cols_cell,
                                 n_neighbors = 15L)
saveRDS(umap_cell_meth, "argelaguet_umap_cell_meth.rds")

plot_umap(umap_cell_meth, "lineage_class",
          title    = "Cell-level UMAP - avg. methylation features",
          subtitle = "per-cell avg. methylation per annotation as feature vector")
```

<!-- Section dropped: amet has no raw S, so the S-with-meth-regressed-out UMAP (yamet/argelaguet.Rmd lines 988-1033) is gone. -->

## Group-level: adjH across annotations

```{r group_umap_adjH, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h}
adjH_wide_grp <- all_grp_meta %>%
  filter(!is.na(lineage_class)) %>%
  select(stage_san, lineage_san, stage, lineage, lineage_class,
         annotation, median_adjH) %>%
  pivot_wider(names_from = annotation, values_from = median_adjH)

umap_grp_adjH <- run_umap_wide(
  adjH_wide_grp,
  meta_cols   = c("stage_san", "lineage_san", "stage", "lineage", "lineage_class"),
  n_neighbors = 5L
)
saveRDS(umap_grp_adjH, "argelaguet_umap_grp_jsd.rds")

plot_umap(umap_grp_adjH, "lineage_class",
          title    = "Group-level UMAP - jsd features",
          subtitle = "median jsd per annotation as feature vector")
```

```{r group_umap_adjH_stage, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h}
plot_umap(umap_grp_adjH, "stage",
          title    = "Group-level UMAP - jsd features",
          subtitle = "colored by developmental stage")
```

<!-- Section dropped: per-annotation locus-level UMAP from .norm.det.out.gz (yamet/argelaguet.Rmd lines 1062-1179). amet does not export per-locus jsd vectors per group; only per-feature jsd is available, which is already plotted above. -->
