---
title: "Argelaguet 2019 mouse gastrulation: methylation entropy by feature (CpG context)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Izaskun Mallona"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
params:
  features_dir: ""
  win_cell_feature: ""
  win_feature: ""
  win_bed: ""
  manifest: ""
  out_dir: ""
  log_path: ""
  threads: 1
  i_max_lag: 3
---

```{r logging_early, include = FALSE}
source(Sys.getenv("AMET_RENDER_HELPERS"))
param <- amet_setup_render_logging(params$log_path, params$threads)
```

```{r setup}
suppressPackageStartupMessages({
  library(data.table)
  library(readr)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(BiocParallel)
  library(ggrepel)
  library(ggridges)
  library(ComplexHeatmap)
  library(circlize)
  library(uwot)
  library(irlba)
})

repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))
source(file.path(repo_root, "workflow", "scripts", "palettes.R"))
source(file.path(repo_root, "workflow", "scripts", "driver_utils.R"))
source(file.path(repo_root, "workflow", "scripts", "embedding_utils.R"))
```

```{r opts}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 5,
  fig.height = 5,
  cache = FALSE,
  include = TRUE,
  fig.path = "argelaguet_plots/",
  dev = c("png", "svg"),
  dpi = 500,
  cache.lazy = FALSE,
  warning = TRUE,
  message = TRUE
)
```

Mouse (mm10 / GRCm38), CpG context. TET TKO cells (Plate11-16) excluded.
Only cells passing methylation QC included.

```{r helpers}
ann_labels <- c(
  "genes" = "Genes",
  "promoters" = "Promoters",
  "lines" = "LINEs",
  "sines" = "SINEs",
  "h3k4me3" = "H3K4me3 (ENCODE)",
  "h3k9me3" = "H3K9me3 (ENCODE)",
  "h3k27me3" = "H3K27me3 (ENCODE)",
  "h3k4me1" = "H3K4me1 (ENCODE)",
  "h3k27ac" = "H3K27ac (ENCODE)",
  ## gastrulation-specific from GSE125318 (E7.5 ChIP-seq)
  "enh-E75-Ect" = "Enh E7.5 Ect",
  "enh-E75-End" = "Enh E7.5 End",
  "enh-E75-Mes" = "Enh E7.5 Mes",
  "enh-E75-union" = "Enh E7.5 union",
  "h3k4me3-E75-Ect" = "H3K4me3 E7.5 Ect",
  "h3k4me3-E75-End" = "H3K4me3 E7.5 End",
  "h3k4me3-E75-Mes" = "H3K4me3 E7.5 Mes",
  "h3k4me3-E75-common" = "H3K4me3 E7.5 common",
  ## ESC marks from ENCODE
  "esc-p300" = "ESC p300",
  "esc-dhs" = "ESC DHS"
)

harmonize_levels <- function(df) {
  missing_ann <- setdiff(unique(df$annotation), names(ann_labels))
  if (length(missing_ann) > 0)
    warning("unknown annotations: ", paste(missing_ann, collapse = ", "))
  df$annotation <- factor(df$annotation,
                           levels = names(ann_labels),
                           labels = unname(ann_labels))
  if ("stage" %in% names(df))
    df$stage <- factor(df$stage, levels = sort(unique(df$stage)))
  df
}

## Sanitize metadata strings the same way the smk does for wildcard values.
sanitize <- function(x) gsub("[ ._]", "-", x)

## Recover annotation prefix from amet feature_id (<annotation>_<index>).
recover_annotation <- function(fid) sub("_\\d+$", "", fid)

## Filename parsers:
##   filename = "<annotation>_<stage>_<lineage>.cell_feature.tsv.gz"
##   annotation may itself contain hyphens; stage and lineage are the trailing
##   two underscore-separated tokens before the extension.
get_annotation <- function(fn) sub("^(.*)_[^_]+_[^_]+\\..*$", "\\1", fn)
get_stage <- function(fn) sub("^.*_([^_]+)_[^_]+\\..*$", "\\1", fn)
get_lineage <- function(fn) sub("^.*_([^_.]+)\\.[^_]*$", "\\1", fn)
```

```{r load_metadata}
man <- fread(params$manifest)
meta_cells <- as.data.frame(man) %>%
  mutate(
    stage_san = sanitize(stage),
    lineage_san = sanitize(lineage10x)
  )

meta_grp <- meta_cells %>%
  group_by(stage_san, lineage_san) %>%
  summarise(
    stage = first(stage),
    lineage = first(lineage10x),
    lineage_class = {
      if ("lineage10x_2" %in% names(meta_cells)) {
        vals <- lineage10x_2[!is.na(lineage10x_2) & lineage10x_2 != ""]
        if (length(vals) == 0) NA_character_ else first(vals)
      } else NA_character_
    },
    n_meta = n(),
    .groups = "drop"
  )

cat("Groups (stage x lineage):", nrow(meta_grp), "\n")
cat("Stages:", paste(sort(unique(meta_grp$stage)), collapse = ", "), "\n")
cat("Lineage classes:", paste(sort(unique(meta_grp$lineage_class)), collapse = ", "), "\n")
```

```{r load_amet}
## Each amet run produced one TSV pair per (annotation, stage, lineage)
## combo. Enumerate files, parse the wildcard values from filenames,
## concatenate.
features_dir <- params$features_dir
cf_files <- list.files(features_dir,
                       pattern = "\\.cell_feature\\.tsv\\.gz$",
                       full.names = TRUE)
fe_files <- list.files(features_dir,
                       pattern = "\\.feature\\.tsv\\.gz$",
                       full.names = TRUE)
fe_files <- fe_files[!grepl("\\.cell_feature\\.tsv\\.gz$", fe_files)]

## Aggregate per cell at read time: feat_cf collapses from O(cells * features
## * strata * annotations) rows to O(cells * strata * annotations). Every
## downstream plot/embedding already averages over features per cell, so no
## information is lost.
## amet can emit a near-empty .feature.tsv.gz (just gzip overhead, no rows or
## header) when a combo has zero groups passing min_cells_per_group; the
## cell-level file may also be tiny if every cell got filtered out. Guard
## both readers so a single dropped combo doesn't kill the whole report.
read_combo_cf <- function(fp) {
  fn <- basename(fp)
  base <- sub("\\.cell_feature\\.tsv\\.gz$", "", fn)
  if (file.info(fp)$size < 30) {
    message("[load_amet] skipping near-empty cell_feature file: ", fn)
    return(NULL)
  }
  dt <- tryCatch(fread(fp, select = c("cell_id", "mean_meth", "i_total")),
                 error = function(e) {
                   message("[load_amet] failed to read ", fn, ": ", conditionMessage(e))
                   NULL
                 })
  if (is.null(dt) || nrow(dt) == 0L) return(NULL)
  agg <- dt[, .(mean_meth = mean(mean_meth, na.rm = TRUE),
                i_total = mean(i_total, na.rm = TRUE)),
            by = cell_id]
  agg[, annotation := get_annotation(paste0(base, ".x"))]
  agg[, stage_san := get_stage(paste0(base, ".x"))]
  agg[, lineage_san := get_lineage(paste0(base, ".x"))]
  agg
}

read_combo_fe <- function(fp) {
  fn <- basename(fp)
  base <- sub("\\.feature\\.tsv\\.gz$", "", fn)
  if (file.info(fp)$size < 30) {
    message("[load_amet] skipping near-empty feature file: ", fn)
    return(NULL)
  }
  dt <- tryCatch(fread(fp),
                 error = function(e) {
                   message("[load_amet] failed to read ", fn, ": ", conditionMessage(e))
                   NULL
                 })
  if (is.null(dt) || nrow(dt) == 0L) return(NULL)
  dt[, annotation := get_annotation(paste0(base, ".x"))]
  dt[, stage_san := get_stage(paste0(base, ".x"))]
  dt[, lineage_san := get_lineage(paste0(base, ".x"))]
  dt
}

feat_cf <- rbindlist(lapply(cf_files, read_combo_cf), fill = TRUE)
feat_fe <- rbindlist(lapply(fe_files, read_combo_fe), fill = TRUE)

cat("loaded", length(cf_files), "cell_feature TSVs and",
    length(fe_files), "feature TSVs from", features_dir, "\n")

## Attach stage, lineage, lineage_class by stratum only (one row per
## (stage_san, lineage_san), not per cell). Update by reference so we don't
## copy feat_cf.
meta_grp_dt <- as.data.table(meta_grp)[, .(stage_san, lineage_san,
                                            stage, lineage, lineage_class)]
feat_cf[meta_grp_dt, on = .(stage_san, lineage_san),
        `:=`(stage = i.stage,
             lineage = i.lineage,
             lineage_class = i.lineage_class)]
```

# i_total per cell {.tabset .tabset-pills}

Per-cell within-cell mutual information score from amet's per-cell-per-feature
output (amet `i_total`).

```{r import_i_total}
i_total_long <- as.data.frame(feat_cf)
i_total_long$avg_meth <- i_total_long$mean_meth
i_total_long <- harmonize_levels(i_total_long)
```

## i_total vs methylation

```{r scatter_i_total, fig.width = ng_fig_size(5, 9)$w, fig.height = ng_fig_size(5, 9)$h}
ggplot(i_total_long %>% filter(!is.na(lineage_class)),
       aes(x = avg_meth, y = i_total, color = lineage_class)) +
  geom_point(size = 0.3, alpha = 0.15) +
  scale_color_manual(values = argelaguet_lineage_class_pal, na.value = "grey70") +
  facet_grid(annotation ~ stage) +
  labs(x = "average methylation", y = expression(i[total]),
       color = "lineage class") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

## i_total distributions

```{r ridges_i_total, fig.width = ng_fig_size(4, 9)$w, fig.height = ng_fig_size(4, 9)$h}
ggplot(i_total_long %>% filter(!is.na(lineage_class)),
       aes(x = i_total, y = lineage, fill = lineage_class)) +
  geom_density_ridges(alpha = 0.6, scale = 0.85) +
  scale_fill_manual(values = argelaguet_lineage_class_pal) +
  facet_grid(annotation ~ stage, scales = "free") +
  labs(x = expression(i[total])) +
  theme_ng() +
  theme(legend.position = "none")
```

# jsd per feature {.tabset .tabset-pills}

Per-feature multi-distribution Jensen-Shannon divergence across cells in a
group (amet `jsd`).

```{r import_jsd}
## Aggregate amet's per-feature jsd to one median per (annotation,
## stage_san, lineage_san) so the downstream group_medians join doesn't
## explode.
jsd_long <- as.data.frame(
  as.data.table(feat_fe)[
    , .(median_jsd = median(jsd, na.rm = TRUE),
        median_avg_meth = median(mean_meth_mean, na.rm = TRUE)),
    by = .(annotation, stage_san, lineage_san)
  ]
)
jsd_long <- harmonize_levels(jsd_long)
jsd_long <- jsd_long %>%
  left_join(meta_grp %>% select(stage_san, lineage_san, stage, lineage,
                                 lineage_class),
            by = c("stage_san", "lineage_san"))
```

## Median jsd vs methylation

```{r scatter_jsd, fig.width = ng_fig_size(4, 9)$w, fig.height = ng_fig_size(4, 9)$h}
ggplot(jsd_long %>% filter(!is.na(lineage_class)),
       aes(x = median_avg_meth, y = median_jsd, color = lineage_class)) +
  geom_point(size = 1.5, alpha = 0.8) +
  scale_color_manual(values = argelaguet_lineage_class_pal, na.value = "grey70") +
  facet_grid(annotation ~ stage) +
  labs(x = "median avg. methylation", y = "jsd",
       color = "lineage class") +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  theme_ng()
```

# Assembly {.tabset .tabset-pills}

i_total is per-cell. jsd is per-feature per-group.

```{r group_medians}
saveRDS(
  list(i_total_long = i_total_long,
       jsd_long = jsd_long),
  file = "argelaguet_entropy.rds"
)

i_total_grp <- i_total_long %>%
  group_by(annotation, stage_san, lineage_san) %>%
  summarise(median_i_total = median(i_total, na.rm = TRUE),
            median_meth = median(avg_meth, na.rm = TRUE),
            .groups = "drop")

all_grp <- i_total_grp %>%
  left_join(jsd_long %>% select(annotation, stage_san, lineage_san,
                                median_jsd),
            by = c("annotation", "stage_san", "lineage_san"))

all_grp_meta <- all_grp %>%
  left_join(meta_grp, by = c("stage_san", "lineage_san"))

saveRDS(
  list(all_grp = all_grp,
       meta_grp = meta_grp,
       all_grp_meta = all_grp_meta),
  file = "argelaguet_groups_meta.rds"
)

cat("Groups with entropy data:", nrow(all_grp_meta), "\n")
```

## i_total vs jsd

```{r i_total_vs_jsd, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(4, 1)$h}
df_cmp <- all_grp_meta %>% filter(!is.na(lineage_class))
if (nrow(df_cmp) > 0) {
  ggplot(df_cmp, aes(x = median_jsd, y = median_i_total, color = annotation)) +
    geom_point(size = 1.5, alpha = 0.8) +
    scale_color_manual(values = argelaguet_annotation_pal) +
    facet_grid(. ~ lineage_class) +
    labs(x = "median jsd", y = expression("median " * i[total]),
         color = "annotation") +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme_ng()
}
```

# Time course {.tabset .tabset-pills}

Entropy trajectories across developmental stages. Each point is the median for one
(stage, lineage) group. The gastrulation time series spans E4.5 through E7.5, covering
the emergence of epiblast, primitive streak, and the three germ layers. Entropy at
enhancers and active chromatin marks is expected to rise in progenitor states and
fall upon lineage commitment.

```{r timecourse_data}
stage_order <- sort(unique(all_grp_meta$stage))
tc <- all_grp_meta %>%
  filter(!is.na(lineage_class), !is.na(stage)) %>%
  mutate(stage = factor(stage, levels = stage_order))
```

## i_total by stage and lineage class

```{r timecourse_i_total, fig.width = ng_fig_size(length(stage_order), 9)$w, fig.height = ng_fig_size(length(stage_order), 9)$h}
ggplot(tc, aes(x = stage, y = median_i_total,
               color = lineage_class, group = interaction(lineage_san, annotation))) +
  geom_line(alpha = 0.5, linewidth = 0.4) +
  geom_point(size = 1.2, alpha = 0.8) +
  scale_color_manual(values = argelaguet_lineage_class_pal) +
  facet_wrap(~ annotation, ncol = 3, scales = "free_y") +
  labs(x = "developmental stage", y = expression("median " * i[total]),
       color = "lineage class") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1,
                                                   linewidth = 1))) +
  theme_ng()
```

## jsd by stage and lineage class

```{r timecourse_jsd, fig.width = ng_fig_size(length(stage_order), 9)$w, fig.height = ng_fig_size(length(stage_order), 9)$h}
ggplot(tc, aes(x = stage, y = median_jsd,
               color = lineage_class, group = interaction(lineage_san, annotation))) +
  geom_line(alpha = 0.5, linewidth = 0.4) +
  geom_point(size = 1.2, alpha = 0.8) +
  scale_color_manual(values = argelaguet_lineage_class_pal) +
  facet_wrap(~ annotation, ncol = 3, scales = "free_y") +
  labs(x = "developmental stage", y = "median jsd",
       color = "lineage class") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1,
                                                   linewidth = 1))) +
  theme_ng()
```

## Methylation by stage and lineage class

```{r timecourse_meth, fig.width = ng_fig_size(length(stage_order), 9)$w, fig.height = ng_fig_size(length(stage_order), 9)$h}
ggplot(tc, aes(x = stage, y = median_meth,
               color = lineage_class, group = interaction(lineage_san, annotation))) +
  geom_line(alpha = 0.5, linewidth = 0.4) +
  geom_point(size = 1.2, alpha = 0.8) +
  scale_color_manual(values = argelaguet_lineage_class_pal) +
  facet_wrap(~ annotation, ncol = 3, scales = "free_y") +
  labs(x = "developmental stage", y = "median avg. methylation",
       color = "lineage class") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1,
                                                   linewidth = 1))) +
  theme_ng()
```

# Enhancers vs promoters {.tabset .tabset-pills}

Comparing entropy at active enhancer marks (H3K27ac, H3K4me1) against promoter marks
(H3K4me3, Promoters). The gastrulation paper reports that meso- and endoderm-defining
enhancers undergo concerted demethylation upon germ-layer specification; i_total
at these marks captures the heterogeneity of that process. H3K27ac and H3K4me1
here come from ENCODE bulk mm10 data, not gastrulation-specific ChIP.

```{r enh_vs_prom_data}
enh_marks <- c("Enh E7.5 Ect", "Enh E7.5 End", "Enh E7.5 Mes",
               "Enh E7.5 union", "ESC p300")
prom_marks <- c("H3K4me3 E7.5 common", "Promoters",
                "H3K4me3 (ENCODE)")

ep_df <- all_grp_meta %>%
  filter(annotation %in% c(enh_marks, prom_marks),
         !is.na(lineage_class)) %>%
  mutate(
    mark_type = case_when(
      annotation %in% enh_marks  ~ "enhancer",
      annotation %in% prom_marks ~ "promoter"
    ),
    stage = factor(stage, levels = stage_order)
  )
```

## i_total enhancers vs promoters across stage

```{r enh_vs_prom_i_total, fig.width = ng_fig_size(length(stage_order), 2)$w, fig.height = ng_fig_size(length(stage_order), 2)$h}
ggplot(ep_df, aes(x = stage, y = median_i_total,
                   color = lineage_class, shape = mark_type,
                   group = interaction(lineage_san, annotation))) +
  geom_line(alpha = 0.4, linewidth = 0.4) +
  geom_point(size = 1.5, alpha = 0.85) +
  scale_color_manual(values = argelaguet_lineage_class_pal) +
  scale_shape_manual(values = mark_type_shapes) +
  facet_grid(mark_type ~ lineage_class, scales = "free_y") +
  labs(x = "developmental stage", y = expression("median " * i[total]),
       color = "lineage class", shape = "mark type") +
  guides(
    color = guide_legend(override.aes = list(size = 2.5, alpha = 1, linewidth = 1)),
    shape = guide_legend(override.aes = list(size = 2.5))
  ) +
  theme_ng()
```

## jsd enhancers vs promoters across stage

```{r enh_vs_prom_jsd, fig.width = ng_fig_size(length(stage_order), 2)$w, fig.height = ng_fig_size(length(stage_order), 2)$h}
ggplot(ep_df, aes(x = stage, y = median_jsd,
                   color = lineage_class, shape = mark_type,
                   group = interaction(lineage_san, annotation))) +
  geom_line(alpha = 0.4, linewidth = 0.4) +
  geom_point(size = 1.5, alpha = 0.85) +
  scale_color_manual(values = argelaguet_lineage_class_pal) +
  scale_shape_manual(values = mark_type_shapes) +
  facet_grid(mark_type ~ lineage_class, scales = "free_y") +
  labs(x = "developmental stage", y = "median jsd",
       color = "lineage class", shape = "mark type") +
  guides(
    color = guide_legend(override.aes = list(size = 2.5, alpha = 1, linewidth = 1)),
    shape = guide_legend(override.aes = list(size = 2.5))
  ) +
  theme_ng()
```

## Paired scatter: i_total enhancer vs i_total promoter

Each point is one (stage, lineage) group. Groups above the diagonal have higher
within-cell entropy at enhancers than at promoters.

```{r enh_prom_scatter, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h}
ep_wide <- all_grp_meta %>%
  filter(annotation %in% c("Enh E7.5 union", "Promoters"),
         !is.na(lineage_class)) %>%
  select(stage_san, lineage_san, lineage_class, stage, annotation, median_i_total) %>%
  pivot_wider(names_from = annotation, values_from = median_i_total) %>%
  rename(i_total_enh = `Enh E7.5 union`, i_total_prom = Promoters) %>%
  mutate(stage = factor(stage, levels = stage_order))

if (all(c("i_total_enh", "i_total_prom") %in% names(ep_wide))) {
  ggplot(ep_wide, aes(x = i_total_prom, y = i_total_enh,
                       color = lineage_class, shape = stage)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                color = "grey60", linewidth = 0.4) +
    geom_point(size = 2, alpha = 0.8) +
    scale_color_manual(values = argelaguet_lineage_class_pal) +
    scale_shape_manual(values = seq_along(levels(ep_wide$stage))) +
    labs(x = expression("median " * i[total] * " at promoters"),
         y = expression("median " * i[total] * " at E7.5 enhancers (union)"),
         color = "lineage class", shape = "stage") +
    guides(
      color = guide_legend(override.aes = list(size = 3)),
      shape = guide_legend(override.aes = list(size = 3))
    ) +
    theme_ng()
}
```

# Variation across lineages {.tabset .tabset-pills}

SD and range of group medians per annotation, broken down by lineage class.

```{r variation_helpers}
safe_range <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  diff(range(x))
}
```

```{r variation_lineage, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(4, 1)$h}
var_lineage <- all_grp_meta %>%
  filter(!is.na(lineage_class)) %>%
  group_by(annotation, lineage_class) %>%
  summarise(
    i_total_sd = sd(median_i_total, na.rm = TRUE),
    jsd_sd = sd(median_jsd, na.rm = TRUE),
    i_total_range = safe_range(median_i_total),
    jsd_range = safe_range(median_jsd),
    .groups = "drop"
  )

if (any(!is.na(var_lineage$jsd_sd))) {
  ggplot(var_lineage, aes(x = jsd_sd, y = i_total_sd,
                          color = annotation, label = annotation)) +
    geom_point(size = 3) +
    geom_text_repel(size = 3) +
    scale_color_manual(values = argelaguet_annotation_pal) +
    facet_grid(. ~ lineage_class) +
    labs(x = "SD of jsd across groups",
         y = expression("SD of " * i[total] * " across groups")) +
    theme_ng()
}
```

# Driver categorization

For each annotation, we compute the SD of group-level median jsd (across-cell
heterogeneity) and median i_total (within-cell heterogeneity) across all
(stage, lineage) groups. If one SD is at least 1.5x the other, that entropy
component dominates. Annotations where both SDs fall below the 30th percentile
are labelled "neither".

```{r drivers, fig.width = ng_fig_size(3, 3)$w, fig.height = ng_fig_size(3, 3)$h}
driver_df <- categorize_drivers(all_grp_meta, "lineage_class")

cat("Driver table:\n")
print(table(driver_df$driver))

plot_driver_scatter(driver_df,
                    x_label = "SD of jsd across lineage class groups",
                    y_label = expression("SD of " * i[total] * " across lineage class groups"))
```

# Heatmaps {.tabset .tabset-pills}

Rows = lineage (clustered). Columns = annotation. Row bars show lineage class.

```{r heatmap_helpers}
make_heatmap <- function(df, value_col, title,
                          palette = c("navy", "white", "firebrick")) {
  wide <- df %>%
    select(lineage, annotation, lineage_class,
           value = all_of(value_col)) %>%
    filter(!is.na(lineage_class)) %>%
    group_by(lineage, annotation, lineage_class) %>%
    summarise(value = median(value, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = annotation, values_from = value)

  if (nrow(wide) < 2) {
    message("Too few rows for heatmap")
    return(invisible(NULL))
  }

  mat <- wide %>% select(-lineage, -lineage_class) %>% as.matrix()
  rownames(mat) <- wide$lineage

  ## Drop rows / columns that are entirely NA so hclust gets a finite distance.
  keep_rows <- rowSums(!is.na(mat)) >= 2
  keep_cols <- colSums(!is.na(mat)) >= 2
  mat <- mat[keep_rows, keep_cols, drop = FALSE]
  wide <- wide[keep_rows, , drop = FALSE]

  if (nrow(mat) < 2 || ncol(mat) < 2) {
    message("Skipping heatmap: not enough non-NA rows/cols")
    return(invisible(NULL))
  }

  breaks <- c(min(mat, na.rm = TRUE),
              mean(range(mat, na.rm = TRUE)),
              max(mat, na.rm = TRUE))
  if (any(!is.finite(breaks)) || breaks[1] == breaks[3]) {
    message("Skipping heatmap: degenerate data (all-NA or constant matrix)")
    return(invisible(NULL))
  }

  row_ann <- rowAnnotation(
    lineage_class = wide$lineage_class,
    col = list(lineage_class = argelaguet_lineage_class_pal),
    annotation_legend_param = list(
      lineage_class = list(title = "lineage class")
    )
  )

  col_fun <- colorRamp2(breaks, palette)

  Heatmap(mat,
          name = title,
          col = col_fun,
          row_split = wide$lineage_class,
          cluster_rows = TRUE,
          cluster_columns = FALSE,
          show_row_names = TRUE,
          show_column_names = TRUE,
          row_names_gp = gpar(fontsize = 7),
          right_annotation = row_ann,
          column_title = title,
          na_col = "grey90")
}
```

## i_total

```{r heatmap_i_total, fig.width = 10, fig.height = 8}
heat_df_i_total <- i_total_long %>%
  group_by(annotation, stage_san, lineage_san, lineage, lineage_class) %>%
  summarise(median_i_total = median(i_total, na.rm = TRUE), .groups = "drop")

make_heatmap(heat_df_i_total, "median_i_total", "median i_total",
             palette = c("navy", "white", "firebrick"))
```

## jsd

```{r heatmap_jsd, fig.width = 10, fig.height = 8}
heat_df_jsd <- jsd_long %>%
  group_by(annotation, stage_san, lineage_san, lineage, lineage_class) %>%
  summarise(median_jsd = median(median_jsd, na.rm = TRUE), .groups = "drop")

make_heatmap(heat_df_jsd, "median_jsd", "median jsd",
             palette = c("navy", "white", "firebrick"))
```

# Cell-level UMAP {.tabset .tabset-pills}

```{r umap_helpers}

plot_umap <- function(df, color_col, title, subtitle = NULL) {
  if (is.null(df) || nrow(df) == 0) {
    message("No UMAP result for: ", title)
    return(invisible(NULL))
  }
  ggplot(df %>% filter(!is.na(.data[[color_col]])),
         aes(x = UMAP1, y = UMAP2, color = .data[[color_col]])) +
    geom_point(size = 1.5, alpha = 0.7) +
    scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
    labs(title = title, subtitle = subtitle,
         color = color_col) +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme_ng()
}
```

```{r cell_umap_load}
## Build a wide cells x annotation matrix from amet's per-cell-per-feature
## i_total.
cells_i_total_long <- as.data.frame(feat_cf)[, c("cell_id", "stage_san",
                                                 "lineage_san", "annotation",
                                                 "i_total")] %>%
  rename(value = i_total) %>%
  group_by(cell_id, stage_san, lineage_san, annotation) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop")
cells_i_total_long$annotation <- factor(cells_i_total_long$annotation,
                                        levels = names(ann_labels),
                                        labels = unname(ann_labels))

cells_meth_long <- as.data.frame(feat_cf)[, c("cell_id", "stage_san",
                                               "lineage_san", "annotation",
                                               "mean_meth")] %>%
  rename(value = mean_meth) %>%
  group_by(cell_id, stage_san, lineage_san, annotation) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop")
cells_meth_long$annotation <- factor(cells_meth_long$annotation,
                                      levels = names(ann_labels),
                                      labels = unname(ann_labels))

meta_cols_cell <- c("stage_san", "lineage_san", "cell_id",
                    "stage", "lineage", "lineage_class")

cells_i_total_wide <- pivot_wider(cells_i_total_long,
                                  id_cols = c(cell_id, stage_san, lineage_san),
                                  names_from = annotation,
                                  values_from = value) %>%
  left_join(meta_grp %>% select(stage_san, lineage_san, stage, lineage,
                                 lineage_class),
            by = c("stage_san", "lineage_san"))

cells_meth_wide <- pivot_wider(cells_meth_long,
                                id_cols = c(cell_id, stage_san, lineage_san),
                                names_from = annotation,
                                values_from = value) %>%
  left_join(meta_grp %>% select(stage_san, lineage_san, stage, lineage,
                                 lineage_class),
            by = c("stage_san", "lineage_san"))

saveRDS(list(cells_i_total_wide = cells_i_total_wide,
             cells_meth_wide = cells_meth_wide),
        "argelaguet_cell_matrices.rds")
```

## Cell-level: i_total across annotations

```{r cell_umap_i_total, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h}
umap_cell_i_total <- run_umap_wide(cells_i_total_wide, meta_cols_cell,
                                   n_neighbors = 15L)
saveRDS(umap_cell_i_total, "argelaguet_umap_cell_i_total.rds")

plot_umap(umap_cell_i_total, "lineage_class",
          title = "Cell-level UMAP - i_total features",
          subtitle = "per-cell i_total per annotation as feature vector")
```

## Cell-level: i_total colored by stage

```{r cell_umap_i_total_stage, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h}
plot_umap(umap_cell_i_total, "stage",
          title = "Cell-level UMAP - i_total features",
          subtitle = "colored by developmental stage")
```

## Cell-level: avg methylation across annotations

```{r cell_umap_meth, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h}
umap_cell_meth <- run_umap_wide(cells_meth_wide, meta_cols_cell,
                                 n_neighbors = 15L)
saveRDS(umap_cell_meth, "argelaguet_umap_cell_meth.rds")

plot_umap(umap_cell_meth, "lineage_class",
          title = "Cell-level UMAP - avg. methylation features",
          subtitle = "per-cell avg. methylation per annotation as feature vector")
```

## Group-level: jsd across annotations

```{r group_umap_jsd, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h}
jsd_wide_grp <- all_grp_meta %>%
  filter(!is.na(lineage_class)) %>%
  select(stage_san, lineage_san, stage, lineage, lineage_class,
         annotation, median_jsd) %>%
  pivot_wider(names_from = annotation, values_from = median_jsd)

umap_grp_jsd <- run_umap_wide(
  jsd_wide_grp,
  meta_cols = c("stage_san", "lineage_san", "stage", "lineage", "lineage_class"),
  n_neighbors = 5L
)
saveRDS(umap_grp_jsd, "argelaguet_umap_grp_jsd.rds")

plot_umap(umap_grp_jsd, "lineage_class",
          title = "Group-level UMAP - jsd features",
          subtitle = "median jsd per annotation as feature vector")
```

```{r group_umap_jsd_stage, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h}
plot_umap(umap_grp_jsd, "stage",
          title = "Group-level UMAP - jsd features",
          subtitle = "colored by developmental stage")
```
