---
title: "amet on Argelaguet 2019 mouse gastrulation scNMT-seq"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    number_sections: true
    df_print: default
params:
  feat_cell_feature: ""
  feat_feature: ""
  feat_bed: ""
  win_cell_feature: ""
  win_feature: ""
  win_bed: ""
  manifest: ""
  out_dir: ""
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 5, fig.height = 4)

suppressPackageStartupMessages({
    library(data.table)
    library(ggplot2)
    library(patchwork)
    library(dplyr)
    library(tidyr)
})

repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))

panels_dir <- file.path(params$out_dir, "panels")
dir.create(panels_dir, showWarnings = FALSE, recursive = TRUE)
```

# Inputs and design

Per-feature analysis uses gastrulation-specific BEDs (H3K27ac distal peaks for E7.5 ectoderm / endoderm / mesoderm and ESC p300). Tiled-genome analysis covers chr19 in fixed-size windows for cell-level summaries and embeddings.

Score mapping:
amet `i_total` (within-cell regularity) replaces yamet's adjS.
amet `jsd` (across-cell heterogeneity) replaces yamet's adjH.

```{r load}
man <- fread(params$manifest)
feat_cf <- fread(params$feat_cell_feature)
feat_fe <- fread(params$feat_feature)
win_cf <- fread(params$win_cell_feature)
win_fe <- fread(params$win_feature)
feat_bed <- fread(params$feat_bed, header = FALSE,
                  col.names = c("chrom", "start", "end", "feature_id"))

# feature_id is <annotation>_<index>; strip trailing _<digits> to recover annotation.
recover_annotation <- function(fid) sub("_\\d+$", "", fid)
feat_bed[, annotation := recover_annotation(feature_id)]

meta_cols <- c("cell_id", "group", "stage", "embryo")
man <- man[, ..meta_cols]

feat_cf[, annotation := recover_annotation(feature_id)]
feat_cf <- merge(feat_cf, man[, .(cell_id, stage, embryo)],
                 by = "cell_id", all.x = TRUE)
feat_fe[, annotation := recover_annotation(feature_id)]
win_cf <- merge(win_cf, man[, .(cell_id, stage, embryo)],
                by = "cell_id", all.x = TRUE)
```

Sample sizes:

```{r sizes}
cat("cells:", nrow(man), "\n")
cat("groups:", uniqueN(man$group), "\n")
cat("stages:", paste(sort(unique(man$stage)), collapse = ", "), "\n")
cat("feature regions:", uniqueN(feat_bed$feature_id), "across",
    uniqueN(feat_bed$annotation), "annotation classes\n")
cat("genome windows:", uniqueN(win_cf$feature_id), "\n")
```

```{r per_group}
man[, .N, by = group]
```

# Per-feature analysis (regulatory regions)

## i_total per lineage and feature class

```{r itotal_feature, fig.width = 7, fig.height = 4.5}
fcf <- feat_cf[!is.na(i_total)]
p_itotal_box <- ggplot(fcf, aes(x = group, y = i_total)) +
    geom_boxplot(aes(fill = group), outlier.size = 0.3, linewidth = 0.3,
                 alpha = 0.7, show.legend = FALSE) +
    facet_wrap(~ annotation, nrow = 1) +
    labs(x = NULL, y = "i_total") +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

save_eval(p_itotal_box,
          fcf[, .(cell_id, group, stage, annotation, feature_id, i_total)],
          file.path(panels_dir, "feat_itotal_by_group"),
          width_mm = 160, height_mm = 70)
p_itotal_box
```

## i_total vs mean methylation (p-decoupling check)

```{r itotal_vs_meth, fig.width = 7, fig.height = 4.5}
p_decouple <- ggplot(fcf, aes(x = mean_meth, y = i_total, colour = group)) +
    geom_point(size = 0.6, alpha = 0.6) +
    facet_wrap(~ annotation, nrow = 1) +
    labs(x = "mean methylation", y = "i_total") +
    theme_ng()
save_eval(p_decouple,
          fcf[, .(cell_id, group, annotation, mean_meth, i_total)],
          file.path(panels_dir, "feat_itotal_vs_meth"),
          width_mm = 160, height_mm = 70)
p_decouple
```

## i_total distribution per stage and lineage

```{r itotal_by_stage, fig.width = 7, fig.height = 4.5}
fcf_pl <- fcf[!is.na(stage)]
if (nrow(fcf_pl) > 0) {
    p_stage <- ggplot(fcf_pl, aes(x = stage, y = i_total, fill = group)) +
        geom_boxplot(outlier.size = 0.3, linewidth = 0.3, alpha = 0.6) +
        facet_wrap(~ annotation, nrow = 1) +
        labs(x = NULL, y = "i_total") +
        theme_ng() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
    save_eval(p_stage,
              fcf_pl[, .(cell_id, group, stage, annotation, i_total)],
              file.path(panels_dir, "feat_itotal_by_stage"),
              width_mm = 160, height_mm = 70)
    print(p_stage)
} else {
    cat("No stage column -- skipping\n")
}
```

## jsd per (annotation, group)

```{r jsd_box, fig.width = 7, fig.height = 4.5}
ffe <- feat_fe[!is.na(jsd)]
if (nrow(ffe) > 0) {
    p_jsd <- ggplot(ffe, aes(x = group, y = jsd, fill = group)) +
        geom_boxplot(outlier.size = 0.3, linewidth = 0.3, alpha = 0.7,
                     show.legend = FALSE) +
        facet_wrap(~ annotation, nrow = 1) +
        labs(x = NULL, y = "jsd") +
        theme_ng() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
    save_eval(p_jsd,
              ffe[, .(annotation, feature_id, group, n_cells, jsd)],
              file.path(panels_dir, "feat_jsd_by_group"),
              width_mm = 160, height_mm = 70)
    print(p_jsd)
} else {
    cat("All jsd values are NA at the configured min-cells-per-group threshold.\n")
}
```

## i_total vs jsd at the per-feature level

```{r itotal_vs_jsd, fig.width = 5, fig.height = 4.5}
fe_summary <- fcf[, .(median_i_total = median(i_total, na.rm = TRUE)),
                  by = .(feature_id, annotation, group)]
fe_summary <- merge(fe_summary,
                    feat_fe[, .(feature_id, group, jsd, n_cells)],
                    by = c("feature_id", "group"), all = FALSE)
fe_summary <- fe_summary[!is.na(jsd) & !is.na(median_i_total)]
if (nrow(fe_summary) > 0) {
    p_assembly <- ggplot(fe_summary,
                         aes(x = jsd, y = median_i_total, colour = annotation)) +
        geom_point(size = 0.7, alpha = 0.8) +
        labs(x = "jsd (per feature)",
             y = "median i_total (per feature, per group)",
             colour = "annotation") +
        theme_ng()
    save_eval(p_assembly, fe_summary,
              file.path(panels_dir, "feat_itotal_vs_jsd"),
              width_mm = 80, height_mm = 70)
    print(p_assembly)
} else {
    cat("Not enough overlapping non-NA jsd and i_total to plot.\n")
}
```

# Window-level analysis (chr19 tiles)

## Mean methylation per cell, by stage

```{r win_mean_meth, fig.width = 5, fig.height = 4.5}
wcf <- win_cf[!is.na(mean_meth)]
cell_meth <- wcf[, .(mean_meth = mean(mean_meth, na.rm = TRUE),
                     n_windows = .N), by = .(cell_id, group, stage)]
if (nrow(cell_meth) > 0 && uniqueN(cell_meth$stage) > 0) {
    p_wmeth <- ggplot(cell_meth[!is.na(stage)],
                      aes(x = stage, y = mean_meth, fill = stage)) +
        geom_boxplot(outlier.size = 0.3, linewidth = 0.3,
                     alpha = 0.7, show.legend = FALSE) +
        labs(x = NULL, y = "mean methylation per cell") +
        theme_ng() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
    save_eval(p_wmeth, cell_meth,
              file.path(panels_dir, "win_meanmeth_by_stage"),
              width_mm = 80, height_mm = 70)
    print(p_wmeth)
}
```

## i_total per cell, by lineage

```{r win_itotal_cell, fig.width = 5, fig.height = 4.5}
wcf_keep <- win_cf[!is.na(i_total)]
cell_it <- wcf_keep[, .(mean_i_total = mean(i_total, na.rm = TRUE),
                        n_windows = .N), by = .(cell_id, group, stage)]
if (nrow(cell_it) > 0) {
    p_wit <- ggplot(cell_it, aes(x = group, y = mean_i_total, fill = group)) +
        geom_boxplot(outlier.size = 0.3, linewidth = 0.3,
                     alpha = 0.7, show.legend = FALSE) +
        labs(x = NULL, y = "mean i_total per cell") +
        theme_ng() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
    save_eval(p_wit, cell_it,
              file.path(panels_dir, "win_itotal_per_cell"),
              width_mm = 80, height_mm = 70)
    print(p_wit)
}
```

## i_total vs methylation per cell

```{r win_itotal_vs_meth, fig.width = 5, fig.height = 4.5}
cell_full <- merge(cell_meth, cell_it[, .(cell_id, mean_i_total)],
                   by = "cell_id", all = TRUE)
cell_full <- cell_full[!is.na(mean_i_total) & !is.na(mean_meth)]
if (nrow(cell_full) > 0) {
    p_wcorr <- ggplot(cell_full,
                      aes(x = mean_meth, y = mean_i_total, colour = group)) +
        geom_point(size = 0.7, alpha = 0.8) +
        labs(x = "mean methylation per cell",
             y = "mean i_total per cell") +
        theme_ng()
    save_eval(p_wcorr, cell_full,
              file.path(panels_dir, "win_itotal_vs_meth"),
              width_mm = 80, height_mm = 70)
    print(p_wcorr)
}
```

## Cell embedding from window i_total matrix

PCA of the cell-by-window i_total matrix. UMAP is added when there are at least 10 windows with sufficient cell coverage. Below that threshold the matrix is too low-rank for UMAP to be meaningful.

```{r embedding, fig.width = 7, fig.height = 4.5}
mat <- dcast(wcf_keep, cell_id ~ feature_id, value.var = "i_total",
             fun.aggregate = mean)
mat_cells <- mat$cell_id
mat <- as.matrix(mat[, -1])
rownames(mat) <- mat_cells
mat <- mat[, colSums(!is.na(mat)) >= 0.5 * nrow(mat), drop = FALSE]
mat[is.na(mat)] <- 0  # mean impute (only after dropping low-coverage windows)

if (ncol(mat) >= 2 && nrow(mat) >= 3) {
    pcs <- prcomp(mat, scale. = FALSE)
    emb <- data.table(cell_id = rownames(mat),
                      PC1 = pcs$x[, 1],
                      PC2 = pcs$x[, 2])
    emb <- merge(emb, man[, .(cell_id, group, stage)], by = "cell_id")

    p_pca_lin <- ggplot(emb, aes(x = PC1, y = PC2, colour = group)) +
        geom_point(size = 1.5, alpha = 0.9) +
        labs(title = "PCA, coloured by lineage", colour = "lineage") +
        theme_ng()
    p_pca_stg <- ggplot(emb[!is.na(stage)],
                        aes(x = PC1, y = PC2, colour = stage)) +
        geom_point(size = 1.5, alpha = 0.9) +
        labs(title = "PCA, coloured by stage", colour = "stage") +
        theme_ng()
    save_eval(p_pca_lin, emb, file.path(panels_dir, "win_pca_lineage"),
              width_mm = 80, height_mm = 70)
    save_eval(p_pca_stg, emb[!is.na(stage)],
              file.path(panels_dir, "win_pca_stage"),
              width_mm = 80, height_mm = 70)
    print(p_pca_lin + p_pca_stg)
} else {
    cat("Not enough windows with coverage for PCA. Have:", ncol(mat), "windows,",
        nrow(mat), "cells.\n")
}
```

# Combined paper figure

```{r paper_figure, fig.width = 9, fig.height = 6}
panels <- list()
if (exists("p_itotal_box")) panels$itotal_feat <- p_itotal_box +
    theme(axis.text.x = element_text(size = 6))
if (exists("p_jsd")) panels$jsd_feat <- p_jsd +
    theme(axis.text.x = element_text(size = 6))
if (exists("p_wit")) panels$itotal_win <- p_wit +
    theme(axis.text.x = element_text(size = 6))
if (exists("p_pca_lin")) panels$pca <- p_pca_lin

if (length(panels) >= 2) {
    combined <- Reduce(`+`, panels) + plot_layout(ncol = 2)
    ggsave(file.path(panels_dir, "paper_figure.pdf"),
           combined, width = 220, height = 150, units = "mm",
           device = grDevices::cairo_pdf)
    combined
}
```

# Session

```{r session}
sessionInfo()
```
