---
title: "CRC by feature (no windows)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Izaskun Mallona"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
params:
  features_dir: ""
  windows_dir: ""
  win_bed: ""
  manifest: ""
  out_dir: ""
---

```{r, setup}
suppressPackageStartupMessages({
  library(data.table)
  library(readr)
  library(ggplot2)
  library(knitr)
  library(ComplexHeatmap)
  library(viridis)
  library(BiocParallel)
  library(dplyr)
  library(tidyr)
  library(ggrepel)
})

repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))
source(file.path(repo_root, "workflow", "scripts", "palettes.R"))
source(file.path(repo_root, "workflow", "scripts", "driver_utils.R"))
param <- SerialParam()
```

```{r}
opts_chunk$set(
  fig.width = 5,
  fig.height = 5,
  cache = FALSE,
  include = TRUE,
  fig.path = "crc_plots/",
  dev = c("png", "svg"),
  dpi = 500,
  cache.lazy = FALSE,
  warning = TRUE,
  message = TRUE
)
```

These are amet feature TSVs (per (subcat, cat, patient, location) combo). Per-cell `i_total` from cell_feature.tsv.gz replaces yamet's per-cell sampen_norm; per-feature `jsd` from feature.tsv.gz replaces yamet's per-group adjH.

```{r, import_short_reports}
## Filename parsers (mirror yamet/crc.Rmd):
##   feat: <subcat>_<cat>_<patient>_<location>.{cell_feature,feature}.tsv.gz
get_subcat   <- function(fn) sub("^([^_]+)_[^_]+_[^_]+_[^_]+\\..*$", "\\1", fn)
get_cat      <- function(fn) sub("^[^_]+_([^_]+)_[^_]+_[^_]+\\..*$", "\\1", fn)
get_patient  <- function(fn) sub("^.*_([^_]+)_[^_]+\\..*$",          "\\1", fn)
get_location <- function(fn) sub("^.*_([^_.]+)\\.[^_]*$",            "\\1", fn)

read_combo_feat <- function(fp, value_kind) {
  fn <- basename(fp)
  base <- sub(paste0("\\.", value_kind, "\\.tsv\\.gz$"), "", fn)
  bx <- paste0(base, ".x")
  sc <- get_subcat(bx); ct <- get_cat(bx)
  pt <- get_patient(bx); lc <- get_location(bx)
  dt <- fread(fp)
  dt[, subcat   := sc]
  dt[, cat      := ct]
  dt[, patient  := pt]
  dt[, location := lc]
  dt
}

cf_files <- list.files(params$features_dir,
                       pattern = "\\.cell_feature\\.tsv\\.gz$",
                       full.names = TRUE)
fe_files <- list.files(params$features_dir,
                       pattern = "\\.feature\\.tsv\\.gz$",
                       full.names = TRUE)
fe_files <- fe_files[!grepl("\\.cell_feature\\.tsv\\.gz$", fe_files)]

feat_cf <- rbindlist(lapply(cf_files, read_combo_feat, value_kind = "cell_feature"),
                    fill = TRUE)
feat_fe <- rbindlist(lapply(fe_files, read_combo_feat, value_kind = "feature"),
                    fill = TRUE)

cat("loaded", length(cf_files), "cell_feature TSVs and",
    length(fe_files), "feature TSVs from", params$features_dir, "\n")

## yamet's `files` long table holds one row per (cell, annotation, patient,
## location) with columns sampen + avg_meth. Rebuild from amet's per-cell
## i_total (renamed to sampen) and mean_meth (renamed to avg_meth) so the
## downstream chunks copy verbatim from yamet.
files <- as.data.frame(feat_cf)
files$annotation <- files$subcat
files$sampen     <- files$i_total
files$avg_meth   <- files$mean_meth
files <- files[, c("sampen", "avg_meth", "annotation", "patient", "location")]

## prettifies and sorts categorical vars
harmonize_levels <- function(df) {
    pretty_hmm <- c(
        "0_Enhancer"        = "0 Enh.",
        "1_Transcribed"     = "1 Transc. ",
        "2_Enhancer"        = "2 Enh.",
        "3_Quiescent"       = "3 Quies.",
        "4_Transcribed"     = "4 Transc.",
        "5_RegPermissive"   = "5 Reg. perm.",
        "6_LowConfidence"   = "6 Low conf.",
        "7_RegPermissive"   = "7 Reg. perm.",
        "8_Quiescent"       = "8 Quies.",
        "9_ConstitutiveHet" = "9 Cons. het.",
        "10_Quiescent"      = "10 Quies.",
        "11_Promoter"       = "11 Prom.",
        "12_Promoter"       = "12 Prom.",
        "13_ConstitutiveHet"= "13 Cons. het"
    )

    pretty_other <- c(
        "cpgIslandExt"      = "CpGi",
        "H3K4me3"           = "H3K4me3",
        "genes"             = "Genes",
        "H3K27me3"          = "H3K27me3",
        "H3K9me3"           = "H3K9me3",
        "laminb1"           = "Lamin B1",
        "pmds"              = "PMDs",
        "lines"             = "LINEs",
        "sines"             = "SINEs",
        "hmds"              = "HMDs",
        "crc01_lost_scna"   = "SCNA loss",
        "crc01_nc_scna"     = "SCNA neutral",
        "crc01_gain_scna"   = "SCNA gain")

    all_levels <- c(pretty_hmm, pretty_other)

    missing <- setdiff(unique(df$annotation), names(all_levels))
    if (length(missing) > 0) {
        stop(missing)
    }


  df$annotation <- factor(df$annotation,
                          levels = names(all_levels),
                          labels = unname(all_levels))

  df$patient  <- as.factor(df$patient)
  df$location <- factor(df$location, levels = c("NC","PT","LN","ML","MP","MO"))

  df
}

files <- harmonize_levels(files)
table(files$annotation, useNA = 'always')

```

<!-- Section dropped: amet ships only methylation-stable i_total; the raw sampen scatter and per-patient/per-location plots in yamet/crc.Rmd lines 215-417 (Bulk unnormalized sampen tabset) are gone. -->

# adjS - methylation-adjusted sample entropy {.tabset .tabset-pills}

That is, not by locus but by cell norm sampEns (i_total) (that are calculated according to some regionset anyway). amet's i_total is the single per-cell-per-feature within-cell mutual information score; the same data already loaded above are used here directly.

```{r, import_short_reports_norm}
## In yamet there is a separate read pass for the sampen_norm column. amet's
## i_total is that score, already in `files`, so just reuse the same data
## frame here.
```

## Scatters by location and annotation

```{r}
ggplot(
  files,
  aes(y = sampen, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation) +
  labs(
    x = "avg. methylation",
    y = expression(i[total])
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Scatters by location and annotation and patient

```{r, fig.width = 8, fig.height = 20}
ggplot(
  files,
  aes(y = sampen, x = avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_grid(annotation ~ patient) +
  labs(
    x = "avg. methylation",
    y = expression(i[total])
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Scatters by location and annotation and patient (shorter)

```{r, fig.width = 8, fig.height = 6}
ggplot(
  subset(files, annotation %in% c("Lamin B1", "H3K27me3", "H3K9me3")),
  aes(y = sampen, x = avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_grid(annotation ~ patient) +
  labs(
    x = "avg. methylation",
    y = expression(i[total])
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```


## Segmentations {.tabset .tabset-pills}

### Scatters by location and annotation

```{r}
ggplot(
    subset(files, grepl("^[0-9]{1,2}", annotation)),
  aes(y = sampen, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "avg. methylation",
    y = expression(i[total])
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```


## Nonsegmentations {.tabset .tabset-pills}

### Scatters by location and annotation

```{r}
ggplot(
    subset(files, !grepl("^[0-9]{1,2}", annotation)),
  aes(y = sampen, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "avg. methylation",
    y = expression(i[total])
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Others  {.tabset .tabset-pills}

### i_total boxes by location and annotation - crc01

```{r, fig.width = 9, fig.height = 5}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = sampen, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = expression(i[total])) +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### meth boxes by location and annotation - crc01

```{r, fig.width = 9, fig.height = 5}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = avg_meth, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "avg. methylation") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### Range plots - by compartment

```{r, fig.width = 8, fig.height = 8}
ranges <- as.data.frame(files %>%
   filter(patient == "CRC01") %>%
   group_by(compartment = annotation, location) %>%
   summarise(
     sampen_mean = mean(sampen, na.rm = TRUE),
     sampen_min  = min(sampen, na.rm = TRUE),
     sampen_max  = max(sampen, na.rm = TRUE),
     meth_mean   = mean(avg_meth, na.rm = TRUE),
     meth_min    = min(avg_meth, na.rm = TRUE),
     meth_max    = max(avg_meth, na.rm = TRUE)))

ggplot(ranges, aes(color = location)) +
  ## horizontal segment: range of methylations for that location/compartment
  geom_segment(
    aes(x = meth_min, xend = meth_max, y = sampen_mean, yend = sampen_mean),
    linewidth = 0.6, alpha = 1
  ) +
  # vertical segment: range of i_total for that location/compartment
  geom_segment(
    aes(x = meth_mean, xend = meth_mean, y = sampen_min, yend = sampen_max),
    linewidth = 0.6, alpha = 1
  ) +
  # center point
  geom_point(
    aes(x = meth_mean, y = sampen_mean),
    shape = 3, size = 1.2
  ) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "DNA methylation", y = expression(i[total])) +
  facet_wrap(~ compartment) +
  theme_ng()

```

```{r}
adjsampens <- files
```

## adjS within methylation bins {.tabset .tabset-pills}

Boxplots of i_total stratified by 0.1-wide average methylation bins, colored by
location. Persistent differences within bins show that i_total captures
heterogeneity beyond average methylation level.

### All annotations

```{r adjS_meth_bins_all, fig.width = ng_fig_size(5, 6)$w, fig.height = ng_fig_size(5, 6)$h}
adjsampens %>%
  mutate(meth_bin = cut(avg_meth, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>%
  filter(!is.na(meth_bin), !is.na(sampen)) %>%
  ggplot(aes(x = meth_bin, y = sampen, fill = location)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "avg. methylation bin", y = expression(i[total])) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### CRC01 only

```{r adjS_meth_bins_crc01, fig.width = ng_fig_size(5, 6)$w, fig.height = ng_fig_size(5, 6)$h}
adjsampens %>%
  filter(patient == "CRC01") %>%
  mutate(meth_bin = cut(avg_meth, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>%
  filter(!is.na(meth_bin), !is.na(sampen)) %>%
  ggplot(aes(x = meth_bin, y = sampen, fill = location)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "avg. methylation bin", y = expression(i[total])) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

<!-- Section dropped: amet has no raw Shannon entropy. yamet/crc.Rmd lines 694-804 (Unnormalized Shannon's MEDIANS DOWNSAMPLED tabset) is gone. -->

# Meth-normalized Shannons (the amet way) MEDIANS {.tabset .tabset-pills}

amet's per-feature `jsd` from feature.tsv.gz replaces yamet's per-group adjH from `.norm.det.out.gz`. yamet's adjshannons holds one row per (annotation, patient, location); amet's feature TSV has many rows per group, so aggregate to the same shape with `median(jsd)`.

```{r, eval = TRUE}
adjshannons <- as.data.frame(
  as.data.table(feat_fe)[
    , .(median_shannon  = median(jsd, na.rm = TRUE),
        median_avg_meth = median(mean_meth_mean, na.rm = TRUE)),
    by = .(annotation = subcat, patient, location)
  ]
)

adjshannons <- harmonize_levels(adjshannons)
```

## Scatters by location and annotation

```{r}
ggplot(
  adjshannons,
  aes(y = median_shannon, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(adjshannons$patient)) +
  facet_wrap(~annotation) +
  labs(
    x = "median average methylation",
    y = "median jsd"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Scatters by location and annotation and patient

```{r}
ggplot(
  adjshannons,
  aes(y = median_shannon, x = median_avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  labs(
    x = "median average methylation",
    y = "median jsd"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```



## Segmentations {.tabset .tabset-pills}

### Scatters by location and annotation

```{r}
ggplot(
    subset(adjshannons, grepl("^[0-9]{1,2}", annotation)),
  aes(y = median_shannon, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(adjshannons$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "median avg. methylation",
    y = "median jsd"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```


## Nonsegmentations {.tabset .tabset-pills}

### Scatters by location and annotation

```{r}
ggplot(
    subset(adjshannons, !grepl("^[0-9]{1,2}", annotation)),
  aes(y = median_shannon, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(adjshannons$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "median avg. methylation",
    y = "median jsd"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

# Attempt to assemble these all

amet ships only i_total (per cell) and jsd (per group), not raw S/H. So `sampens` and `shannons` (the unnormalized variants) do not exist; only `adjsampens` and `adjshannons` carry data.

```{r}
str(adjshannons)
str(adjsampens)
```

```{r save_for_emanuel}
saveRDS(list(adjshannons = adjshannons,
             adjsampens = adjsampens),
        file = "crc_entropy_summaries.rds")
```


```{r summary_heatmap, fig.width = 10, fig.height = 8}
library(ComplexHeatmap)
library(circlize)

make_hm_mat <- function(df, value_col) {
  wide <- df %>%
    group_by(annotation, location) %>%
    summarise(v = median(.data[[value_col]], na.rm = TRUE), .groups = "drop") %>%
    tidyr::pivot_wider(names_from = location, values_from = v)
  mat <- as.matrix(wide[, -1])
  rownames(mat) <- wide$annotation
  mat
}

hm_adjS <- make_hm_mat(adjsampens, "sampen")
hm_adjH <- make_hm_mat(adjshannons, "median_shannon")

col_adjS <- colorRamp2(
  quantile(hm_adjS, c(0.05, 0.5, 0.95), na.rm = TRUE),
  c("#2166AC", "white", "#B2182B"))
col_adjH <- colorRamp2(
  quantile(hm_adjH, c(0.05, 0.5, 0.95), na.rm = TRUE),
  c("#4DAC26", "white", "#D01C8B"))

draw(
  Heatmap(hm_adjS, name = "i_total", col = col_adjS,
          cluster_columns = FALSE,
          row_names_gp = gpar(fontsize = 7),
          column_names_gp = gpar(fontsize = 7)) +
  Heatmap(hm_adjH, name = "jsd", col = col_adjH,
          cluster_rows = FALSE, cluster_columns = FALSE,
          show_row_names = FALSE,
          column_names_gp = gpar(fontsize = 7)),
  column_title = "Median i_total and jsd per annotation x location"
)
```



```{r, fig.width = 9, fig.height = 12}

library(dplyr)

adj_df <- adjsampens %>%
  left_join(adjshannons,
            by = c("annotation", "patient", "location"),
            suffix = c("_sampen", "_shannon"))

head(adj_df)
ggplot(
    adj_df,
    aes(x = median_shannon, y = sampen,
        color = location, shape = patient)
) +
    scale_shape_manual(values = 1:nlevels(adj_df$patient)) +
    geom_point(size = 0.8) +
    facet_wrap(~annotation, ncol = 4) +
    labs(
        x = "median jsd",
        y = expression(i[total])
    ) +
  theme_ng()

```

<!-- Section dropped: amet has no raw Shannon entropy / unnormalized sampen, so the second scatter (sampens vs shannons) in yamet/crc.Rmd lines 1019-1039 is gone. -->

```{r}
variation <- adj_df %>%
    group_by(annotation) %>%
    summarise(
        shannon_range = max(median_shannon, na.rm = TRUE) -
            min(median_shannon, na.rm = TRUE),
        sampen_range  = max(sampen, na.rm = TRUE) -
            min(sampen, na.rm = TRUE),
        shannon_sd = sd(median_shannon, na.rm = TRUE),
        sampen_sd  = sd(sampen, na.rm = TRUE)
    )

head(variation)

ggplot(variation, aes(x = shannon_range, y = sampen_range, label = annotation)) +
    geom_point() +
    ggrepel::geom_text_repel() +
    labs(
        x = "jsd range aggregating all biopsy sites",
        y = expression(i[total] * " range aggregating all biopsy sites"),
        ) +
  theme_ng()


variation_loc <- adj_df %>%
    group_by(annotation, location) %>%
    summarise(
        shannon_median = median(median_shannon, na.rm = TRUE),
        sampen_median  = median(sampen, na.rm = TRUE),
        shannon_sd = sd(median_shannon, na.rm = TRUE),
        sampen_sd  = sd(sampen, na.rm = TRUE),
        shannon_range = max(median_shannon, na.rm = TRUE) -
            min(median_shannon, na.rm = TRUE),
        sampen_range  = max(sampen, na.rm = TRUE) -
            min(sampen, na.rm = TRUE)
    )

ggplot(variation_loc,
       aes(x = shannon_median, y = sampen_median,
           color = location)) +
    geom_point() +
    facet_wrap(~annotation) +
    theme_ng() +
    labs(
         x = "median jsd across biopsy sites",
         y = expression("median " * i[total] * " across biopsy sites"))

ggplot(variation_loc,
       aes(x = shannon_sd, y = sampen_sd,
           color = location)) +
    geom_point() +
    facet_wrap(~annotation) +
    theme_ng() +
    labs(
         x = "sd jsd across biopsy sites",
         y = expression("sd " * i[total] * " across biopsy sites"))


ggplot(variation_loc,
       aes(x = shannon_range, y = sampen_range,
           color = location)) +
    geom_point() +
    facet_wrap(~annotation) +
    theme_ng() +
    labs(
         x = "jsd range across biopsy sites",
         y = expression(i[total] * " range across biopsy sites"))


```


# Driver categorization

For each annotation, we compute the SD of group-level median jsd (across-cell
heterogeneity) and median i_total (within-cell heterogeneity) across biopsy
locations. If one SD is at least 1.5x the other, that entropy component
dominates. Annotations where both SDs fall below the 30th percentile are
labelled "neither".

```{r drivers, fig.width = 5, fig.height = 5}
crc_grp_for_driver <- adj_df %>%
  rename(median_adjH = median_shannon, median_adjS = sampen)

driver_df <- categorize_drivers(crc_grp_for_driver, "location")

cat("Driver table:\n")
print(table(driver_df$driver))

saveRDS(driver_df, file = "crc_driver_sd_range.rds")

plot_driver_scatter(driver_df,
                    x_label = "SD of jsd across biopsy locations",
                    y_label = expression("SD of " * i[total] * " across biopsy locations"))
```


# Session

```{r, cache = FALSE}
sessionInfo()
```
