---
title: "CRC by feature (no windows)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Izaskun Mallona"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
params:
  features_dir: ""
  windows_dir: ""
  win_bed: ""
  manifest: ""
  out_dir: ""
  log_path: ""
  threads: 1
  i_max_lag: 3
---

```{r logging_early, include = FALSE}
source(Sys.getenv("AMET_RENDER_HELPERS"))
param <- amet_setup_render_logging(params$log_path, params$threads)
```

```{r, setup}
suppressPackageStartupMessages({
  library(data.table)
  library(readr)
  library(ggplot2)
  library(knitr)
  library(ComplexHeatmap)
  library(viridis)
  library(BiocParallel)
  library(dplyr)
  library(tidyr)
  library(ggrepel)
})

repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))
source(file.path(repo_root, "workflow", "scripts", "palettes.R"))
source(file.path(repo_root, "workflow", "scripts", "driver_utils.R"))
```

```{r}
opts_chunk$set(
  fig.width = 5,
  fig.height = 5,
  cache = FALSE,
  error = TRUE,
  include = TRUE,
  fig.path = "crc_plots/",
  dev = c("png", "svg"),
  dpi = 500,
  cache.lazy = FALSE,
  warning = TRUE,
  message = TRUE
)
```

These are amet feature TSVs (per (subcat, cat, patient, location) combo). Per-cell `i_total` from cell_feature.tsv.gz captures within-cell heterogeneity; per-feature `jsd` from feature.tsv.gz captures across-cell heterogeneity.

```{r, import_short_reports}
## Filename layout: <subcat>_<cat>_<patient>_<location>.{cell_feature,feature}.tsv.gz
## subcat may itself contain underscores (e.g., 0_Enhancer, crc01_nc_scna), so
## we pin location, patient and cat to the last three tokens and let subcat
## absorb the rest.
parse_combo_base <- function(base) {
  parts <- strsplit(base, "_", fixed = TRUE)[[1]]
  n <- length(parts)
  stopifnot(n >= 4)
  list(
    subcat = paste(parts[seq_len(n - 3)], collapse = "_"),
    cat = parts[n - 2],
    patient = parts[n - 1],
    location = parts[n]
  )
}

## Aggregate per cell at read time: feat_cf goes from O(cells * features
## * strata) rows to O(cells * strata). Downstream plots average per cell.
## amet can emit a near-empty .feature.tsv.gz when a combo has no groups
## passing min_cells_per_group; the cell-level file may also be tiny. Guard
## both readers so a dropped combo doesn't kill the whole report.
read_combo_cf <- function(fp) {
  fn <- basename(fp)
  base <- sub("\\.cell_feature\\.tsv\\.gz$", "", fn)
  meta <- parse_combo_base(base)
  if (file.info(fp)$size < 30) {
    message("[load_amet] skipping near-empty cell_feature file: ", fn)
    return(NULL)
  }
  dt <- tryCatch(fread(fp, select = c("cell_id", "mean_meth", "i_total")),
                 error = function(e) {
                   message("[load_amet] failed to read ", fn, ": ", conditionMessage(e))
                   NULL
                 })
  if (is.null(dt) || nrow(dt) == 0L) return(NULL)
  agg <- dt[, .(mean_meth = mean(mean_meth, na.rm = TRUE),
                i_total = mean(i_total, na.rm = TRUE)),
            by = cell_id]
  agg[, subcat := meta$subcat]
  agg[, cat := meta$cat]
  agg[, patient := meta$patient]
  agg[, location := meta$location]
  agg
}

read_combo_fe <- function(fp) {
  fn <- basename(fp)
  base <- sub("\\.feature\\.tsv\\.gz$", "", fn)
  meta <- parse_combo_base(base)
  if (file.info(fp)$size < 30) {
    message("[load_amet] skipping near-empty feature file: ", fn)
    return(NULL)
  }
  dt <- tryCatch(fread(fp),
                 error = function(e) {
                   message("[load_amet] failed to read ", fn, ": ", conditionMessage(e))
                   NULL
                 })
  if (is.null(dt) || nrow(dt) == 0L) return(NULL)
  dt[, subcat := meta$subcat]
  dt[, cat := meta$cat]
  dt[, patient := meta$patient]
  dt[, location := meta$location]
  dt
}

cf_files <- list.files(params$features_dir,
                       pattern = "\\.cell_feature\\.tsv\\.gz$",
                       full.names = TRUE)
fe_files <- list.files(params$features_dir,
                       pattern = "\\.feature\\.tsv\\.gz$",
                       full.names = TRUE)
fe_files <- fe_files[!grepl("\\.cell_feature\\.tsv\\.gz$", fe_files)]

feat_cf <- rbindlist(lapply(cf_files, read_combo_cf), fill = TRUE)
feat_fe <- rbindlist(lapply(fe_files, read_combo_fe), fill = TRUE)

cat("loaded", length(cf_files), "cell_feature TSVs and",
    length(fe_files), "feature TSVs from", params$features_dir, "\n")

## Long table `files` holds one row per (cell, annotation, patient,
## location) with columns i_total + avg_meth, built from amet's per-cell
## Keep an `avg_meth` alias for chunks that still use that column name.
files <- as.data.frame(feat_cf)
files$annotation <- files$subcat
files$avg_meth <- files$mean_meth
files <- files[, c("i_total", "avg_meth", "annotation", "patient", "location")]

## prettifies and sorts categorical vars
harmonize_levels <- function(df) {
    pretty_hmm <- c(
        "0_Enhancer"        = "0 Enh.",
        "1_Transcribed"     = "1 Transc. ",
        "2_Enhancer"        = "2 Enh.",
        "3_Quiescent"       = "3 Quies.",
        "4_Transcribed"     = "4 Transc.",
        "5_RegPermissive"   = "5 Reg. perm.",
        "6_LowConfidence"   = "6 Low conf.",
        "7_RegPermissive"   = "7 Reg. perm.",
        "8_Quiescent"       = "8 Quies.",
        "9_ConstitutiveHet" = "9 Cons. het.",
        "10_Quiescent"      = "10 Quies.",
        "11_Promoter"       = "11 Prom.",
        "12_Promoter"       = "12 Prom.",
        "13_ConstitutiveHet"= "13 Cons. het"
    )

    pretty_other <- c(
        "cpgIslandExt"      = "CpGi",
        "H3K4me3"           = "H3K4me3",
        "genes"             = "Genes",
        "H3K27me3"          = "H3K27me3",
        "H3K9me3"           = "H3K9me3",
        "laminb1"           = "Lamin B1",
        "pmds"              = "PMDs",
        "lines"             = "LINEs",
        "sines"             = "SINEs",
        "hmds"              = "HMDs",
        "crc01_lost_scna"   = "SCNA loss",
        "crc01_nc_scna"     = "SCNA neutral",
        "crc01_gain_scna"   = "SCNA gain")

    all_levels <- c(pretty_hmm, pretty_other)

    missing <- setdiff(unique(df$annotation), names(all_levels))
    if (length(missing) > 0) {
        stop(missing)
    }


  df$annotation <- factor(df$annotation,
                          levels = names(all_levels),
                          labels = unname(all_levels))

  df$patient <- as.factor(df$patient)
  df$location <- factor(df$location, levels = c("NC","PT","LN","ML","MP","MO"))

  df
}

files <- harmonize_levels(files)
table(files$annotation, useNA = 'always')

```

# i_total per cell {.tabset .tabset-pills}

amet's `i_total` is the per-cell-per-feature within-cell mutual information score from `cell_feature.tsv.gz`. The same `files` data frame loaded above is reused here directly.

```{r, import_short_reports_norm}
## i_total is already in `files`, so reuse the same data frame here.
```

## Scatters by location and annotation

```{r}
ggplot(
  files,
  aes(y = i_total, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation) +
  labs(
    x = "avg. methylation",
    y = expression(i[total])
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Scatters by location and annotation and patient

```{r, fig.width = 8, fig.height = 20}
ggplot(
  files,
  aes(y = i_total, x = avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_grid(annotation ~ patient) +
  labs(
    x = "avg. methylation",
    y = expression(i[total])
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Scatters by location and annotation and patient (shorter)

```{r, fig.width = 8, fig.height = 6}
ggplot(
  subset(files, annotation %in% c("Lamin B1", "H3K27me3", "H3K9me3")),
  aes(y = i_total, x = avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_grid(annotation ~ patient) +
  labs(
    x = "avg. methylation",
    y = expression(i[total])
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```


## Segmentations {.tabset .tabset-pills}

### Scatters by location and annotation

```{r}
ggplot(
    subset(files, grepl("^[0-9]{1,2}", annotation)),
  aes(y = i_total, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "avg. methylation",
    y = expression(i[total])
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```


## Nonsegmentations {.tabset .tabset-pills}

### Scatters by location and annotation

```{r}
ggplot(
    subset(files, !grepl("^[0-9]{1,2}", annotation)),
  aes(y = i_total, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "avg. methylation",
    y = expression(i[total])
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Others  {.tabset .tabset-pills}

### i_total boxes by location and annotation - crc01

```{r, fig.width = 9, fig.height = 5}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = i_total, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = expression(i[total])) +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### meth boxes by location and annotation - crc01

```{r, fig.width = 9, fig.height = 5}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = avg_meth, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "avg. methylation") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### Range plots - by compartment

```{r, fig.width = 8, fig.height = 8}
ranges <- as.data.frame(files %>%
   filter(patient == "CRC01") %>%
   group_by(compartment = annotation, location) %>%
   summarise(
     i_total_mean = mean(i_total, na.rm = TRUE),
     i_total_min = min(i_total, na.rm = TRUE),
     i_total_max = max(i_total, na.rm = TRUE),
     meth_mean = mean(avg_meth, na.rm = TRUE),
     meth_min = min(avg_meth, na.rm = TRUE),
     meth_max = max(avg_meth, na.rm = TRUE)))

ggplot(ranges, aes(color = location)) +
  ## horizontal segment: range of methylations for that location/compartment
  geom_segment(
    aes(x = meth_min, xend = meth_max, y = i_total_mean, yend = i_total_mean),
    linewidth = 0.6, alpha = 1
  ) +
  # vertical segment: range of i_total for that location/compartment
  geom_segment(
    aes(x = meth_mean, xend = meth_mean, y = i_total_min, yend = i_total_max),
    linewidth = 0.6, alpha = 1
  ) +
  # center point
  geom_point(
    aes(x = meth_mean, y = i_total_mean),
    shape = 3, size = 1.2
  ) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "DNA methylation", y = expression(i[total])) +
  facet_wrap(~ compartment) +
  theme_ng()

```

```{r}
i_total_long <- files
```

## i_total within methylation bins {.tabset .tabset-pills}

Boxplots of i_total stratified by 0.1-wide average methylation bins, colored by
location. Persistent differences within bins show that i_total captures
heterogeneity beyond average methylation level.

### All annotations

```{r i_total_meth_bins_all, fig.width = ng_fig_size(5, 6)$w, fig.height = ng_fig_size(5, 6)$h}
i_total_long %>%
  mutate(meth_bin = cut(avg_meth, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>%
  filter(!is.na(meth_bin), !is.na(i_total)) %>%
  ggplot(aes(x = meth_bin, y = i_total, fill = location)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "avg. methylation bin", y = expression(i[total])) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### CRC01 only

```{r i_total_meth_bins_crc01, fig.width = ng_fig_size(5, 6)$w, fig.height = ng_fig_size(5, 6)$h}
i_total_long %>%
  filter(patient == "CRC01") %>%
  mutate(meth_bin = cut(avg_meth, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>%
  filter(!is.na(meth_bin), !is.na(i_total)) %>%
  ggplot(aes(x = meth_bin, y = i_total, fill = location)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "avg. methylation bin", y = expression(i[total])) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

# Per-group median jsd {.tabset .tabset-pills}

amet's per-feature `jsd` from feature.tsv.gz captures across-cell heterogeneity. amet's feature TSV has many rows per group, so aggregate to one row per (annotation, patient, location) with `median(jsd)`.

```{r, eval = TRUE}
jsd_median_long <- as.data.frame(
  as.data.table(feat_fe)[
    , .(median_jsd = median(jsd, na.rm = TRUE),
        median_avg_meth = median(mean_meth_mean, na.rm = TRUE)),
    by = .(annotation = subcat, patient, location)
  ]
)

jsd_median_long <- harmonize_levels(jsd_median_long)
```

## Scatters by location and annotation

```{r}
ggplot(
  jsd_median_long,
  aes(y = median_jsd, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(jsd_median_long$patient)) +
  facet_wrap(~annotation) +
  labs(
    x = "median average methylation",
    y = "median jsd"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Scatters by location and annotation and patient

```{r}
ggplot(
  jsd_median_long,
  aes(y = median_jsd, x = median_avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  labs(
    x = "median average methylation",
    y = "median jsd"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```



## Segmentations {.tabset .tabset-pills}

### Scatters by location and annotation

```{r}
ggplot(
    subset(jsd_median_long, grepl("^[0-9]{1,2}", annotation)),
  aes(y = median_jsd, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(jsd_median_long$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "median avg. methylation",
    y = "median jsd"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```


## Nonsegmentations {.tabset .tabset-pills}

### Scatters by location and annotation

```{r}
ggplot(
    subset(jsd_median_long, !grepl("^[0-9]{1,2}", annotation)),
  aes(y = median_jsd, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(jsd_median_long$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "median avg. methylation",
    y = "median jsd"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

# Attempt to assemble these all

amet ships i_total (per cell) and jsd (per group). The long tables `i_total_long` and `jsd_median_long` carry per-cell and per-group data respectively.

```{r}
str(jsd_median_long)
str(i_total_long)
```

```{r save_for_emanuel}
saveRDS(list(jsd_median_long = jsd_median_long,
             i_total_long = i_total_long),
        file = "crc_entropy_summaries.rds")
```


```{r summary_heatmap, fig.width = 10, fig.height = 8}
library(ComplexHeatmap)
library(circlize)

make_hm_mat <- function(df, value_col) {
  wide <- df %>%
    group_by(annotation, location) %>%
    summarise(v = median(.data[[value_col]], na.rm = TRUE), .groups = "drop") %>%
    tidyr::pivot_wider(names_from = location, values_from = v)
  mat <- as.matrix(wide[, -1])
  rownames(mat) <- wide$annotation
  mat
}

hm_i_total <- make_hm_mat(i_total_long, "i_total")
hm_jsd <- make_hm_mat(jsd_median_long, "median_jsd")

safe_ramp <- function(mat, cols) {
  vals <- mat[is.finite(mat)]
  if (length(vals) == 0 || length(unique(vals)) < 2) return(NULL)
  q <- quantile(vals, c(0.05, 0.5, 0.95))
  if (length(unique(q)) < length(q)) q <- range(vals) |> (\(r) c(r[1], mean(r), r[2]))()
  if (length(unique(q)) < 3) return(NULL)
  colorRamp2(q, cols)
}

col_i_total <- safe_ramp(hm_i_total, c("#2166AC", "white", "#B2182B"))
col_jsd <- safe_ramp(hm_jsd, c("#4DAC26", "white", "#D01C8B"))

if (!is.null(col_i_total) && !is.null(col_jsd) &&
    nrow(hm_i_total) > 0 && nrow(hm_jsd) > 0) {
  draw(
    Heatmap(hm_i_total, name = "i_total", col = col_i_total,
            cluster_columns = FALSE,
            row_names_gp = gpar(fontsize = 7),
            column_names_gp = gpar(fontsize = 7)) +
    Heatmap(hm_jsd, name = "jsd", col = col_jsd,
            cluster_rows = FALSE, cluster_columns = FALSE,
            show_row_names = FALSE,
            column_names_gp = gpar(fontsize = 7)),
    column_title = "Median i_total and jsd per annotation x location"
  )
} else {
  message("summary_heatmap: insufficient variation in heatmap matrices, skipping draw")
}
```



```{r, fig.width = 9, fig.height = 12}

library(dplyr)

adj_df <- i_total_long %>%
  left_join(jsd_median_long,
            by = c("annotation", "patient", "location"),
            suffix = c("_i_total", "_jsd"))

head(adj_df)
ggplot(
    adj_df,
    aes(x = median_jsd, y = i_total,
        color = location, shape = patient)
) +
    scale_shape_manual(values = 1:nlevels(adj_df$patient)) +
    geom_point(size = 0.8) +
    facet_wrap(~annotation, ncol = 4) +
    labs(
        x = "median jsd",
        y = expression(i[total])
    ) +
  theme_ng()

```


```{r}
variation <- adj_df %>%
    group_by(annotation) %>%
    summarise(
        jsd_range = max(median_jsd, na.rm = TRUE) -
            min(median_jsd, na.rm = TRUE),
        i_total_range = max(i_total, na.rm = TRUE) -
            min(i_total, na.rm = TRUE),
        jsd_sd = sd(median_jsd, na.rm = TRUE),
        i_total_sd = sd(i_total, na.rm = TRUE)
    )

head(variation)

ggplot(variation, aes(x = jsd_range, y = i_total_range, label = annotation)) +
    geom_point() +
    ggrepel::geom_text_repel() +
    labs(
        x = "jsd range aggregating all biopsy sites",
        y = expression(i[total] * " range aggregating all biopsy sites"),
        ) +
  theme_ng()


variation_loc <- adj_df %>%
    group_by(annotation, location) %>%
    summarise(
        jsd_median = median(median_jsd, na.rm = TRUE),
        i_total_median = median(i_total, na.rm = TRUE),
        jsd_sd = sd(median_jsd, na.rm = TRUE),
        i_total_sd = sd(i_total, na.rm = TRUE),
        jsd_range = max(median_jsd, na.rm = TRUE) -
            min(median_jsd, na.rm = TRUE),
        i_total_range = max(i_total, na.rm = TRUE) -
            min(i_total, na.rm = TRUE)
    )

ggplot(variation_loc,
       aes(x = jsd_median, y = i_total_median,
           color = location)) +
    geom_point() +
    facet_wrap(~annotation) +
    theme_ng() +
    labs(
         x = "median jsd across biopsy sites",
         y = expression("median " * i[total] * " across biopsy sites"))

ggplot(variation_loc,
       aes(x = jsd_sd, y = i_total_sd,
           color = location)) +
    geom_point() +
    facet_wrap(~annotation) +
    theme_ng() +
    labs(
         x = "sd jsd across biopsy sites",
         y = expression("sd " * i[total] * " across biopsy sites"))


ggplot(variation_loc,
       aes(x = jsd_range, y = i_total_range,
           color = location)) +
    geom_point() +
    facet_wrap(~annotation) +
    theme_ng() +
    labs(
         x = "jsd range across biopsy sites",
         y = expression(i[total] * " range across biopsy sites"))


```


# Driver categorization

For each annotation, we compute the SD of group-level median jsd (across-cell
heterogeneity) and median i_total (within-cell heterogeneity) across biopsy
locations. If one SD is at least 1.5x the other, that entropy component
dominates. Annotations where both SDs fall below the 30th percentile are
labelled "neither".

```{r drivers, fig.width = 5, fig.height = 5}
crc_grp_for_driver <- adj_df %>%
  rename(median_i_total = i_total)

driver_df <- categorize_drivers(crc_grp_for_driver, "location")

cat("Driver table:\n")
print(table(driver_df$driver))

saveRDS(driver_df, file = "crc_driver_sd_range.rds")

if (nrow(driver_df) > 0 && any(is.finite(driver_df$jsd_sd)) && any(is.finite(driver_df$i_total_sd))) {
  plot_driver_scatter(driver_df,
                      x_label = "SD of jsd across biopsy locations",
                      y_label = expression("SD of " * i[total] * " across biopsy locations"))
} else {
  message("drivers: no finite SDs, skipping driver scatter")
}
```


# Session

```{r, cache = FALSE}
sessionInfo()
```
