---
title: "amet on Bian 2018 CRC scTrioSeq2"
author: "Izaskun Mallona"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    number_sections: true
    df_print: default
params:
  feat_cell_feature: ""
  feat_feature: ""
  feat_bed: ""
  win_cell_feature: ""
  win_feature: ""
  win_bed: ""
  manifest: ""
  out_dir: ""
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 5, fig.height = 4)

suppressPackageStartupMessages({
    library(data.table)
    library(ggplot2)
    library(ggridges)
    library(ggrepel)
    library(viridis)
    library(patchwork)
    library(dplyr)
    library(tidyr)
    library(uwot)
})

repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))

panels_dir <- file.path(params$out_dir, "panels")
feat_dir <- file.path(panels_dir, "feature")
win_dir <- file.path(panels_dir, "window")
dir.create(feat_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(win_dir, showWarnings = FALSE, recursive = TRUE)

location_levels <- c("NC", "PT", "LN", "ML", "MP", "MO")
loc_pal <- c(NC = "#4DAF4A", PT = "#E41A1C", LN = "#377EB8",
             ML = "#984EA3", MP = "#FF7F00", MO = "#A65628")
```

# Inputs and design

CRC prototype run: chr19, two patients (CRC01, CRC02), three biopsy locations (NC = normal colon, PT = primary tumour, LN = lymph node metastasis). Per-cell raw files are scTrioSeq2 singleC tables read by amet's bismark parser.

The report is split into two analyses with **disjoint inputs and disjoint feature definitions**:

- **Feature-based**: amet outputs over CpG islands on chr19. One feature = one CGI. This section answers questions about regulatory regions: which annotation classes are most variable, which contrasts (PT vs NC, LN vs NC) carry the strongest signal, and how regularity (`i_total`) and across-cell heterogeneity (`jsd`) co-vary.
- **Window-based**: amet outputs over fixed-size genome tiles on chr19. One feature = one window. Used for cell-level summaries, embeddings, and unbiased landscape views.

Score mapping: amet `i_total` (within-cell mutual information across CpG lags) replaces yamet's adjS; amet `jsd` (across-cell heterogeneity in lag-1 2-mer histograms) replaces yamet's adjH.

```{r load}
man <- fread(params$manifest)
feat_cf <- fread(params$feat_cell_feature)
feat_fe <- fread(params$feat_feature)
win_cf <- fread(params$win_cell_feature)
win_fe <- fread(params$win_feature)
feat_bed <- fread(params$feat_bed, header = FALSE,
                  col.names = c("chrom", "start", "end", "feature_id"))
win_bed <- fread(params$win_bed, header = FALSE,
                 col.names = c("chrom", "start", "end", "feature_id"))

man[, location := factor(location, levels = location_levels)]
man[, group := location]

feat_cf <- merge(feat_cf, man[, .(cell_id, patient, location)],
                 by = "cell_id", all.x = TRUE)
feat_fe[, location := factor(group, levels = location_levels)]

win_cf <- merge(win_cf, man[, .(cell_id, patient, location)],
                by = "cell_id", all.x = TRUE)
win_fe[, location := factor(group, levels = location_levels)]
```

# Sample sizes and coverage

```{r sizes, fig.width = 6, fig.height = 3}
sizes_dt <- man[, .(cells = .N), by = .(patient, location)][order(patient, location)]
print(sizes_dt)

p_n <- ggplot(sizes_dt, aes(x = location, y = cells, fill = patient)) +
    geom_col(position = position_dodge(width = 0.85), width = 0.8) +
    labs(x = NULL, y = "cells") +
    theme_ng()
save_eval(p_n, sizes_dt, file.path(panels_dir, "ncells_per_group"),
          width_mm = 90, height_mm = 60)
p_n
```

```{r coverage_diag, fig.width = 7, fig.height = 3.5}
cov_feat <- feat_cf[, .(n_features = .N,
                        median_n_cpgs = median(n_cpgs, na.rm = TRUE)),
                    by = .(cell_id, location)]
cov_win <- win_cf[, .(n_windows = .N,
                      median_n_cpgs = median(n_cpgs, na.rm = TRUE)),
                  by = .(cell_id, location)]
cov_long <- rbind(
    cov_feat[, .(cell_id, location, n = n_features, scope = "feature (CGI)")],
    cov_win[,  .(cell_id, location, n = n_windows,  scope = "window")]
)
p_cov <- ggplot(cov_long, aes(x = location, y = n, fill = location)) +
    geom_violin(alpha = 0.6, scale = "width") +
    geom_boxplot(width = 0.18, outlier.size = 0.3, fill = "white") +
    scale_fill_manual(values = loc_pal) +
    facet_wrap(~ scope, scales = "free_y") +
    labs(x = NULL, y = "covered features per cell") +
    theme_ng() +
    theme(legend.position = "none")
save_eval(p_cov, cov_long, file.path(panels_dir, "coverage_per_cell"),
          width_mm = 130, height_mm = 70)
p_cov
```

# Feature-based analysis: CpG islands on chr19

Each feature = one CGI. Score: `i_total` (within-cell, per cell × CGI), `jsd` (across-cell, per CGI × group). Coverage: `n_cpgs` per cell × CGI; `mean_meth` per cell × CGI.

```{r feat_filter}
fcf <- feat_cf[!is.na(i_total)]
ffe <- feat_fe[!is.na(jsd)]
```

## Within-cell regularity (i_total)

### Per-location distribution

```{r feat_itotal_box, fig.width = 5, fig.height = 4}
p_itotal_box <- ggplot(fcf, aes(x = location, y = i_total, fill = location)) +
    geom_violin(alpha = 0.55, scale = "width", linewidth = 0.3) +
    geom_boxplot(width = 0.18, outlier.size = 0.3, fill = "white",
                 linewidth = 0.3) +
    scale_fill_manual(values = loc_pal) +
    labs(x = NULL, y = expression(i[total])) +
    theme_ng() + theme(legend.position = "none")
save_eval(p_itotal_box,
          fcf[, .(cell_id, location, patient, feature_id, i_total)],
          file.path(feat_dir, "itotal_by_location"))
p_itotal_box
```

### Stratified by patient

```{r feat_itotal_pat, fig.width = 7, fig.height = 4}
p_itotal_pat <- ggplot(fcf, aes(x = location, y = i_total, fill = location)) +
    geom_boxplot(outlier.size = 0.3, linewidth = 0.3, alpha = 0.7) +
    scale_fill_manual(values = loc_pal) +
    facet_wrap(~ patient) +
    labs(x = NULL, y = expression(i[total])) +
    theme_ng() + theme(legend.position = "none")
save_eval(p_itotal_pat,
          fcf[, .(cell_id, location, patient, i_total)],
          file.path(feat_dir, "itotal_by_location_patient"),
          width_mm = 130, height_mm = 70)
p_itotal_pat
```

### Density ridges per location

```{r feat_itotal_ridge, fig.width = 5, fig.height = 4}
p_ridge <- ggplot(fcf, aes(x = i_total, y = location, fill = location)) +
    geom_density_ridges(alpha = 0.7, scale = 1.2, linewidth = 0.3) +
    scale_fill_manual(values = loc_pal) +
    labs(x = expression(i[total]), y = NULL) +
    theme_ng() + theme(legend.position = "none", aspect.ratio = NULL)
save_eval(p_ridge, fcf[, .(location, i_total)],
          file.path(feat_dir, "itotal_ridge_location"))
p_ridge
```

### Mean per cell, per location

```{r feat_itotal_per_cell, fig.width = 5, fig.height = 4}
cell_summary_feat <- fcf[, .(mean_i_total = mean(i_total, na.rm = TRUE),
                             median_i_total = median(i_total, na.rm = TRUE),
                             mean_meth = mean(mean_meth, na.rm = TRUE),
                             n_features = .N),
                         by = .(cell_id, patient, location)]

p_cell_it <- ggplot(cell_summary_feat,
                    aes(x = location, y = mean_i_total, fill = location)) +
    geom_boxplot(outlier.size = 0.3, linewidth = 0.3, alpha = 0.75) +
    geom_jitter(width = 0.18, size = 0.6, alpha = 0.7) +
    scale_fill_manual(values = loc_pal) +
    labs(x = NULL, y = expression("mean " * i[total] * " per cell")) +
    theme_ng() + theme(legend.position = "none")
save_eval(p_cell_it, cell_summary_feat,
          file.path(feat_dir, "itotal_per_cell"))
p_cell_it
```

## Methylation diagnostics

```{r feat_meth_box, fig.width = 5, fig.height = 4}
p_meth_box <- ggplot(cell_summary_feat,
                     aes(x = location, y = mean_meth, fill = location)) +
    geom_boxplot(outlier.size = 0.3, linewidth = 0.3, alpha = 0.75) +
    scale_fill_manual(values = loc_pal) +
    labs(x = NULL, y = "mean methylation per cell (CGIs)") +
    theme_ng() + theme(legend.position = "none")
save_eval(p_meth_box, cell_summary_feat,
          file.path(feat_dir, "meanmeth_per_cell"))
p_meth_box
```

## Decoupling: i_total vs mean methylation

```{r feat_decouple_pt, fig.width = 5, fig.height = 4}
p_decouple_pt <- ggplot(fcf,
                        aes(x = mean_meth, y = i_total, colour = location)) +
    geom_point(size = 0.5, alpha = 0.5) +
    geom_smooth(method = "loess", se = FALSE, linewidth = 0.4) +
    scale_colour_manual(values = loc_pal) +
    labs(x = "mean methylation (per cell, per CGI)",
         y = expression(i[total]),
         colour = "location") +
    theme_ng()
save_eval(p_decouple_pt,
          fcf[, .(cell_id, location, mean_meth, i_total)],
          file.path(feat_dir, "itotal_vs_meth_scatter"))
p_decouple_pt
```

```{r feat_decouple_facet, fig.width = 7, fig.height = 4}
p_decouple_pat <- p_decouple_pt + facet_wrap(~ patient)
save_eval(p_decouple_pat,
          fcf[, .(cell_id, location, patient, mean_meth, i_total)],
          file.path(feat_dir, "itotal_vs_meth_by_patient"),
          width_mm = 140, height_mm = 70)
p_decouple_pat
```

```{r feat_decouple_per_cell, fig.width = 5, fig.height = 4}
p_per_cell_dec <- ggplot(cell_summary_feat,
                         aes(x = mean_meth, y = mean_i_total, colour = location,
                             shape = patient)) +
    geom_point(size = 1.6, alpha = 0.9) +
    scale_colour_manual(values = loc_pal) +
    labs(x = "mean methylation per cell",
         y = expression("mean " * i[total] * " per cell")) +
    theme_ng()
save_eval(p_per_cell_dec, cell_summary_feat,
          file.path(feat_dir, "itotal_vs_meth_per_cell"))
p_per_cell_dec
```

## Across-cell heterogeneity (jsd)

### Per-location distribution

```{r feat_jsd_box, fig.width = 5, fig.height = 4}
if (nrow(ffe) > 0) {
    p_jsd_box <- ggplot(ffe, aes(x = location, y = jsd, fill = location)) +
        geom_violin(alpha = 0.55, scale = "width", linewidth = 0.3) +
        geom_boxplot(width = 0.18, outlier.size = 0.3, fill = "white",
                     linewidth = 0.3) +
        scale_fill_manual(values = loc_pal) +
        labs(x = NULL, y = "jsd") +
        theme_ng() + theme(legend.position = "none")
    save_eval(p_jsd_box, ffe[, .(feature_id, location, n_cells, jsd)],
              file.path(feat_dir, "jsd_by_location"))
    p_jsd_box
} else {
    cat("All jsd values NA at the configured min-cells-per-group threshold.\n")
}
```

### Density ridges

```{r feat_jsd_ridge, fig.width = 5, fig.height = 4}
if (exists("p_jsd_box")) {
    p_jsd_ridge <- ggplot(ffe, aes(x = jsd, y = location, fill = location)) +
        geom_density_ridges(alpha = 0.7, scale = 1.1, linewidth = 0.3) +
        scale_fill_manual(values = loc_pal) +
        labs(x = "jsd", y = NULL) +
        theme_ng() + theme(legend.position = "none", aspect.ratio = NULL)
    save_eval(p_jsd_ridge, ffe[, .(location, jsd)],
              file.path(feat_dir, "jsd_ridge_location"))
    p_jsd_ridge
}
```

## Joint structure: i_total vs jsd at the feature level

```{r feat_joint, fig.width = 5, fig.height = 4}
fe_joint <- fcf[, .(median_i_total = median(i_total, na.rm = TRUE)),
                by = .(feature_id, location)]
fe_joint <- merge(fe_joint, ffe[, .(feature_id, location, jsd, n_cells)],
                  by = c("feature_id", "location"))
fe_joint <- fe_joint[!is.na(jsd) & !is.na(median_i_total)]

p_joint <- ggplot(fe_joint,
                  aes(x = jsd, y = median_i_total, colour = location)) +
    geom_point(size = 0.7, alpha = 0.85) +
    scale_colour_manual(values = loc_pal) +
    labs(x = "jsd (per CGI, per location)",
         y = expression("median " * i[total] * " (per CGI, per location)")) +
    theme_ng()
save_eval(p_joint, fe_joint, file.path(feat_dir, "itotal_vs_jsd_per_feature"))
p_joint
```

## Heatmap: median i_total per (CGI, location)

```{r feat_heatmap_it, fig.width = 5, fig.height = 7}
hm <- dcast(fcf[, .(median_i_total = median(i_total, na.rm = TRUE)),
                by = .(feature_id, location)],
            feature_id ~ location, value.var = "median_i_total")
hm_mat <- as.matrix(hm[, -1])
rownames(hm_mat) <- hm$feature_id
hm_mat <- hm_mat[rowSums(!is.na(hm_mat)) >= 2, , drop = FALSE]
hm_mat <- hm_mat[order(rowMeans(hm_mat, na.rm = TRUE)), , drop = FALSE]
hm_long <- as.data.table(reshape2::melt(hm_mat, varnames = c("feature_id", "location")))
hm_long[, feature_id := factor(feature_id, levels = rownames(hm_mat))]

p_hm_it <- ggplot(hm_long, aes(x = location, y = feature_id, fill = value)) +
    geom_tile() +
    scale_fill_viridis(option = "magma", na.value = "grey90") +
    labs(x = NULL, y = "CpG island", fill = expression(i[total])) +
    theme_ng() +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
          aspect.ratio = NULL)
save_eval(p_hm_it, hm_long, file.path(feat_dir, "heatmap_median_itotal"),
          width_mm = 90, height_mm = 140)
p_hm_it
```

## Heatmap: jsd per (CGI, location)

```{r feat_heatmap_jsd, fig.width = 5, fig.height = 7}
if (nrow(ffe) > 0) {
    jsd_w <- dcast(ffe, feature_id ~ location, value.var = "jsd")
    jsd_mat <- as.matrix(jsd_w[, -1])
    rownames(jsd_mat) <- jsd_w$feature_id
    jsd_mat <- jsd_mat[rowSums(!is.na(jsd_mat)) >= 1, , drop = FALSE]
    jsd_mat <- jsd_mat[order(rowMeans(jsd_mat, na.rm = TRUE)), , drop = FALSE]
    jsd_long <- as.data.table(reshape2::melt(jsd_mat, varnames = c("feature_id", "location")))
    jsd_long[, feature_id := factor(feature_id, levels = rownames(jsd_mat))]

    p_hm_jsd <- ggplot(jsd_long, aes(x = location, y = feature_id, fill = value)) +
        geom_tile() +
        scale_fill_viridis(option = "viridis", na.value = "grey90") +
        labs(x = NULL, y = "CpG island", fill = "jsd") +
        theme_ng() +
        theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              aspect.ratio = NULL)
    save_eval(p_hm_jsd, jsd_long, file.path(feat_dir, "heatmap_jsd"),
              width_mm = 90, height_mm = 140)
    p_hm_jsd
}
```

## Disease contrasts: PT vs NC, LN vs NC

```{r feat_contrasts, fig.width = 8, fig.height = 4}
contrast_dt <- list()
for (lvl in c("PT", "LN")) {
    if (!all(c("NC", lvl) %in% levels(droplevels(fe_joint$location)))) next
    a <- fe_joint[location == "NC", .(feature_id, NC = median_i_total, NC_jsd = jsd)]
    b <- fe_joint[location == lvl,
                  .(feature_id, target = median_i_total, target_jsd = jsd)]
    j <- merge(a, b, by = "feature_id")
    j[, contrast := paste0(lvl, " - NC")]
    contrast_dt[[lvl]] <- j
}
contrast_dt <- rbindlist(contrast_dt, fill = TRUE)

if (nrow(contrast_dt)) {
    p_contrast_it <- ggplot(contrast_dt, aes(x = NC, y = target, colour = contrast)) +
        geom_abline(slope = 1, intercept = 0, colour = "grey60",
                    linetype = "dashed", linewidth = 0.3) +
        geom_point(size = 0.6, alpha = 0.6) +
        facet_wrap(~ contrast) +
        labs(x = expression("median " * i[total] * " (NC)"),
             y = expression("median " * i[total] * " (target)"),
             colour = NULL) +
        theme_ng() + theme(legend.position = "none")
    save_eval(p_contrast_it, contrast_dt,
              file.path(feat_dir, "contrast_itotal_scatter"),
              width_mm = 150, height_mm = 70)
    p_contrast_it
}
```

```{r feat_contrasts_delta, fig.width = 6, fig.height = 4}
if (nrow(contrast_dt)) {
    contrast_dt[, delta_it := target - NC]
    contrast_dt[, delta_jsd := target_jsd - NC_jsd]

    p_delta_hist <- ggplot(contrast_dt, aes(x = delta_it, fill = contrast)) +
        geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
        geom_vline(xintercept = 0, linetype = "dashed",
                   colour = "grey40", linewidth = 0.3) +
        labs(x = expression(Delta * i[total] * " (target - NC)"), y = "CGIs") +
        theme_ng() + theme(legend.position = "top")
    save_eval(p_delta_hist, contrast_dt,
              file.path(feat_dir, "contrast_itotal_delta_hist"),
              width_mm = 110, height_mm = 70)
    p_delta_hist
}
```

```{r feat_contrasts_jsd, fig.width = 6, fig.height = 4}
if (nrow(contrast_dt) && any(!is.na(contrast_dt$delta_jsd))) {
    p_delta_jsd <- ggplot(contrast_dt[!is.na(delta_jsd)],
                          aes(x = delta_jsd, fill = contrast)) +
        geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
        geom_vline(xintercept = 0, linetype = "dashed",
                   colour = "grey40", linewidth = 0.3) +
        labs(x = expression(Delta * "jsd (target - NC)"), y = "CGIs") +
        theme_ng() + theme(legend.position = "top")
    save_eval(p_delta_jsd, contrast_dt[!is.na(delta_jsd)],
              file.path(feat_dir, "contrast_jsd_delta_hist"),
              width_mm = 110, height_mm = 70)
    p_delta_jsd
}
```

## Top variable CGIs

```{r feat_top_var, fig.width = 6, fig.height = 4}
var_dt <- fcf[, .(sd_i_total = sd(i_total, na.rm = TRUE),
                  median_i_total = median(i_total, na.rm = TRUE),
                  n = .N),
              by = feature_id][n >= 5]
top_var <- head(var_dt[order(-sd_i_total)], 20)

p_top_var <- ggplot(top_var,
                    aes(x = reorder(feature_id, sd_i_total), y = sd_i_total)) +
    geom_col(fill = "#444444") +
    coord_flip() +
    labs(x = NULL, y = expression("sd(" * i[total] * ") across cells")) +
    theme_ng() + theme(axis.text.y = element_text(size = 6),
                       aspect.ratio = NULL)
save_eval(p_top_var, top_var, file.path(feat_dir, "top_variable_cgi"),
          width_mm = 110, height_mm = 80)
p_top_var
```

# Window-based analysis: chr19 genome tiles

Each feature = one fixed-size window. Same scores (`i_total`, `jsd`) but coverage is dense and uniform, so this section drives cell-level summaries and embeddings rather than per-region biology.

```{r win_filter}
wcf <- win_cf[!is.na(i_total)]
wfe <- win_fe[!is.na(jsd)]
```

## Per-cell coverage of windows

```{r win_coverage, fig.width = 6, fig.height = 4}
win_cov <- win_cf[, .(n_windows = .N), by = .(cell_id, location, patient)]
p_win_cov <- ggplot(win_cov, aes(x = location, y = n_windows, fill = location)) +
    geom_violin(alpha = 0.55, scale = "width", linewidth = 0.3) +
    geom_jitter(width = 0.15, size = 0.5, alpha = 0.6) +
    scale_fill_manual(values = loc_pal) +
    labs(x = NULL, y = "covered windows per cell") +
    theme_ng() + theme(legend.position = "none")
save_eval(p_win_cov, win_cov, file.path(win_dir, "coverage_windows_per_cell"))
p_win_cov
```

## Mean methylation per cell

```{r win_meanmeth_box, fig.width = 5, fig.height = 4}
cell_meth <- win_cf[!is.na(mean_meth),
                    .(mean_meth = mean(mean_meth, na.rm = TRUE),
                      n_windows = .N),
                    by = .(cell_id, patient, location)]
p_wmeth <- ggplot(cell_meth, aes(x = location, y = mean_meth, fill = location)) +
    geom_violin(alpha = 0.55, scale = "width", linewidth = 0.3) +
    geom_boxplot(width = 0.18, outlier.size = 0.3, fill = "white",
                 linewidth = 0.3) +
    scale_fill_manual(values = loc_pal) +
    labs(x = NULL, y = "mean methylation per cell (windows)") +
    theme_ng() + theme(legend.position = "none")
save_eval(p_wmeth, cell_meth, file.path(win_dir, "meanmeth_per_cell"))
p_wmeth
```

## Mean i_total per cell

```{r win_itotal_box, fig.width = 5, fig.height = 4}
cell_it <- wcf[, .(mean_i_total = mean(i_total, na.rm = TRUE),
                   median_i_total = median(i_total, na.rm = TRUE),
                   n_windows = .N),
               by = .(cell_id, patient, location)]
p_wit <- ggplot(cell_it, aes(x = location, y = mean_i_total, fill = location)) +
    geom_violin(alpha = 0.55, scale = "width", linewidth = 0.3) +
    geom_boxplot(width = 0.18, outlier.size = 0.3, fill = "white",
                 linewidth = 0.3) +
    scale_fill_manual(values = loc_pal) +
    labs(x = NULL, y = expression("mean " * i[total] * " per cell")) +
    theme_ng() + theme(legend.position = "none")
save_eval(p_wit, cell_it, file.path(win_dir, "mean_itotal_per_cell"))
p_wit
```

## Methylation vs i_total per cell

```{r win_per_cell_corr, fig.width = 5, fig.height = 4}
cell_full <- merge(cell_meth, cell_it[, .(cell_id, mean_i_total, median_i_total)],
                   by = "cell_id")
p_wcorr <- ggplot(cell_full,
                  aes(x = mean_meth, y = mean_i_total,
                      colour = location, shape = patient)) +
    geom_point(size = 1.6, alpha = 0.9) +
    scale_colour_manual(values = loc_pal) +
    labs(x = "mean methylation per cell",
         y = expression("mean " * i[total] * " per cell")) +
    theme_ng()
save_eval(p_wcorr, cell_full, file.path(win_dir, "itotal_vs_meth_per_cell"))
p_wcorr
```

## Window-level i_total ridges per location

```{r win_itotal_ridge, fig.width = 5, fig.height = 4}
p_win_ridge <- ggplot(wcf, aes(x = i_total, y = location, fill = location)) +
    geom_density_ridges(alpha = 0.7, scale = 1.2, linewidth = 0.3) +
    scale_fill_manual(values = loc_pal) +
    labs(x = expression(i[total] * " per (cell, window)"), y = NULL) +
    theme_ng() + theme(legend.position = "none", aspect.ratio = NULL)
save_eval(p_win_ridge, wcf[, .(location, i_total)],
          file.path(win_dir, "itotal_ridge_location"))
p_win_ridge
```

## Window-level jsd ridges per location

```{r win_jsd_ridge, fig.width = 5, fig.height = 4}
if (nrow(wfe)) {
    p_win_jsd_ridge <- ggplot(wfe, aes(x = jsd, y = location, fill = location)) +
        geom_density_ridges(alpha = 0.7, scale = 1.1, linewidth = 0.3) +
        scale_fill_manual(values = loc_pal) +
        labs(x = "jsd per (window, location)", y = NULL) +
        theme_ng() + theme(legend.position = "none", aspect.ratio = NULL)
    save_eval(p_win_jsd_ridge, wfe[, .(location, jsd)],
              file.path(win_dir, "jsd_ridge_location"))
    p_win_jsd_ridge
}
```

## Cell embedding: UMAP on cell-by-window i_total

```{r win_build_mat}
mat <- dcast(wcf, cell_id ~ feature_id, value.var = "i_total",
             fun.aggregate = mean)
mat_cells <- mat$cell_id
mat <- as.matrix(mat[, -1])
rownames(mat) <- mat_cells
mat <- mat[, colSums(!is.na(mat)) >= 0.5 * nrow(mat), drop = FALSE]
mat[is.na(mat)] <- 0
cat("embedding matrix:", nrow(mat), "cells x", ncol(mat), "windows\n")
```

```{r win_umap, fig.width = 9, fig.height = 4}
umd <- NULL
if (nrow(mat) >= 5 && ncol(mat) >= 5) {
    set.seed(1L)
    n_neigh <- max(2L, min(15L, nrow(mat) - 1L))
    um <- tryCatch(uwot::umap(mat, n_neighbors = n_neigh, min_dist = 0.3,
                              n_components = 2, verbose = FALSE),
                   error = function(e) { cat(conditionMessage(e), "\n"); NULL })
    if (!is.null(um)) {
        umd <- data.table(cell_id = rownames(mat),
                          UMAP1 = um[, 1], UMAP2 = um[, 2])
        umd <- merge(umd, man[, .(cell_id, patient, location)], by = "cell_id")

        p_um_loc <- ggplot(umd, aes(x = UMAP1, y = UMAP2, colour = location,
                                     shape = patient)) +
            geom_point(size = 1.8, alpha = 0.95) +
            scale_colour_manual(values = loc_pal) +
            labs(title = "UMAP: location") + theme_ng()
        p_um_pat <- ggplot(umd, aes(x = UMAP1, y = UMAP2, colour = patient,
                                     shape = location)) +
            geom_point(size = 1.8, alpha = 0.95) +
            labs(title = "UMAP: patient") + theme_ng()
        save_eval(p_um_loc, umd, file.path(win_dir, "umap_location"))
        save_eval(p_um_pat, umd, file.path(win_dir, "umap_patient"))
        print(p_um_loc + p_um_pat)
    } else {
        cat("UMAP failed; skipping.\n")
    }
}
```

## UMAP coloured by mean methylation

```{r win_umap_meth, fig.width = 5, fig.height = 4}
if (!is.null(umd)) {
    umd_full <- merge(umd, cell_full[, .(cell_id, mean_meth, mean_i_total)],
                      by = "cell_id")
    p_um_meth <- ggplot(umd_full,
                        aes(x = UMAP1, y = UMAP2, colour = mean_meth)) +
        geom_point(size = 1.8, alpha = 0.95) +
        scale_colour_viridis(option = "magma") +
        labs(colour = "mean meth") +
        theme_ng()
    save_eval(p_um_meth, umd_full, file.path(win_dir, "umap_meth"))
    p_um_meth
}
```

## UMAP coloured by mean i_total

```{r win_umap_it, fig.width = 5, fig.height = 4}
if (!is.null(umd)) {
    p_um_it <- ggplot(umd_full,
                      aes(x = UMAP1, y = UMAP2, colour = mean_i_total)) +
        geom_point(size = 1.8, alpha = 0.95) +
        scale_colour_viridis(option = "viridis") +
        labs(colour = expression("mean " * i[total])) +
        theme_ng()
    save_eval(p_um_it, umd_full, file.path(win_dir, "umap_itotal"))
    p_um_it
}
```

## Top variable windows

```{r win_top_var, fig.width = 6, fig.height = 4}
win_var <- wcf[, .(sd_i_total = sd(i_total, na.rm = TRUE),
                   median_i_total = median(i_total, na.rm = TRUE),
                   n = .N),
               by = feature_id][n >= 5]
top_win <- head(win_var[order(-sd_i_total)], 20)
p_win_top <- ggplot(top_win,
                    aes(x = reorder(feature_id, sd_i_total), y = sd_i_total)) +
    geom_col(fill = "#444444") +
    coord_flip() +
    labs(x = NULL, y = expression("sd(" * i[total] * ") across cells")) +
    theme_ng() +
    theme(axis.text.y = element_text(size = 6), aspect.ratio = NULL)
save_eval(p_win_top, top_win, file.path(win_dir, "top_variable_windows"),
          width_mm = 110, height_mm = 80)
p_win_top
```

## Per-window jsd ranked

```{r win_jsd_rank, fig.width = 6, fig.height = 4}
if (nrow(wfe)) {
    jsd_rank <- wfe[, .(median_jsd = median(jsd, na.rm = TRUE)),
                    by = feature_id]
    setorder(jsd_rank, -median_jsd)
    jsd_rank[, rank := .I]
    p_jsd_rank <- ggplot(jsd_rank, aes(x = rank, y = median_jsd)) +
        geom_line(linewidth = 0.4) +
        labs(x = "rank", y = "median jsd across locations") +
        theme_ng()
    save_eval(p_jsd_rank, jsd_rank, file.path(win_dir, "jsd_rank"))
    p_jsd_rank
}
```

## Per-window i_total location heatmap (top variable)

```{r win_heatmap_top, fig.width = 5, fig.height = 6}
top_ids <- top_win$feature_id
hm_w <- dcast(wcf[feature_id %in% top_ids,
                  .(median_i = median(i_total, na.rm = TRUE)),
                  by = .(feature_id, location)],
              feature_id ~ location, value.var = "median_i")
hm_w_mat <- as.matrix(hm_w[, -1])
rownames(hm_w_mat) <- hm_w$feature_id
hm_w_long <- as.data.table(reshape2::melt(hm_w_mat,
                                          varnames = c("feature_id", "location")))
p_win_hm <- ggplot(hm_w_long, aes(x = location, y = feature_id, fill = value)) +
    geom_tile() +
    scale_fill_viridis(option = "magma", na.value = "grey90") +
    labs(x = NULL, y = "window", fill = expression(i[total])) +
    theme_ng() +
    theme(axis.text.y = element_text(size = 6), aspect.ratio = NULL)
save_eval(p_win_hm, hm_w_long, file.path(win_dir, "heatmap_top_variable_windows"),
          width_mm = 100, height_mm = 130)
p_win_hm
```

# Combined paper figure

```{r paper, fig.width = 11, fig.height = 8}
panels <- list(
    A = if (exists("p_itotal_box")) p_itotal_box + labs(title = "A. i_total / CGI") else NULL,
    B = if (exists("p_jsd_box")) p_jsd_box + labs(title = "B. jsd / CGI") else NULL,
    C = if (exists("p_decouple_pt")) p_decouple_pt + labs(title = "C. decoupling / CGI") else NULL,
    D = if (exists("p_wit")) p_wit + labs(title = "D. mean i_total per cell / windows") else NULL,
    E = if (exists("p_um_loc")) p_um_loc + labs(title = "E. UMAP / location") else NULL,
    F = if (exists("p_um_meth")) p_um_meth + labs(title = "F. UMAP / methylation") else NULL
)
panels <- Filter(Negate(is.null), panels)
if (length(panels) >= 2) {
    combined <- Reduce(`+`, panels) + plot_layout(ncol = 3)
    ggsave(file.path(panels_dir, "paper_figure.pdf"),
           combined, width = 260, height = 170, units = "mm",
           device = grDevices::cairo_pdf)
    combined
}
```

# Session

```{r session}
sessionInfo()
```
