---
title: "CRC: cell embeddings from epigenetic heterogeneity"
author: "Izaskun Mallona"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
params:
  features_dir: ""
  windows_dir: ""
  win_bed: ""
  manifest: ""
  out_dir: ""
  corrected_sce: ""
  log_path: ""
---

```{r logging_early, include = FALSE}
source(Sys.getenv("AMET_RENDER_HELPERS"))
amet_setup_render_logging(params$log_path)
```

```{r setup}
suppressPackageStartupMessages({
  library(readr)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(BiocParallel)
  library(uwot)
  library(irlba)
  library(matrixStats)
  library(SingleCellExperiment)
  library(SummarizedExperiment)
  library(scater)
  library(cluster)
  library(ComplexHeatmap)
  library(circlize)
  library(ggrepel)
  library(viridis)
  library(cowplot)
})

repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))
source(file.path(repo_root, "workflow", "scripts", "palettes.R"))
source(file.path(repo_root, "workflow", "scripts", "embedding_utils.R"))
param <- SerialParam()
```

```{r opts}
knitr::opts_chunk$set(
  echo       = TRUE,
  fig.width  = ng_fig_size(1, 1)$w,
  fig.height = ng_fig_size(1, 1)$h,
  cache      = TRUE,
  include    = TRUE,
  fig.path   = "crc_embeddings_plots/",
  dev        = c("png", "svg"),
  dpi        = 500,
  cache.lazy = FALSE,
  warning    = TRUE,
  message    = TRUE
)
```

Two complementary feature spaces are compared for cell embedding:

1. **Per-window**: cells embedded using i_total across all genomic tiles (high-
   dimensional, spatially resolved). HVW selection applied before PCA and UMAP.
2. **Per-annotation**: amet's window BED carries no per-window annotation columns,
   so the per-annotation feature space and per-annotation HVW -> PCA -> UMAP loop
   are dropped here.

amet exports only `i_total` (within-cell mutual information across CpG lags) and
methylation. There is no raw sampen analogue, so any cell-level S_cellmeth
correction is not available.

```{r constants}
loc_levels <- c("NC", "PT", "LN", "ML", "MP", "MO")

N_HVW <- 1000  ## top variable windows to retain before PCA
N_PCS <- 10    ## PCs to compute
```

# Load data

```{r load_sce}
corrected_sce_path <- if (nzchar(params$corrected_sce))
                       params$corrected_sce else
                       file.path(params$out_dir, "sce_windows_colon_corrected.rds")
windows_sce <- readRDS(corrected_sce_path)
colData(windows_sce)$location <- factor(colData(windows_sce)$location, levels = loc_levels)

cat("Cells:", ncol(windows_sce), "\n")
cat("Windows:", nrow(windows_sce), "\n")
cat("Assays:", paste(assayNames(windows_sce), collapse = ", "), "\n")
cat("Locations:\n")
print(table(colData(windows_sce)$location, colData(windows_sce)$patient, useNA = "ifany"))
```

```{r filter_impute}
na_frac <- rowMeans(is.na(assay(windows_sce, "sampen")))
windows_sce <- windows_sce[na_frac < 0.3, ]
cat("Windows after NA filter:", nrow(windows_sce), "\n")
```

<!-- Section dropped: S_cellmeth assay (cell-mean-meth regressed out from raw S) requires a raw sampen assay that amet does not export. -->

# Shared helpers

```{r helpers}
## run_embedding() and sil_score() are provided by embedding_utils.R.
hvw_pca_umap <- function(mat, n_hvw = N_HVW, n_pcs = N_PCS) {
  run_embedding(mat, n_hvf = n_hvw, n_pcs = n_pcs,
                n_neighbors = 15L, min_dist = 0.3, seed = 42L)
}

plot_pair <- function(coords, sce_obj, title_prefix) {
  df <- data.frame(
    umap1    = coords[, 1], umap2 = coords[, 2],
    location = colData(sce_obj)$location,
    patient  = colData(sce_obj)$patient
  )
  p_loc <- ggplot(df, aes(x = umap1, y = umap2, color = location)) +
    geom_point(size = 0.6, alpha = 0.7) +
    scale_color_manual(values = crc_location_pal) +
    labs(title = paste(title_prefix, "- location"), x = "UMAP 1", y = "UMAP 2") +
    guides(x = guide_x_nolap(),
           color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
    theme_ng()
  p_pat <- ggplot(df, aes(x = umap1, y = umap2, color = patient)) +
    geom_point(size = 0.6, alpha = 0.7) +
    scale_color_manual(values = crc_patient_pal) +
    labs(title = paste(title_prefix, "- patient"), x = "UMAP 1", y = "UMAP 2") +
    guides(x = guide_x_nolap(),
           color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
    theme_ng()
  plot_grid(p_loc, p_pat, nrow = 1)
}

## Named list mapping assay names to display labels. amet exports only
## i_total (sampen), the meth-corrected i_total (sampen_corrected, "adjS"),
## and methylation. Raw S and S_cellmeth are not available.
assay_map <- c(
  "sampen"           = "i_total",
  "sampen_corrected" = "adjS",
  "meth"             = "methylation"
)
```

# Per-window cell embeddings (HVW -> PCA -> UMAP)

HVW selection and NA handling strategy (same across all reports):

- Zero is a valid measurement (unmethylated window / zero entropy). NA means
  no sequencing coverage at that window in that cell.
- Stage 1 (preferred): only windows with complete coverage across all cells
  (0 NAs) are eligible for HVW selection. Variance is computed on these
  complete windows; the top `r N_HVW` are selected. All cells are retained.
- Stage 2 (sparse fallback): when fewer than 10 complete windows exist,
  HVWs are selected from windows covered in up to 30% of cells using
  coverage-weighted variance (score = var * coverage_fraction). Cells missing
  at the selected HVWs are dropped.

Top `r N_HVW` most variable windows are selected per assay before PCA and UMAP.

```{r win_hvw_pca_umap}
win_embeds <- lapply(names(assay_map), function(akey) {
  hvw_pca_umap(as.matrix(assay(windows_sce, akey)))
})
names(win_embeds) <- assay_map

failed <- names(which(sapply(win_embeds, is.null)))
if (length(failed) > 0) warning("run_embedding returned NULL for: ", paste(failed, collapse = ", "))

cat("\nHVF selection summary (stage 1 = complete features; stage 2 = sparse fallback):\n")
for (nm in names(win_embeds)) {
  em <- win_embeds[[nm]]
  if (is.null(em)) {
    cat(nm, ": NULL\n")
  } else {
    n_cells_used <- sum(em$kept_cols)
    cat(nm, ": HVF =", length(em$hvf_idx),
        "| cells used =", n_cells_used, "/", ncol(windows_sce),
        if (n_cells_used < ncol(windows_sce)) "(sparse fallback)" else "(all cells)",
        "\n")
  }
}

```

```{r save_win_embeds_debug, cache = FALSE}
saveRDS(list(col_data = as.data.frame(colData(windows_sce)),
             row_data = as.data.frame(rowData(windows_sce)),
             assay_names = assayNames(windows_sce),
             sce_dim = dim(windows_sce),
             win_embeds = win_embeds, assay_map = assay_map),
        file.path(params$out_dir, "crc_embeddings_debug.rds"),
        compress = "xz")
```

## By location {.tabset .tabset-pills}

```{r win_loc_S, fig.width = ng_fig_size(2, 1)$w, fig.height = ng_fig_size(2, 1)$h}
### i_total
if (!is.null(win_embeds[["i_total"]])) print(plot_pair(win_embeds[["i_total"]]$umap, windows_sce[, win_embeds[["i_total"]]$kept_cols], "i_total (windows)"))
```

```{r win_loc_adjS, fig.width = ng_fig_size(2, 1)$w, fig.height = ng_fig_size(2, 1)$h}
### adjS
if (!is.null(win_embeds[["adjS"]])) print(plot_pair(win_embeds[["adjS"]]$umap, windows_sce[, win_embeds[["adjS"]]$kept_cols], "adjS (windows)"))
```

<!-- Section dropped: S_cellmeth panel needs a raw S assay that amet does not export. -->

```{r win_loc_meth, fig.width = ng_fig_size(2, 1)$w, fig.height = ng_fig_size(2, 1)$h}
### methylation
if (!is.null(win_embeds[["methylation"]])) print(plot_pair(win_embeds[["methylation"]]$umap, windows_sce[, win_embeds[["methylation"]]$kept_cols], "methylation (windows)"))
```

## Multi-panel comparison (location)

```{r win_panel, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(4, 1)$h}
ok_embeds <- Filter(Negate(is.null), win_embeds)
df_win_panel <- bind_rows(lapply(names(ok_embeds), function(lbl) {
  em <- ok_embeds[[lbl]]
  data.frame(umap1 = em$umap[, 1],
             umap2 = em$umap[, 2],
             location = colData(windows_sce[, em$kept_cols])$location,
             assay = lbl)
}))
df_win_panel$assay <- factor(df_win_panel$assay, levels = assay_map)

ggplot(df_win_panel, aes(x = umap1, y = umap2, color = location)) +
  geom_point(size = 0.4, alpha = 0.6) +
  scale_color_manual(values = crc_location_pal) +
  facet_wrap(~ assay, nrow = 1) +
  labs(x = "UMAP 1", y = "UMAP 2", title = "Per-window embeddings (HVW -> PCA -> UMAP)") +
  guides(x = guide_x_nolap(),
         color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

## PCA (first two PCs, per assay)

```{r win_pca_plot, fig.width = ng_fig_size(4, 2)$w, fig.height = ng_fig_size(4, 2)$h}
df_win_pca <- bind_rows(lapply(names(ok_embeds), function(lbl) {
  em <- ok_embeds[[lbl]]
  cd <- colData(windows_sce[, em$kept_cols])
  data.frame(PC1 = em$pca[, 1], PC2 = em$pca[, 2],
             location = cd$location,
             patient  = cd$patient,
             assay    = lbl)
}))
df_win_pca$assay <- factor(df_win_pca$assay, levels = assay_map)

ggplot(df_win_pca, aes(x = PC1, y = PC2, color = location)) +
  geom_point(size = 0.5, alpha = 0.6) +
  scale_color_manual(values = crc_location_pal) +
  facet_grid(assay ~ .) +
  labs(x = "PC 1", y = "PC 2", title = "PCA: per-window HVW") +
  guides(x = guide_x_nolap(),
         color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

## Variance explained

```{r win_varexp, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(2, 1)$h}
ve_win <- bind_rows(lapply(names(assay_map), function(akey) {
  lbl <- assay_map[[akey]]
  if (is.null(win_embeds[[lbl]])) return(NULL)
  mat <- as.matrix(assay(windows_sce, akey))
  hvw_mat <- mat[win_embeds[[lbl]]$hvf_idx, , drop = FALSE]
  rownames(hvw_mat) <- NULL
  tmp_sce <- SingleCellExperiment(
    assays = list(x = hvw_mat),
    colData = colData(windows_sce)
  )
  ve <- getVarianceExplained(tmp_sce, exprs_values = "x",
                             variables = c("location", "patient"))
  data.frame(assay = lbl,
             location = median(ve[, "location"], na.rm = TRUE),
             patient = median(ve[, "patient"], na.rm = TRUE))
}))
ve_win_long <- pivot_longer(ve_win, c("location", "patient"),
                            names_to = "variable", values_to = "median_ve")
ve_win_long$assay <- factor(ve_win_long$assay, levels = assay_map)

ggplot(ve_win_long, aes(x = assay, y = median_ve, fill = variable)) +
  geom_col(position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = c("location" = "#0072B2", "patient" = "#E69F00")) +
  labs(x = NULL, y = "median % variance explained",
       title = "Variance explained (per-window HVW)") +
  theme_ng()

```

```{r save_win_varexp, cache = FALSE}
saveRDS(list(win_embeds = win_embeds, assay_map = assay_map,
             sce_dim = dim(windows_sce)),
        file.path(params$out_dir, "crc_win_varexp_debug.rds"),
        compress = "xz")
write.csv(ve_win_long, file.path(params$out_dir, "crc_win_varexp.csv"),
          row.names = FALSE)
```

## Silhouette scores

```{r win_sil}
sil_win <- data.frame(
  assay = names(ok_embeds),
  silhouette = sapply(names(ok_embeds), function(lbl)
    sil_score(ok_embeds[[lbl]]$umap,
              colData(windows_sce[, ok_embeds[[lbl]]$kept_cols])$location)),
  feature_space = "windows"
)
print(sil_win)
```

# Per-annotation cell embeddings (mean i_total as features)

amet's window BED has only (chrom, start, end, feature_id) columns; no per-window
genomic-feature annotation matrix is exported. The per-annotation feature space
and the per-annotation HVW -> PCA -> UMAP loop require that matrix and so are
not portable here.

<!-- Section dropped: Per-annotation cell embeddings / Per-annotation HVW -> PCA -> UMAP require a per-window annotation matrix that amet's window BED does not provide. -->

# Summary: all feature spaces and metrics

```{r combined_sil, fig.width = ng_fig_size(4, 3)$w, fig.height = ng_fig_size(4, 1)$h}
sil_all <- sil_win
sil_all$assay         <- factor(sil_all$assay, levels = assay_map)
sil_all$feature_space <- factor(sil_all$feature_space, levels = c("windows"))

ggplot(sil_all, aes(x = assay, y = silhouette, fill = feature_space)) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c("windows" = "#0072B2")) +
  labs(x = NULL, y = "mean silhouette (location clusters)",
       title = "Location separation: per-window embeddings") +
  theme_ng()
```

# Per-cell entropy summaries

```{r cell_summary_df}
cell_df <- data.frame(
  location = colData(windows_sce)$location,
  patient = colData(windows_sce)$patient,
  mean_S = colMeans(assay(windows_sce, "sampen"), na.rm = TRUE),
  mean_adjS = colMeans(assay(windows_sce, "sampen_corrected"), na.rm = TRUE),
  mean_meth = colMeans(assay(windows_sce, "meth"), na.rm = TRUE)
)
```

```{r save_cell_summary, cache = FALSE}
write.csv(cell_df, file.path(params$out_dir, "crc_per_cell_summary.csv"),
          row.names = FALSE)
```

```{r violins, fig.width = ng_fig_size(6, 1)$w, fig.height = ng_fig_size(6, 3)$h}
cell_long <- pivot_longer(cell_df, c("mean_S", "mean_adjS", "mean_meth"),
                          names_to = "metric", values_to = "value")
cell_long$metric <- factor(cell_long$metric,
                           levels = c("mean_S", "mean_adjS", "mean_meth"))

ggplot(cell_long, aes(x = location, y = value, fill = location)) +
  geom_violin(trim = FALSE, scale = "width", alpha = 0.7) +
  geom_boxplot(width = 0.1, outlier.size = 0.3, alpha = 0.6) +
  scale_fill_manual(values = crc_location_pal) +
  facet_grid(metric ~ patient, scales = "free_y") +
  labs(x = "biopsy location", y = "per-cell mean") +
  theme_ng() +
  theme(legend.position = "none", axis.text.x = element_text(size = 6))
```

```{r scatter_S_meth, fig.width = ng_fig_size(6, 1)$w, fig.height = ng_fig_size(6, 1)$h}
ggplot(cell_df, aes(x = mean_meth, y = mean_S, color = location)) +
  geom_point(size = 0.5, alpha = 0.6) +
  geom_smooth(aes(group = location), method = "lm", se = FALSE, linewidth = 0.5) +
  scale_color_manual(values = crc_location_pal) +
  facet_wrap(~ patient, nrow = 1) +
  labs(x = "cell mean methylation", y = expression("cell mean " * i[total])) +
  guides(x = guide_x_nolap(),
         color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

```{r scatter_adjS_meth, fig.width = ng_fig_size(6, 1)$w, fig.height = ng_fig_size(6, 1)$h}
ggplot(cell_df, aes(x = mean_meth, y = mean_adjS, color = location)) +
  geom_point(size = 0.5, alpha = 0.6) +
  geom_smooth(aes(group = location), method = "lm", se = FALSE, linewidth = 0.5) +
  scale_color_manual(values = crc_location_pal) +
  facet_wrap(~ patient, nrow = 1) +
  labs(x = "cell mean methylation", y = "cell mean adjS") +
  guides(x = guide_x_nolap(),
         color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

# Session info

```{r session}
sessionInfo()
```
