---
title: "CRC by genomic tiles, SCE-based"
author: "Izaskun Mallona"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
params:
  features_dir: ""
  windows_dir: ""
  win_bed: ""
  manifest: ""
  out_dir: ""
  windows_sce: ""
  de: ""
  corrected_sce: ""
  windows_annotation: ""
  log_path: ""
  threads: 1
  i_max_lag: 3
---

```{r logging_early, include = FALSE}
source(Sys.getenv("AMET_RENDER_HELPERS"))
param <- amet_setup_render_logging(params$log_path, params$threads)
```

```{r, setup}
suppressPackageStartupMessages({
  library(data.table)
  library(readr)
  library(ggplot2)
  library(knitr)
  library(ComplexHeatmap)
  library(viridis)
  library(tidyr)
  library(Matrix)
  library(limma)
  library(BiocParallel)

  library(dplyr)

  library(SingleCellExperiment)
  library(SummarizedExperiment)

  library(pheatmap)

  library(reshape2)
  library(scater)
})
```

```{r}
repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))
source(file.path(repo_root, "workflow", "scripts", "palettes.R"))
source(file.path(repo_root, "workflow", "scripts", "diff_testing.R"))

```

```{r}
opts_chunk$set(
  fig.width = 5,
  fig.height = 5,
  cache = FALSE,
  error = TRUE,
  include = TRUE,
  fig.path = "crc_windows_plots_sce/",
  dev = c("png", "svg"),
  cache.lazy = FALSE,
  warning = TRUE,
  message = TRUE
)
```


```{r}
windows_sce_path <- if (nzchar(params$windows_sce))
                      params$windows_sce else
                      file.path(params$out_dir, "sce_windows_colon.rds")
de_path <- if (nzchar(params$de))
             params$de else
             file.path(params$out_dir, "de_list.rds")
windows_sce <- readRDS(windows_sce_path)
de <- readRDS(de_path)
```

Let's get some basic stats about numbers of up/down, per comparison

```{r}
summarize_contrast <- function(df, alpha = 0.05) {
  if (is.null(df) || nrow(df) == 0) {
    return(tibble(total_significant = 0L, upregulated = 0L, downregulated = 0L))
  }
  sig <- df %>% filter(!is.na(adj_p), adj_p < alpha)
  tibble(
    total_significant = nrow(sig),
    upregulated = sum(sig$moderated_t >= 0, na.rm = TRUE),
    downregulated = sum(sig$moderated_t < 0, na.rm = TRUE)
  )
}

summary_table <- lapply(names(de), function(name) {
  out <- summarize_contrast(de[[name]]$coefs_df)
  cbind(contrast = name, out)
}) %>% bind_rows()

summary_table

df_long <- melt(
  summary_table,
  id.vars = "contrast",
  measure.vars = c("upregulated", "downregulated"),
  variable.name = "direction",
  value.name = "count"
)

ggplot(df_long, aes(x = contrast, y = count, fill = direction)) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_text(aes(label = count),
            position = position_dodge(width = 0.8),
            vjust = -0.3, size = 2.5) +
  scale_y_continuous(trans = "log1p",
                     breaks = c(0, 1, 10, 100, 1000, 10000),
                     labels = c(0, 1, 10, 100, 1000, 10000)) +
  scale_fill_manual(values = c("upregulated" = "#1b9e77",
                               "downregulated" = "#d95f02")) +
  theme_ng() +
  theme(axis.text = element_text(size = 7),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Differentially entropy windows per contrast",
       y = "Number of windows (log1p scale)",
       x = "Contrast")

```

Filter DE contrasts to those having NC as baseline.

```{r}
de <- de[intersect(c('pt_vs_nc', 'ln_vs_nc', 'ml_vs_nc', 'mp_vs_nc'), names(de))]


summary_table <- lapply(names(de), function(name) {
  out <- summarize_contrast(de[[name]]$coefs_df)
  cbind(contrast = name, out)
}) %>% bind_rows()

summary_table

df_long <- melt(
  summary_table,
  id.vars = "contrast",
  measure.vars = c("upregulated", "downregulated"),
  variable.name = "direction",
  value.name = "count"
)

ggplot(df_long, aes(x = contrast, y = count, fill = direction)) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_text(aes(label = count),
            position = position_dodge(width = 0.8),
            vjust = -0.3, size = 2.5) +
  scale_y_continuous(trans = "log1p",
                     breaks = c(0, 1, 10, 100, 1000, 10000),
                     labels = c(0, 1, 10, 100, 1000, 10000)) +
  scale_fill_manual(values = c("upregulated" = "#1b9e77",
                               "downregulated" = "#d95f02")) +
  theme_ng() +
  theme(axis.text = element_text(size = 7),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Differentially entropy windows per contrast",
       y = "Number of windows (log1p scale)",
       x = "Contrast")
```



For plotting entropies: I cannot use the residuals of the fits above. But I can anyway fit a model on i_total vs avg meth and use the residuals, without patient or location covars, to do general plots. Sort of poor woman's corrected i_total.

**TODO:** the `i_total_resid` assay built below (regression residuals) is *not* the same quantity as `i_norm` (the analytical `i_total / (k_max * H(p_hat))` used in simulations and eval scripts). Both decouple `i_total` from methylation, but via different math. See `TODO.md` -- harmonize on one canonical decoupling strategy across the dataset Rmds, the simulations, and the eval scripts.

```{r}
formula <- "i_total ~ meth + I(meth^2)"

str(assay(windows_sce, 'i_total'))

rowwise_lm_residuals <- function(i) {
  df <- data.frame(
    i_total = assay(windows_sce, 'i_total')[i, ],
    meth = assay(windows_sce, 'meth')[i, ]
  )
  ## na.action = na.exclude keeps residuals aligned to df rows (NA for skipped).
  fit <- try(lm(formula, data = df, na.action = na.exclude), silent = TRUE)
  if (inherits(fit, "try-error")) return(rep(NA_real_, nrow(df)))
  as.numeric(residuals(fit))
}

res_mat <- do.call(rbind, bplapply(seq_len(nrow(windows_sce)), rowwise_lm_residuals,
                                   BPPARAM = param))

rownames(res_mat) <- rownames(windows_sce)
colnames(res_mat) <- colnames(windows_sce)

assay(windows_sce, "i_total_resid") <- res_mat

colData(windows_sce)$location <- factor(
  colData(windows_sce)$location,
  levels = c("NC", "PT", "LN", "ML", "MP", "MO"))
```




```{r}

df <- reshape2::melt(as.matrix(assay(windows_sce, "i_total_resid")[,1:min(200, ncol(windows_sce))]))
ggplot(df, aes(value)) + geom_density(fill="steelblue", alpha=0.5) +
  labs(title="Distribution of corrected i_total values (downsampled)") +
    theme_ng()
```


```{r, fig.width = 8, fig.height = 4}

df_violin <- data.frame(
  value = colMeans(assay(windows_sce, "i_total_resid")),
  location = colData(windows_sce)$location,
  patient = colData(windows_sce)$patient
)

ggplot(df_violin, aes(x = location, y = value)) +
  geom_violin(trim = FALSE, scale = "width") +
  geom_boxplot(width = 0.1, outlier.size = 0.5) +
    facet_wrap(~patient) +
  theme_ng() +
  labs(y = expression("Mean " * i[total] * " corrected"),
       title = "Distribution per location and patient")
```

```{r, fig.width = 8, fig.height = 4}
df_cells <- data.frame(
  mean_i_total = colMeans(assay(windows_sce, "i_total")),
  mean_i_total_resid = colMeans(assay(windows_sce, "i_total_resid")),
  location = colData(windows_sce)$location,
  patient = colData(windows_sce)$patient
)

ggplot(df_cells, aes(x = mean_i_total, y = mean_i_total_resid, color = location)) +
  geom_point(alpha = 0.8) +
  theme_ng() +
    facet_wrap(~patient) +
  labs(x = expression("mean " * i[total]),
       y = expression("mean " * i[total] * " corrected"),
       title = "mean i_total vs mean i_total (corrected), per cell")

```

Row annotation is loaded from the windows_annotation TSV and merged onto the
SCE's rowData. Annotation columns are binarised: any non-zero fractional
overlap becomes 1.

```{r load_window_annotation_sce}
load_window_annotation_sce <- function(path) {
  if (is.null(path) || !nzchar(path) || !file.exists(path)) {
    return(data.frame())
  }
  tab <- tryCatch(
    as.data.frame(fread(path, header = TRUE, sep = "\t")),
    error = function(e) {
      message("annotation read failed: ", conditionMessage(e))
      data.frame()
    })
  if (nrow(tab) == 0) return(tab)
  tab$region <- sprintf("%s:%s-%s", tab$chrom, tab$start, tab$end)
  rownames(tab) <- tab$region
  drop_cols <- c("chrom", "start", "end", "feature_id", "region")
  tab[, setdiff(colnames(tab), drop_cols), drop = FALSE]
}

annot <- load_window_annotation_sce(params$windows_annotation)
annotation_mat_ok <- ncol(annot) > 0 && nrow(annot) > 0

if (annotation_mat_ok) {
  rd_existing <- as.data.frame(rowData(windows_sce))
  rd_new <- annot[match(rownames(windows_sce), rownames(annot)), , drop = FALSE]
  rownames(rd_new) <- rownames(windows_sce)
  keep_existing <- setdiff(colnames(rd_existing), colnames(rd_new))
  if (length(keep_existing) > 0) {
    rd_new <- cbind(rd_new, rd_existing[, keep_existing, drop = FALSE])
  }
  rd_bin <- as.data.frame(lapply(rd_new, function(x) {
    if (is.numeric(x)) as.integer(!is.na(x) & x > 0) else x
  }))
  colnames(rd_bin) <- paste0(colnames(rd_new), "_bin")
  rd_combined <- cbind(rd_new, rd_bin)
  rownames(rd_combined) <- rownames(windows_sce)
  rowData(windows_sce) <- rd_combined
}

cat("annotation columns merged:", ncol(annot),
    "| binarised columns:", if (annotation_mat_ok) sum(grepl("_bin$",
      colnames(rowData(windows_sce)))) else 0, "\n")
```

Tag windows with the up/down significance from each differential-entropy contrast.

```{r tag_de_directions}
for (contrast in names(de)) {
  cf <- de[[contrast]]$coefs_df
  if (is.null(cf) || nrow(cf) == 0) next
  up <- cf$region[!is.na(cf$adj_p) & cf$adj_p < 0.05 & cf$moderated_t >= 0]
  down <- cf$region[!is.na(cf$adj_p) & cf$adj_p < 0.05 & cf$moderated_t < 0]
  rowData(windows_sce)[, paste0(contrast, "_up_bin")] <- as.integer(
    rownames(windows_sce) %in% up)
  rowData(windows_sce)[, paste0(contrast, "_down_bin")] <- as.integer(
    rownames(windows_sce) %in% down)
}
```

```{r save_corrected_sce, cache = FALSE, error = FALSE}
corrected_path <- if (nzchar(params$corrected_sce))
                    params$corrected_sce else
                    file.path(params$out_dir, "sce_windows_colon_corrected.rds")
dir.create(dirname(corrected_path), showWarnings = FALSE, recursive = TRUE)
saveRDS(windows_sce, corrected_path)
```

## Per-cell mean i_total_resid by annotation overlap (CRC01)

Per cell, average corrected i_total over windows split by whether the feature
is present (binary overlap == 1) or absent. CRC01 only so the SCNA columns are
biologically meaningful.

```{r mean_resid_by_annotation, fig.width = 11, fig.height = 4}
if (annotation_mat_ok && "CRC01" %in% colData(windows_sce)$patient) {
  crc01 <- windows_sce[, windows_sce$patient == "CRC01"]
  assay_mat <- assay(crc01, "i_total_resid")

  bin_features <- c("pmds_pmd_bin", "hmds_pmd_bin",
                    "H3K27me3_chip_bin", "H3K9me3_chip_bin", "H3K4me3_chip_bin",
                    "laminb1_lad_bin", "genes_genes_bin",
                    "cpgIslandExt_cpgIslandExt_bin",
                    "crc01_gain_scna_scna_bin", "crc01_lost_scna_scna_bin")
  bin_features <- intersect(bin_features, colnames(rowData(crc01)))

  if (length(bin_features) > 0) {
    cells <- colnames(crc01)
    df_list <- lapply(bin_features, function(bf) {
      bin <- rowData(crc01)[[bf]]
      m0 <- colMeans(assay_mat[bin == 0, , drop = FALSE], na.rm = TRUE)
      m1 <- colMeans(assay_mat[bin == 1, , drop = FALSE], na.rm = TRUE)
      data.frame(
        cell = rep(cells, 2),
        bin = rep(c(0, 1), each = length(cells)),
        mean_i_total_resid = c(m0, m1),
        feature = bf,
        stringsAsFactors = FALSE
      )
    })
    coldata_crc01 <- as.data.frame(colData(crc01))
    coldata_crc01$cell <- colnames(crc01)
    df_crc01 <- bind_rows(df_list) %>%
      left_join(coldata_crc01[, c("cell", "location", "patient")],
                by = "cell")
    df_crc01$bin <- factor(df_crc01$bin, levels = c(0, 1))

    print(
      ggplot(df_crc01, aes(x = location, y = mean_i_total_resid,
                           fill = factor(bin))) +
        geom_violin(trim = FALSE, position = position_dodge(width = 0.8),
                    alpha = 0.6) +
        geom_jitter(aes(color = factor(bin)),
                    position = position_jitterdodge(jitter.width = 0.3,
                                                    dodge.width = 0.8),
                    size = 0.5, alpha = 0.5) +
        facet_wrap(~ feature, nrow = 2) +
        labs(x = "biopsy location", y = "i_total_resid (cell mean)",
             fill = "genomic context", color = "genomic context") +
        theme_ng(base_size = 14)
    )
  }
}
```

## Annotation overlap of differential-entropy windows (CRC01)

```{r annotation_overlap_de, fig.width = 11, fig.height = 4}
if (annotation_mat_ok && "CRC01" %in% colData(windows_sce)$patient) {
  crc01 <- windows_sce[, windows_sce$patient == "CRC01"]
  rd <- as.data.frame(rowData(crc01))
  contrast_features <- grep("_(up|down)_bin$", colnames(rd), value = TRUE)
  non_hmm <- c("pmds_pmd_bin", "hmds_pmd_bin",
               "H3K27me3_chip_bin", "H3K9me3_chip_bin", "H3K4me3_chip_bin",
               "laminb1_lad_bin", "genes_genes_bin",
               "cpgIslandExt_cpgIslandExt_bin",
               "crc01_gain_scna_scna_bin", "crc01_lost_scna_scna_bin")
  non_hmm <- intersect(non_hmm, colnames(rd))
  hmm_features <- grep("_hmm_bin$", colnames(rd), value = TRUE)

  build_prop <- function(annot_features) {
    if (length(annot_features) == 0 || length(contrast_features) == 0)
      return(data.frame())
    grid <- expand.grid(annotation = annot_features,
                        contrast = contrast_features,
                        stringsAsFactors = FALSE)
    grid$proportion <- vapply(seq_len(nrow(grid)), function(i) {
      ann_v <- rd[[grid$annotation[i]]]
      ct_v <- as.logical(rd[[grid$contrast[i]]])
      total <- sum(ct_v, na.rm = TRUE)
      if (total == 0) NA_real_ else sum(ann_v == 1 & ct_v, na.rm = TRUE) / total
    }, numeric(1))
    grid <- grid[!is.na(grid$proportion), ]
    grid$direction <- ifelse(grepl("_up_", grid$contrast), "Up", "Down")
    grid$contrast_clean <- gsub("_(up|down)_bin$", "", grid$contrast)
    grid$annotation <- gsub("_bin$", "", grid$annotation)
    grid
  }

  plot_overlap <- function(df, title) {
    ggplot(df, aes(x = contrast_clean, y = annotation)) +
      geom_point(aes(size = proportion, color = direction)) +
      scale_size(range = c(1, 6)) +
      scale_color_manual(values = c("Up" = "#d95f02", "Down" = "#1b9e77")) +
      facet_wrap(~ direction, nrow = 1) +
      labs(x = "contrast", y = "annotation", size = "proportion",
           color = "direction", title = title) +
      theme_ng() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            axis.text = element_text(size = 7))
  }

  prop_non_hmm <- build_prop(non_hmm)
  if (nrow(prop_non_hmm) > 0)
    print(plot_overlap(prop_non_hmm,
                       "Annotation overlap in diff i_total windows (non-HMM)"))
  prop_hmm <- build_prop(hmm_features)
  if (nrow(prop_hmm) > 0)
    print(plot_overlap(prop_hmm,
                       "Annotation overlap in diff i_total windows (HMM)"))
}
```

# Aggregate by rowdata categories

Per-cell median i_total_resid grouped by each binary annotation overlap column,
appended back to colData.

```{r aggregate_rowdata_categories, fig.width = 9, fig.height = 9}
if (annotation_mat_ok) {
  expr <- assay(windows_sce, "i_total_resid")
  rd_bin_only <- as.data.frame(rowData(windows_sce))[, grep("_bin$",
    colnames(rowData(windows_sce))), drop = FALSE]
  rd_mat <- as.matrix(rd_bin_only)

  cell_ann_median <- matrix(NA_real_, nrow = ncol(expr), ncol = ncol(rd_mat),
                            dimnames = list(colnames(expr),
                                            paste0(colnames(rd_mat), "_median")))
  for (j in seq_len(ncol(rd_mat))) {
    idx <- rd_mat[, j] == 1
    if (sum(idx, na.rm = TRUE) > 0) {
      cell_ann_median[, j] <- apply(expr[idx, , drop = FALSE], 2,
                                    median, na.rm = TRUE)
    }
  }

  colData(windows_sce) <- cbind(
    colData(windows_sce),
    as.data.frame(cell_ann_median)
  )

  df <- as.data.frame(colData(windows_sce))
  ann <- grep("_bin_median$", colnames(df), value = TRUE)
  if (length(ann) > 0) {
    plot_df <- data.frame(
      value = unlist(df[ann]),
      annotation = rep(ann, each = nrow(df)),
      location = rep(df$location, times = length(ann)),
      stringsAsFactors = FALSE
    )
    print(
      ggplot(plot_df, aes(x = location, y = value)) +
        geom_boxplot(outlier.size = 0.5) +
        facet_wrap(~ annotation, scales = "free_y") +
        labs(x = "Tissue type",
             y = "i_total_resid (median per annotation)",
             title = "Per-cell i_total_resid across annotations and biopsy location") +
        theme_ng() +
        theme(strip.text = element_text(size = 7),
              axis.text.x = element_text(angle = 45, hjust = 1))
    )
  }

  saveRDS(windows_sce, corrected_path)
}
```

```{r}
sessionInfo()
```
