---
title: "CRC by genomic tiles, SCE-based"
author: "Izaskun Mallona"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
params:
  features_dir: ""
  windows_dir: ""
  win_bed: ""
  manifest: ""
  out_dir: ""
  windows_sce: ""
  de: ""
  corrected_sce: ""
---

```{r, setup}
suppressPackageStartupMessages({
  library(readr)
  library(ggplot2)
  library(knitr)
  library(ComplexHeatmap)
  library(viridis)
  library(tidyr)
  library(Matrix)
  library(limma)
  library(BiocParallel)

  library(dplyr)

  library(SingleCellExperiment)
  library(SummarizedExperiment)

  library(pheatmap)

  library(reshape2)
  library(scater)
})
```

```{r}
repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))
source(file.path(repo_root, "workflow", "scripts", "palettes.R"))
source(file.path(repo_root, "workflow", "scripts", "diff_testing.R"))

param <- SerialParam()
```

```{r}
opts_chunk$set(
  fig.width = 5,
  fig.height = 5,
  cache = TRUE,
  include = TRUE,
  fig.path = "crc_windows_plots_sce/",
  dev = c("png", "svg"),
  cache.lazy = FALSE,
  warning = TRUE,
  message = TRUE
)
```


```{r}
windows_sce <- readRDS(params$windows_sce)
de          <- readRDS(params$de)
```

Let's get some basic stats about numbers of up/down, per comparison

```{r}
summarize_contrast <- function(df, alpha = 0.05) {
  sig <- df %>% filter(adj_p < alpha)
  tibble(
    total_significant = nrow(sig),
    upregulated   = sum(sig$moderated_t >= 0, na.rm = TRUE),
    downregulated = sum(sig$moderated_t < 0, na.rm = TRUE)
  )
}

summary_table <- lapply(names(de), function(name) {
  out <- summarize_contrast(de[[name]]$coefs_df)
  cbind(contrast = name, out)
}) %>% bind_rows()

summary_table

df_long <- melt(
  summary_table,
  id.vars = "contrast",
  measure.vars = c("upregulated", "downregulated"),
  variable.name = "direction",
  value.name = "count"
)

ggplot(df_long, aes(x = contrast, y = count, fill = direction)) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_text(aes(label = count),
            position = position_dodge(width = 0.8),
            vjust = -0.3, size = 2.5) +
  scale_y_continuous(trans = "log1p",
                     breaks = c(0, 1, 10, 100, 1000, 10000),
                     labels = c(0, 1, 10, 100, 1000, 10000)) +
  scale_fill_manual(values = c("upregulated" = "#1b9e77",
                               "downregulated" = "#d95f02")) +
  theme_ng() +
  theme(axis.text = element_text(size = 7),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Differentially entropy windows per contrast",
       y = "Number of windows (log1p scale)",
       x = "Contrast")

```

Filter DE contrasts to those having NC as baseline.

```{r}
de <- de[c('pt_vs_nc', 'ln_vs_nc', 'ml_vs_nc', 'mp_vs_nc')]


summary_table <- lapply(names(de), function(name) {
  out <- summarize_contrast(de[[name]]$coefs_df)
  cbind(contrast = name, out)
}) %>% bind_rows()

summary_table

df_long <- melt(
  summary_table,
  id.vars = "contrast",
  measure.vars = c("upregulated", "downregulated"),
  variable.name = "direction",
  value.name = "count"
)

ggplot(df_long, aes(x = contrast, y = count, fill = direction)) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_text(aes(label = count),
            position = position_dodge(width = 0.8),
            vjust = -0.3, size = 2.5) +
  scale_y_continuous(trans = "log1p",
                     breaks = c(0, 1, 10, 100, 1000, 10000),
                     labels = c(0, 1, 10, 100, 1000, 10000)) +
  scale_fill_manual(values = c("upregulated" = "#1b9e77",
                               "downregulated" = "#d95f02")) +
  theme_ng() +
  theme(axis.text = element_text(size = 7),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Differentially entropy windows per contrast",
       y = "Number of windows (log1p scale)",
       x = "Contrast")
```



For plotting entropies: I cannot use the residuals of the fits above. But I can anyway fit a model on i_total vs avg meth and use the residuals, without patient or location covars, to do general plots. Sort of poor woman's corrected i_total.

```{r}
formula <- "sampen ~ meth + I(meth^2)"

str(assay(windows_sce, 'sampen'))

rowwise_lm_residuals <- function(i) {
    df <- data.frame(
        sampen  = assay(windows_sce, 'sampen')[i, ],
        meth    = assay(windows_sce, 'meth')[i, ])

    fit <- try(lm(formula, data = df), silent = TRUE)
    if (inherits(fit, "try-error")) return(rep(NA, nrow(df)))
    res <- residuals(fit)
    return(res)
}

res_mat <- do.call(rbind, bplapply(seq_len(nrow(windows_sce)), rowwise_lm_residuals,
                                   BPPARAM = param))

rownames(res_mat) <- rownames(windows_sce)
colnames(res_mat) <- colnames(windows_sce)

assay(windows_sce, "sampen_corrected") <- res_mat

colData(windows_sce)$location <- factor(
  colData(windows_sce)$location,
  levels = c("NC", "PT", "LN", "ML", "MP", "MO"))
```




```{r}

df <- reshape2::melt(as.matrix(assay(windows_sce, "sampen_corrected")[,1:min(200, ncol(windows_sce))]))
ggplot(df, aes(value)) + geom_density(fill="steelblue", alpha=0.5) +
  labs(title="Distribution of corrected i_total values (downsampled)") +
    theme_ng()
```


```{r, fig.width = 8, fig.height = 4}

df_violin <- data.frame(
  value = colMeans(assay(windows_sce, "sampen_corrected")),
  location = colData(windows_sce)$location,
  patient  = colData(windows_sce)$patient
)

ggplot(df_violin, aes(x = location, y = value)) +
  geom_violin(trim = FALSE, scale = "width") +
  geom_boxplot(width = 0.1, outlier.size = 0.5) +
    facet_wrap(~patient) +
  theme_ng() +
  labs(y = expression("Mean " * i[total] * " corrected"),
       title = "Distribution per location and patient")
```

```{r, fig.width = 8, fig.height = 4}
df_cells <- data.frame(
  mean_sampen = colMeans(assay(windows_sce, "sampen")),
  mean_sampen_corrected = colMeans(assay(windows_sce, "sampen_corrected")),
  location = colData(windows_sce)$location,
  patient = colData(windows_sce)$patient
)

ggplot(df_cells, aes(x = mean_sampen, y = mean_sampen_corrected, color = location)) +
  geom_point(alpha = 0.8) +
  theme_ng() +
    facet_wrap(~patient) +
  labs(x = expression("mean " * i[total]),
       y = expression("mean " * i[total] * " corrected"),
       title = "mean i_total vs mean i_total (corrected), per cell")

```

Row annotation (rowData) is re-read from the annotation file here because the SCE was built from a partial read in crc_windows.Rmd; the full annotation is loaded and assigned below.

amet's window BED has only (chrom, start, end, feature_id) and so carries no per-window genomic-feature columns. The annotation reader is replaced with a no-op; binarized annotation columns and their downstream plots are not portable.


```{r}
## The annotation merge / binarization pipeline cannot run on amet's window
## BED. Save the corrected SCE as-is.

corrected_path <- if (nzchar(params$corrected_sce))
                    params$corrected_sce else
                    file.path(params$out_dir, "sce_windows_colon_corrected.rds")
dir.create(dirname(corrected_path), showWarnings = FALSE, recursive = TRUE)
saveRDS(windows_sce, corrected_path)
```

<!-- Section dropped: per-cell averages over binarized genomic-feature columns (genes, H3K27me3, SCNAs, HMM states, ...) contrasting up/down DE windows. amet's window BED carries none of these columns, so these plots cannot be reproduced. -->

# Aggregate by rowdata categories

amet's window BED does not provide per-window genomic feature columns, so the aggregation (per-cell median i_total per genomic category, residual entropy after removing annotation effects) is dropped here.

<!-- Section dropped: Aggregate by rowdata categories / per-cell residuals require binarized rowData annotation columns that amet's window BED does not provide. -->

```{r}
sessionInfo()
```
