---
title: "CRC by genomic tiles, aggregated and unaggregated single cells"
author: "Izaskun Mallona"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
params:
  features_dir: ""
  windows_dir: ""
  win_bed: ""
  manifest: ""
  out_dir: ""
  log_path: ""
---

```{r logging_early, include = FALSE}
if (nzchar(params$log_path)) {
  log_con <- file(params$log_path, open = "at")
  sink(log_con)
  sink(log_con, type = "message")
}
```

```{r, setup}
suppressPackageStartupMessages({
  library(data.table)
  library(readr)
  library(ggplot2)
  library(knitr)
  library(ComplexHeatmap)
  library(viridis)
  library(tidyr)
  library(Matrix)
  library(limma)
  library(BiocParallel)
  library(circlize)

  library(dplyr)
  library(purrr)
  library(uwot)

  library(SingleCellExperiment)
  library(SummarizedExperiment)

  library(UpSetR)
  library(pheatmap)
  library(igraph)
  library(ggraph)
})

repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))
source(file.path(repo_root, "workflow", "scripts", "palettes.R"))
source(file.path(repo_root, "workflow", "scripts", "driver_utils.R"))
source(file.path(repo_root, "workflow", "scripts", "diff_testing.R"))

param <- SerialParam()
```

```{r}
opts_chunk$set(
  fig.width = 5,
  fig.height = 5,
  cache = TRUE,
  error = TRUE,
  include = TRUE,
  fig.path = "crc_windows_plots/",
  dev = c("png", "svg"),
  cache.lazy = FALSE,
  warning = TRUE,
  message = TRUE
)

## render on error hook
knitr::knit_hooks$set(error = function(x, options) {
  knitr::knit_exit()
})
```


# QC: tile annotations  {.tabset .tabset-pills}

To double check annotations are ok. amet's window BED has only feature_id (no per-window genomic-feature columns), so this section is skipped.

```{r, fig.width = 8, fig.height = 5}
## amet's win_bed is just (chrom, start, end, feature_id) with no per-window
## annotation matrix. Show the bed shape for sanity.
win_bed <- fread(params$win_bed, header = FALSE,
                 col.names = c("chrom", "start", "end", "feature_id"))
cat("windows BED:", nrow(win_bed), "rows\n")
print(head(win_bed))
```

```{r read_shannons}
## amet exports per-feature jsd (feature.tsv.gz) and per-cell-per-feature
## i_total (cell_feature.tsv.gz). Build a long `shannons` data.frame plus a
## per-(patient, location) `sampens_list` of sparse matrices keyed by
## region = "<chr>:<start>-<end>".

get_patient  <- function(fn) sub("^([^_]+)_[^_]+\\..*$", "\\1", fn)
get_location <- function(fn) sub("^[^_]+_([^_.]+)\\..*$", "\\1", fn)

cf_files <- list.files(params$windows_dir,
                       pattern = "\\.cell_feature\\.tsv\\.gz$",
                       full.names = TRUE)
fe_files <- list.files(params$windows_dir,
                       pattern = "\\.feature\\.tsv\\.gz$",
                       full.names = TRUE)
fe_files <- fe_files[!grepl("\\.cell_feature\\.tsv\\.gz$", fe_files)]

bed_dt <- as.data.table(win_bed)
bed_dt[, region := sprintf("%s:%s-%s", chrom, start, end)]
fid2region <- setNames(bed_dt$region, bed_dt$feature_id)

regex_combo <- "([^_]+)_([^_.]+)\\..*$"

process_reports <- function(cf_fp) {
  fn <- basename(cf_fp)
  patient  <- sub(regex_combo, "\\1", fn)
  location <- sub(regex_combo, "\\2", fn)

  fe_fp <- file.path(dirname(cf_fp),
                     sub("\\.cell_feature\\.tsv\\.gz$", ".feature.tsv.gz", fn))

  cf <- fread(cf_fp)
  cf[, region := fid2region[feature_id]]
  cf <- cf[!is.na(region)]

  if (file.exists(fe_fp)) {
    fe <- fread(fe_fp)
    fe[, region := fid2region[feature_id]]
    fe <- fe[!is.na(region)]
    dtm <- data.frame(
      shannon  = fe$jsd,
      avg_meth = fe$mean_meth_mean,
      patient  = patient,
      location = location,
      window   = fe$region,
      stringsAsFactors = FALSE
    )
  } else {
    dtm <- data.frame()
  }

  rows <- sort(unique(cf$region))
  cols <- sort(unique(cf$cell_id))
  cf[, ri := match(region, rows)]
  cf[, ci := match(cell_id, cols)]
  se <- sparseMatrix(i = cf$ri, j = cf$ci, x = cf$i_total,
                     dims = c(length(rows), length(cols)),
                     dimnames = list(rows, cols))

  ms <- sparseMatrix(i = cf$ri, j = cf$ci, x = cf$mean_meth,
                     dims = c(length(rows), length(cols)),
                     dimnames = list(rows, cols))

  list(
    shannon = dtm,
    sampen = list(
      file = cf_fp,
      patient = patient,
      location = location,
      sampen = se,
      meth = ms
    )
  )
}
```

```{r}
all_reports <- bplapply(cf_files, process_reports, BPPARAM = param)

shannons <- do.call(rbind, lapply(all_reports, \(x) x$shannon))
shannons$patient  <- as.factor(shannons$patient)
shannons$location <- factor(shannons$location,
                            levels = c("NC","PT","LN","ML","MP","MO"))
rownames(shannons) <- paste(shannons$patient, shannons$location,
                            seq_len(nrow(shannons)), sep = "_")

sampens_list <- lapply(all_reports, \(x) x$sampen)
```


# Single cell visualization/ UMAP and SCE generation {.tabset .tabset-pills}

```{r}
flatten_entry <- function(entry, filter) {
  mat <- as.matrix(entry$sampen)
  df <- as.data.frame(t(mat))
  df$cell <- colnames(entry$sampen)
  df$patient <- entry$patient
  df$location <- entry$location
  df
}

df_all <- map_dfr(sampens_list, flatten_entry)

features <- df_all %>% select(starts_with("chr"))

# drop windows with >20% NAs
keep_cols <- colMeans(is.na(features)) < 0.2
features <- features[, keep_cols]

# drop cells with >20% NAs
keep_rows <- rowMeans(is.na(features)) < 0.2
features <- features[keep_rows, ]
df_all_filtered <- df_all[keep_rows, ]

## drop any remaining NAs to get a complete matrix for UMAP
keep_cols2 <- colSums(is.na(features)) == 0
keep_rows2 <- rowSums(is.na(features[, keep_cols2])) == 0
features <- features[keep_rows2, keep_cols2]
df_all_filtered <- df_all_filtered[keep_rows2, ]

set.seed(123)
umap_coords <- umap(features, n_neighbors = 15, min_dist = 0.3, metric = "euclidean")

plot_df <- cbind(as.data.frame(umap_coords),
                 cell = df_all_filtered$cell,
                 patient = df_all_filtered$patient,
                 location = df_all_filtered$location)
```

## UMAP by location (i_total)

```{r, fig.width = 5, fig.height = 5}
ggplot(plot_df, aes(x = V1, y = V2, color = location)) +
  geom_point(alpha = 1) +
  theme_ng()
```


## UMAP by patient (i_total)

```{r, fig.width = 5, fig.height = 5}
ggplot(plot_df, aes(x = V1, y = V2, color = patient)) +
  geom_point(alpha = 1) +
  theme_ng()
```

And this is with all features; what would it happen if focusing in changing windows only?

## Violins {.tabset .tabset-pills}

### Mean i_total

```{r}
cell_summary <- features %>%
  as.data.frame() %>%
  mutate(mean_sampen = rowMeans(., na.rm = TRUE),
         var_sampen  = apply(., 1, var, na.rm = TRUE)) %>%
  select(mean_sampen, var_sampen) %>%
  bind_cols(df_all_filtered %>% select(cell, patient, location))

ggplot(cell_summary, aes(x = location, y = mean_sampen, fill = location)) +
  geom_violin(trim = FALSE, scale = "width") +
  geom_boxplot(width = 0.1, outlier.size = 0.5, alpha = 0.5) +
  facet_wrap(~patient) +
  theme_ng() +
  labs(title = "Per-cell mean i_total by patient and location",
       x = "Location", y = expression("Mean " * i[total]))
```

### Var i_total

```{r}
ggplot(cell_summary, aes(x = location, y = var_sampen, fill = location)) +
  geom_violin(trim = FALSE, scale = "width") +
  geom_boxplot(width = 0.1, outlier.size = 0.5, alpha = 0.5) +
  facet_wrap(~patient) +
  theme_ng() +
  labs(title = "Per-cell i_total variance by patient and location",
       x = "Location", y = expression("Variance of " * i[total]))

```

## Export SCE

Export a SCE with NA filtering.

```{r}

assay_list <- list(sampen = t(as.matrix(features)))

# Add summary stats
df_all_filtered <- df_all_filtered |>
   mutate(
       mean_sampen = rowMeans(features, na.rm = TRUE),
       var_sampen  = apply(features, 1, var, na.rm = TRUE))

col_data <- data.frame(df_all_filtered[,c('cell', 'patient', 'location', 'mean_sampen', 'var_sampen')])
## amet's window BED has only (chrom, start, end, feature_id), no per-window
## genomic feature annotation. Build a placeholder rowData with the region
## coordinates so the SCE constructor accepts it.
row_data <- data.frame(region = colnames(features),
                       row.names = colnames(features))

windows_sce <- SingleCellExperiment(
  assays = assay_list,
  colData = col_data,
  rowData = row_data
)

# add assay with DNA meth as well; amet's cell_feature.tsv.gz already carries
# mean_meth per (cell, window) so we transpose the per-combo matrices, glue
# them along the cell axis, and align rows/cols with the existing sampen assay.
meth_mats <- lapply(sampens_list, function(x) x$meth)
meth_mat  <- do.call(cbind, meth_mats)
rm(meth_mats)

meth_mat <- meth_mat[rownames(windows_sce), colnames(assay(windows_sce, "sampen"))]

stopifnot(
  identical(rownames(assay(windows_sce, "sampen")), rownames(meth_mat)),
  identical(colnames(assay(windows_sce, "sampen")), colnames(meth_mat)))

assay(windows_sce, "meth") <- meth_mat

rm(meth_mat)
```

```{r save_sce_windows, cache = FALSE, error = FALSE}
stopifnot(exists("windows_sce"))
sce_path <- file.path(params$out_dir, "sce_windows_colon.rds")
dir.create(dirname(sce_path), showWarnings = FALSE, recursive = TRUE)
saveRDS(windows_sce, file = sce_path)
```

# QC  {.tabset .tabset-pills}

Back to the SCE-free plotting/reports

```{r}
sampens_list <- bplapply(sampens_list, \(x) {
  x$median <- apply(
    x$sampen, 1, function(x) median(x, na.rm = TRUE)
  )
  x$range <- suppressWarnings(
    apply(x$sampen, 1, function(x) max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
  )
  x
}, BPPARAM = param)
```


## By location - jsd

```{r, fig.width = 10, fig.height = 10}
par(mfrow = c(3, 2), pty = "s")
for (loc in levels(shannons$location)) {
  tmp <- shannons[shannons$location == loc, ]
  plot(shannon ~ avg_meth, col = as.numeric(as.factor(tmp$location)), pch = ".", data = tmp, main = loc)
}
```


## By patient (colored by location as before) - jsd

```{r, fig.width = 10, fig.height = 10}
par(mfrow = c(3, 3), pty = "s")
for (patient in levels(shannons$patient)) {
  tmp <- shannons[shannons$patient == patient, ]
  plot(shannon ~ avg_meth, col = as.numeric(as.factor(tmp$location)), pch = ".", data = tmp, main = patient[1])
}
```

## By patient and location - jsd

```{r, fig.width = 8, fig.height = 11}
par(mfrow = c(4, 4), pty = "s")
for (patient in levels(shannons$patient)) {
  tmp <- shannons[shannons$patient == patient, ]
  for (loc in unique(tmp$location)) {
    tmp <- shannons[shannons$patient == patient & shannons$location == loc, ]
    plot(shannon ~ avg_meth,
      col = as.numeric(as.factor(tmp$location)),
      pch = ".", data = tmp, main = paste(loc, patient[1])
    )
  }
}
```

# Stats

## Original by Atreya

We reread everything from scratch (!); todo streamline.

```{r}
process_reports2 <- function(cf_fp) {
  fn <- basename(cf_fp)
  patient  <- sub(regex_combo, "\\1", fn)
  location <- sub(regex_combo, "\\2", fn)

  fe_fp <- file.path(dirname(cf_fp),
                     sub("\\.cell_feature\\.tsv\\.gz$", ".feature.tsv.gz", fn))

  cf <- fread(cf_fp)
  cf[, region := fid2region[feature_id]]
  cf <- cf[!is.na(region)]

  fe <- if (file.exists(fe_fp)) fread(fe_fp) else data.table()
  if (nrow(fe) > 0) {
    fe[, region := fid2region[feature_id]]
    fe <- fe[!is.na(region)]
  }

  rows <- sort(unique(cf$region))
  cols <- sort(unique(cf$cell_id))
  cf[, ri := match(region, rows)]
  cf[, ci := match(cell_id, cols)]
  se <- sparseMatrix(i = cf$ri, j = cf$ci, x = cf$i_total,
                     dims = c(length(rows), length(cols)),
                     dimnames = list(rows, cols))
  ms <- sparseMatrix(i = cf$ri, j = cf$ci, x = cf$mean_meth,
                     dims = c(length(rows), length(cols)),
                     dimnames = list(rows, cols))

  out <- cf[, .(cell_meth = mean(mean_meth, na.rm = TRUE)), by = cell_id]
  setorder(out, cell_id)

  shannon <- if (nrow(fe) > 0)
    fe[, .(shannon = jsd, avg_meth = mean_meth_mean)] else
    data.table(shannon = numeric(0), avg_meth = numeric(0))

  list(
    shannon = shannon,
    sampen = se, meth = ms, cell_meth = out$cell_meth
  )
}

all_data <- bplapply(cf_files, process_reports2, BPPARAM = param)
sampens <- do.call(cbind, lapply(all_data, \(x) x$sampen))
meths   <- do.call(cbind, lapply(all_data, \(x) x$meth))


meta <- data.frame(
  colname = colnames(sampens),
  patient = sub(regex_combo, "\\1", basename(cf_files))[
    rep(seq_along(cf_files), sapply(all_data, \(x) ncol(x$sampen)))],
  subloc  = sub(regex_combo, "\\2", basename(cf_files))[
    rep(seq_along(cf_files), sapply(all_data, \(x) ncol(x$sampen)))],
  cell_meth = unlist(lapply(all_data, \(x) x$cell_meth), recursive = FALSE)
)
meta$loc <- substr(meta$subloc, 1, 2)
meta$loc <- relevel(factor(meta$loc), ref = "NC")
meta <- meta[order(as.integer(meta$loc), meta$subloc, meta$patient), ]
sampens <- sampens[, meta$colname]
meths <- meths[, meta$colname]

groups <- unique(meta[c("subloc", "patient")])
groups <- groups[
  order(substr(groups$subloc, 1, 2), groups$subloc, groups$patient),
]
sub_sampens <- do.call(cbind, lapply(seq_len(nrow(groups)), \(i) {
  cols <- meta$colname[
    meta$patient == groups$patient[i] & meta$subloc == groups$subloc[i]
  ]
  rowMeans(sampens[, cols, drop = FALSE], na.rm = TRUE)
}))
colnames(sub_sampens) <- paste(groups$subloc, groups$patient, sep = "_")

sub_meths <- do.call(cbind, lapply(seq_len(nrow(groups)), \(i) {
  cols <- meta$colname[
    meta$patient == groups$patient[i] & meta$subloc == groups$subloc[i]
  ]
  rowMeans(meths[, cols, drop = FALSE], na.rm = TRUE)
}))
colnames(sub_meths) <- paste(groups$subloc, groups$patient, sep = "_")
```

So these data group entropies/i_total from different biopsy:patient pairs


```{r}
row_na_frac <- \(x) mean(is.na(x)) >= 0.5
rows_too_sparse <- apply(sub_sampens, 1, row_na_frac) |
  apply(sub_meths, 1, row_na_frac)

sub_sampens <- sub_sampens[!rows_too_sparse, , drop = FALSE]
sub_meths <- sub_meths[!rows_too_sparse, , drop = FALSE]
```



```{r, fig.width = 5, fig.height = 5}
c02_nc <- data.frame(y = sub_sampens[, "NC_CRC02"], x = sub_meths[, "NC_CRC02"])
c02_nc <- na.omit(c02_nc)
sampled_data <- if (nrow(c02_nc) > 5000)
                  c02_nc[sample(nrow(c02_nc), 5000), ] else c02_nc

ggplot(sampled_data, aes(x = x, y = y)) +
  geom_point(alpha = 0.6, size = 0.6) +
  labs(
    title = "Adjacent normal in CRC02",
    x = "average methylation",
    y = expression(i[total])
  ) +
  theme_ng()
```


Let's do a window-wise i_total vs meth with loc and patient, regardless of the annotation of the window. Mind the quadratic meth term (not sure what for).

```{r,  fig.width = 5, fig.height = 5}
rowwise_lm <- function(i) {
  df <- data.frame(
    sampen = sub_sampens[i, ],
    meth = sub_meths[i, ],
    loc = factor(substr(groups$subloc, 1, 2)),
    patient = groups$patient
  )
  df$loc <- relevel(df$loc, ref = "NC")

  fit <- try(lm(sampen ~ meth + I(meth^2) + loc + patient, data = df), silent = T)
  s <- summary(fit)$coefficients

  if (!("locPT" %in% rownames(s))) {
    return(rep(NA, 5))
  }

  c(
    estimate  = s["locPT", "Estimate"],
    std_error = s["locPT", "Std. Error"],
    t_value   = s["locPT", "t value"],
    p_value   = s["locPT", "Pr(>|t|)"],
    df        = df.residual(fit)
  )
}

coefs_list <- bplapply(seq_len(nrow(sub_sampens)), rowwise_lm, BPPARAM = param)

coefs_df <- do.call(rbind, coefs_list)
colnames(coefs_df) <- c("estimate", "std_error", "t_value", "p_value", "df")
coefs_df <- as.data.frame(coefs_df)


coefs_df[] <- lapply(coefs_df, as.numeric)

valid <- complete.cases(coefs_df)
coefs_valid <- coefs_df[valid, ]

s2 <- coefs_valid$std_error^2
df_resid <- coefs_valid$df
squeezed <- squeezeVar(var = s2, df = df_resid)

moderated_t <- coefs_valid$estimate / sqrt(squeezed$var.post)
moderated_p <- 2 * pt(-abs(moderated_t), df = squeezed$df.prior + df_resid)
adj_p <- p.adjust(moderated_p, method = "BH")

coefs_valid$moderated_t <- moderated_t
coefs_valid$moderated_p <- moderated_p
coefs_valid$adj_p <- adj_p

coefs_df$moderated_t <- NA
coefs_df$moderated_p <- NA
coefs_df$adj_p <- NA
coefs_df[
  valid, c("moderated_t", "moderated_p", "adj_p")
] <- coefs_valid[, c("moderated_t", "moderated_p", "adj_p")]


coefs_df$region <- rownames(sub_sampens)

saveRDS(object = coefs_df, 'atreyas_coefs.rds')

table(coefs_df$adj_p < 0.05)
hist(coefs_df$adj_p)

sorted_idx <- order(coefs_df$adj_p, na.last = NA)

top_2k_idx <- sorted_idx[1:min(2000, length(sorted_idx))]

top_entropy <- sub_sampens[top_2k_idx, ]
top_meth <- sub_meths[top_2k_idx, ]
```


## Heatmap including diff entropic (Atreya's)

Rows are restricted to the top 2000 windows by BH-adjusted p-value from the PT vs NC
comparison (meth + location + patient model), then further to complete cases across all
biopsy groups. Clustering all ~300k windows genome-wide is not tractable.
Columns are biopsy groups (patient x location), annotated at the top.

amet's window BED carries no per-window genomic feature columns, so the
right-hand row annotation is a placeholder showing significance only.

```{r, fig.width = 16, fig.height = 10}
stopifnot(all(rownames(top_entropy) == rownames(top_meth)))

cc_top <- complete.cases(top_entropy) & complete.cases(top_meth)
top_entropy_cc <- top_entropy[cc_top, ]
top_meth_cc <- top_meth[cc_top, ]
coefs_df_top <- coefs_df[top_2k_idx[cc_top], ]

row.hc <- hclust(dist(top_entropy_cc), method = "ward.D2")

column_ha <- HeatmapAnnotation(
  location = substr(groups$subloc, 1, 2),
  patient = groups$patient
)

row_ha <- rowAnnotation(tile_annotation = as.matrix(data.frame(
  diff_sampen_atreya = coefs_df_top$adj_p < 0.05
)))

Heatmap(top_entropy_cc,
  name = expression(i[total]),
  top_annotation = column_ha,
  right_annotation = row_ha,
  show_row_names = FALSE,
  show_column_names = FALSE,
  cluster_columns = FALSE,
  cluster_rows = row.hc
) + Heatmap(top_meth_cc,
  name = "DNA methylation",
  col = colorRamp2(c(0, 1), c("white", "black")),
  show_row_names = FALSE,
  show_column_names = FALSE,
  cluster_columns = FALSE,
  cluster_rows = row.hc
)
```

# H vs S driver categorization {.tabset .tabset-pills}

amet exports per-feature `jsd` (across-cell heterogeneity). adjS analogue
(per-window median i_total) is computed from `sampens_list`. Per-annotation
R2 cannot be built without a per-window annotation matrix, so this section
reports only the global adjS / jsd long tables.

```{r read_adj_shannons}
adj_shannons_long <- shannons %>%
  transmute(window, adjH = shannon, patient, location)
adj_shannons_long$location <- factor(adj_shannons_long$location,
                                     levels = c("NC", "PT", "LN", "ML", "MP", "MO"))

cat("adjH rows:", nrow(adj_shannons_long),
    "| groups:", length(unique(paste(adj_shannons_long$patient,
                                     adj_shannons_long$location))), "\n")
```

```{r build_adj_sampens_long}
adj_sampens_long <- lapply(sampens_list, function(s) {
  data.frame(
    window   = names(s$median),
    adjS     = s$median,
    patient  = s$patient,
    location = s$location,
    stringsAsFactors = FALSE
  )
}) %>% bind_rows()

adj_sampens_long$location <- factor(adj_sampens_long$location,
                                    levels = c("NC", "PT", "LN", "ML", "MP", "MO"))

cat("adjS rows:", nrow(adj_sampens_long), "\n")
```

```{r r2_driver_windows}
saveRDS(list(adj_shannons_long = adj_shannons_long,
             adj_sampens_long  = adj_sampens_long),
        file = "crc_windows_driver_r2.rds")
```

## adjH vs adjS median per location

```{r adjH_adjS_medians, fig.width = 5, fig.height = 5}
hm_win <- inner_join(
  adj_shannons_long %>%
    group_by(annotation = "all windows", location) %>%
    summarise(adjH = median(adjH, na.rm = TRUE), .groups = "drop"),
  adj_sampens_long %>%
    group_by(annotation = "all windows", location) %>%
    summarise(adjS = median(adjS, na.rm = TRUE), .groups = "drop"),
  by = c("annotation", "location")
)

ggplot(hm_win, aes(x = adjH, y = adjS, color = location, label = location)) +
  geom_point(size = 3) +
  scale_color_manual(values = crc_location_pal) +
  ggrepel::geom_text_repel(size = 3) +
  labs(x = "median jsd", y = expression("median " * i[total]),
       title = "Global jsd vs i_total per biopsy location") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
```

# Stats / planned comparisons {.tabset .tabset-pills}

## PT vs NC


```{r}
de <- list()
```

```{r ptnc, fig.width = 5, fig.height = 5}
de$pt_vs_nc <- diff_entropy_test(
  sub_sampens = sub_sampens,
  sub_meths   = sub_meths,
  groups      = groups,
  formula     = "sampen ~ meth + I(meth^2) + loc + patient",
  loc_levels  = c("PT","NC"),
  ref_level   = "NC",
  contrast    = "locPT",
  param       = param,
  top_n       = 2000,
  out_file    = "pt_vs_nc_coefs.rds"
)

head(de$pt_vs_nc$coefs_df)
hist(coefs_df$adj_p)
table(de$pt_vs_nc$coefs_df$adj_p < 0.05)
```

## LN vs PT


```{r, fig.width = 5, fig.height = 5}
de$ln_vs_pt <- diff_entropy_test(
  sub_sampens = sub_sampens,
  sub_meths   = sub_meths,
  groups      = groups,
  formula     = "sampen ~ meth + I(meth^2) + loc + patient",
  loc_levels  = c("PT","LN"),
  ref_level   = "PT",
  contrast    = "locLN",
  param       = param,
  top_n       = 2000,
  out_file    = "ln_vs_pt_coefs.rds"
)

head(de$ln_vs_pt$coefs_df)
hist(de$ln_vs_pt$coefs_df$p_value)
table(de$ln_vs_pt$coefs_df$adj_p < 0.05)
```

## LN vs NC

```{r lnnc, fig.width = 5, fig.height = 5}
de$ln_vs_nc <- diff_entropy_test(
  sub_sampens = sub_sampens,
  sub_meths   = sub_meths,
  groups      = groups,
  formula     = "sampen ~ meth + I(meth^2) + loc + patient",
  loc_levels  = c("NC","LN"),
  ref_level   = "NC",
  contrast    = "locLN",
  param       = param,
  top_n       = 2000,
  out_file    = "ln_vs_nc_coefs.rds"
)

hist(de$ln_vs_nc$coefs_df$p_value)
table(de$ln_vs_nc$coefs_df$adj_p < 0.05)
```

## ML vs LN

```{r mlln, fig.width = 5, fig.height = 5}
de$ml_vs_ln <- diff_entropy_test(
  sub_sampens = sub_sampens,
  sub_meths   = sub_meths,
  groups      = groups,
  formula     = "sampen ~ meth + I(meth^2) + loc + patient",
  loc_levels  = c("ML","LN"),
  ref_level   = "LN",
  contrast    = "locML",
  param       = param,
  top_n       = 2000,
  out_file    = "ml_vs_ln_coefs.rds"
)

hist(de$ml_vs_ln$coefs_df$p_value)
table(de$ml_vs_ln$coefs_df$adj_p < 0.05)
```

## ML vs NC


```{r mlnc, fig.width = 5, fig.height = 5}
de$ml_vs_nc <- diff_entropy_test(
  sub_sampens = sub_sampens,
  sub_meths   = sub_meths,
  groups      = groups,
  formula     = "sampen ~ meth + I(meth^2) + loc + patient",
  loc_levels  = c("ML","NC"),
  ref_level   = "NC",
  contrast    = "locML",
  param       = param,
  top_n       = 2000,
  out_file    = "ml_vs_nc_coefs.rds"
)

hist(de$ml_vs_nc$coefs_df$p_value)
table(de$ml_vs_nc$coefs_df$adj_p < 0.05)
```

## MP vs ML

```{r mpml, fig.width = 5, fig.height = 5}
de$mp_vs_ml <- diff_entropy_test(
  sub_sampens = sub_sampens,
  sub_meths   = sub_meths,
  groups      = groups,
  formula     = "sampen ~ meth + I(meth^2) + loc + patient",
  loc_levels  = c("MP","ML"),
  ref_level   = "ML",
  contrast    = "locMP",
  param       = param,
  top_n       = 2000,
  out_file    = "mp_vs_ml_coefs.rds"
)

hist(de$mp_vs_ml$coefs_df$p_value)
table(de$mp_vs_ml$coefs_df$adj_p < 0.05)
```

## MP vs NC

```{r mpnc, fig.width = 5, fig.height = 5}
de$mp_vs_nc <- diff_entropy_test(
  sub_sampens = sub_sampens,
  sub_meths   = sub_meths,
  groups      = groups,
  formula     = "sampen ~ meth + I(meth^2) + loc + patient",
  loc_levels  = c("MP","NC"),
  ref_level   = "NC",
  contrast    = "locMP",
  param       = param,
  top_n       = 2000,
  out_file    = "mp_vs_nc_coefs.rds"
)

hist(de$mp_vs_nc$coefs_df$p_value)
table(de$mp_vs_nc$coefs_df$adj_p < 0.05)
```

```{r save_de_list, cache = FALSE, error = FALSE}
stopifnot(exists("de"))
saveRDS(de, file = file.path(params$out_dir, "de_list.rds"))
```

# Plotting {.tabset .tabset-pills}

amet's window BED has no per-window genomic-feature annotation matrix; UpSetR / Jaccard-graph / pheatmap chunks require that matrix and so are not portable. They are dropped here.

<!-- Section dropped: Plotting / Gains and losses of entropy / UpSetR / Graph Jaccards / Pheatmap Jaccards / Up and down-entropic separately / GO require a per-window annotation matrix that amet does not export. -->

# Session

```{r}
sessionInfo()
```
