---
title: "CRC by genomic tiles, aggregated and unaggregated single cells"
author: "Izaskun Mallona"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
params:
  features_dir: ""
  windows_dir: ""
  win_bed: ""
  manifest: ""
  out_dir: ""
  windows_annotation: ""
  log_path: ""
  threads: 1
  i_max_lag: 3
---

```{r logging_early, include = FALSE}
source(Sys.getenv("AMET_RENDER_HELPERS"))
param <- amet_setup_render_logging(params$log_path, params$threads)
```

```{r, setup}
suppressPackageStartupMessages({
  library(data.table)
  library(readr)
  library(ggplot2)
  library(knitr)
  library(ComplexHeatmap)
  library(viridis)
  library(tidyr)
  library(Matrix)
  library(limma)
  library(BiocParallel)
  library(circlize)

  library(dplyr)
  library(purrr)
  library(uwot)

  library(SingleCellExperiment)
  library(SummarizedExperiment)

  library(UpSetR)
  library(pheatmap)
  library(igraph)
  library(ggraph)
})

repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))
source(file.path(repo_root, "workflow", "scripts", "palettes.R"))
source(file.path(repo_root, "workflow", "scripts", "driver_utils.R"))
source(file.path(repo_root, "workflow", "scripts", "diff_testing.R"))

```

```{r}
opts_chunk$set(
  fig.width = 5,
  fig.height = 5,
  cache = FALSE,
  error = TRUE,
  include = TRUE,
  fig.path = "crc_windows_plots/",
  dev = c("png", "svg"),
  cache.lazy = FALSE,
  warning = TRUE,
  message = TRUE
)

```


# QC: tile annotations  {.tabset .tabset-pills}

The per-window annotation matrix is built upstream by `crc_combine_window_annotations`:
one column per (subcat, cat) pair holding the fraction of window bases covered by
the annotation BED. Empty path or read error degrades gracefully to an empty
matrix so downstream chunks can skip.

```{r, fig.width = 8, fig.height = 5}
win_bed <- fread(params$win_bed, header = FALSE,
                 col.names = c("chrom", "start", "end", "feature_id"))
cat("windows BED:", nrow(win_bed), "rows\n")
print(head(win_bed))

load_window_annotation <- function(path, win_bed) {
  if (is.null(path) || !nzchar(path) || !file.exists(path)) {
    return(data.frame(row.names = sprintf(
      "%s:%s-%s", win_bed$chrom, win_bed$start, win_bed$end)))
  }
  tab <- tryCatch(
    as.data.frame(fread(path, header = TRUE, sep = "\t")),
    error = function(e) {
      message("annotation read failed: ", conditionMessage(e))
      NULL
    })
  if (is.null(tab) || nrow(tab) == 0) {
    return(data.frame(row.names = sprintf(
      "%s:%s-%s", win_bed$chrom, win_bed$start, win_bed$end)))
  }
  tab$region <- sprintf("%s:%s-%s", tab$chrom, tab$start, tab$end)
  rownames(tab) <- tab$region
  drop_cols <- c("chrom", "start", "end", "feature_id", "region")
  tab[, setdiff(colnames(tab), drop_cols), drop = FALSE]
}

annot <- load_window_annotation(params$windows_annotation, win_bed)
cat("annotation columns:", ncol(annot),
    "| rows:", nrow(annot), "\n")
if (ncol(annot) > 0 && nrow(annot) > 0) {
  hc_sample <- head(annot, 5e4)
  hc_sample <- hc_sample[, sapply(hc_sample, function(x)
    is.numeric(x) && sum(!is.na(x)) > 0), drop = FALSE]
  if (ncol(hc_sample) >= 2)
    try(plot(hclust(dist(t(hc_sample))),
             main = "annotation columns dendrogram"),
        silent = TRUE)
}
```

```{r read_jsd_long}
## amet exports per-feature jsd (feature.tsv.gz) and per-cell-per-feature
## i_total (cell_feature.tsv.gz). Build a long `jsd_long` data.frame plus a
## per-(patient, location) `i_total_list` of sparse matrices keyed by
## region = "<chr>:<start>-<end>".

get_patient <- function(fn) sub("^([^_]+)_[^_]+\\..*$", "\\1", fn)
get_location <- function(fn) sub("^[^_]+_([^_.]+)\\..*$", "\\1", fn)

cf_files <- list.files(params$windows_dir,
                       pattern = "\\.cell_feature\\.tsv\\.gz$",
                       full.names = TRUE)
fe_files <- list.files(params$windows_dir,
                       pattern = "\\.feature\\.tsv\\.gz$",
                       full.names = TRUE)
fe_files <- fe_files[!grepl("\\.cell_feature\\.tsv\\.gz$", fe_files)]

bed_dt <- as.data.table(win_bed)
bed_dt[, region := sprintf("%s:%s-%s", chrom, start, end)]
fid2region <- setNames(bed_dt$region, bed_dt$feature_id)

regex_combo <- "([^_]+)_([^_.]+)\\..*$"

process_reports <- function(cf_fp) {
  fn <- basename(cf_fp)
  patient <- sub(regex_combo, "\\1", fn)
  location <- sub(regex_combo, "\\2", fn)

  fe_fp <- file.path(dirname(cf_fp),
                     sub("\\.cell_feature\\.tsv\\.gz$", ".feature.tsv.gz", fn))

  cf <- fread(cf_fp)
  cf[, region := fid2region[feature_id]]
  cf <- cf[!is.na(region)]

  if (file.exists(fe_fp)) {
    fe <- fread(fe_fp)
    fe[, region := fid2region[feature_id]]
    fe <- fe[!is.na(region)]
    dtm <- data.frame(
      jsd = fe$jsd,
      avg_meth = fe$mean_meth_mean,
      patient = patient,
      location = location,
      window = fe$region,
      stringsAsFactors = FALSE
    )
  } else {
    dtm <- data.frame()
  }

  rows <- sort(unique(cf$region))
  cols <- sort(unique(cf$cell_id))
  cf[, ri := match(region, rows)]
  cf[, ci := match(cell_id, cols)]
  se <- sparseMatrix(i = cf$ri, j = cf$ci, x = cf$i_total,
                     dims = c(length(rows), length(cols)),
                     dimnames = list(rows, cols))

  ms <- sparseMatrix(i = cf$ri, j = cf$ci, x = cf$mean_meth,
                     dims = c(length(rows), length(cols)),
                     dimnames = list(rows, cols))

  list(
    jsd = dtm,
    i_total = list(
      file = cf_fp,
      patient = patient,
      location = location,
      i_total = se,
      meth = ms
    )
  )
}
```

```{r}
all_reports <- bplapply(cf_files, process_reports, BPPARAM = param)

jsd_long <- do.call(rbind, lapply(all_reports, \(x) x$jsd))
jsd_long$patient <- as.factor(jsd_long$patient)
jsd_long$location <- factor(jsd_long$location,
                            levels = c("NC","PT","LN","ML","MP","MO"))
rownames(jsd_long) <- paste(jsd_long$patient, jsd_long$location,
                            seq_len(nrow(jsd_long)), sep = "_")

i_total_list <- lapply(all_reports, \(x) x$i_total)
```


# Single cell visualization/ UMAP and SCE generation {.tabset .tabset-pills}

```{r}
flatten_entry <- function(entry, filter) {
  mat <- as.matrix(entry$i_total)
  df <- as.data.frame(t(mat))
  df$cell <- colnames(entry$i_total)
  df$patient <- entry$patient
  df$location <- entry$location
  df
}

df_all <- map_dfr(i_total_list, flatten_entry)

features <- df_all %>% select(starts_with("chr"))

# drop windows with >20% NAs
keep_cols <- colMeans(is.na(features)) < 0.2
features <- features[, keep_cols]

# drop cells with >20% NAs
keep_rows <- rowMeans(is.na(features)) < 0.2
features <- features[keep_rows, ]
df_all_filtered <- df_all[keep_rows, ]

## drop any remaining NAs to get a complete matrix for UMAP
keep_cols2 <- colSums(is.na(features)) == 0
keep_rows2 <- rowSums(is.na(features[, keep_cols2])) == 0
features <- features[keep_rows2, keep_cols2]
df_all_filtered <- df_all_filtered[keep_rows2, ]

set.seed(123)
umap_coords <- umap(features, n_neighbors = 15, min_dist = 0.3, metric = "euclidean")

plot_df <- cbind(as.data.frame(umap_coords),
                 cell = df_all_filtered$cell,
                 patient = df_all_filtered$patient,
                 location = df_all_filtered$location)
```

## UMAP by location (i_total)

```{r, fig.width = 5, fig.height = 5}
ggplot(plot_df, aes(x = V1, y = V2, color = location)) +
  geom_point(alpha = 1) +
  theme_ng()
```


## UMAP by patient (i_total)

```{r, fig.width = 5, fig.height = 5}
ggplot(plot_df, aes(x = V1, y = V2, color = patient)) +
  geom_point(alpha = 1) +
  theme_ng()
```

And this is with all features; what would it happen if focusing in changing windows only?

## Violins {.tabset .tabset-pills}

### Mean i_total

```{r}
cell_summary <- features %>%
  as.data.frame() %>%
  mutate(mean_i_total = rowMeans(., na.rm = TRUE),
         var_i_total = apply(., 1, var, na.rm = TRUE)) %>%
  select(mean_i_total, var_i_total) %>%
  bind_cols(df_all_filtered %>% select(cell, patient, location))

ggplot(cell_summary, aes(x = location, y = mean_i_total, fill = location)) +
  geom_violin(trim = FALSE, scale = "width") +
  geom_boxplot(width = 0.1, outlier.size = 0.5, alpha = 0.5) +
  facet_wrap(~patient) +
  theme_ng() +
  labs(title = "Per-cell mean i_total by patient and location",
       x = "Location", y = expression("Mean " * i[total]))
```

### Var i_total

```{r}
ggplot(cell_summary, aes(x = location, y = var_i_total, fill = location)) +
  geom_violin(trim = FALSE, scale = "width") +
  geom_boxplot(width = 0.1, outlier.size = 0.5, alpha = 0.5) +
  facet_wrap(~patient) +
  theme_ng() +
  labs(title = "Per-cell i_total variance by patient and location",
       x = "Location", y = expression("Variance of " * i[total]))

```

## Export SCE

Export a SCE with NA filtering.

```{r}

assay_list <- list(i_total = t(as.matrix(features)))

# Add summary stats
df_all_filtered <- df_all_filtered |>
   mutate(
       mean_i_total = rowMeans(features, na.rm = TRUE),
       var_i_total = apply(features, 1, var, na.rm = TRUE))

col_data <- data.frame(df_all_filtered[,c('cell', 'patient', 'location', 'mean_i_total', 'var_i_total')])
## rowData carries per-window annotation overlap fractions when available;
## otherwise just the region id so the SCE constructor accepts it.
row_data <- data.frame(region = colnames(features),
                       row.names = colnames(features))
if (ncol(annot) > 0) {
  row_data <- cbind(row_data,
                    annot[match(rownames(row_data), rownames(annot)), ,
                          drop = FALSE])
}

windows_sce <- SingleCellExperiment(
  assays = assay_list,
  colData = col_data,
  rowData = row_data
)

# add assay with DNA meth as well; amet's cell_feature.tsv.gz already carries
# mean_meth per (cell, window) so we transpose the per-combo matrices, glue
# them along the cell axis, and align rows/cols with the existing i_total assay.
meth_mats <- lapply(i_total_list, function(x) x$meth)
meth_mat <- do.call(cbind, meth_mats)
rm(meth_mats)

meth_mat <- meth_mat[rownames(windows_sce), colnames(assay(windows_sce, "i_total"))]

stopifnot(
  identical(rownames(assay(windows_sce, "i_total")), rownames(meth_mat)),
  identical(colnames(assay(windows_sce, "i_total")), colnames(meth_mat)))

assay(windows_sce, "meth") <- meth_mat

rm(meth_mat)
```

```{r save_sce_windows, cache = FALSE, error = FALSE}
stopifnot(exists("windows_sce"))
sce_path <- file.path(params$out_dir, "sce_windows_colon.rds")
dir.create(dirname(sce_path), showWarnings = FALSE, recursive = TRUE)
saveRDS(windows_sce, file = sce_path)
message("save_sce_windows: wrote ", sce_path, " (",
        file.size(sce_path), " bytes, dim ",
        paste(dim(windows_sce), collapse = "x"), ")")
```

# QC  {.tabset .tabset-pills}

Back to the SCE-free plotting/reports

```{r}
i_total_list <- bplapply(i_total_list, \(x) {
  x$median <- apply(
    x$i_total, 1, function(x) median(x, na.rm = TRUE)
  )
  x$range <- suppressWarnings(
    apply(x$i_total, 1, function(x) max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
  )
  x
}, BPPARAM = param)
```


## By location - jsd

```{r, fig.width = 10, fig.height = 10}
par(mfrow = c(3, 2), pty = "s")
for (loc in levels(jsd_long$location)) {
  tmp <- jsd_long[jsd_long$location == loc, ]
  if (nrow(tmp) == 0) next
  plot(jsd ~ avg_meth, col = as.numeric(as.factor(tmp$location)), pch = ".", data = tmp, main = loc)
}
```


## By patient (colored by location as before) - jsd

```{r, fig.width = 10, fig.height = 10}
par(mfrow = c(3, 3), pty = "s")
for (patient in levels(jsd_long$patient)) {
  tmp <- jsd_long[jsd_long$patient == patient, ]
  if (nrow(tmp) == 0) next
  plot(jsd ~ avg_meth, col = as.numeric(as.factor(tmp$location)), pch = ".", data = tmp, main = patient[1])
}
```

## By patient and location - jsd

```{r, fig.width = 8, fig.height = 11}
par(mfrow = c(4, 4), pty = "s")
for (patient in levels(jsd_long$patient)) {
  tmp <- jsd_long[jsd_long$patient == patient, ]
  for (loc in unique(tmp$location)) {
    tmp <- jsd_long[jsd_long$patient == patient & jsd_long$location == loc, ]
    if (nrow(tmp) == 0) next
    plot(jsd ~ avg_meth,
      col = as.numeric(as.factor(tmp$location)),
      pch = ".", data = tmp, main = paste(loc, patient[1])
    )
  }
}
```

# Stats

## Manual PT vs NC contrast

Reread everything from scratch; this duplicates what `diff_entropy_test()` produces for the `pt_vs_nc` contrast below. Kept for now until the downstream heatmap is rewired.

```{r}
process_reports2 <- function(cf_fp) {
  fn <- basename(cf_fp)
  patient <- sub(regex_combo, "\\1", fn)
  location <- sub(regex_combo, "\\2", fn)

  fe_fp <- file.path(dirname(cf_fp),
                     sub("\\.cell_feature\\.tsv\\.gz$", ".feature.tsv.gz", fn))

  cf <- fread(cf_fp)
  cf[, region := fid2region[feature_id]]
  cf <- cf[!is.na(region)]

  fe <- if (file.exists(fe_fp)) fread(fe_fp) else data.table()
  if (nrow(fe) > 0) {
    fe[, region := fid2region[feature_id]]
    fe <- fe[!is.na(region)]
  }

  rows <- sort(unique(cf$region))
  cols <- sort(unique(cf$cell_id))
  cf[, ri := match(region, rows)]
  cf[, ci := match(cell_id, cols)]
  se <- sparseMatrix(i = cf$ri, j = cf$ci, x = cf$i_total,
                     dims = c(length(rows), length(cols)),
                     dimnames = list(rows, cols))
  ms <- sparseMatrix(i = cf$ri, j = cf$ci, x = cf$mean_meth,
                     dims = c(length(rows), length(cols)),
                     dimnames = list(rows, cols))

  out <- cf[, .(cell_meth = mean(mean_meth, na.rm = TRUE)), by = cell_id]
  setorder(out, cell_id)

  jsd_dt <- if (nrow(fe) > 0)
    fe[, .(jsd = jsd, avg_meth = mean_meth_mean)] else
    data.table(jsd = numeric(0), avg_meth = numeric(0))

  list(
    jsd = jsd_dt,
    i_total = se, meth = ms, cell_meth = out$cell_meth
  )
}

all_data <- bplapply(cf_files, process_reports2, BPPARAM = param)
i_total_mat <- do.call(cbind, lapply(all_data, \(x) x$i_total))
meths <- do.call(cbind, lapply(all_data, \(x) x$meth))


meta <- data.frame(
  colname = colnames(i_total_mat),
  patient = sub(regex_combo, "\\1", basename(cf_files))[
    rep(seq_along(cf_files), sapply(all_data, \(x) ncol(x$i_total)))],
  subloc = sub(regex_combo, "\\2", basename(cf_files))[
    rep(seq_along(cf_files), sapply(all_data, \(x) ncol(x$i_total)))],
  cell_meth = unlist(lapply(all_data, \(x) x$cell_meth), recursive = FALSE)
)
meta$loc <- substr(meta$subloc, 1, 2)
meta$loc <- relevel(factor(meta$loc), ref = "NC")
meta <- meta[order(as.integer(meta$loc), meta$subloc, meta$patient), ]
i_total_mat <- i_total_mat[, meta$colname]
meths <- meths[, meta$colname]

groups <- unique(meta[c("subloc", "patient")])
groups <- groups[
  order(substr(groups$subloc, 1, 2), groups$subloc, groups$patient),
]
sub_i_total <- do.call(cbind, lapply(seq_len(nrow(groups)), \(i) {
  cols <- meta$colname[
    meta$patient == groups$patient[i] & meta$subloc == groups$subloc[i]
  ]
  rowMeans(i_total_mat[, cols, drop = FALSE], na.rm = TRUE)
}))
colnames(sub_i_total) <- paste(groups$subloc, groups$patient, sep = "_")

sub_meth <- do.call(cbind, lapply(seq_len(nrow(groups)), \(i) {
  cols <- meta$colname[
    meta$patient == groups$patient[i] & meta$subloc == groups$subloc[i]
  ]
  rowMeans(meths[, cols, drop = FALSE], na.rm = TRUE)
}))
colnames(sub_meth) <- paste(groups$subloc, groups$patient, sep = "_")
```

So these data group entropies/i_total from different biopsy:patient pairs


```{r}
row_na_frac <- \(x) mean(is.na(x)) >= 0.5
rows_too_sparse <- apply(sub_i_total, 1, row_na_frac) |
  apply(sub_meth, 1, row_na_frac)

sub_i_total <- sub_i_total[!rows_too_sparse, , drop = FALSE]
sub_meth <- sub_meth[!rows_too_sparse, , drop = FALSE]
```



```{r, fig.width = 5, fig.height = 5}
nc_cols <- grep("^NC_", colnames(sub_i_total), value = TRUE)
if (length(nc_cols) > 0) {
  pick <- nc_cols[1]
  nc_df <- data.frame(y = sub_i_total[, pick], x = sub_meth[, pick])
  nc_df <- na.omit(nc_df)
  if (nrow(nc_df) > 0) {
    sampled_data <- if (nrow(nc_df) > 5000) nc_df[sample(nrow(nc_df), 5000), ] else nc_df
    print(
      ggplot(sampled_data, aes(x = x, y = y)) +
        geom_point(alpha = 0.6, size = 0.6) +
        labs(
          title = paste("Adjacent normal in", sub("^NC_", "", pick)),
          x = "average methylation",
          y = expression(i[total])
        ) +
        theme_ng()
    )
  } else {
    message("no finite NC rows in ", pick, "; skipping NC scatter")
  }
} else {
  message("no NC columns in sub_i_total; skipping NC scatter")
}
```


Let's do a window-wise i_total vs meth with loc and patient, regardless of the annotation of the window. Mind the quadratic meth term (not sure what for).

```{r,  fig.width = 5, fig.height = 5}
rowwise_lm <- function(i) {
  df <- data.frame(
    i_total = sub_i_total[i, ],
    meth = sub_meth[i, ],
    loc = factor(substr(groups$subloc, 1, 2)),
    patient = groups$patient
  )
  df$loc <- relevel(df$loc, ref = "NC")

  fit <- try(lm(i_total ~ meth + I(meth^2) + loc + patient, data = df), silent = T)
  s <- summary(fit)$coefficients

  if (!("locPT" %in% rownames(s))) {
    return(rep(NA, 5))
  }

  c(
    estimate = s["locPT", "Estimate"],
    std_error = s["locPT", "Std. Error"],
    t_value = s["locPT", "t value"],
    p_value = s["locPT", "Pr(>|t|)"],
    df = df.residual(fit)
  )
}

coefs_list <- bplapply(seq_len(nrow(sub_i_total)), rowwise_lm, BPPARAM = param)

coefs_df <- do.call(rbind, coefs_list)
colnames(coefs_df) <- c("estimate", "std_error", "t_value", "p_value", "df")
coefs_df <- as.data.frame(coefs_df)


coefs_df[] <- lapply(coefs_df, as.numeric)

valid <- complete.cases(coefs_df)
coefs_valid <- coefs_df[valid, ]

s2 <- coefs_valid$std_error^2
df_resid <- coefs_valid$df
squeezed <- squeezeVar(var = s2, df = df_resid)

moderated_t <- coefs_valid$estimate / sqrt(squeezed$var.post)
moderated_p <- 2 * pt(-abs(moderated_t), df = squeezed$df.prior + df_resid)
adj_p <- p.adjust(moderated_p, method = "BH")

coefs_valid$moderated_t <- moderated_t
coefs_valid$moderated_p <- moderated_p
coefs_valid$adj_p <- adj_p

coefs_df$moderated_t <- NA
coefs_df$moderated_p <- NA
coefs_df$adj_p <- NA
coefs_df[
  valid, c("moderated_t", "moderated_p", "adj_p")
] <- coefs_valid[, c("moderated_t", "moderated_p", "adj_p")]


coefs_df$region <- rownames(sub_i_total)

saveRDS(object = coefs_df, 'pt_vs_nc_manual_coefs.rds')

table(coefs_df$adj_p < 0.05)
hist(coefs_df$adj_p)

sorted_idx <- order(coefs_df$adj_p, na.last = NA)

top_2k_idx <- sorted_idx[1:min(2000, length(sorted_idx))]

top_entropy <- sub_i_total[top_2k_idx, ]
top_meth <- sub_meth[top_2k_idx, ]
```


## Heatmap including differentially entropic windows

Rows are restricted to the top 2000 windows by BH-adjusted p-value from the PT vs NC
comparison (meth + location + patient model), then further to complete cases across all
biopsy groups. Clustering all ~300k windows genome-wide is not tractable.
Columns are biopsy groups (patient x location), annotated at the top.

Right-hand row annotation carries the differential-entropy flag plus the
binarised per-window genomic-feature overlaps when those columns are present
in the SCE rowData.

```{r, fig.width = 16, fig.height = 10}
stopifnot(all(rownames(top_entropy) == rownames(top_meth)))

cc_top <- complete.cases(top_entropy) & complete.cases(top_meth)
top_entropy_cc <- top_entropy[cc_top, , drop = FALSE]
top_meth_cc <- top_meth[cc_top, , drop = FALSE]
coefs_df_top <- coefs_df[top_2k_idx[cc_top], ]

if (nrow(top_entropy_cc) >= 2) {
  row.hc <- hclust(dist(top_entropy_cc), method = "ward.D2")

  column_ha <- HeatmapAnnotation(
    location = substr(groups$subloc, 1, 2),
    patient = groups$patient
  )

  row_ann_df <- data.frame(
    diff_i_total_sig = coefs_df_top$adj_p < 0.05
  )
  if (ncol(annot) > 0 && nrow(annot) > 0) {
    overlay <- annot[coefs_df_top$region, , drop = FALSE]
    overlay[] <- lapply(overlay, function(x) {
      if (is.numeric(x)) as.integer(!is.na(x) & x > 0) else x
    })
    row_ann_df <- cbind(row_ann_df, overlay)
  }
  row_ha <- rowAnnotation(tile_annotation = as.matrix(row_ann_df))

  draw(Heatmap(top_entropy_cc,
    name = "i_total",
    top_annotation = column_ha,
    right_annotation = row_ha,
    show_row_names = FALSE,
    show_column_names = FALSE,
    cluster_columns = FALSE,
    cluster_rows = row.hc
  ) + Heatmap(top_meth_cc,
    name = "DNA methylation",
    col = colorRamp2(c(0, 1), c("white", "black")),
    show_row_names = FALSE,
    show_column_names = FALSE,
    cluster_columns = FALSE,
    cluster_rows = row.hc
  ))
} else {
  message("differential heatmap: fewer than 2 complete-case windows, skipping")
}
```

# jsd vs i_total per-window summaries {.tabset .tabset-pills}

amet exports per-feature `jsd` (across-cell heterogeneity) and per-window
median `i_total` (within-cell heterogeneity). Per-annotation R2 of location
on each summary statistic is reported when the annotation matrix is present.

```{r build_jsd_median_long}
jsd_median_long <- jsd_long %>%
  transmute(window, jsd_median = jsd, patient, location)
jsd_median_long$location <- factor(jsd_median_long$location,
                                   levels = c("NC", "PT", "LN", "ML", "MP", "MO"))

cat("jsd_median rows:", nrow(jsd_median_long),
    "| groups:", length(unique(paste(jsd_median_long$patient,
                                     jsd_median_long$location))), "\n")
```

```{r build_i_total_median_long}
i_total_median_long <- lapply(i_total_list, function(s) {
  data.frame(
    window = names(s$median),
    i_total_median = s$median,
    patient = s$patient,
    location = s$location,
    stringsAsFactors = FALSE
  )
}) %>% bind_rows()

i_total_median_long$location <- factor(i_total_median_long$location,
                                    levels = c("NC", "PT", "LN", "ML", "MP", "MO"))

cat("i_total_median rows:", nrow(i_total_median_long), "\n")
```

```{r r2_driver_windows}
r2_safe_win <- function(y, grp) {
  df <- data.frame(y = y, g = grp)
  df <- df[!is.na(df$y), ]
  if (nrow(df) < 3 || length(unique(df$g)) < 2) return(NA_real_)
  tryCatch(summary(lm(y ~ g, data = df))$r.squared,
           error = function(e) NA_real_)
}

if (ncol(annot) > 0 && nrow(annot) > 0) {
  ann_cols_win <- colnames(annot)[sapply(annot, function(x)
    is.numeric(x) && any(!is.na(x) & x > 0))]
  common_win <- intersect(jsd_median_long$window, i_total_median_long$window)
  common_win <- intersect(common_win, rownames(annot))

  r2_drivers_win <- purrr::compact(lapply(ann_cols_win, function(a) {
    members <- rownames(annot)[!is.na(annot[[a]]) & annot[[a]] > 0]
    members <- intersect(members, common_win)
    if (length(members) < 10) return(NULL)
    sub_J <- jsd_median_long[jsd_median_long$window %in% members, ]
    sub_I <- i_total_median_long[i_total_median_long$window %in% members, ]
    data.frame(
      annotation = a,
      r2_jsd_median = r2_safe_win(sub_J$jsd_median, sub_J$location),
      r2_i_total_median = r2_safe_win(sub_I$i_total_median, sub_I$location),
      n_windows = length(members)
    )
  })) %>% bind_rows()
} else {
  r2_drivers_win <- data.frame(annotation = character(),
                               r2_jsd_median = numeric(),
                               r2_i_total_median = numeric(),
                               n_windows = integer())
}

saveRDS(list(jsd_median_long = jsd_median_long,
             i_total_median_long = i_total_median_long,
             r2_drivers_win = r2_drivers_win),
        file = "crc_windows_driver_r2.rds")

r2_drivers_win
```

## jsd_median vs i_total_median median per location

```{r adjH_adjS_medians, fig.width = 5, fig.height = 5}
hm_win <- inner_join(
  jsd_median_long %>%
    group_by(annotation = "all windows", location) %>%
    summarise(jsd_median = median(jsd_median, na.rm = TRUE), .groups = "drop"),
  i_total_median_long %>%
    group_by(annotation = "all windows", location) %>%
    summarise(i_total_median = median(i_total_median, na.rm = TRUE), .groups = "drop"),
  by = c("annotation", "location")
)

ggplot(hm_win, aes(x = jsd_median, y = i_total_median, color = location, label = location)) +
  geom_point(size = 3) +
  scale_color_manual(values = crc_location_pal) +
  ggrepel::geom_text_repel(size = 3) +
  labs(x = "median jsd", y = expression("median " * i[total]),
       title = "Global jsd vs i_total per biopsy location") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
```

# Stats / planned comparisons {.tabset .tabset-pills}

## PT vs NC


```{r}
de <- list()
```

```{r ptnc, fig.width = 5, fig.height = 5}
de$pt_vs_nc <- diff_entropy_test(
  sub_i_total = sub_i_total,
  sub_meth = sub_meth,
  groups = groups,
  formula = "i_total ~ meth + I(meth^2) + loc + patient",
  loc_levels = c("PT","NC"),
  ref_level = "NC",
  contrast = "locPT",
  param = param,
  top_n = 2000,
  out_file = "pt_vs_nc_coefs.rds"
)

head(de$pt_vs_nc$coefs_df)
if (any(is.finite(de$pt_vs_nc$coefs_df$adj_p))) hist(de$pt_vs_nc$coefs_df$adj_p)
table(de$pt_vs_nc$coefs_df$adj_p < 0.05)
```

## LN vs PT


```{r, fig.width = 5, fig.height = 5}
de$ln_vs_pt <- diff_entropy_test(
  sub_i_total = sub_i_total,
  sub_meth = sub_meth,
  groups = groups,
  formula = "i_total ~ meth + I(meth^2) + loc + patient",
  loc_levels = c("PT","LN"),
  ref_level = "PT",
  contrast = "locLN",
  param = param,
  top_n = 2000,
  out_file = "ln_vs_pt_coefs.rds"
)

head(de$ln_vs_pt$coefs_df)
hist(de$ln_vs_pt$coefs_df$p_value)
table(de$ln_vs_pt$coefs_df$adj_p < 0.05)
```

## LN vs NC

```{r lnnc, fig.width = 5, fig.height = 5}
de$ln_vs_nc <- diff_entropy_test(
  sub_i_total = sub_i_total,
  sub_meth = sub_meth,
  groups = groups,
  formula = "i_total ~ meth + I(meth^2) + loc + patient",
  loc_levels = c("NC","LN"),
  ref_level = "NC",
  contrast = "locLN",
  param = param,
  top_n = 2000,
  out_file = "ln_vs_nc_coefs.rds"
)

hist(de$ln_vs_nc$coefs_df$p_value)
table(de$ln_vs_nc$coefs_df$adj_p < 0.05)
```

## ML vs LN

```{r mlln, fig.width = 5, fig.height = 5}
de$ml_vs_ln <- diff_entropy_test(
  sub_i_total = sub_i_total,
  sub_meth = sub_meth,
  groups = groups,
  formula = "i_total ~ meth + I(meth^2) + loc + patient",
  loc_levels = c("ML","LN"),
  ref_level = "LN",
  contrast = "locML",
  param = param,
  top_n = 2000,
  out_file = "ml_vs_ln_coefs.rds"
)

hist(de$ml_vs_ln$coefs_df$p_value)
table(de$ml_vs_ln$coefs_df$adj_p < 0.05)
```

## ML vs NC


```{r mlnc, fig.width = 5, fig.height = 5}
de$ml_vs_nc <- diff_entropy_test(
  sub_i_total = sub_i_total,
  sub_meth = sub_meth,
  groups = groups,
  formula = "i_total ~ meth + I(meth^2) + loc + patient",
  loc_levels = c("ML","NC"),
  ref_level = "NC",
  contrast = "locML",
  param = param,
  top_n = 2000,
  out_file = "ml_vs_nc_coefs.rds"
)

hist(de$ml_vs_nc$coefs_df$p_value)
table(de$ml_vs_nc$coefs_df$adj_p < 0.05)
```

## MP vs ML

```{r mpml, fig.width = 5, fig.height = 5}
de$mp_vs_ml <- diff_entropy_test(
  sub_i_total = sub_i_total,
  sub_meth = sub_meth,
  groups = groups,
  formula = "i_total ~ meth + I(meth^2) + loc + patient",
  loc_levels = c("MP","ML"),
  ref_level = "ML",
  contrast = "locMP",
  param = param,
  top_n = 2000,
  out_file = "mp_vs_ml_coefs.rds"
)

hist(de$mp_vs_ml$coefs_df$p_value)
table(de$mp_vs_ml$coefs_df$adj_p < 0.05)
```

## MP vs NC

```{r mpnc, fig.width = 5, fig.height = 5}
de$mp_vs_nc <- diff_entropy_test(
  sub_i_total = sub_i_total,
  sub_meth = sub_meth,
  groups = groups,
  formula = "i_total ~ meth + I(meth^2) + loc + patient",
  loc_levels = c("MP","NC"),
  ref_level = "NC",
  contrast = "locMP",
  param = param,
  top_n = 2000,
  out_file = "mp_vs_nc_coefs.rds"
)

hist(de$mp_vs_nc$coefs_df$p_value)
table(de$mp_vs_nc$coefs_df$adj_p < 0.05)
```

```{r save_de_list, cache = FALSE, error = FALSE}
stopifnot(exists("de"))
de_path <- file.path(params$out_dir, "de_list.rds")
message("save_de_list: de has ", length(de), " entries (",
        paste(names(de), collapse = ", "), ")")
saveRDS(de, file = de_path)
message("save_de_list: wrote ", de_path, " (", file.size(de_path), " bytes)")
```

# Plotting {.tabset .tabset-pills}

UpSetR / Jaccard / pheatmap blocks on (annotation x differential-entropy)
co-occurrence. All gated on the annotation matrix being available.

```{r build_full_annot}
full_annot_ok <- ncol(annot) > 0 && nrow(annot) > 0
if (full_annot_ok) {
  full_annot <- annot
  ## conserved/no-change SCNA is dropped because it is the implicit baseline.
  full_annot <- full_annot[, setdiff(colnames(full_annot),
                                      "crc01_nc_scna"), drop = FALSE]
  for (contrast in names(de)) {
    coef_tab <- de[[contrast]]$coefs_df
    tmp <- merge(data.frame(region = rownames(full_annot)),
                 coef_tab[, c("region", "moderated_t", "adj_p")],
                 by = "region", all.x = TRUE, sort = FALSE)
    full_annot[[paste0(contrast, "__mod_t")]] <- tmp$moderated_t
    full_annot[[paste0(contrast, "__signif")]] <- !is.na(tmp$adj_p) & tmp$adj_p < 0.05
  }
} else {
  full_annot <- data.frame()
  message("annotation matrix unavailable, skipping UpSetR / Jaccard chunks")
}
```

## UpSetR

```{r upset, fig.width = 16, fig.height = 10}
if (full_annot_ok) {
  mat <- full_annot
  mat <- mat[, !grepl("__mod_t$", colnames(mat)), drop = FALSE]
  signif_cols <- grep("__signif$", names(mat), value = TRUE)

  mat[sapply(mat, is.logical)] <- lapply(mat[sapply(mat, is.logical)], as.integer)
  rows_with_signif_de <- rowSums(mat[, signif_cols, drop = FALSE], na.rm = TRUE) > 0
  mat <- mat[rows_with_signif_de, , drop = FALSE]

  if (nrow(mat) > 0 && ncol(mat) > 0) {
    mat_bin <- as.data.frame((mat > 0) * 1)
    upset(mat_bin,
          nsets = ncol(mat_bin),
          nintersects = 30,
          order.by = "freq")
  } else {
    message("upset: no significant rows, skipping")
  }
} else {
  message("upset: annotation matrix unavailable")
}
```

## Graph Jaccards

```{r jaccard_graph}
if (full_annot_ok && exists("mat_bin") && ncol(mat_bin) >= 2) {
  jaccard <- function(x, y) {
    u <- sum(x | y)
    if (u == 0) NA_real_ else sum(x & y) / u
  }
  assoc <- outer(names(mat_bin), names(mat_bin),
                 Vectorize(function(a, b) jaccard(mat_bin[[a]], mat_bin[[b]])))
  dimnames(assoc) <- list(names(mat_bin), names(mat_bin))
  assoc[is.na(assoc)] <- 0
  g <- graph_from_adjacency_matrix(assoc, mode = "undirected",
                                   weighted = TRUE, diag = FALSE)
  V(g)$is_signif <- grepl("__signif$", V(g)$name)
  V(g)$color <- ifelse(V(g)$is_signif, "tomato", "skyblue")

  coords <- layout_with_fr(g, niter = 1e4)
  w <- E(g)$weight
  edge_cols <- colorRampPalette(c("lightgrey", "darkblue"))(100)
  edge_cols_alpha <- if (length(w) > 0 && diff(range(w)) > 0) {
    a <- (w - min(w)) / diff(range(w))
    paste0(edge_cols[cut(w, breaks = 100)], sprintf("%02X", as.integer(a * 255)))
  } else {
    rep(edge_cols[1], length(w))
  }
  plot(g, vertex.size = 7, vertex.label.cex = 0.8,
       vertex.color = V(g)$color, layout = coords,
       edge.width = E(g)$weight * 10,
       edge.color = edge_cols_alpha)
} else {
  message("jaccard graph: insufficient annotation matrix or no significant rows")
}
```

## Pheatmap Jaccards

```{r jaccard_pheatmap, fig.width = 8, fig.height = 8}
if (full_annot_ok && exists("assoc") && nrow(assoc) >= 2) {
  assoc_pheat <- assoc
  assoc_pheat[!is.finite(assoc_pheat)] <- 0
  pheatmap(assoc_pheat,
           clustering_distance_rows = "euclidean",
           clustering_distance_cols = "euclidean",
           main = "Feature co-annotation (jaccard)")
}
```

## Annotation overlap in differential-entropy windows (HMM and non-HMM)

```{r annotation_overlap_dotplots, fig.width = 11, fig.height = 4}
if (full_annot_ok) {
  rd_full <- as.data.frame(full_annot)
  signif_cols <- grep("__signif$", names(rd_full), value = TRUE)
  modt_cols <- grep("__mod_t$", names(rd_full), value = TRUE)
  annot_features <- setdiff(names(rd_full), c(signif_cols, modt_cols))

  rd_bin <- as.data.frame(lapply(rd_full[, annot_features, drop = FALSE],
                                  function(x) {
    if (is.numeric(x)) as.integer(!is.na(x) & x > 0) else as.integer(as.logical(x))
  }))

  build_prop_df <- function(annot_set) {
    grid <- expand.grid(annotation = annot_set,
                        contrast = signif_cols,
                        direction = c("Up", "Down"),
                        stringsAsFactors = FALSE)
    grid$proportion <- vapply(seq_len(nrow(grid)), function(i) {
      ann <- grid$annotation[i]
      ct <- grid$contrast[i]
      dir <- grid$direction[i]
      sig_flag <- as.logical(rd_full[[ct]])
      mod_t <- rd_full[[sub("__signif$", "__mod_t", ct)]]
      pos <- if (dir == "Up") mod_t >= 0 else mod_t < 0
      total <- sum(sig_flag & pos, na.rm = TRUE)
      if (total == 0) return(NA_real_)
      hits <- sum(rd_bin[[ann]] == 1 & sig_flag & pos, na.rm = TRUE)
      hits / total
    }, numeric(1))
    grid$contrast_clean <- gsub("__signif$", "", grid$contrast)
    grid[!is.na(grid$proportion), ]
  }

  hmm_features <- grep("_hmm$", annot_features, value = TRUE)
  non_hmm <- setdiff(annot_features, hmm_features)

  if (length(non_hmm) > 0) {
    prop_non_hmm <- build_prop_df(non_hmm)
    print(
      ggplot(prop_non_hmm, aes(x = contrast_clean, y = annotation)) +
        geom_point(aes(size = proportion, color = direction)) +
        scale_size(range = c(1, 6)) +
        scale_color_manual(values = c("Up" = "#d95f02", "Down" = "#1b9e77")) +
        facet_wrap(~ direction, nrow = 1) +
        labs(x = "contrast", y = "annotation",
             title = "Annotation overlap in diff i_total windows (non-HMM)") +
        theme_ng() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1),
              axis.text = element_text(size = 7))
    )
  }
  if (length(hmm_features) > 0) {
    prop_hmm <- build_prop_df(hmm_features)
    print(
      ggplot(prop_hmm, aes(x = contrast_clean, y = annotation)) +
        geom_point(aes(size = proportion, color = direction)) +
        scale_size(range = c(1, 6)) +
        scale_color_manual(values = c("Up" = "#d95f02", "Down" = "#1b9e77")) +
        facet_wrap(~ direction, nrow = 1) +
        labs(x = "contrast", y = "annotation",
             title = "Annotation overlap in diff i_total windows (HMM)") +
        theme_ng() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1),
              axis.text = element_text(size = 7))
    )
  }
}
```

# Session

```{r}
sessionInfo()
```
