---
title: "Ecker 2021 mouse brain: methylation entropy by feature (CpG context)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Izaskun Mallona"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
params:
  features_dir: ""
  win_cell_feature: ""
  win_feature: ""
  win_bed: ""
  manifest: ""
  out_dir: ""
  log_path: ""
  threads: 1
  i_max_lag: 3
---

```{r logging_early, include = FALSE}
source(Sys.getenv("AMET_RENDER_HELPERS"))
param <- amet_setup_render_logging(params$log_path, params$threads)
```

```{r setup}
suppressPackageStartupMessages({
  library(data.table)
  library(readr)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(BiocParallel)
  library(ggrepel)
  library(ggridges)
  library(ComplexHeatmap)
  library(circlize)
  library(uwot)
  library(irlba)
})

repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))
source(file.path(repo_root, "workflow", "scripts", "palettes.R"))
source(file.path(repo_root, "workflow", "scripts", "embedding_utils.R"))
source(file.path(repo_root, "workflow", "scripts", "driver_utils.R"))
```

```{r opts}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 5,
  fig.height = 5,
  cache = FALSE,
  include = TRUE,
  fig.path = "ecker_plots/",
  dev = c("png", "svg"),
  dpi = 500,
  cache.lazy = FALSE,
  warning = TRUE,
  message = TRUE
)
```

Mouse (mm10), CpG context only.

```{r helpers}
## filenames: {region}_{sub_type}.{annotation}.*.gz
## amet writes one triplet per --features BED; the label is the BED basename
## (the annotation). region, sub_type and annotation contain no underscores
## or dots.
get_annotation <- function(fn) sub("^[^_.]+_[^_.]+\\.([^_.]+)\\..*$", "\\1", fn)
## Slab axis (e.g. 2C/3C/4B/5D), the first filename token.
get_region <- function(fn) sub("^([^_.]+)_[^_.]+\\..*$", "\\1", fn)
get_sub_type <- function(fn) sub("^[^_.]+_([^_.]+)\\..*$", "\\1", fn)

ann_labels <- c(
  "genes"    = "Genes",
  "promoters"= "Promoters",
  "lines"    = "LINEs",
  "sines"    = "SINEs",
  "h3k4me3"  = "H3K4me3",
  "h3k9me3"  = "H3K9me3",
  "h3k27me3" = "H3K27me3",
  "h3k4me1"  = "H3K4me1",
  "h3k27ac"  = "H3K27ac"
)

harmonize_levels <- function(df) {
  missing_ann <- setdiff(unique(df$annotation), names(ann_labels))
  if (length(missing_ann) > 0)
    warning("unknown annotations: ", paste(missing_ann, collapse = ", "))
  df$annotation <- factor(df$annotation,
                           levels = names(ann_labels),
                           labels = unname(ann_labels))
  if ("region" %in% names(df)) df$region <- factor(df$region)
  if ("sub_type"   %in% names(df)) df$sub_type <- factor(df$sub_type)
  df
}

## _sanitize for Ecker: gsub(" ", "-", x).
sanitize <- function(x) gsub(" ", "-", x)
```

```{r load_metadata_early}
## Load cell-level metadata early so cell_class / major_type are available
## in all data-loading chunks below. amet's manifest carries region,
## sub_type, cell_class, major_type per cell (see make_manifest_ecker.py).
man <- fread(params$manifest)
meta_cells <- as.data.frame(man) %>%
  mutate(
    region = sanitize(as.character(region)),
    sub_type = sanitize(as.character(sub_type))
  )

meta_grp <- meta_cells %>%
  group_by(region, sub_type) %>%
  summarise(
    cell_class = if ("cell_class" %in% names(meta_cells))
                   first(na.omit(cell_class)) else NA_character_,
    major_type = if ("major_type" %in% names(meta_cells))
                   first(na.omit(major_type)) else NA_character_,
    n_meta = n(),
    .groups = "drop"
  )

cat("meta_grp rows:", nrow(meta_grp), "\n")
cat("cell_class levels:", paste(sort(unique(meta_grp$cell_class)), collapse = ", "), "\n")
```

```{r load_amet}
## amet emits one TSV pair per (annotation, region, sub_type) stratum.
## Enumerate, parse the wildcard values from filenames, concatenate.
features_dir <- params$features_dir
cf_files <- list.files(features_dir,
                       pattern = "\\.cell_feature\\.tsv\\.gz$",
                       full.names = TRUE)
fe_files <- list.files(features_dir,
                       pattern = "\\.feature\\.tsv\\.gz$",
                       full.names = TRUE)
fe_files <- fe_files[!grepl("\\.cell_feature\\.tsv\\.gz$", fe_files)]

## Aggregate per cell at read time so feat_cf collapses from O(cells *
## features * strata * annotations) rows to O(cells * strata * annotations).
## Every downstream plot/embedding averages over features per cell.
## Same guard as in argelaguet.Rmd: amet may produce a near-empty
## .feature.tsv.gz when a combo has no groups passing min_cells_per_group.
read_combo_cf <- function(fp) {
  fn <- basename(fp)
  base <- sub("\\.cell_feature\\.tsv\\.gz$", "", fn)
  bx <- paste0(base, ".x")
  if (file.info(fp)$size < 30) {
    message("[load_amet] skipping near-empty cell_feature file: ", fn)
    return(NULL)
  }
  dt <- tryCatch(fread(fp, select = c("cell_id", "mean_meth", "i_total")),
                 error = function(e) {
                   message("[load_amet] failed to read ", fn, ": ", conditionMessage(e))
                   NULL
                 })
  if (is.null(dt) || nrow(dt) == 0L) return(NULL)
  agg <- dt[, .(mean_meth = mean(mean_meth, na.rm = TRUE),
                i_total = mean(i_total, na.rm = TRUE)),
            by = cell_id]
  agg[, annotation := get_annotation(bx)]
  agg[, region := get_region(bx)]
  agg[, sub_type := get_sub_type(bx)]
  agg
}

read_combo_fe <- function(fp) {
  fn <- basename(fp)
  base <- sub("\\.feature\\.tsv\\.gz$", "", fn)
  bx <- paste0(base, ".x")
  if (file.info(fp)$size < 30) {
    message("[load_amet] skipping near-empty feature file: ", fn)
    return(NULL)
  }
  dt <- tryCatch(fread(fp),
                 error = function(e) {
                   message("[load_amet] failed to read ", fn, ": ", conditionMessage(e))
                   NULL
                 })
  if (is.null(dt) || nrow(dt) == 0L) return(NULL)
  dt[, annotation := get_annotation(bx)]
  dt[, region := get_region(bx)]
  dt[, sub_type := get_sub_type(bx)]
  dt
}

feat_cf <- rbindlist(lapply(cf_files, read_combo_cf), fill = TRUE)
feat_fe <- rbindlist(lapply(fe_files, read_combo_fe), fill = TRUE)

cat("loaded", length(cf_files), "cell_feature TSVs and",
    length(fe_files), "feature TSVs from", features_dir, "\n")

## Attach cell_class / major_type per stratum (one row per
## (region, sub_type), not per cell). Update by reference.
meta_grp_dt <- as.data.table(meta_grp)[, .(region, sub_type,
                                            cell_class, major_type)]
feat_cf[meta_grp_dt, on = .(region, sub_type),
        `:=`(cell_class = i.cell_class, major_type = i.major_type)]
```


# Within-cell i_total {.tabset .tabset-pills}

Per-cell within-cell mutual-information score from amet's per-cell-per-feature
output (`i_total`).

```{r import_i_total}
i_total_long <- as.data.frame(feat_cf)
i_total_long$avg_meth <- i_total_long$mean_meth
i_total_long <- harmonize_levels(i_total_long)
head(i_total_long)
```

## i_total vs methylation

```{r scatter_i_total, fig.width = 10, fig.height = 15}
ggplot(i_total_long %>% filter(!is.na(cell_class)),
       aes(x = avg_meth, y = i_total, color = major_type)) +
  geom_point(size = 0.4, alpha = 0.15) +
  scale_color_manual(values = ecker_major_type_pal, na.value = "grey70") +
  facet_grid(annotation ~ cell_class) +
  labs(x = "average methylation", y = expression(i[total]), color = "cell type") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

## Distributions

```{r violin_i_total, fig.width = 10, fig.height = 15}
ggplot(i_total_long %>% filter(!is.na(cell_class)),
       aes(x = major_type, y = i_total, fill = major_type)) +
  geom_violin(alpha = 0.5, draw_quantiles = 0.5) +
  scale_fill_manual(values = ecker_major_type_pal) +
  facet_grid(annotation ~ cell_class, scales = "free_y") +
  labs(y = expression(i[total]), fill = "cell type") +
  theme_ng() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

```{r ridges_i_total, fig.width = 10, fig.height = 15}
ggplot(i_total_long %>% filter(!is.na(cell_class)),
       aes(x = i_total, y = major_type, fill = major_type)) +
  geom_density_ridges(alpha = 0.6, scale = 0.85) +
  scale_fill_manual(values = ecker_major_type_pal) +
  facet_grid(annotation ~ cell_class, scales = "free") +
  labs(x = expression(i[total])) +
  theme_ng() +
  theme(legend.position = "none")
```


# Across-cell jsd medians {.tabset .tabset-pills}

Per-feature jsd from amet's feature TSV (`jsd`, the multi-distribution
Jensen-Shannon divergence across cells in a group).

```{r import_jsd}
## Aggregate amet's per-feature jsd to one median per (annotation,
## region, sub_type) so the downstream group_medians join doesn't explode.
jsd_medians <- as.data.frame(
  as.data.table(feat_fe)[
    , .(median_jsd = median(jsd, na.rm = TRUE),
        median_avg_meth = median(mean_meth_mean, na.rm = TRUE)),
    by = .(annotation, region, sub_type)
  ]
)
jsd_medians <- harmonize_levels(jsd_medians)
jsd_medians <- jsd_medians %>%
  left_join(meta_grp %>% select(region, sub_type, cell_class, major_type),
            by = c("region", "sub_type"))
head(jsd_medians)
```

## Median jsd vs methylation

```{r scatter_jsd, fig.width = 10, fig.height = 15}
ggplot(jsd_medians %>% filter(!is.na(cell_class)),
       aes(x = median_avg_meth, y = median_jsd, color = major_type)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_manual(values = ecker_major_type_pal, na.value = "grey70") +
  facet_grid(annotation ~ cell_class) +
  labs(x = "median avg. methylation", y = "median jsd", color = "cell type") +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  theme_ng()
```

# Assembly

i_total is per-cell-per-feature. jsd is per-feature per-group.

```{r str_check}
str(i_total_long)
str(jsd_medians)
```

```{r save_rds}
saveRDS(
  list(i_total_long = i_total_long,
       jsd_medians = jsd_medians),
  file = "ecker_entropy.rds"
)
```

Aggregate per-cell i_total to per-group medians so it is comparable to jsd.

```{r group_medians}
i_total_grp <- i_total_long %>%
  group_by(annotation, region, sub_type) %>%
  summarise(median_i_total = median(i_total, na.rm = TRUE),
            median_meth = median(avg_meth, na.rm = TRUE),
            .groups = "drop")

all_grp <- i_total_grp %>%
  left_join(jsd_medians %>% select(annotation, region, sub_type, median_jsd),
            by = c("annotation", "region", "sub_type"))

head(all_grp)
```

```{r load_metadata}
## Attach CellClass / MajorType to the group-level table.
all_grp_meta <- all_grp %>%
  mutate(region = as.character(region),
         sub_type = as.character(sub_type)) %>%
  left_join(meta_grp, by = c("region", "sub_type"))

saveRDS(
  list(all_grp = all_grp,
       meta_grp = meta_grp,
       all_grp_meta = all_grp_meta,
       meta_cells = meta_cells),
  file = "ecker_groups_meta.rds"
)

cat("all_grp_meta rows:", nrow(all_grp_meta), "\n")
cat("cell_class non-NA:", sum(!is.na(all_grp_meta$cell_class)), "\n")
cat("median_jsd non-NA:", sum(!is.na(all_grp_meta$median_jsd)), "\n")
```

## i_total vs jsd

```{r i_total_vs_jsd, fig.width = 13, fig.height = 4}
df_cmp <- all_grp_meta %>% filter(!is.na(cell_class))
if (nrow(df_cmp) > 0) {
  ggplot(df_cmp, aes(x = median_jsd, y = median_i_total, color = annotation)) +
    geom_point(size = 1.5, alpha = 0.8) +
    scale_color_manual(values = ecker_annotation_pal) +
    facet_grid(. ~ cell_class) +
    labs(x = "median jsd", y = expression("median " * i[total]),
         color = "annotation") +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme_ng()
} else {
  message("No data with cell_class assigned - check metadata join")
}
```


# Variation across cell types {.tabset .tabset-pills}

SD and range of group medians per annotation, broken down by cell class.


```{r variation_helpers}
safe_range <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  diff(range(x))
}
```

```{r variation_class, fig.width = 13, fig.height = 4}
var_class <- all_grp_meta %>%
  filter(!is.na(cell_class)) %>%
  group_by(annotation, cell_class) %>%
  summarise(
    i_total_sd = sd(median_i_total, na.rm = TRUE),
    jsd_sd = sd(median_jsd, na.rm = TRUE),
    i_total_range = safe_range(median_i_total),
    jsd_range = safe_range(median_jsd),
    .groups = "drop"
  )

if (nrow(var_class) > 0 && any(!is.na(var_class$jsd_sd))) {
  ggplot(var_class, aes(x = jsd_sd, y = i_total_sd,
                         color = annotation, label = annotation)) +
    geom_point(size = 3) +
    geom_text_repel(size = 3) +
    scale_color_manual(values = ecker_annotation_pal) +
    facet_grid(. ~ cell_class) +
    labs(x = "SD of jsd across cell sub-types",
         y = expression("SD of " * i[total] * " across cell sub-types")) +
    theme_ng()
} else {
  message("No variation data available")
}
```

# Driver categorization

For each annotation we compute the SD of group-level median jsd (across-cell
heterogeneity) and median i_total (within-cell heterogeneity) across all
(region, sub_type) groups. If one SD is at least 1.5x the other, that
component dominates. Annotations where both SDs fall below the 30th percentile
are labelled "neither".

```{r drivers, fig.width = 5, fig.height = 5}
driver_df <- categorize_drivers(all_grp_meta, "cell_class")

cat("Driver table:\n")
print(table(driver_df$driver))

plot_driver_scatter(driver_df,
                    x_label = "SD of jsd across cell class groups",
                    y_label = expression("SD of " * i[total] * " across cell class groups"))
```

# Heatmaps of median entropy per group {.tabset .tabset-pills}

Rows = sub-type (clustered). Columns = annotation. Row annotation bars show cell class and major type.

```{r heatmap_helpers}
## ComplexHeatmap clusters rows within each row_split slice with hclust, which
## errors on NA/NaN/Inf distances. jsd matrices have sub-types with no value
## for some annotations. Impute non-finite cells to the slice mean (0 if the
## slice has no finite value) for clustering only; the displayed matrix keeps
## its NAs.
cluster_rows_na_safe <- function(m) {
  finite_vals <- m[is.finite(m)]
  fill <- if (length(finite_vals) > 0) mean(finite_vals) else 0
  m[!is.finite(m)] <- fill
  hclust(dist(m))
}

make_heatmap <- function(df, value_col, title,
                          palette = c("navy", "white", "firebrick")) {
  ## wide matrix: rows = sub_type, cols = annotation
  wide <- df %>%
    select(sub_type, annotation, cell_class, major_type,
           value = all_of(value_col)) %>%
    filter(!is.na(cell_class)) %>%
    group_by(sub_type, annotation, cell_class, major_type) %>%
    summarise(value = median(value, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = annotation, values_from = value)

  if (nrow(wide) < 2) { message("Too few rows for heatmap"); return(invisible(NULL)) }

  mat <- wide %>% select(-sub_type, -cell_class, -major_type) %>% as.matrix()
  rownames(mat) <- wide$sub_type

  ## row annotations
  row_ann <- rowAnnotation(
    cell_class = wide$cell_class,
    major_type = wide$major_type,
    col = list(cell_class = ecker_cell_class_pal, major_type = ecker_major_type_pal),
    annotation_legend_param = list(
      cell_class = list(title = "cell class"),
      major_type = list(title = "major type")
    )
  )

  col_fun <- colorRamp2(
    c(min(mat, na.rm = TRUE),
      mean(range(mat, na.rm = TRUE)),
      max(mat, na.rm = TRUE)),
    palette
  )

  Heatmap(mat,
          name = title,
          col = col_fun,
          row_split = wide$cell_class,
          cluster_rows = cluster_rows_na_safe,
          cluster_columns = FALSE,
          show_row_names = TRUE,
          show_column_names = TRUE,
          row_names_gp = gpar(fontsize = 7),
          right_annotation = row_ann,
          column_title = title,
          na_col = "grey90")
}
```

## i_total

```{r heatmap_i_total, fig.width = 10, fig.height = 8}
heat_df_i_total <- i_total_long %>%
  group_by(annotation, region, sub_type, cell_class, major_type) %>%
  summarise(median_i_total = median(i_total, na.rm = TRUE), .groups = "drop")

make_heatmap(heat_df_i_total, "median_i_total", "median i_total",
             palette = c("navy", "white", "firebrick"))
```

## jsd

```{r heatmap_jsd, fig.width = 10, fig.height = 8}
heat_df_jsd <- jsd_medians %>%
  group_by(annotation, region, sub_type, cell_class, major_type) %>%
  summarise(median_jsd = median(median_jsd, na.rm = TRUE), .groups = "drop")

make_heatmap(heat_df_jsd, "median_jsd", "median jsd",
             palette = c("navy", "white", "firebrick"))
```

<!-- Dropped: per-locus entropy distributions per group. Per-feature jsd (above) is the analog. -->

# UMAP: cell-type separability {.tabset .tabset-pills}

```{r umap_helpers}
plot_umap_class <- function(df, title, subtitle = NULL) {
  if (is.null(df) || nrow(df) == 0) {
    message("No UMAP result for: ", title)
    return(invisible(NULL))
  }
  if (all(is.na(df$cell_class))) {
    message("All cell_class values NA in UMAP for: ", title,
            " - check metadata join")
    return(invisible(NULL))
  }
  ggplot(df %>% filter(!is.na(cell_class)),
         aes(x = UMAP1, y = UMAP2, color = cell_class)) +
    geom_point(size = 2, alpha = 0.8) +
    scale_color_manual(values = ecker_cell_class_pal) +
    labs(title = title, subtitle = subtitle, color = "cell class") +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme_ng()
}
```

## Cell-level: i_total across annotations

```{r cell_umap_load}
## Build a wide cells x annotation matrix from amet's per-cell-per-feature
## i_total.
cells_i_total_long <- as.data.frame(feat_cf)[, c("cell_id", "region",
                                                  "sub_type", "annotation",
                                                  "i_total")] %>%
  rename(value = i_total) %>%
  group_by(cell_id, region, sub_type, annotation) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop")
cells_i_total_long$annotation <- factor(cells_i_total_long$annotation,
                                         levels = names(ann_labels),
                                         labels = unname(ann_labels))

cells_meth_long <- as.data.frame(feat_cf)[, c("cell_id", "region",
                                               "sub_type", "annotation",
                                               "mean_meth")] %>%
  rename(value = mean_meth) %>%
  group_by(cell_id, region, sub_type, annotation) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop")
cells_meth_long$annotation <- factor(cells_meth_long$annotation,
                                      levels = names(ann_labels),
                                      labels = unname(ann_labels))

cells_i_total_wide <- pivot_wider(cells_i_total_long,
                                   id_cols = c(cell_id, region, sub_type),
                                   names_from = annotation,
                                   values_from = value) %>%
  left_join(meta_grp %>% select(region, sub_type, cell_class, major_type),
            by = c("region", "sub_type"))

cells_meth_wide <- pivot_wider(cells_meth_long,
                                id_cols = c(cell_id, region, sub_type),
                                names_from = annotation,
                                values_from = value) %>%
  left_join(meta_grp %>% select(region, sub_type, cell_class, major_type),
            by = c("region", "sub_type"))

saveRDS(list(cells_i_total_wide = cells_i_total_wide,
             cells_meth_wide = cells_meth_wide),
        "ecker_cell_matrices.rds")

cat("Cells (i_total):", nrow(cells_i_total_wide), "\n")
cat("Cell class breakdown:\n")
print(table(cells_i_total_wide$cell_class, useNA = "always"))
```

```{r cell_umap_i_total, fig.width = 5, fig.height = 5}
meta_cols_cell <- c("cell_id", "region", "sub_type", "cell_class", "major_type")

umap_cell_i_total <- run_umap_wide(cells_i_total_wide, meta_cols_cell, n_neighbors = 15L)
saveRDS(umap_cell_i_total, "ecker_umap_cell_i_total.rds")

plot_umap_class(umap_cell_i_total,
                title = "Cell-level UMAP - i_total features",
                subtitle = "per-cell i_total per annotation as feature vector")
```

## Cell-level: avg_meth across annotations

```{r cell_umap_meth, fig.width = 5, fig.height = 5}
umap_cell_meth <- run_umap_wide(cells_meth_wide, meta_cols_cell, n_neighbors = 15L)
saveRDS(umap_cell_meth, "ecker_umap_cell_meth.rds")

plot_umap_class(umap_cell_meth,
                title = "Cell-level UMAP - avg. methylation features",
                subtitle = "per-cell avg. methylation per annotation as feature vector")
```

## Group-level: jsd across annotations

```{r group_umap_jsd, fig.width = 5, fig.height = 5}
jsd_wide_grp <- all_grp_meta %>%
  filter(!is.na(cell_class)) %>%
  select(region, sub_type, cell_class, major_type, annotation, median_jsd) %>%
  pivot_wider(names_from = annotation, values_from = median_jsd)

umap_grp_jsd <- run_umap_wide(jsd_wide_grp,
                              meta_cols = c("region", "sub_type",
                                            "cell_class", "major_type"),
                              n_neighbors = 5L)
saveRDS(umap_grp_jsd, "ecker_umap_grp_jsd.rds")

plot_umap_class(umap_grp_jsd,
                title = "Group-level UMAP - jsd features",
                subtitle = "median jsd per annotation as feature vector")
```

## Group-level: avg_meth across annotations

```{r group_umap_meth, fig.width = 5, fig.height = 5}
meth_wide_grp <- all_grp_meta %>%
  filter(!is.na(cell_class)) %>%
  select(region, sub_type, cell_class, major_type, annotation, median_meth) %>%
  pivot_wider(names_from = annotation, values_from = median_meth)

umap_grp_meth <- run_umap_wide(meth_wide_grp,
                                meta_cols = c("region", "sub_type",
                                              "cell_class", "major_type"),
                                n_neighbors = 5L)
plot_umap_class(umap_grp_meth,
                title = "Group-level UMAP - avg. methylation features",
                subtitle = "median avg. methylation per annotation as feature vector")
```

<!-- Dropped: per-locus jsd UMAP. -->

<!-- Paper UMAP overlay omitted: manifest lacks paper L1UMAP coords. -->

# Session

```{r session, cache = FALSE}
sessionInfo()
```
