---
title: "Ecker: single-cell genomic windows summaries"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
params:
  features_dir: ""
  win_cell_feature: ""
  win_feature: ""
  win_bed: ""
  windows_annotation: ""
  manifest: ""
  out_dir: ""
  log_path: ""
  threads: 1
  i_max_lag: 3
---

```{r logging_early, include = FALSE}
source(Sys.getenv("AMET_RENDER_HELPERS"))
param <- amet_setup_render_logging(params$log_path, params$threads)
```

```{r load_window_annotation}
## Per-window annotation overlap (fraction). Rows are aligned to windows
## by feature_id; the matrix is binarised below as <annotation>_bin.
## Sparse-proto guard: missing or empty file leaves ann_mat = NULL and
## downstream annotation chunks skip cleanly.
suppressPackageStartupMessages({
  library(data.table)
})
data.table::setDTthreads(params$threads)

ann_mat <- NULL
ann_bin <- NULL
ann_cols <- character(0)
ann_path <- params$windows_annotation
if (is.character(ann_path) && nzchar(ann_path) && file.exists(ann_path)) {
  ann_tbl <- tryCatch(
    fread(ann_path, sep = "\t", header = TRUE, nThread = params$threads),
    error = function(e) {
      message("[ecker_windows] failed to read windows_annotation: ", conditionMessage(e))
      NULL
    })
  if (!is.null(ann_tbl) && nrow(ann_tbl) > 0L &&
      "feature_id" %in% colnames(ann_tbl)) {
    keep <- setdiff(colnames(ann_tbl), c("chrom", "start", "end", "feature_id"))
    if (length(keep) > 0L) {
      ann_mat <- as.data.frame(ann_tbl[, c("feature_id", keep), with = FALSE])
      rownames(ann_mat) <- ann_mat$feature_id
      ann_cols <- keep
      ann_bin <- as.data.frame(lapply(ann_mat[, ann_cols, drop = FALSE],
                                      function(x) as.integer(x > 0)))
      colnames(ann_bin) <- paste0(ann_cols, "_bin")
      rownames(ann_bin) <- ann_mat$feature_id
    }
  }
}
cat("Annotation columns:", length(ann_cols),
    "| windows annotated:", if (is.null(ann_mat)) 0L else nrow(ann_mat), "\n")
```

```{r setup}
suppressPackageStartupMessages({
  library(data.table)
  library(readr)
  library(dplyr)
  library(ggplot2)
  library(BiocParallel)
})

repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))
source(file.path(repo_root, "workflow", "scripts", "palettes.R"))


knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  cache = FALSE,
  fig.path = "ecker_windows_plots/"
)
```

## Cell selection upstream of this Rmd:
##   Per (region, sub_type) stratum, cells are picked by source-TAR size
##   (proxy for observed-CpG coverage; tar contains one allc tsv.gz per cell)
##   with plate-stratified round-robin so a single sequencing plate cannot
##   dominate the pick. Cap = max_cells_per_combo (config/datasets.yaml:
##   prototype.cells_per_group in proto, full.max_cells_per_combo otherwise).

```{r load_meta}
man <- fread(params$manifest)
cat("Cells in manifest:", nrow(man), "\n")
cat("Cell classes:", length(unique(man$cell_class)), "\n")
cat("Major types:", length(unique(man$major_type)), "\n")
```

```{r load_amet_windows}
## amet's window-scope output: two long tables. per-cell-per-window i_total
## + mean_meth, and per-window-per-group jsd (loaded but not used in this
## Rmd, kept for symmetry with the per-feature reports).
win_cf <- fread(params$win_cell_feature)
keep_cols <- intersect(c("cell_id", "cell_class", "major_type", "region", "sub_type"),
                       colnames(man))
win_cf <- merge(win_cf, man[, ..keep_cols],
                by = "cell_id", all.x = TRUE)

cat("Cell-window rows:", nrow(win_cf), "\n")
cat("Unique windows:", uniqueN(win_cf$feature_id), "\n")
cat("Unique cells:", uniqueN(win_cf$cell_id), "\n")
cat("Cell classes in data:\n")
print(table(win_cf$cell_class, useNA = "ifany"))
cat("Major types in data:\n")
print(table(win_cf$major_type, useNA = "ifany"))
```

```{r filter_na}
## NA filter: drop windows missing in > 30% of cells. Applied here on the
## long table.
n_cells <- uniqueN(win_cf$cell_id)
win_cov <- win_cf[, .(cov_frac = .N / n_cells), by = feature_id]
keep_win <- win_cov[cov_frac >= 0.7, feature_id]
win_cf <- win_cf[feature_id %in% keep_win]
cat("Windows after NA filter (>= 70% cells covered):", uniqueN(win_cf$feature_id), "\n")
```

<!-- amet i_total is per cell, from cell_feature.tsv.gz. No meth correction here. -->

# QC

```{r cell_qc_df}
cell_df <- win_cf[, .(
  mean_i_total = mean(i_total,  na.rm = TRUE),
  mean_meth = mean(mean_meth, na.rm = TRUE)
), by = .(cell_id, cell_class, major_type)]

write.csv(cell_df, "ecker_windows_per_cell_summary.csv", row.names = FALSE)
```


```{r itotal_by_class, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(4, 1)$h}
ggplot(cell_df, aes(x = cell_class, y = mean_i_total, fill = cell_class)) +
  geom_violin(trim = FALSE, scale = "width") +
  geom_boxplot(width = 0.1, outlier.size = 0.5, alpha = 0.5) +
  scale_fill_manual(values = ecker_cell_class_pal) +
  theme_ng(base_size = 8) +
  theme(legend.position = "none") +
  labs(y = expression("mean " * i[total]),
       title = expression("Per-cell mean " * i[total] * " by cell class"))
```

```{r meth_vs_itotal, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(4, 1)$h}
ggplot(cell_df[is.finite(mean_meth) & is.finite(mean_i_total)],
       aes(x = mean_meth, y = mean_i_total, color = cell_class)) +
  geom_point(size = 0.4, alpha = 0.5) +
  scale_color_manual(values = ecker_cell_class_pal) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1), ncol = 2)) +
  theme_ng(base_size = 8) +
  labs(x = "mean methylation", y = expression("mean " * i[total]),
       title = expression("Global " * i[total] * " vs methylation per cell"))
```

# Per-annotation breakdowns

Per cell, average i_total across windows that do (1) or do not (0) overlap
each annotation. Stratifies the within-cell score by genomic context (PMD-like
H3K9me3, H3K27me3, gene bodies, etc.).

```{r per_cell_by_annotation, fig.width = ng_fig_size(4, 2)$w, fig.height = ng_fig_size(4, 2)$h}
if (is.null(ann_bin) || ncol(ann_bin) == 0L) {
  cat("Annotation matrix empty; skipping per-annotation breakdowns.\n")
} else {
  common_wins <- intersect(unique(win_cf$feature_id), rownames(ann_bin))
  if (length(common_wins) == 0L) {
    cat("No window overlap between i_total table and annotation matrix; skipping.\n")
  } else {
    win_sub <- win_cf[feature_id %in% common_wins]
    ann_bin_sub <- ann_bin[common_wins, , drop = FALSE]

    df_list <- lapply(colnames(ann_bin_sub), function(bf) {
      bin <- ann_bin_sub[[bf]]
      names(bin) <- rownames(ann_bin_sub)
      tmp <- win_sub[, .(
        mean_i_total = mean(i_total, na.rm = TRUE),
        mean_meth = mean(mean_meth, na.rm = TRUE)
      ), by = .(cell_id, feature_id)]
      tmp[, bin := bin[feature_id]]
      out <- tmp[!is.na(bin), .(
        mean_i_total = mean(mean_i_total, na.rm = TRUE),
        mean_meth = mean(mean_meth, na.rm = TRUE)
      ), by = .(cell_id, bin)]
      out[, annotation := sub("_bin$", "", bf)]
      out
    })

    df_ann <- rbindlist(df_list)
    df_ann <- merge(df_ann, man[, ..keep_cols], by = "cell_id", all.x = TRUE)
    df_ann$bin <- factor(df_ann$bin, levels = c(0, 1))

    write.csv(df_ann, "ecker_windows_per_cell_by_annotation.csv", row.names = FALSE)

    print(
      ggplot(df_ann[is.finite(mean_i_total)],
             aes(x = cell_class, y = mean_i_total, fill = bin)) +
        geom_boxplot(outlier.size = 0.2, position = position_dodge(width = 0.8)) +
        scale_fill_manual(values = c("0" = "#bbbbbb", "1" = "#d95f02"),
                          name = "overlaps annotation") +
        facet_wrap(~ annotation, ncol = 3) +
        theme_ng(base_size = 7) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
              legend.position = "bottom") +
        labs(x = NULL, y = expression("mean " * i[total]),
             title = expression("Per-cell mean " * i[total] *
                                " by annotation overlap (binarised)"))
    )
  }
}
```

```{r jsd_by_annotation, fig.width = ng_fig_size(4, 2)$w, fig.height = ng_fig_size(4, 2)$h}
## Across-cell JSD aggregated by binarised annotation overlap, per group
## (cell_class).
if (is.null(ann_bin) || ncol(ann_bin) == 0L) {
  cat("Annotation matrix empty; skipping per-annotation JSD aggregation.\n")
} else if (!file.exists(params$win_feature)) {
  cat("win_feature missing; skipping JSD aggregation.\n")
} else {
  win_feat <- fread(params$win_feature)
  if (!"jsd" %in% colnames(win_feat) || nrow(win_feat) == 0L) {
    cat("No jsd column in win_feature; skipping.\n")
  } else {
    keep_feat <- intersect(win_feat$feature_id, rownames(ann_bin))
    if (length(keep_feat) == 0L) {
      cat("No window overlap between jsd table and annotation matrix; skipping.\n")
    } else {
      jf <- win_feat[feature_id %in% keep_feat]
      ann_bin_sub <- ann_bin[keep_feat, , drop = FALSE]

      jsd_df <- rbindlist(lapply(colnames(ann_bin_sub), function(bf) {
        bin <- ann_bin_sub[[bf]]
        names(bin) <- rownames(ann_bin_sub)
        tmp <- copy(jf)
        tmp[, bin := bin[feature_id]]
        tmp[, annotation := sub("_bin$", "", bf)]
        tmp[!is.na(bin), .(jsd, bin, annotation,
                          group = if ("group" %in% colnames(tmp)) group else "all")]
      }))
      jsd_df$bin <- factor(jsd_df$bin, levels = c(0, 1))

      write.csv(jsd_df, "ecker_windows_jsd_by_annotation.csv", row.names = FALSE)

      print(
        ggplot(jsd_df, aes(x = annotation, y = jsd, fill = bin)) +
          geom_boxplot(outlier.size = 0.2, position = position_dodge(width = 0.8)) +
          scale_fill_manual(values = c("0" = "#bbbbbb", "1" = "#1b9e77"),
                            name = "overlaps annotation") +
          theme_ng(base_size = 7) +
          theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
                legend.position = "bottom") +
          labs(x = NULL, y = "jsd",
               title = "Per-window JSD by annotation overlap (binarised)")
      )
    }
  }
}
```

```{r session}
sessionInfo()
```
