---
title: "Ecker: cell embeddings from genomic windows entropy"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
params:
  features_dir: ""
  win_cell_feature: ""
  win_h5: ""
  win_feature: ""
  win_bed: ""
  windows_annotation: ""
  manifest: ""
  out_dir: ""
  log_path: ""
  threads: 1
  i_max_lag: 3
---

```{r logging_early, include = FALSE}
source(Sys.getenv("AMET_RENDER_HELPERS"))
param <- amet_setup_render_logging(params$log_path, params$threads)
```

```{r load_window_annotation}
suppressPackageStartupMessages({
  library(data.table)
})
data.table::setDTthreads(params$threads)

ann_mat <- NULL
ann_bin <- NULL
ann_cols <- character(0)
ann_path <- params$windows_annotation
if (is.character(ann_path) && nzchar(ann_path) && file.exists(ann_path)) {
  ann_tbl <- tryCatch(
    fread(ann_path, sep = "\t", header = TRUE, nThread = params$threads),
    error = function(e) {
      message("[ecker_embeddings] failed to read windows_annotation: ", conditionMessage(e))
      NULL
    })
  if (!is.null(ann_tbl) && nrow(ann_tbl) > 0L &&
      "feature_id" %in% colnames(ann_tbl)) {
    keep <- setdiff(colnames(ann_tbl), c("chrom", "start", "end", "feature_id"))
    if (length(keep) > 0L) {
      ann_mat <- as.data.frame(ann_tbl[, c("feature_id", keep), with = FALSE])
      rownames(ann_mat) <- ann_mat$feature_id
      ann_cols <- keep
      ann_bin <- as.data.frame(lapply(ann_mat[, ann_cols, drop = FALSE],
                                      function(x) as.integer(x > 0)))
      colnames(ann_bin) <- paste0(ann_cols, "_bin")
      rownames(ann_bin) <- ann_mat$feature_id
    }
  }
}
cat("Annotation columns:", length(ann_cols),
    "| windows annotated:", if (is.null(ann_mat)) 0L else nrow(ann_mat), "\n")
```

```{r setup}
suppressPackageStartupMessages({
  library(data.table)
  library(readr)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(BiocParallel)
  library(HDF5Array)
  library(uwot)
  library(irlba)
  library(matrixStats)

  library(cluster)
  library(viridis)
  library(viridisLite)
  library(cowplot)
  library(ggrastr)
})

repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))
source(file.path(repo_root, "workflow", "scripts", "palettes.R"))
source(file.path(repo_root, "workflow", "scripts", "embedding_utils.R"))
source(file.path(repo_root, "workflow", "scripts", "windows_h5.R"))


knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  cache = FALSE,
  fig.path = "ecker_embeddings_plots/"
)
```

```{r constants}
N_HVW <- 1000
N_PCS <- 10

## amet exposes i_total per cell per window and mean_meth per cell per window;
## both are carried forward as separate assays for the embeddings.
assay_map <- c(
  "i_total" = "i_total",
  "meth"    = "methylation"
)
```

## Cell selection and NA-aware embedding

The cells embedded here are every cell in the dataset manifest (cells.tsv). The windows run applies no per-stratum cap, unlike the per-feature runs, which keep at most max_cells_per_combo cells per (region, sub_type). Using all cells keeps the embedding at atlas scale instead of collapsing each cell type to a handful of representatives, at the cost of preserving any residual plate effect (cell cycle, batch) rather than averaging it out.

scBS-seq methylation and per-cell entropy values have a meaningful zero
(unmethylated CpG, zero local entropy) which is distinct from missing coverage.
Imputing NAs to zero would bias both the variance signal and the embedding
geometry, so the pipeline keeps NAs as NAs throughout: HVF selection scores
features by variance times coverage fraction, and when too few windows have
complete coverage the PCA step uses NIPALS (`pcaMethods`) which iterates over
the observed entries without filling in missing values. Cells are dropped only
when their per-cell NA fraction is above the QC threshold; once they pass that
filter they are never dropped to make a feature block dense.

# Load data

```{r load_amet}
## amet's window-scope output is a per-(cell, window) long table too large to
## read whole; an upstream rule pivots it into an HDF5 store. load_windows_h5
## returns realized windows x cells matrices, one per assay.
wd <- load_windows_h5(params$win_h5)
man <- fread(params$manifest)

assays_list <- list(
  i_total = wd$i_total,
  meth = wd$meth
)

col_data <- merge(data.frame(cell_id = wd$cell_id, stringsAsFactors = FALSE),
                  as.data.frame(man), by = "cell_id", all.x = TRUE, sort = FALSE)
rownames(col_data) <- col_data$cell_id
## Align col_data rows to the matrix column order.
col_data <- col_data[wd$cell_id, , drop = FALSE]
col_data$cell_class <- factor(col_data$cell_class)
col_data$major_type <- factor(col_data$major_type)
col_data$region <- factor(col_data$region)
col_data$sub_type <- factor(col_data$sub_type)

cc_lvls <- sort(unique(as.character(col_data$cell_class)))

cat("Cells:", nrow(col_data), "\n")
cat("Windows:", nrow(assays_list$i_total), "\n")
cat("Assays:", paste(names(assays_list), collapse = ", "), "\n")
cat("Cell classes:\n")
print(table(col_data$cell_class, useNA = "ifany"))
```

# Shared helpers

```{r helpers}
hvw_pca_umap <- function(mat) {
  ## run_embedding needs an in-memory matrix, so the HDF5-backed assay is
  ## realized here. TODO: check if memory is too greedy at this realization.
  run_embedding(as.matrix(mat), n_hvf = N_HVW, n_pcs = N_PCS,
                n_neighbors = 15L, min_dist = 0.3, seed = 42L)
}

plot_umap_cellclass <- function(coords, kept_cols, title_prefix) {
  df <- data.frame(
    umap1 = coords[, 1], umap2 = coords[, 2],
    cell_class = col_data$cell_class[kept_cols]
  )
  ggplot(df, aes(x = umap1, y = umap2, color = cell_class)) +
    geom_point_rast(size = 0.3, alpha = 0.5, raster.dpi = 300) +
    scale_color_manual(values = ecker_cell_class_pal) +
    labs(title = paste(title_prefix, "- cell class"),
         x = "UMAP 1", y = "UMAP 2") +
    guides(color = guide_legend(override.aes = list(size = 2, alpha = 1), ncol = 2)) +
    theme_ng()
}

plot_umap_majortype <- function(coords, kept_cols, title_prefix) {
  df <- data.frame(
    umap1 = coords[, 1], umap2 = coords[, 2],
    major_type = col_data$major_type[kept_cols]
  )
  ggplot(df, aes(x = umap1, y = umap2, color = major_type)) +
    geom_point_rast(size = 0.3, alpha = 0.5, raster.dpi = 300) +
    scale_color_manual(values = ecker_major_type_pal) +
    labs(title = paste(title_prefix, "- major type"),
         x = "UMAP 1", y = "UMAP 2") +
    guides(color = guide_legend(override.aes = list(size = 2, alpha = 1), ncol = 2)) +
    theme_ng()
}
```

# Per-window embeddings (HVW -> PCA -> UMAP)

HVW selection and NA handling: see `embedding_utils.R`.
Top `r N_HVW` most variable windows retained per assay before PCA and UMAP.

```{r win_hvw_pca_umap}
win_embeds <- lapply(names(assay_map), function(akey) {
  if (!akey %in% names(assays_list)) return(NULL)
  hvw_pca_umap(assays_list[[akey]])
})
names(win_embeds) <- assay_map

failed <- names(which(sapply(win_embeds, is.null)))
if (length(failed) > 0)
  warning("run_embedding returned NULL for: ", paste(failed, collapse = ", "))

cat("\nHVF selection summary:\n")
for (nm in names(win_embeds)) {
  em <- win_embeds[[nm]]
  if (is.null(em)) {
    cat(nm, ": NULL\n")
  } else {
    cat(nm, ": HVF =", length(em$hvf_idx),
        "| cells =", sum(em$kept_cols), "/", nrow(col_data), "\n")
  }
}

saveRDS(
  list(col_data = col_data,
       n_windows = nrow(assays_list$i_total),
       n_cells = nrow(col_data),
       assay_names = names(assays_list),
       win_embeds = win_embeds,
       assay_map = assay_map),
  "ecker_umap_windows_i_total.rds", compress = "xz"
)
```

```{r per_cell_summary}
per_cell_summary <- data.frame(
  cell_id = col_data$cell_id,
  cell_class = col_data$cell_class,
  major_type = col_data$major_type,
  sub_type = col_data$sub_type,
  mean_meth = colMeans(assays_list$meth,    na.rm = TRUE),
  mean_i_total = colMeans(assays_list$i_total, na.rm = TRUE),
  stringsAsFactors = FALSE
)
write.csv(per_cell_summary, "ecker_embeddings_per_cell_summary.csv", row.names = FALSE)
```

## By cell class {.tabset .tabset-pills}


```{r win_cc_i_total, fig.width = ng_fig_size(2, 1, panel_mm = 60)$w, fig.height = ng_fig_size(2, 1, panel_mm = 60)$h}
### i_total
if (!is.null(win_embeds[["i_total"]])) {
  p1 <- plot_umap_cellclass(win_embeds[["i_total"]]$umap,
                            win_embeds[["i_total"]]$kept_cols, "i_total")
  p2 <- plot_umap_majortype(win_embeds[["i_total"]]$umap,
                            win_embeds[["i_total"]]$kept_cols, "i_total")
  print(plot_grid(p1, p2, nrow = 1))
}
```

```{r win_cc_meth, fig.width = ng_fig_size(2, 1, panel_mm = 60)$w, fig.height = ng_fig_size(2, 1, panel_mm = 60)$h}
### methylation
if (!is.null(win_embeds[["methylation"]])) {
  p1 <- plot_umap_cellclass(win_embeds[["methylation"]]$umap,
                            win_embeds[["methylation"]]$kept_cols, "methylation")
  p2 <- plot_umap_majortype(win_embeds[["methylation"]]$umap,
                            win_embeds[["methylation"]]$kept_cols, "methylation")
  print(plot_grid(p1, p2, nrow = 1))
}
```

## Two-panel comparison (cell class)

```{r win_panel_cc, fig.width = ng_fig_size(2, 1, panel_mm = 55)$w, fig.height = ng_fig_size(2, 1, panel_mm = 55)$h}
ok_embeds <- Filter(Negate(is.null), win_embeds)
df_panel <- bind_rows(lapply(names(ok_embeds), function(lbl) {
  em <- ok_embeds[[lbl]]
  data.frame(umap1 = em$umap[, 1], umap2 = em$umap[, 2],
             cell_class = col_data$cell_class[em$kept_cols],
             assay = lbl)
}))
df_panel$assay <- factor(df_panel$assay, levels = assay_map)

ggplot(df_panel, aes(x = umap1, y = umap2, color = cell_class)) +
  geom_point_rast(size = 0.2, alpha = 0.5, raster.dpi = 300) +
  scale_color_manual(values = ecker_cell_class_pal) +
  facet_wrap(~ assay, nrow = 1) +
  labs(x = "UMAP 1", y = "UMAP 2",
       title = "Per-window embeddings (HVW -> PCA -> UMAP)") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1), ncol = 2)) +
  theme_ng()
```

## Variance explained

```{r win_varexp, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(2, 1)$h}
ve_win <- bind_rows(lapply(names(assay_map), function(akey) {
  lbl <- assay_map[[akey]]
  if (is.null(win_embeds[[lbl]]) || !akey %in% names(assays_list)) return(NULL)
  kept <- win_embeds[[lbl]]$kept_cols
  hvw_mat <- as.matrix(
    assays_list[[akey]][win_embeds[[lbl]]$hvf_idx, kept, drop = FALSE])
  data.frame(
    assay = lbl,
    cell_class = median(row_variance_explained(hvw_mat, col_data$cell_class[kept]), na.rm = TRUE),
    sub_type = median(row_variance_explained(hvw_mat, col_data$sub_type[kept]),   na.rm = TRUE))
}))

ve_long <- pivot_longer(ve_win, c("cell_class", "sub_type"),
                        names_to = "variable", values_to = "median_ve")
ve_long$assay <- factor(ve_long$assay, levels = assay_map)

write.csv(ve_long, "ecker_win_varexp.csv", row.names = FALSE)

ggplot(ve_long, aes(x = assay, y = median_ve, fill = variable)) +
  geom_col(position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = c("cell_class" = "#0072B2", "sub_type" = "#E69F00")) +
  labs(x = NULL, y = "median % variance explained",
       title = "Variance explained (per-window HVW)") +
  theme_ng()
```

## Embedding diagnostics

Why does per-window variance explained stay low? Candidates are high NA
fractions per window, few HVFs surviving QC, cells dropped by the sparse
fallback, and per-HVF variance being low in absolute terms. The diagnostic
below logs these quantities per assay to `ecker_embedding_diagnostics.csv`
and a compact text summary.

```{r embedding_diagnostic}
diag_rows <- list()

na_stats <- function(mat) {
  na_frac_row <- rowMeans(is.na(mat))
  na_frac_col <- colMeans(is.na(mat))
  list(
    row_na_median = median(na_frac_row, na.rm = TRUE),
    row_na_p90 = quantile(na_frac_row, 0.9, na.rm = TRUE),
    col_na_median = median(na_frac_col, na.rm = TRUE),
    col_na_p90 = quantile(na_frac_col, 0.9, na.rm = TRUE),
    n_complete_rows = sum(na_frac_row == 0, na.rm = TRUE)
  )
}

for (akey in names(assay_map)) {
  lbl <- assay_map[[akey]]
  if (!akey %in% names(assays_list)) next
  em <- win_embeds[[lbl]]
  if (is.null(em)) next

  mat <- assays_list[[akey]]
  ns <- na_stats(mat)

  hvf_mat <- as.matrix(mat[em$hvf_idx, em$kept_cols, drop = FALSE])
  hvf_row_var <- matrixStats::rowVars(hvf_mat, na.rm = TRUE)

  diag_rows[[lbl]] <- data.frame(
    assay = lbl,
    stage = em$stage,
    n_cells_total = nrow(col_data),
    n_cells_used = sum(em$kept_cols),
    cells_dropped_fraction = 1 - sum(em$kept_cols) / nrow(col_data),
    n_windows_total = nrow(mat),
    n_complete_rows = ns$n_complete_rows,
    row_na_median = ns$row_na_median,
    row_na_p90 = as.numeric(ns$row_na_p90),
    col_na_median = ns$col_na_median,
    col_na_p90 = as.numeric(ns$col_na_p90),
    n_hvf = length(em$hvf_idx),
    hvf_var_median = median(hvf_row_var, na.rm = TRUE),
    hvf_var_q25 = quantile(hvf_row_var, 0.25, na.rm = TRUE),
    hvf_var_q75 = quantile(hvf_row_var, 0.75, na.rm = TRUE),
    stringsAsFactors = FALSE
  )
}

diag_df <- bind_rows(diag_rows)
write.csv(diag_df, "ecker_embedding_diagnostics.csv", row.names = FALSE)

diag_lines <- c(
  "Ecker embedding diagnostic summary",
  paste0("Total cells: ", nrow(col_data)),
  paste0("Total windows: ", nrow(assays_list$i_total)),
  ""
)
for (lbl in diag_df$assay) {
  r <- diag_df[diag_df$assay == lbl, ]
  diag_lines <- c(diag_lines, sprintf(
    "[%s] stage=%s | cells used=%d/%d (%.1f%% dropped) | HVF=%d | complete windows=%d | row NA median=%.3f | HVF row variance median=%.4f",
    lbl, r$stage,
    r$n_cells_used, r$n_cells_total,
    100 * r$cells_dropped_fraction,
    r$n_hvf, r$n_complete_rows, r$row_na_median, r$hvf_var_median
  ))
}
writeLines(diag_lines, "ecker_embedding_diagnostics.txt")

cat(paste(diag_lines, collapse = "\n"), "\n")
```

## Silhouette scores

```{r win_sil, fig.width = ng_fig_size(2, 1)$w, fig.height = ng_fig_size(1, 1)$h}
sil_win <- data.frame(
  assay = names(ok_embeds),
  silhouette_cell_class = sapply(names(ok_embeds), function(lbl)
    sil_score(ok_embeds[[lbl]]$umap,
              col_data$cell_class[ok_embeds[[lbl]]$kept_cols])),
  silhouette_sub_region = sapply(names(ok_embeds), function(lbl)
    sil_score(ok_embeds[[lbl]]$umap,
              col_data$region[ok_embeds[[lbl]]$kept_cols]))
)

sil_long <- pivot_longer(sil_win, c("silhouette_cell_class", "silhouette_sub_region"),
                         names_to = "grouping", values_to = "silhouette")
sil_long$grouping <- sub("silhouette_", "", sil_long$grouping)
sil_long$assay <- factor(sil_long$assay, levels = assay_map)

ggplot(sil_long, aes(x = assay, y = silhouette, fill = grouping)) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c("cell_class" = "#0072B2", "region" = "#E69F00")) +
  labs(x = NULL, y = "mean silhouette",
       title = "Cluster separation by assay") +
  theme_ng()
```

# Per-cell entropy summaries

```{r cell_df}
cell_df <- data.frame(
  cell_class = col_data$cell_class,
  region = col_data$region,
  mean_i_total = colMeans(assays_list$i_total, na.rm = TRUE),
  mean_meth = colMeans(assays_list$meth,    na.rm = TRUE)
)
```

```{r violins, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(4, 2)$h}
cell_long <- pivot_longer(cell_df, c("mean_i_total", "mean_meth"),
                          names_to = "metric", values_to = "value")

ggplot(cell_long, aes(x = cell_class, y = value, fill = cell_class)) +
  geom_violin(trim = FALSE, scale = "width", alpha = 0.7) +
  geom_boxplot(width = 0.1, outlier.size = 0.3, alpha = 0.6) +
  scale_fill_manual(values = ecker_cell_class_pal) +
  facet_grid(metric ~ ., scales = "free_y") +
  labs(x = "cell class", y = "per-cell mean") +
  theme_ng() +
  theme(legend.position = "none")
```

# Per-locus embeddings (windows as observations)

Each window is embedded from its per-cell i_total vector. Points are
coloured by binarised annotation overlap. Cells with too few covered
windows or windows with too many missing cells are dropped in
`run_embedding`. Annotation-driven panels are skipped if `ann_bin`
is empty.

```{r locus_umap_i_total_data}
i_total_window_embed <- NULL
if (nrow(assays_list$i_total) > 0L) {
  ## Realize the HDF5-backed assay for run_embedding.
  ## TODO: check if memory is too greedy at this realization.
  i_total_window_embed <- tryCatch(
    run_embedding(as.matrix(t(assays_list$i_total)), n_hvf = N_HVW, n_pcs = N_PCS,
                  n_neighbors = 15L, min_dist = 0.3, seed = 42L),
    error = function(e) {
      message("[locus_umap_i_total] failed: ", conditionMessage(e)); NULL })
}
```

```{r locus_umap_i_total_plot, fig.width = ng_fig_size(3, 1, panel_mm = 55)$w, fig.height = ng_fig_size(3, 1, panel_mm = 55)$h}
if (is.null(i_total_window_embed)) {
  cat("Per-locus i_total UMAP unavailable.\n")
} else if (is.null(ann_bin) || ncol(ann_bin) == 0L) {
  cat("Annotation matrix empty; rendering per-locus i_total UMAP without annotation overlay.\n")
  df <- data.frame(UMAP1 = i_total_window_embed$umap[, 1],
                   UMAP2 = i_total_window_embed$umap[, 2])
  print(ggplot(df, aes(UMAP1, UMAP2)) +
          geom_point_rast(size = 0.15, alpha = 0.4, raster.dpi = 300) +
          labs(title = "per-locus i_total UMAP") + theme_ng())
} else {
  win_names <- rownames(assays_list$i_total)[i_total_window_embed$kept_cols]
  ann_aligned <- ann_bin[match(win_names, rownames(ann_bin)), , drop = FALSE]
  df_base <- data.frame(UMAP1 = i_total_window_embed$umap[, 1],
                        UMAP2 = i_total_window_embed$umap[, 2],
                        feature_id = win_names)
  panels <- lapply(colnames(ann_aligned), function(cn) {
    df <- df_base
    df$overlap <- factor(ann_aligned[[cn]], levels = c(0, 1))
    ggplot(df, aes(UMAP1, UMAP2, color = overlap)) +
      geom_point_rast(size = 0.12, alpha = 0.55, raster.dpi = 300) +
      scale_color_manual(values = c("0" = "#bbbbbb", "1" = "#d95f02"),
                         na.value = "#eeeeee", drop = FALSE) +
      labs(title = sub("_bin$", "", cn)) +
      theme_ng(base_size = 7) +
      theme(legend.position = "none",
            plot.title = element_text(size = 6.5))
  })
  print(plot_grid(plotlist = panels, ncol = 3))
}
```

```{r locus_umap_i_norm_data}
## Canonical i_norm: i_total / (k_max * H(mean_meth)) per (window, cell).
## H is binary Shannon entropy. k_max comes from amet's --i-max-lag via params.
## Realizes the HDF5-backed assays for the per-column i_norm loop and the
## embedding. TODO: check if memory is too greedy at this realization.
i_total_mat <- as.matrix(assays_list$i_total)
meth_mat <- as.matrix(assays_list$meth)
i_norm_mat <- matrix(NA_real_, nrow(i_total_mat), ncol(i_total_mat),
                     dimnames = dimnames(i_total_mat))
for (j in seq_len(ncol(i_total_mat))) {
  i_norm_mat[, j] <- compute_i_norm(i_total_mat[, j], meth_mat[, j],
                                    params$i_max_lag)
}

i_norm_window_embed <- NULL
if (nrow(i_norm_mat) > 0L && sum(is.finite(i_norm_mat)) > 0L) {
  i_norm_window_embed <- tryCatch(
    run_embedding(t(i_norm_mat), n_hvf = N_HVW, n_pcs = N_PCS,
                  n_neighbors = 15L, min_dist = 0.3, seed = 42L),
    error = function(e) {
      message("[locus_umap_i_norm] failed: ", conditionMessage(e)); NULL })
}
```

```{r locus_umap_i_norm_plot, fig.width = ng_fig_size(3, 1, panel_mm = 55)$w, fig.height = ng_fig_size(3, 1, panel_mm = 55)$h}
if (is.null(i_norm_window_embed)) {
  cat("Per-locus i_norm UMAP unavailable.\n")
} else if (is.null(ann_bin) || ncol(ann_bin) == 0L) {
  df <- data.frame(UMAP1 = i_norm_window_embed$umap[, 1],
                   UMAP2 = i_norm_window_embed$umap[, 2])
  print(ggplot(df, aes(UMAP1, UMAP2)) +
          geom_point_rast(size = 0.15, alpha = 0.4, raster.dpi = 300) +
          labs(title = "per-locus i_norm UMAP") + theme_ng())
} else {
  win_names <- rownames(i_norm_mat)[i_norm_window_embed$kept_cols]
  ann_aligned <- ann_bin[match(win_names, rownames(ann_bin)), , drop = FALSE]
  df_base <- data.frame(UMAP1 = i_norm_window_embed$umap[, 1],
                        UMAP2 = i_norm_window_embed$umap[, 2],
                        feature_id = win_names)
  panels <- lapply(colnames(ann_aligned), function(cn) {
    df <- df_base
    df$overlap <- factor(ann_aligned[[cn]], levels = c(0, 1))
    ggplot(df, aes(UMAP1, UMAP2, color = overlap)) +
      geom_point_rast(size = 0.12, alpha = 0.55, raster.dpi = 300) +
      scale_color_manual(values = c("0" = "#bbbbbb", "1" = "#0072B2"),
                         na.value = "#eeeeee", drop = FALSE) +
      labs(title = sub("_bin$", "", cn)) +
      theme_ng(base_size = 7) +
      theme(legend.position = "none",
            plot.title = element_text(size = 6.5))
  })
  print(plot_grid(plotlist = panels, ncol = 3))
}
```

```{r locus_umap_jsd_data}
## Per-locus jsd UMAP: each window has one jsd per group; embed windows in
## group-space. Requires the windows-feature TSV (one row per (window, group)).
jsd_window_embed <- NULL
jsd_window_names <- character(0)
if (file.exists(params$win_feature)) {
  jsd_tbl <- tryCatch(
    fread(params$win_feature, nThread = params$threads),
    error = function(e) { message("[locus_umap_jsd] read failed: ", conditionMessage(e)); NULL })
  if (!is.null(jsd_tbl) && "jsd" %in% colnames(jsd_tbl) &&
      "feature_id" %in% colnames(jsd_tbl) && "group" %in% colnames(jsd_tbl)) {
    jsd_wide <- tryCatch(
      dcast(jsd_tbl, feature_id ~ group, value.var = "jsd", fun.aggregate = mean),
      error = function(e) { message("[locus_umap_jsd] dcast failed: ", conditionMessage(e)); NULL })
    if (!is.null(jsd_wide) && nrow(jsd_wide) > 0L && ncol(jsd_wide) > 1L) {
      jsd_window_names <- jsd_wide$feature_id
      jsd_mat <- as.matrix(jsd_wide[, -1, with = FALSE])
      jsd_mat[!is.finite(jsd_mat)] <- NA
      rownames(jsd_mat) <- jsd_window_names
      ## run_embedding expects features x cells; we want windows as
      ## observations, so groups become "features" via transpose.
      jsd_window_embed <- tryCatch(
        run_embedding(t(jsd_mat), n_hvf = N_HVW,
                      n_pcs = max(2L, min(N_PCS, ncol(jsd_mat) - 1L)),
                      n_neighbors = 15L, min_dist = 0.3, seed = 42L),
        error = function(e) {
          message("[locus_umap_jsd] embed failed: ", conditionMessage(e)); NULL })
    }
  }
}
```

```{r locus_umap_jsd_plot, fig.width = ng_fig_size(3, 1, panel_mm = 55)$w, fig.height = ng_fig_size(3, 1, panel_mm = 55)$h}
if (is.null(jsd_window_embed)) {
  cat("Per-locus jsd UMAP unavailable.\n")
} else if (is.null(ann_bin) || ncol(ann_bin) == 0L) {
  df <- data.frame(UMAP1 = jsd_window_embed$umap[, 1],
                   UMAP2 = jsd_window_embed$umap[, 2])
  print(ggplot(df, aes(UMAP1, UMAP2)) +
          geom_point_rast(size = 0.15, alpha = 0.4, raster.dpi = 300) +
          labs(title = "per-locus jsd UMAP") + theme_ng())
} else {
  win_names <- jsd_window_names[jsd_window_embed$kept_cols]
  ann_aligned <- ann_bin[match(win_names, rownames(ann_bin)), , drop = FALSE]
  df_base <- data.frame(UMAP1 = jsd_window_embed$umap[, 1],
                        UMAP2 = jsd_window_embed$umap[, 2],
                        feature_id = win_names)
  panels <- lapply(colnames(ann_aligned), function(cn) {
    df <- df_base
    df$overlap <- factor(ann_aligned[[cn]], levels = c(0, 1))
    ggplot(df, aes(UMAP1, UMAP2, color = overlap)) +
      geom_point_rast(size = 0.12, alpha = 0.55, raster.dpi = 300) +
      scale_color_manual(values = c("0" = "#bbbbbb", "1" = "#1b9e77"),
                         na.value = "#eeeeee", drop = FALSE) +
      labs(title = sub("_bin$", "", cn)) +
      theme_ng(base_size = 7) +
      theme(legend.position = "none",
            plot.title = element_text(size = 6.5))
  })
  print(plot_grid(plotlist = panels, ncol = 3))
}
```

```{r session}
sessionInfo()
```
