---
title: "amet on Liu 2021 mouse brain CEMBA (MOp)"
author: "Izaskun Mallona"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    number_sections: true
    df_print: default
params:
  feat_cell_feature: ""
  feat_feature: ""
  feat_bed: ""
  win_cell_feature: ""
  win_feature: ""
  win_bed: ""
  manifest: ""
  out_dir: ""
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 5, fig.height = 4)

suppressPackageStartupMessages({
    library(data.table)
    library(ggplot2)
    library(ggridges)
    library(ggrepel)
    library(viridis)
    library(patchwork)
    library(dplyr)
    library(tidyr)
    library(uwot)
})

repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))

panels_dir <- file.path(params$out_dir, "panels")
feat_dir <- file.path(panels_dir, "feature")
win_dir <- file.path(panels_dir, "window")
dir.create(feat_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(win_dir, showWarnings = FALSE, recursive = TRUE)
```

# Inputs and design

CEMBA mouse motor cortex (MOp) prototype: chr19, four cell types from the 4587-cell atlas. Per-cell raw files are methylpy allc tsv.gz, read by amet's allc parser via per-tar extraction.

The report is split into two analyses with disjoint feature definitions:

- **Feature-based**: amet outputs over genes on chr19. One feature = one gene locus. Used to look at CpG-context regularity inside transcribed loci, contrasted across cell types.
- **Window-based**: amet outputs over fixed-size genome tiles on chr19. One feature = one window. Used for cell-level summaries and embeddings without locus-bias.

Score mapping: amet `i_total` (within-cell mutual information across CpG lags) for adjS-like questions; amet `jsd` (across-cell heterogeneity) for adjH-like questions.

```{r load}
man <- fread(params$manifest)
feat_cf <- fread(params$feat_cell_feature)
feat_fe <- fread(params$feat_feature)
win_cf <- fread(params$win_cell_feature)
win_fe <- fread(params$win_feature)
feat_bed <- fread(params$feat_bed, header = FALSE,
                  col.names = c("chrom", "start", "end", "feature_id"))
win_bed <- fread(params$win_bed, header = FALSE,
                 col.names = c("chrom", "start", "end", "feature_id"))

# group is sub_type. Pass-through extras (cell_class, major_type) are kept if
# present on the manifest.
extra_cols <- intersect(c("cell_class", "major_type", "region", "sub_region"),
                        colnames(man))
keep <- c("cell_id", "group", extra_cols)
man <- man[, ..keep]

feat_cf <- merge(feat_cf, man, by = c("cell_id", "group"), all.x = TRUE)
win_cf <- merge(win_cf, man, by = c("cell_id", "group"), all.x = TRUE)
group_levels <- sort(unique(man$group))
ng <- length(group_levels)
celltype_pal <- setNames(viridis::viridis(ng, end = 0.95), group_levels)
```

# Sample sizes and coverage

```{r sizes, fig.width = 6, fig.height = 3.5}
sz <- man[, .(cells = .N), by = group][order(group)]
print(sz)

p_n <- ggplot(sz, aes(x = group, y = cells, fill = group)) +
    geom_col(width = 0.7) +
    scale_fill_manual(values = celltype_pal) +
    labs(x = NULL, y = "cells") +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")
save_eval(p_n, sz, file.path(panels_dir, "ncells_per_group"),
          width_mm = 90, height_mm = 60)
p_n
```

```{r coverage, fig.width = 7, fig.height = 3.5}
cov_feat <- feat_cf[, .(n_features = .N,
                        median_n_cpgs = median(n_cpgs, na.rm = TRUE)),
                    by = .(cell_id, group)]
cov_win <- win_cf[, .(n_windows = .N,
                      median_n_cpgs = median(n_cpgs, na.rm = TRUE)),
                  by = .(cell_id, group)]
cov_long <- rbind(
    cov_feat[, .(cell_id, group, n = n_features, scope = "feature (gene)")],
    cov_win[,  .(cell_id, group, n = n_windows,  scope = "window")]
)
p_cov <- ggplot(cov_long, aes(x = group, y = n, fill = group)) +
    geom_violin(alpha = 0.6, scale = "width") +
    geom_boxplot(width = 0.18, outlier.size = 0.3, fill = "white") +
    scale_fill_manual(values = celltype_pal) +
    facet_wrap(~ scope, scales = "free_y") +
    labs(x = NULL, y = "covered features per cell") +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")
save_eval(p_cov, cov_long, file.path(panels_dir, "coverage_per_cell"),
          width_mm = 130, height_mm = 70)
p_cov
```

# Feature-based analysis: gene loci on chr19

```{r feat_filter}
fcf <- feat_cf[!is.na(i_total)]
ffe <- feat_fe[!is.na(jsd)]
ffe[, group := factor(group, levels = group_levels)]
```

## Within-cell regularity (i_total)

### Per-cell-type distribution

```{r feat_itotal_box, fig.width = 6, fig.height = 4}
p_itotal_box <- ggplot(fcf, aes(x = group, y = i_total, fill = group)) +
    geom_violin(alpha = 0.55, scale = "width", linewidth = 0.3) +
    geom_boxplot(width = 0.18, outlier.size = 0.3, fill = "white",
                 linewidth = 0.3) +
    scale_fill_manual(values = celltype_pal) +
    labs(x = NULL, y = expression(i[total])) +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")
save_eval(p_itotal_box, fcf[, .(cell_id, group, feature_id, i_total)],
          file.path(feat_dir, "itotal_by_group"),
          width_mm = 110, height_mm = 70)
p_itotal_box
```

### Density ridges per cell type

```{r feat_itotal_ridge, fig.width = 6, fig.height = 4}
p_ridge <- ggplot(fcf, aes(x = i_total, y = group, fill = group)) +
    geom_density_ridges(alpha = 0.7, scale = 1.2, linewidth = 0.3) +
    scale_fill_manual(values = celltype_pal) +
    labs(x = expression(i[total]), y = NULL) +
    theme_ng() +
    theme(legend.position = "none", aspect.ratio = NULL)
save_eval(p_ridge, fcf[, .(group, i_total)],
          file.path(feat_dir, "itotal_ridge_group"),
          width_mm = 110, height_mm = 70)
p_ridge
```

### Mean per cell

```{r feat_itotal_per_cell, fig.width = 6, fig.height = 4}
cell_summary_feat <- fcf[, .(mean_i_total = mean(i_total, na.rm = TRUE),
                             median_i_total = median(i_total, na.rm = TRUE),
                             mean_meth = mean(mean_meth, na.rm = TRUE),
                             n_features = .N),
                         by = .(cell_id, group)]
p_cell_it <- ggplot(cell_summary_feat,
                    aes(x = group, y = mean_i_total, fill = group)) +
    geom_boxplot(outlier.size = 0.3, linewidth = 0.3, alpha = 0.75) +
    geom_jitter(width = 0.18, size = 0.6, alpha = 0.7) +
    scale_fill_manual(values = celltype_pal) +
    labs(x = NULL, y = expression("mean " * i[total] * " per cell")) +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")
save_eval(p_cell_it, cell_summary_feat,
          file.path(feat_dir, "itotal_per_cell"),
          width_mm = 110, height_mm = 70)
p_cell_it
```

## Methylation per cell type

```{r feat_meth, fig.width = 6, fig.height = 4}
p_meth <- ggplot(cell_summary_feat,
                 aes(x = group, y = mean_meth, fill = group)) +
    geom_boxplot(outlier.size = 0.3, linewidth = 0.3, alpha = 0.75) +
    scale_fill_manual(values = celltype_pal) +
    labs(x = NULL, y = "mean methylation per cell (gene loci)") +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")
save_eval(p_meth, cell_summary_feat,
          file.path(feat_dir, "meanmeth_per_cell"),
          width_mm = 110, height_mm = 70)
p_meth
```

## Decoupling: i_total vs methylation

```{r feat_decouple, fig.width = 6, fig.height = 4}
p_dec <- ggplot(fcf, aes(x = mean_meth, y = i_total, colour = group)) +
    geom_point(size = 0.5, alpha = 0.5) +
    geom_smooth(method = "loess", se = FALSE, linewidth = 0.4) +
    scale_colour_manual(values = celltype_pal) +
    labs(x = "mean methylation (per cell, per gene)",
         y = expression(i[total]),
         colour = "cell type") +
    theme_ng()
save_eval(p_dec, fcf[, .(cell_id, group, mean_meth, i_total)],
          file.path(feat_dir, "itotal_vs_meth_scatter"),
          width_mm = 110, height_mm = 70)
p_dec
```

```{r feat_decouple_per_cell, fig.width = 5, fig.height = 4}
p_dec_pc <- ggplot(cell_summary_feat,
                   aes(x = mean_meth, y = mean_i_total, colour = group)) +
    geom_point(size = 1.6, alpha = 0.9) +
    scale_colour_manual(values = celltype_pal) +
    labs(x = "mean methylation per cell",
         y = expression("mean " * i[total] * " per cell")) +
    theme_ng()
save_eval(p_dec_pc, cell_summary_feat,
          file.path(feat_dir, "itotal_vs_meth_per_cell"))
p_dec_pc
```

## Across-cell heterogeneity (jsd)

### Per-cell-type distribution

```{r feat_jsd_box, fig.width = 6, fig.height = 4}
if (nrow(ffe)) {
    p_jsd_box <- ggplot(ffe, aes(x = group, y = jsd, fill = group)) +
        geom_violin(alpha = 0.55, scale = "width", linewidth = 0.3) +
        geom_boxplot(width = 0.18, outlier.size = 0.3, fill = "white",
                     linewidth = 0.3) +
        scale_fill_manual(values = celltype_pal) +
        labs(x = NULL, y = "jsd") +
        theme_ng() +
        theme(axis.text.x = element_text(angle = 30, hjust = 1),
              legend.position = "none")
    save_eval(p_jsd_box, ffe[, .(feature_id, group, n_cells, jsd)],
              file.path(feat_dir, "jsd_by_group"),
              width_mm = 110, height_mm = 70)
    p_jsd_box
}
```

### Density ridges

```{r feat_jsd_ridge, fig.width = 6, fig.height = 4}
if (exists("p_jsd_box")) {
    p_jsd_ridge <- ggplot(ffe, aes(x = jsd, y = group, fill = group)) +
        geom_density_ridges(alpha = 0.7, scale = 1.1, linewidth = 0.3) +
        scale_fill_manual(values = celltype_pal) +
        labs(x = "jsd", y = NULL) +
        theme_ng() +
        theme(legend.position = "none", aspect.ratio = NULL)
    save_eval(p_jsd_ridge, ffe[, .(group, jsd)],
              file.path(feat_dir, "jsd_ridge_group"),
              width_mm = 110, height_mm = 70)
    p_jsd_ridge
}
```

## Joint structure: i_total vs jsd per gene

```{r feat_joint, fig.width = 5, fig.height = 4}
fe_joint <- fcf[, .(median_i_total = median(i_total, na.rm = TRUE)),
                by = .(feature_id, group)]
fe_joint <- merge(fe_joint, ffe[, .(feature_id, group, jsd, n_cells)],
                  by = c("feature_id", "group"))
fe_joint <- fe_joint[!is.na(jsd) & !is.na(median_i_total)]
p_joint <- ggplot(fe_joint,
                  aes(x = jsd, y = median_i_total, colour = group)) +
    geom_point(size = 0.7, alpha = 0.85) +
    scale_colour_manual(values = celltype_pal) +
    labs(x = "jsd (per gene, per cell type)",
         y = expression("median " * i[total]),
         colour = "cell type") +
    theme_ng()
save_eval(p_joint, fe_joint, file.path(feat_dir, "itotal_vs_jsd_per_feature"))
p_joint
```

## Heatmap: median i_total per (gene, cell type)

```{r feat_heatmap_it, fig.width = 5, fig.height = 6}
hm <- dcast(fcf[, .(median_i_total = median(i_total, na.rm = TRUE)),
                by = .(feature_id, group)],
            feature_id ~ group, value.var = "median_i_total")
hm_mat <- as.matrix(hm[, -1])
rownames(hm_mat) <- hm$feature_id
hm_mat <- hm_mat[rowSums(!is.na(hm_mat)) >= 2, , drop = FALSE]
hm_mat <- hm_mat[order(rowMeans(hm_mat, na.rm = TRUE)), , drop = FALSE]
hm_long <- as.data.table(reshape2::melt(hm_mat,
                                        varnames = c("feature_id", "group")))
hm_long[, feature_id := factor(feature_id, levels = rownames(hm_mat))]

p_hm_it <- ggplot(hm_long, aes(x = group, y = feature_id, fill = value)) +
    geom_tile() +
    scale_fill_viridis(option = "magma", na.value = "grey90") +
    labs(x = NULL, y = "gene", fill = expression(i[total])) +
    theme_ng() +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
          axis.text.x = element_text(angle = 30, hjust = 1),
          aspect.ratio = NULL)
save_eval(p_hm_it, hm_long, file.path(feat_dir, "heatmap_median_itotal"),
          width_mm = 90, height_mm = 130)
p_hm_it
```

## Heatmap: jsd per (gene, cell type)

```{r feat_heatmap_jsd, fig.width = 5, fig.height = 6}
if (nrow(ffe)) {
    jsd_w <- dcast(ffe, feature_id ~ group, value.var = "jsd")
    jsd_mat <- as.matrix(jsd_w[, -1])
    rownames(jsd_mat) <- jsd_w$feature_id
    jsd_mat <- jsd_mat[rowSums(!is.na(jsd_mat)) >= 1, , drop = FALSE]
    jsd_mat <- jsd_mat[order(rowMeans(jsd_mat, na.rm = TRUE)), , drop = FALSE]
    jsd_long <- as.data.table(reshape2::melt(jsd_mat,
                                             varnames = c("feature_id", "group")))
    jsd_long[, feature_id := factor(feature_id, levels = rownames(jsd_mat))]

    p_hm_jsd <- ggplot(jsd_long, aes(x = group, y = feature_id, fill = value)) +
        geom_tile() +
        scale_fill_viridis(option = "viridis", na.value = "grey90") +
        labs(x = NULL, y = "gene", fill = "jsd") +
        theme_ng() +
        theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              axis.text.x = element_text(angle = 30, hjust = 1),
              aspect.ratio = NULL)
    save_eval(p_hm_jsd, jsd_long, file.path(feat_dir, "heatmap_jsd"),
              width_mm = 90, height_mm = 130)
    p_hm_jsd
}
```

## Top variable genes

```{r feat_top_var, fig.width = 6, fig.height = 4}
var_dt <- fcf[, .(sd_i_total = sd(i_total, na.rm = TRUE),
                  median_i_total = median(i_total, na.rm = TRUE),
                  n = .N), by = feature_id][n >= 5]
top_var <- head(var_dt[order(-sd_i_total)], 20)

p_top_var <- ggplot(top_var,
                    aes(x = reorder(feature_id, sd_i_total), y = sd_i_total)) +
    geom_col(fill = "#444444") +
    coord_flip() +
    labs(x = NULL, y = expression("sd(" * i[total] * ") across cells")) +
    theme_ng() +
    theme(axis.text.y = element_text(size = 6), aspect.ratio = NULL)
save_eval(p_top_var, top_var, file.path(feat_dir, "top_variable_genes"),
          width_mm = 110, height_mm = 80)
p_top_var
```

# Window-based analysis: chr19 genome tiles

```{r win_filter}
wcf <- win_cf[!is.na(i_total)]
wfe <- win_fe[!is.na(jsd)]
wfe[, group := factor(group, levels = group_levels)]
```

## Per-cell coverage

```{r win_coverage, fig.width = 6, fig.height = 4}
win_cov <- win_cf[, .(n_windows = .N), by = .(cell_id, group)]
p_win_cov <- ggplot(win_cov, aes(x = group, y = n_windows, fill = group)) +
    geom_violin(alpha = 0.55, scale = "width", linewidth = 0.3) +
    geom_jitter(width = 0.15, size = 0.5, alpha = 0.6) +
    scale_fill_manual(values = celltype_pal) +
    labs(x = NULL, y = "covered windows per cell") +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")
save_eval(p_win_cov, win_cov, file.path(win_dir, "coverage_windows_per_cell"),
          width_mm = 110, height_mm = 70)
p_win_cov
```

## Mean methylation per cell

```{r win_meanmeth, fig.width = 6, fig.height = 4}
cell_meth <- win_cf[!is.na(mean_meth),
                    .(mean_meth = mean(mean_meth, na.rm = TRUE),
                      n_windows = .N), by = .(cell_id, group)]
p_wmeth <- ggplot(cell_meth, aes(x = group, y = mean_meth, fill = group)) +
    geom_violin(alpha = 0.55, scale = "width", linewidth = 0.3) +
    geom_boxplot(width = 0.18, outlier.size = 0.3, fill = "white",
                 linewidth = 0.3) +
    scale_fill_manual(values = celltype_pal) +
    labs(x = NULL, y = "mean methylation per cell (windows)") +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")
save_eval(p_wmeth, cell_meth, file.path(win_dir, "meanmeth_per_cell"),
          width_mm = 110, height_mm = 70)
p_wmeth
```

## Mean i_total per cell

```{r win_itotal, fig.width = 6, fig.height = 4}
cell_it <- wcf[, .(mean_i_total = mean(i_total, na.rm = TRUE),
                   median_i_total = median(i_total, na.rm = TRUE),
                   n_windows = .N), by = .(cell_id, group)]
p_wit <- ggplot(cell_it, aes(x = group, y = mean_i_total, fill = group)) +
    geom_violin(alpha = 0.55, scale = "width", linewidth = 0.3) +
    geom_boxplot(width = 0.18, outlier.size = 0.3, fill = "white",
                 linewidth = 0.3) +
    scale_fill_manual(values = celltype_pal) +
    labs(x = NULL, y = expression("mean " * i[total] * " per cell")) +
    theme_ng() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "none")
save_eval(p_wit, cell_it, file.path(win_dir, "mean_itotal_per_cell"),
          width_mm = 110, height_mm = 70)
p_wit
```

## Methylation vs i_total per cell

```{r win_corr, fig.width = 5, fig.height = 4}
cell_full <- merge(cell_meth, cell_it[, .(cell_id, mean_i_total)],
                   by = "cell_id")
p_wcorr <- ggplot(cell_full,
                  aes(x = mean_meth, y = mean_i_total, colour = group)) +
    geom_point(size = 1.6, alpha = 0.9) +
    scale_colour_manual(values = celltype_pal) +
    labs(x = "mean methylation per cell",
         y = expression("mean " * i[total] * " per cell"),
         colour = "cell type") +
    theme_ng()
save_eval(p_wcorr, cell_full, file.path(win_dir, "itotal_vs_meth_per_cell"))
p_wcorr
```

## Window-level i_total and jsd ridges

```{r win_ridges, fig.width = 6, fig.height = 4}
p_w_it_r <- ggplot(wcf, aes(x = i_total, y = group, fill = group)) +
    geom_density_ridges(alpha = 0.7, scale = 1.2, linewidth = 0.3) +
    scale_fill_manual(values = celltype_pal) +
    labs(x = expression(i[total] * " per (cell, window)"), y = NULL) +
    theme_ng() +
    theme(legend.position = "none", aspect.ratio = NULL)
save_eval(p_w_it_r, wcf[, .(group, i_total)],
          file.path(win_dir, "itotal_ridge_group"),
          width_mm = 110, height_mm = 70)
p_w_it_r
```

```{r win_jsd_ridges, fig.width = 6, fig.height = 4}
if (nrow(wfe)) {
    p_w_jsd_r <- ggplot(wfe, aes(x = jsd, y = group, fill = group)) +
        geom_density_ridges(alpha = 0.7, scale = 1.1, linewidth = 0.3) +
        scale_fill_manual(values = celltype_pal) +
        labs(x = "jsd per (window, cell type)", y = NULL) +
        theme_ng() +
        theme(legend.position = "none", aspect.ratio = NULL)
    save_eval(p_w_jsd_r, wfe[, .(group, jsd)],
              file.path(win_dir, "jsd_ridge_group"),
              width_mm = 110, height_mm = 70)
    p_w_jsd_r
}
```

## Cell embedding: PCA

```{r win_build_mat}
mat <- dcast(wcf, cell_id ~ feature_id, value.var = "i_total",
             fun.aggregate = mean)
mat_cells <- mat$cell_id
mat <- as.matrix(mat[, -1])
rownames(mat) <- mat_cells
mat <- mat[, colSums(!is.na(mat)) >= 0.5 * nrow(mat), drop = FALSE]
mat[is.na(mat)] <- 0
cat("embedding matrix:", nrow(mat), "cells x", ncol(mat), "windows\n")
```

```{r win_pca, fig.width = 5, fig.height = 4}
emb <- NULL
if (ncol(mat) >= 2 && nrow(mat) >= 3) {
    pcs <- prcomp(mat, scale. = FALSE)
    emb <- data.table(cell_id = rownames(mat),
                      PC1 = pcs$x[, 1], PC2 = pcs$x[, 2],
                      PC3 = if (ncol(pcs$x) >= 3) pcs$x[, 3] else NA_real_)
    emb <- merge(emb, man[, .(cell_id, group)], by = "cell_id")
    p_pca <- ggplot(emb, aes(x = PC1, y = PC2, colour = group)) +
        geom_point(size = 1.8, alpha = 0.95) +
        scale_colour_manual(values = celltype_pal) +
        labs(colour = "cell type") +
        theme_ng()
    save_eval(p_pca, emb, file.path(win_dir, "pca_group"))
    p_pca
}
```

```{r win_pca_scree, fig.width = 5, fig.height = 4}
if (exists("pcs")) {
    var_explained <- (pcs$sdev^2) / sum(pcs$sdev^2)
    scree <- data.table(PC = seq_along(var_explained),
                        var_explained = var_explained)[1:min(20, .N)]
    p_scree <- ggplot(scree, aes(x = PC, y = var_explained)) +
        geom_col(fill = "#444444") +
        labs(x = "PC", y = "variance explained") +
        theme_ng()
    save_eval(p_scree, scree, file.path(win_dir, "pca_scree"))
    p_scree
}
```

```{r win_pc23, fig.width = 5, fig.height = 4}
if (!is.null(emb) && !all(is.na(emb$PC3))) {
    p_23 <- ggplot(emb, aes(x = PC2, y = PC3, colour = group)) +
        geom_point(size = 1.8, alpha = 0.95) +
        scale_colour_manual(values = celltype_pal) +
        labs(colour = "cell type") +
        theme_ng()
    save_eval(p_23, emb, file.path(win_dir, "pca_pc2_pc3"))
    p_23
}
```

## Cell embedding: UMAP

```{r win_umap, fig.width = 5, fig.height = 4}
if (!is.null(emb) && nrow(mat) >= 5 && ncol(mat) >= 5) {
    set.seed(1L)
    n_neigh <- max(2L, min(15L, nrow(mat) - 1L))
    um <- tryCatch(uwot::umap(mat, n_neighbors = n_neigh, min_dist = 0.3,
                              n_components = 2, verbose = FALSE),
                   error = function(e) { cat(conditionMessage(e), "\n"); NULL })
    if (!is.null(um)) {
        umd <- data.table(cell_id = rownames(mat),
                          UMAP1 = um[, 1], UMAP2 = um[, 2])
        umd <- merge(umd, man[, .(cell_id, group)], by = "cell_id")
        p_um <- ggplot(umd, aes(x = UMAP1, y = UMAP2, colour = group)) +
            geom_point(size = 1.8, alpha = 0.95) +
            scale_colour_manual(values = celltype_pal) +
            labs(colour = "cell type") +
            theme_ng()
        save_eval(p_um, umd, file.path(win_dir, "umap_group"))
        p_um
    }
}
```

## Top variable windows

```{r win_top_var, fig.width = 6, fig.height = 4}
win_var <- wcf[, .(sd_i_total = sd(i_total, na.rm = TRUE),
                   n = .N), by = feature_id][n >= 5]
top_win <- head(win_var[order(-sd_i_total)], 20)
p_win_top <- ggplot(top_win,
                    aes(x = reorder(feature_id, sd_i_total), y = sd_i_total)) +
    geom_col(fill = "#444444") +
    coord_flip() +
    labs(x = NULL, y = expression("sd(" * i[total] * ") across cells")) +
    theme_ng() +
    theme(axis.text.y = element_text(size = 6), aspect.ratio = NULL)
save_eval(p_win_top, top_win, file.path(win_dir, "top_variable_windows"),
          width_mm = 110, height_mm = 80)
p_win_top
```

## Heatmap of top-variable windows by cell type

```{r win_heatmap_top, fig.width = 5, fig.height = 5}
top_ids <- top_win$feature_id
hm_w <- dcast(wcf[feature_id %in% top_ids,
                  .(median_i = median(i_total, na.rm = TRUE)),
                  by = .(feature_id, group)],
              feature_id ~ group, value.var = "median_i")
hm_w_mat <- as.matrix(hm_w[, -1])
rownames(hm_w_mat) <- hm_w$feature_id
hm_w_long <- as.data.table(reshape2::melt(hm_w_mat,
                                          varnames = c("feature_id", "group")))
p_win_hm <- ggplot(hm_w_long, aes(x = group, y = feature_id, fill = value)) +
    geom_tile() +
    scale_fill_viridis(option = "magma", na.value = "grey90") +
    labs(x = NULL, y = "window", fill = expression(i[total])) +
    theme_ng() +
    theme(axis.text.y = element_text(size = 6),
          axis.text.x = element_text(angle = 30, hjust = 1),
          aspect.ratio = NULL)
save_eval(p_win_hm, hm_w_long, file.path(win_dir, "heatmap_top_variable_windows"),
          width_mm = 100, height_mm = 130)
p_win_hm
```

# Combined paper figure

```{r paper, fig.width = 11, fig.height = 7}
panels <- list(
    A = if (exists("p_itotal_box")) p_itotal_box + labs(title = "A. i_total / genes") else NULL,
    B = if (exists("p_jsd_box")) p_jsd_box + labs(title = "B. jsd / genes") else NULL,
    C = if (exists("p_dec")) p_dec + labs(title = "C. decoupling / genes") else NULL,
    D = if (exists("p_wit")) p_wit + labs(title = "D. mean i_total per cell / windows") else NULL,
    E = if (exists("p_pca")) p_pca + labs(title = "E. PCA cells / windows") else NULL,
    F = if (exists("p_um")) p_um + labs(title = "F. UMAP cells / windows") else NULL
)
panels <- Filter(Negate(is.null), panels)
if (length(panels) >= 2) {
    combined <- Reduce(`+`, panels) + plot_layout(ncol = 3)
    ggsave(file.path(panels_dir, "paper_figure.pdf"),
           combined, width = 260, height = 160, units = "mm",
           device = grDevices::cairo_pdf)
    combined
}
```

# Session

```{r session}
sessionInfo()
```
