---
title: "Ecker 2021 mouse brain: methylation entropy by feature (CpG context)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Izaskun Mallona"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
params:
  features_dir: ""
  win_cell_feature: ""
  win_feature: ""
  win_bed: ""
  manifest: ""
  out_dir: ""
---

```{r setup}
suppressPackageStartupMessages({
  library(data.table)
  library(readr)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(BiocParallel)
  library(ggrepel)
  library(ggridges)
  library(ComplexHeatmap)
  library(circlize)
  library(uwot)
  library(irlba)
})

repo_root <- normalizePath(file.path(dirname(knitr::current_input()), "..", ".."))
source(file.path(repo_root, "workflow", "scripts", "plot_theme.R"))
source(file.path(repo_root, "workflow", "scripts", "palettes.R"))
source(file.path(repo_root, "workflow", "scripts", "embedding_utils.R"))
source(file.path(repo_root, "workflow", "scripts", "driver_utils.R"))
param <- SerialParam()
```

```{r opts}
knitr::opts_chunk$set(
  echo       = TRUE,
  fig.width  = 5,
  fig.height = 5,
  cache      = FALSE,
  include    = TRUE,
  fig.path   = "ecker_plots/",
  dev        = c("png", "svg"),
  dpi        = 500,
  cache.lazy = FALSE,
  warning    = TRUE,
  message    = TRUE
)
```

Mouse (mm10), CpG context only.

```{r helpers}
## filenames: {annotation}_{sub_region}_{sub_type}.*.gz
## annotation has no underscores; neither do sub_region or sub_type values
get_annotation <- function(fn) sub("^(.*)_[^_]+_[^_]+\\..*$", "\\1", fn)
get_sub_region <- function(fn) sub("^.*_([^_]+)_[^_]+\\..*$", "\\1", fn)
get_sub_type   <- function(fn) sub("^.*_([^_.]+)\\.[^_]*$",   "\\1", fn)

ann_labels <- c(
  "genes"    = "Genes",
  "promoters"= "Promoters",
  "lines"    = "LINEs",
  "sines"    = "SINEs",
  "h3k4me3"  = "H3K4me3",
  "h3k9me3"  = "H3K9me3",
  "h3k27me3" = "H3K27me3",
  "h3k4me1"  = "H3K4me1",
  "h3k27ac"  = "H3K27ac"
)

harmonize_levels <- function(df) {
  missing_ann <- setdiff(unique(df$annotation), names(ann_labels))
  if (length(missing_ann) > 0)
    warning("unknown annotations: ", paste(missing_ann, collapse = ", "))
  df$annotation <- factor(df$annotation,
                           levels = names(ann_labels),
                           labels = unname(ann_labels))
  if ("sub_region" %in% names(df)) df$sub_region <- factor(df$sub_region)
  if ("sub_type"   %in% names(df)) df$sub_type   <- factor(df$sub_type)
  df
}

## _sanitize for Ecker: gsub(" ", "-", x).
sanitize <- function(x) gsub(" ", "-", x)
```

```{r load_metadata_early}
## Load cell-level metadata early so cell_class / major_type are available
## in all data-loading chunks below. amet's manifest carries sub_region,
## sub_type, cell_class, major_type per cell (see make_manifest_ecker.py).
man <- fread(params$manifest)
meta_cells <- as.data.frame(man) %>%
  mutate(
    sub_region = sanitize(as.character(sub_region)),
    sub_type   = sanitize(as.character(sub_type))
  )

meta_grp <- meta_cells %>%
  group_by(sub_region, sub_type) %>%
  summarise(
    cell_class = if ("cell_class" %in% names(meta_cells))
                   first(na.omit(cell_class)) else NA_character_,
    major_type = if ("major_type" %in% names(meta_cells))
                   first(na.omit(major_type)) else NA_character_,
    n_meta     = n(),
    .groups    = "drop"
  )

cat("meta_grp rows:", nrow(meta_grp), "\n")
cat("cell_class levels:", paste(sort(unique(meta_grp$cell_class)), collapse = ", "), "\n")
```

```{r load_amet}
## amet emits one TSV pair per (annotation, sub_region, sub_type) stratum.
## Enumerate, parse the wildcard values from filenames, concatenate.
features_dir <- params$features_dir
cf_files <- list.files(features_dir,
                       pattern = "\\.cell_feature\\.tsv\\.gz$",
                       full.names = TRUE)
fe_files <- list.files(features_dir,
                       pattern = "\\.feature\\.tsv\\.gz$",
                       full.names = TRUE)
fe_files <- fe_files[!grepl("\\.cell_feature\\.tsv\\.gz$", fe_files)]

## Aggregate per cell at read time so feat_cf collapses from O(cells *
## features * strata * annotations) rows to O(cells * strata * annotations).
## Every downstream plot/embedding averages over features per cell.
read_combo_cf <- function(fp) {
  fn <- basename(fp)
  base <- sub("\\.cell_feature\\.tsv\\.gz$", "", fn)
  bx <- paste0(base, ".x")
  dt <- fread(fp, select = c("cell_id", "mean_meth", "i_total"))
  agg <- dt[, .(mean_meth = mean(mean_meth, na.rm = TRUE),
                i_total = mean(i_total, na.rm = TRUE)),
            by = cell_id]
  agg[, annotation := get_annotation(bx)]
  agg[, sub_region := get_sub_region(bx)]
  agg[, sub_type   := get_sub_type(bx)]
  agg
}

read_combo_fe <- function(fp) {
  fn <- basename(fp)
  base <- sub("\\.feature\\.tsv\\.gz$", "", fn)
  bx <- paste0(base, ".x")
  dt <- fread(fp)
  dt[, annotation := get_annotation(bx)]
  dt[, sub_region := get_sub_region(bx)]
  dt[, sub_type   := get_sub_type(bx)]
  dt
}

feat_cf <- rbindlist(lapply(cf_files, read_combo_cf), fill = TRUE)
feat_fe <- rbindlist(lapply(fe_files, read_combo_fe), fill = TRUE)

cat("loaded", length(cf_files), "cell_feature TSVs and",
    length(fe_files), "feature TSVs from", features_dir, "\n")

## Attach cell_class / major_type per stratum (one row per
## (sub_region, sub_type), not per cell). Update by reference.
meta_grp_dt <- as.data.table(meta_grp)[, .(sub_region, sub_type,
                                            cell_class, major_type)]
feat_cf[meta_grp_dt, on = .(sub_region, sub_type),
        `:=`(cell_class = i.cell_class, major_type = i.major_type)]
```

<!-- Section dropped: amet ships only methylation-stable i_total, no raw sampen analogue, so the S sample-entropy block (import_sampen, scatter_sampen, violin_sampen, ridges_sampen) is not reproduced. -->

# adjS - methylation-adjusted sample entropy {.tabset .tabset-pills}

Per-cell normalised sample entropy from amet's per-cell-per-feature output
(amet `i_total`, the within-cell mutual-information score).

```{r import_adjS}
adjsampens <- as.data.frame(feat_cf)
adjsampens$sampen   <- adjsampens$i_total
adjsampens$avg_meth <- adjsampens$mean_meth
adjsampens <- harmonize_levels(adjsampens)
head(adjsampens)
```

## Entropy vs methylation

```{r scatter_adjS, fig.width = 10, fig.height = 15}
ggplot(adjsampens %>% filter(!is.na(cell_class)),
       aes(x = avg_meth, y = sampen, color = major_type)) +
  geom_point(size = 0.4, alpha = 0.15) +
  scale_color_manual(values = ecker_major_type_pal, na.value = "grey70") +
  facet_grid(annotation ~ cell_class) +
  labs(x = "average methylation", y = expression(i[total]), color = "cell type") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

## Distributions

```{r violin_adjS, fig.width = 10, fig.height = 15}
ggplot(adjsampens %>% filter(!is.na(cell_class)),
       aes(x = major_type, y = sampen, fill = major_type)) +
  geom_violin(alpha = 0.5, draw_quantiles = 0.5) +
  scale_fill_manual(values = ecker_major_type_pal) +
  facet_grid(annotation ~ cell_class, scales = "free_y") +
  labs(y = expression(i[total]), fill = "cell type") +
  theme_ng() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

```{r ridges_adjS, fig.width = 10, fig.height = 15}
ggplot(adjsampens %>% filter(!is.na(cell_class)),
       aes(x = sampen, y = major_type, fill = major_type)) +
  geom_density_ridges(alpha = 0.6, scale = 0.85) +
  scale_fill_manual(values = ecker_major_type_pal) +
  facet_grid(annotation ~ cell_class, scales = "free") +
  labs(x = expression(i[total])) +
  theme_ng() +
  theme(legend.position = "none")
```

<!-- Section dropped: amet ships only methylation-stable jsd, no raw Shannon entropy analogue, so the H Shannon-entropy median block (import_shannon, scatter_shannon) is not reproduced. -->

# adjH (methylation-normalised Shannon entropy) - medians {.tabset .tabset-pills}

Per-feature jsd from amet's feature TSV (amet `jsd`, the multi-distribution
Jensen-Shannon divergence across cells in a group).

```{r import_adjH}
## Aggregate amet's per-feature jsd to one median per (annotation,
## sub_region, sub_type) so the downstream group_medians join doesn't explode.
adjshannons <- as.data.frame(
  as.data.table(feat_fe)[
    , .(median_shannon  = median(jsd, na.rm = TRUE),
        median_avg_meth = median(mean_meth_mean, na.rm = TRUE)),
    by = .(annotation, sub_region, sub_type)
  ]
)
adjshannons <- harmonize_levels(adjshannons)
adjshannons <- adjshannons %>%
  left_join(meta_grp %>% select(sub_region, sub_type, cell_class, major_type),
            by = c("sub_region", "sub_type"))
head(adjshannons)
```

## Median adjH vs methylation

```{r scatter_adjH, fig.width = 10, fig.height = 15}
ggplot(adjshannons %>% filter(!is.na(cell_class)),
       aes(x = median_avg_meth, y = median_shannon, color = major_type)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_manual(values = ecker_major_type_pal, na.value = "grey70") +
  facet_grid(annotation ~ cell_class) +
  labs(x = "median avg. methylation", y = "median jsd", color = "cell type") +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  theme_ng()
```

# Assembly

adjS (i_total) is per-cell. adjH (jsd) is per-feature per-group.

```{r str_check}
str(adjsampens)
str(adjshannons)
```

```{r save_rds}
saveRDS(
  list(adjsampens  = adjsampens,
       adjshannons = adjshannons),
  file = "ecker_entropy.rds"
)
```

Aggregate per-cell adjS to per-group medians so it is comparable to adjH.

```{r group_medians}
adjS_grp <- adjsampens %>%
  group_by(annotation, sub_region, sub_type) %>%
  summarise(median_adjS = median(sampen,    na.rm = TRUE),
            median_meth = median(avg_meth, na.rm = TRUE),
            .groups = "drop")

all_grp <- adjS_grp %>%
  left_join(adjshannons %>% select(annotation, sub_region, sub_type,
                                    median_adjH = median_shannon),
            by = c("annotation", "sub_region", "sub_type"))

head(all_grp)
```

```{r load_metadata}
## Attach CellClass / MajorType to the group-level table.
all_grp_meta <- all_grp %>%
  mutate(sub_region = as.character(sub_region),
         sub_type   = as.character(sub_type)) %>%
  left_join(meta_grp, by = c("sub_region", "sub_type"))

saveRDS(
  list(all_grp      = all_grp,
       meta_grp     = meta_grp,
       all_grp_meta = all_grp_meta,
       meta_cells   = meta_cells),
  file = "ecker_groups_meta.rds"
)

cat("all_grp_meta rows:", nrow(all_grp_meta), "\n")
cat("cell_class non-NA:", sum(!is.na(all_grp_meta$cell_class)), "\n")
cat("median_adjH non-NA:", sum(!is.na(all_grp_meta$median_adjH)), "\n")
```

## adjS vs adjH

```{r adjS_vs_adjH, fig.width = 13, fig.height = 4}
df_cmp <- all_grp_meta %>% filter(!is.na(cell_class))
if (nrow(df_cmp) > 0) {
  ggplot(df_cmp, aes(x = median_adjH, y = median_adjS, color = annotation)) +
    geom_point(size = 1.5, alpha = 0.8) +
    scale_color_manual(values = ecker_annotation_pal) +
    facet_grid(. ~ cell_class) +
    labs(x = "median jsd", y = expression("median " * i[total]),
         color = "annotation") +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme_ng()
} else {
  message("No data with cell_class assigned - check metadata join")
}
```

<!-- Section dropped: amet has no raw S/H, so the S vs H scatter is not reproduced. -->

# Variation across cell types {.tabset .tabset-pills}

SD and range of group medians per annotation, broken down by cell class.

```{r variation_helpers}
safe_range <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  diff(range(x))
}
```

```{r variation_class, fig.width = 13, fig.height = 4}
var_class <- all_grp_meta %>%
  filter(!is.na(cell_class)) %>%
  group_by(annotation, cell_class) %>%
  summarise(
    adjS_sd    = sd(median_adjS, na.rm = TRUE),
    adjH_sd    = sd(median_adjH, na.rm = TRUE),
    adjS_range = safe_range(median_adjS),
    adjH_range = safe_range(median_adjH),
    .groups = "drop"
  )

if (nrow(var_class) > 0 && any(!is.na(var_class$adjH_sd))) {
  ggplot(var_class, aes(x = adjH_sd, y = adjS_sd,
                         color = annotation, label = annotation)) +
    geom_point(size = 3) +
    geom_text_repel(size = 3) +
    scale_color_manual(values = ecker_annotation_pal) +
    facet_grid(. ~ cell_class) +
    labs(x = "SD of jsd across cell sub-types",
         y = expression("SD of " * i[total] * " across cell sub-types")) +
    theme_ng()
} else {
  message("No variation data available")
}
```

# Driver categorization

For each annotation, we compute the SD of group-level median adjH (across-cell
heterogeneity) and median adjS (within-cell heterogeneity) across all
(sub_region, sub_type) groups. If one SD is at least 1.5x the other, that entropy
component dominates. Annotations where both SDs fall below the 30th percentile
are labelled "neither".

```{r drivers, fig.width = 5, fig.height = 5}
driver_df <- categorize_drivers(all_grp_meta, "cell_class")

cat("Driver table:\n")
print(table(driver_df$driver))

plot_driver_scatter(driver_df,
                    x_label = "SD of jsd across cell class groups",
                    y_label = expression("SD of " * i[total] * " across cell class groups"))
```

# Heatmaps of median entropy per group {.tabset .tabset-pills}

Rows = sub-type (clustered). Columns = annotation. Row annotation bars show cell class and major type.

```{r heatmap_helpers}
make_heatmap <- function(df, value_col, title,
                          palette = c("navy", "white", "firebrick")) {
  ## wide matrix: rows = sub_type, cols = annotation
  wide <- df %>%
    select(sub_type, annotation, cell_class, major_type,
           value = all_of(value_col)) %>%
    filter(!is.na(cell_class)) %>%
    group_by(sub_type, annotation, cell_class, major_type) %>%
    summarise(value = median(value, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = annotation, values_from = value)

  if (nrow(wide) < 2) { message("Too few rows for heatmap"); return(invisible(NULL)) }

  mat <- wide %>% select(-sub_type, -cell_class, -major_type) %>% as.matrix()
  rownames(mat) <- wide$sub_type

  ## row annotations
  row_ann <- rowAnnotation(
    cell_class = wide$cell_class,
    major_type = wide$major_type,
    col = list(cell_class = ecker_cell_class_pal, major_type = ecker_major_type_pal),
    annotation_legend_param = list(
      cell_class = list(title = "cell class"),
      major_type = list(title = "major type")
    )
  )

  col_fun <- colorRamp2(
    c(min(mat, na.rm = TRUE),
      mean(range(mat, na.rm = TRUE)),
      max(mat, na.rm = TRUE)),
    palette
  )

  Heatmap(mat,
          name            = title,
          col             = col_fun,
          row_split       = wide$cell_class,
          cluster_rows    = TRUE,
          cluster_columns = FALSE,
          show_row_names  = TRUE,
          show_column_names = TRUE,
          row_names_gp    = gpar(fontsize = 7),
          right_annotation = row_ann,
          column_title    = title,
          na_col          = "grey90")
}
```

## adjS

```{r heatmap_adjS, fig.width = 10, fig.height = 8}
heat_df_adjs <- adjsampens %>%
  group_by(annotation, sub_region, sub_type, cell_class, major_type) %>%
  summarise(median_adjS = median(sampen, na.rm = TRUE), .groups = "drop")

make_heatmap(heat_df_adjs, "median_adjS", "median i_total",
             palette = c("navy", "white", "firebrick"))
```

## adjH

```{r heatmap_adjH, fig.width = 10, fig.height = 8}
heat_df_adjh <- adjshannons %>%
  group_by(annotation, sub_region, sub_type, cell_class, major_type) %>%
  summarise(median_adjH = median(median_shannon, na.rm = TRUE), .groups = "drop")

make_heatmap(heat_df_adjh, "median_adjH", "median jsd",
             palette = c("navy", "white", "firebrick"))
```

<!-- Section dropped: amet has no raw S, so the median S heatmap is not reproduced. -->

<!-- Section dropped: amet does not export per-locus entropy vectors per group, so the locus-level H distributions block is not reproduced. Per-feature jsd is already plotted above. -->

# UMAP: cell-type separability {.tabset .tabset-pills}

```{r umap_helpers}
plot_umap_class <- function(df, title, subtitle = NULL) {
  if (is.null(df) || nrow(df) == 0) {
    message("No UMAP result for: ", title)
    return(invisible(NULL))
  }
  if (all(is.na(df$cell_class))) {
    message("All cell_class values NA in UMAP for: ", title,
            " - check metadata join")
    return(invisible(NULL))
  }
  ggplot(df %>% filter(!is.na(cell_class)),
         aes(x = UMAP1, y = UMAP2, color = cell_class)) +
    geom_point(size = 2, alpha = 0.8) +
    scale_color_manual(values = ecker_cell_class_pal) +
    labs(title = title, subtitle = subtitle, color = "cell class") +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme_ng()
}
```

## Cell-level: adjS across annotations

```{r cell_umap_load}
## Build a wide cells x annotation matrix from amet's per-cell-per-feature
## i_total.
cells_adjS_long <- as.data.frame(feat_cf)[, c("cell_id", "sub_region",
                                               "sub_type", "annotation",
                                               "i_total")] %>%
  rename(value = i_total) %>%
  group_by(cell_id, sub_region, sub_type, annotation) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop")
cells_adjS_long$annotation <- factor(cells_adjS_long$annotation,
                                      levels = names(ann_labels),
                                      labels = unname(ann_labels))

cells_meth_long <- as.data.frame(feat_cf)[, c("cell_id", "sub_region",
                                               "sub_type", "annotation",
                                               "mean_meth")] %>%
  rename(value = mean_meth) %>%
  group_by(cell_id, sub_region, sub_type, annotation) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop")
cells_meth_long$annotation <- factor(cells_meth_long$annotation,
                                      levels = names(ann_labels),
                                      labels = unname(ann_labels))

cells_adjS_wide <- pivot_wider(cells_adjS_long,
                                id_cols     = c(cell_id, sub_region, sub_type),
                                names_from  = annotation,
                                values_from = value) %>%
  left_join(meta_grp %>% select(sub_region, sub_type, cell_class, major_type),
            by = c("sub_region", "sub_type"))

cells_meth_wide <- pivot_wider(cells_meth_long,
                                id_cols     = c(cell_id, sub_region, sub_type),
                                names_from  = annotation,
                                values_from = value) %>%
  left_join(meta_grp %>% select(sub_region, sub_type, cell_class, major_type),
            by = c("sub_region", "sub_type"))

saveRDS(list(cells_adjS_wide = cells_adjS_wide,
             cells_meth_wide = cells_meth_wide),
        "ecker_cell_matrices.rds")

cat("Cells (i_total):", nrow(cells_adjS_wide), "\n")
cat("Cell class breakdown:\n")
print(table(cells_adjS_wide$cell_class, useNA = "always"))
```

```{r cell_umap_adjS, fig.width = 5, fig.height = 5}
meta_cols_cell <- c("cell_id", "sub_region", "sub_type", "cell_class", "major_type")

umap_cell_adjS <- run_umap_wide(cells_adjS_wide, meta_cols_cell, n_neighbors = 15L)
saveRDS(umap_cell_adjS, "ecker_umap_cell_i_total.rds")

plot_umap_class(umap_cell_adjS,
                title    = "Cell-level UMAP - i_total features",
                subtitle = "per-cell i_total per annotation as feature vector")
```

## Cell-level: avg_meth across annotations

```{r cell_umap_meth, fig.width = 5, fig.height = 5}
umap_cell_meth <- run_umap_wide(cells_meth_wide, meta_cols_cell, n_neighbors = 15L)
saveRDS(umap_cell_meth, "ecker_umap_cell_meth.rds")

plot_umap_class(umap_cell_meth,
                title    = "Cell-level UMAP - avg. methylation features",
                subtitle = "per-cell avg. methylation per annotation as feature vector")
```

<!-- Section dropped: amet has no raw S, so the S-with-meth-regressed-out UMAP is not reproduced. -->

## Group-level: adjH across annotations

```{r group_umap_adjH, fig.width = 5, fig.height = 5}
adjH_wide_grp <- all_grp_meta %>%
  filter(!is.na(cell_class)) %>%
  select(sub_region, sub_type, cell_class, major_type, annotation, median_adjH) %>%
  pivot_wider(names_from = annotation, values_from = median_adjH)

umap_grp_adjH <- run_umap_wide(adjH_wide_grp,
                                meta_cols = c("sub_region", "sub_type",
                                              "cell_class", "major_type"),
                                n_neighbors = 5L)
saveRDS(umap_grp_adjH, "ecker_umap_grp_jsd.rds")

plot_umap_class(umap_grp_adjH,
                title    = "Group-level UMAP - jsd features",
                subtitle = "median jsd per annotation as feature vector")
```

## Group-level: avg_meth across annotations

```{r group_umap_meth, fig.width = 5, fig.height = 5}
meth_wide_grp <- all_grp_meta %>%
  filter(!is.na(cell_class)) %>%
  select(sub_region, sub_type, cell_class, major_type, annotation, median_meth) %>%
  pivot_wider(names_from = annotation, values_from = median_meth)

umap_grp_meth <- run_umap_wide(meth_wide_grp,
                                meta_cols = c("sub_region", "sub_type",
                                              "cell_class", "major_type"),
                                n_neighbors = 5L)
plot_umap_class(umap_grp_meth,
                title    = "Group-level UMAP - avg. methylation features",
                subtitle = "median avg. methylation per annotation as feature vector")
```

<!-- Section dropped: per-locus gene-level UMAP from .norm.det.out.gz is not reproduced. amet does not export per-locus jsd vectors per group; only per-feature jsd is available, which is already plotted above. -->

<!-- Section dropped: paper UMAP overlay is not reproduced. amet's manifest does not carry the Ecker paper's L1UMAP coordinates or MajorRegion; revisit if those columns are wired through compose_ecker_metadata.py. -->

# Session

```{r session, cache = FALSE}
sessionInfo()
```
