---
title: "amet simulations report"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
source(file.path("..", "scripts", "theme.R"))
```

# Setup

This report summarises the simulation runs under `results/simulations/`. Each cell is generated from a 2-state Markov chain. Two conditions per marginal methylation level p:

- **iid**: chain with $p_{00} + p_{11} = 1$, no comethylation.
- **structured**: persistent chain (diagonal = `persistence` from `config/sim.yaml`) at the same marginal.

# I_total decoupling from marginal methylation

The headline claim: `I_total` should be near zero for any iid sequence regardless of p, and positive for any structured sequence.

```{r p-decoupling, fig.width = 3, fig.height = 3}
csv_path <- file.path("..", "..", "results", "simulations", "eval", "p_decoupling.csv")
if (file.exists(csv_path)) {
    agg <- read.csv(csv_path)
    knitr::kable(agg, digits = 3, caption = "I_total mean and SD per condition")
} else {
    cat("Run `snakemake -j 4 simulations` first to generate this report.")
}
```

```{r p-decoupling-plot, fig.width = 3, fig.height = 3}
plot_path <- file.path("..", "..", "results", "simulations", "eval", "p_decoupling.pdf")
if (file.exists(plot_path)) {
    knitr::include_graphics(plot_path)
}
```
